### 2017-03-31 manuscript/fig_aggregation.tex
Move paper to git

\begin{figure}
  \def\aggfg#1#2{%
    \rput[b](#1){\newmatrix{#2}{16pt}{\coordstep}}
    \leftlabel{#2}{$\tilde{a}_f$}
    \refs{#2}{\coordstep}{1}
         {\scriptsize  \canonicalref{#2}{\coordstep}}
     \plotmatrix[\coordstep]{\plotframe{upr}}{naf}{2}{1}
  }

  \begin{center}
    \begin{pspicture}(0,-1)(16.5,10)\showgrid
      \let\showgrid\relax
      \rput[tl](0,10){\input{figures/fig4a-unitprocess.fig}}
      \rput[tl](2.7,10){\input{figures/fig4b-foreground.fig}}
      \rput[tl](5.4,10){\input{figures/fig4c-agg-foreground.fig}}
      \rput[tl](8.1,10){\input{figures/fig4d-partial-bg-agg.fig}}
      \rput[tl](11.6,10){\input{figures/fig4e-full-agg.fig}}
      \rput[tl](15.1,10){\input{figures/fig4f-lci.fig}}
    \end{pspicture}
    \caption{Publication structures for different forms of foreground aggregation.}
    \label{fig:aggregation}
  \end{center}
\end{figure}

### 2017-03-31 manuscript/figures/potato-organic.tex
Move paper to git

market for potato, organic [GLO]

{\scriptsize\sffamily
\begin{tabularx}{\textwidth}{|>{\hangindent=2ex}X|c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}|c|}
\hline
(node) Foreground flow \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & \\ 
\hline
(0) potato, organic (kg) [GLO] & \refbox  &  &  &  &  &  &  &  &  & \\ 
(1) potato, organic (kg) [CH] & 0.024 & \refbox  &  &  &  &  &  &  &  & \\ 
(2) potato, organic (kg) [RoW] & 0.976 &  & \refbox  &  &  &  &  &  &  & \\ 
(3) potato seed, organic, at farm (kg) [RoW] &  &  &  & \refbox  &  &  &  & 0.976 &  & \\ 
(4) potato seed, organic, for setting (kg) [CH] &  &  &  &  & \refbox  &  &  &  & 0.024 & \\ 
(5) potato seed, organic, for setting (kg) [RoW] &  &  &  &  &  & \refbox  &  &  & 0.976 & \\ 
(6) potato seed, organic, at farm (kg) [CH] &  &  &  &  &  &  & \refbox  & 0.024 &  & \\ 
(7) potato seed, organic, at farm (kg) [GLO] &  &  &  &  &    1 &    1 &  & \refbox  &  & \\ 
(8) potato seed, organic, for setting (kg) [GLO] &  & 0.11 & 0.11 & 0.16 &  &  & 0.16 &  & \refbox  & \\ 
\hline
Foreground Node Weights $\tilde{x}$ &    1 & 0.024 & 0.976 & 0.128 & 0.00315 & 0.128 & 0.00315 & 0.131 & 0.131& \\ 
\hline
\hline
Background Dependencies \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & $\tilde{a_d}$\\ 
\hline
transport, freight train (metric ton*km) [GLO] & \dependency &  &  &  &  &  &  & \dependency &  & 0.138\\ 
irrigation (m3) [CH] &  & \dependency &  &  &  &  & \dependency &  &  & 0.000426\\ 
tillage, hoeing and earthing-up, potatoes (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.000101\\ 
potato planting (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 5.04e-05\\ 
solid manure loading and spreading, by hydraulic loader and spreader (kg) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.717\\ 
electricity, low voltage (kWh) [CH] &  &  &  &  & \dependency &  &  &  &  & 0.00023\\ 
potato grading (kg) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  &  1.13\\ 
transport, freight, light commercial vehicle (metric ton*km) [GLO] & \dependency &  &  &  &  &  &  & \dependency &  & 0.0258\\ 
tillage, harrowing, by spring tine harrow (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 5.04e-05\\ 
tillage, ploughing (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 5.04e-05\\ 
electricity, low voltage (kWh) [GLO] &  &  &  &  &  & \dependency &  &  &  & 0.00933\\ 
transport, freight, lorry, unspecified (metric ton*km) [GLO] & \dependency &  &  &  &  &  &  & \dependency &  & 0.404\\ 
green manure, organic, until March (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 5.04e-05\\ 
building, multi-storey (m3) [GLO] &  &  &  &  & \dependency & \dependency &  &  &  & 5.24e-06\\ 
liquid manure spreading, by vacuum tanker (m3) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.000565\\ 
application of plant protection product, by field sprayer (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.000222\\ 
transport, freight, sea, transoceanic ship (metric ton*km) [GLO] & \dependency &  &  &  &  &  &  & \dependency &  & 0.525\\ 
potato haulm cutting (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 5.04e-05\\ 
transport, tractor and trailer, agricultural (metric ton*km) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.00113\\ 
harvesting, by complete harvester, potatoes (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 5.04e-05\\ 
tillage, currying, by weeder (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.000101\\ 
transport, freight, inland waterways, barge (metric ton*km) [GLO] & \dependency &  &  &  &  &  &  & \dependency &  & 0.0857\\ 
copper oxide (kg) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.000107\\ 
tillage, harrowing, by rotary harrow (ha) [GLO] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 5.04e-05\\ 
irrigation (m3) [GLO] &  &  & \dependency & \dependency &  &  &  &  &  & 0.0173\\ 
\hline
Foreground Emissions \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & $\tilde{b_f}$\\ 
\hline
Input: Occupation, construction site [natural resource, land] &  &  &  &  & \dependency & \dependency &  &  &  & 2.1e-06\\ 
Input: Transformation, from unspecified [natural resource, land] &  &  &  &  & \dependency & \dependency &  &  &  & 1.05e-06\\ 
Input: Transformation, to industrial area [natural resource, land] &  &  &  &  & \dependency & \dependency &  &  &  & 1.05e-06\\ 
Input: Energy, gross calorific value, in biomass [natural resource, biotic] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  &  3.87\\ 
Output: Phosphate [water, ground-] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 6.25e-06\\ 
Output: Cadmium, ion [water, ground-] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 1.14e-09\\ 
Output: Zinc, ion [water, ground-] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 1.35e-06\\ 
Output: Nitrate [water, ground-] &  & \dependency & \dependency & \dependency &  &  & \dependency &  &  & 0.00696\\ 
$\ldots$ (31 rows omitted) &  &  &  &  &  &  &  &  & \\ 
\hline
\end{tabularx}
}

### 2017-03-31 manuscript/publication-abstract.tex
Move paper to git

\begin{abstract}
  Many of the challenges facing interpretation and knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of life cycle inventory models.  This article presents a framework for LCA study authors to publish inventory models in a structured format based on a mathematical formulation of a \emph{foreground study}.  Using linked semantic data as a foundation, the publication framework offers a mechanism to provide both machine readability and unambiguous human interpretation of published studies.  Implementing the framework requires only that the LCA community provides a way for study authors to make common reference to shared data sources, including background databases, elementary flows, and impact assessment methods, that are widely used in many studies.  The framework contributes to ongoing efforts within industrial ecology to improve the reproducibility and verifiability of scholarly works, and if implemented, plots a course toward distributed, platform-independent computation and validation of LCA results.
\end{abstract}

### 2017-03-31 manuscript/publication-acknowledgments.tex
Move paper to git

  \section*{Acknowledgments}

  This work draws on the efforts of the ``Inventory Model Description and Revision'' discussion group of SETAC North America, whose members include Ben Mourad Amor (University of Sherbrooke), Bill Bernstein (NIST), Paula Bernstein (PR\'e Sustainability), Marcos Esterman (Rochester Institute of Technology), David Evers (Hexion), Karl Haapala (Oregon State University), Troy Hawkins (Eastern Research Group), Wesley Ingwersen (US EPA), Christoph Koffler (Thinkstep), Lise Laurin (EarthShift Global), Antonino Marvuglia (Luxembourg Institute of Science and Technology), David Meyer (US EPA), KC Morris (NIST), Christopher Mutel (Paul Scherrer Institut), Tomas Navarrete (Luxembourg Institute of Science and Technology), Devarajan Ramanujan (Massachusetts Institute of Technology), Barclay Satterfield (Eastman Chemical), and the author.  This work was supported by the US National Science Foundation (CCF-1442966).

### 2017-03-31 manuscript/publication-background.tex
Move paper to git

\section{Supporting Data for Foreground Publication}

The previous section explained how foreground LCA models could be published precisely by authors in a manner that would allow them to be reproduced by readers or data users.  Reproducing study results, however, requires those data users to have access to background LCI data and characterization data as well.  This section discusses those aspects of LCA publishing.

\subsection{Background database publication and aggregation by reference}

The key characteristic of the foreground model is that it describes references to background data sets but explicitly excludes the background data sets themselves.  There are four main reasons for this exclusion:

\begin{enumerate}
\item License.  Most background LCI databases are proprietary (if still "public" under this paper's working definition), and licensing restrictions prohibit data users from republishing them.

\item Attribution.  Foreground data sets describe the system being directly modeled by the study author, and the author's report is thereby the authoritative source of information about them. In contrast, background data sets describe systems that have been modeled by a separate agency.  If the objective of a publication is to allow a reader to accurately interpret or reproduce the result, this aim is better accomplished by allowing the reader to obtain information from the authoritative source than to obtain it second-hand from the study author.

\item Size.  While foreground data sets tend to be sparse, because foreground processes tend to have a limited number of dependencies and emissions, background data sets are full.  Given a strongly connected background database, any emission that appears anywhere in the background will have a nonzero exchange value in every background LCI result.  Because of the scale of information, it is inconvenient to reproduce it all.

\item Invariance.  In current practice, background data sets are prepared infrequently and used in many applications.  The major inventory data providers, Thinkstep and Ecoinvent, generally update their databases approximately once per year.  Many databases do not receive regular updates at all.  It would be redundant and error-prone for authors to reproduce the data sets every time they are used.
\end{enumerate}

Accurate representation of background data sets is still required to meet the objectives of study publication, but for the above reasons it would be inconvenient or inappropriate for a study author to reproduce them in full.  The remedy proposed in this article is for study authors to simply reproduce background data sets \emph{by reference}, naming them with sufficient precision to allow a study reader to access them.  Taking this approach, however, requires that the study reader own a license for any proprietary data used.

One way around this problematic situation is for LCI data providers to offer ``aggregation by reference'' as a service to their licensees.  Under such a service, a study author could populate the entity map with linked references to background data sets, and data providers would allow readers of those studies to compute LCI or LCIA results for the linked data sets on demand.  This system offers considerable advantages over current practice:

\begin{itemize}
  \item Data providers would be situated as the sole authoritative sources for results derived from their data.  This has the dual effects of giving readers confidence in the correctness of the results and ensuring that the data providers continue to occupy an indispensible role in study preparation and review.

 \item Data providers would be able to directly control access to their intellectual property.  A given query for an aggregation result could be answered in different ways depending on the credentials of the user making the query.

 \item Data providers would be liberated from the need to provide database updates \emph{en bloc} to their users.  Individual data sets could be revised and updated internally, and results generated from queries from data users would always contain the latest input data.  Alternatively, studies that made reference to a specific data version could be answered using those specific versions rather than the latest data.

 \item Queries could be answered contextually, with responses varying depending on the geography of the reference, on intermediate flow characteristics, or on other ad hoc criteria.
\end{itemize}

\subsection{Characterization data and LCIA methodology publication}

One potentially significant source of irreproducibility in LCA is the redundant implementation of life cycle impact assessment methodologies by different software providers \citep{Speck_2015,Herrmann_2015}.  This has numerous drawbacks.  First, the reimplementation introduces the possibility for errors and discrepancies.  Second, the labor of revising and updating characterization factors is multiplied because every implementation must be updated separately.  Third, different implementations may be out of sync with each other.  Finally, the reimplementations may themselves not be easily reviewed when studies using them are reviewed.

These shortcomings can be addressed through structured publication of characterization factors using a linked data foundation like the one proposed in this article for inventory models.  Many LCIA methodology providers publish their methods on the Web in machine-readable formats such as spreadsheets, but these publications lack the reference to shared semantic data that would identify distinct flows in common with other implementations. If the information were provided as a ``characterization on demand'' service to LCA software providers, in which LCIA method providers published a list of references to elementary flows denoted with linked data URIs.  Software makers or their users could submit a query to the LCIA method provider for a characterization factor for a given flow, and the provider would respond with the requested information.

This approach would foster the development of software supporting reliable and consistent, platform-independent web-based LCA computations.  {\red The providers would have the ability to write software to automatically validate their in-house characterization factors and keep them up to date.  Study authors and reviewers would be able to use the service to validate structured inventory publications and verify published LCIA results.  
}

### 2017-03-31 manuscript/publication-components.tex
Move paper to git

\subsection{Components of a Study}

The formulation shown in Figure~\ref{fig:foreground} shows four distinct data types in an LCA foreground study:

\begin{itemize}
\item \textbf{Foreground flows}: (Rows and columns of $A_f$; rows of $\tilde{\mathbf{y}}$; columns of $A_d$ or  $B_f$) The foreground flows show how the model is constructed.  The foreground itself is made up of a collection of nodes.  Each node represents both a process and that process's reference product (flow).  The node weight vector $\tilde{\mathbf{x}}$ indicates the activity levels of all foreground nodes within the system boundary resulting from a particular (or canonical) final demand.  It also expresses the magnitudes of product flows emanating from the foreground processes themselves.

\item \textbf{Background processes}: (rows of $A_d$ or $\tilde{\mathbf{a}}_d$; columns of $B_x$) The background processes show how the dependencies of the foreground on the industrial system are resolved to specific inventory data.

 \item \textbf{Elementary flows}: (columns of $\mathbf{e}$, rows of $B_f$ or $\tilde{\mathbf{b}}_f$; rows of $B_x$ or $\tilde{\mathbf{b}}_x$)  Elementary flows cross the nature-technosphere boundary and generate environmental or social impacts.

 \item \textbf{Characterization quantities}: (rows of $E$ or identity of $\mathbf{e}$)  These indicate the impact assessment measurements (or impact category indicators) reported in the study.
\end{itemize}

The collection of foreground nodes and their reference flows expresses a de facto definition of the scope of the model, which is enclosed by an implicit system boundary.  %Most published studies disclose at least the structure of their foreground models by providing a graphical system diagram in the LCA report.  These ``box-and-arrow'' diagrams can be interpreted as graphs and used to produce an adjacency matrix that shows the locations of non-zero elements of Af.  However, the model disclosed in the graph may be a simplified form of the actual model, and the adjacency matrix itself is not sufficient to reproduce the matrix unless the exchange values themselves are also explicitly reported.

The foreground matrices $A_f$, $A_d$, and $B_f$ together include all the study-specific information.  Assuming a data user had access to $B_x$ and $\mathbf{e}$, the three foreground matrices would allow a reader to reproduce the result.  If the foreground data were omitted, the result could be reproduced from the intermediate aggregation results: $\tilde{\mathbf{a}}_d$ and $\tilde{\mathbf{b}}_f$.

### 2017-03-31 manuscript/publication-conclusion.tex
Move paper to git

\section{Conclusion}

An LCA study, as formulated in Eq.~\ref{eqn:study}, can be published in a structured format by providing data users with sufficient information to unambiguously construct the foreground matrices $A_f$, $A_d$, and $B_f$, and also to clearly identify the meaning of the nonzero row and column entries in each matrix.  This can be accomplished by providing sparse, tabular data that make up the matrices, and an entity map which relates the rows and columns to references on the Web.  A study author can provide varying degrees of transparency in reporting their study, but the structured publication format ensures that a reader or data user can interpret the information in the way the author intended.  The entity map also serves as a formal statement of the functional scope of the study.

Before the goal of easily reproducible foreground models can be realized, the linked data foundation must be laid.   The four entity classes found in foreground studies, namely foreground flows, background processes, elementary flows, and characterization quantities, must shift from being recorded locally to being stored on the Semantic Web, where their shared meaning can be curated.  Fortunately, this is easy to do: authoritative sources simply need to publish hyperlinks to their content. Providers of aggregated inventory data should make available static hyperlinks to their content, so that licensees and data users can make shared reference to the same entities, even if the data may not be accessible to the general public. Thinkstep already does this: every process inventory they publish has a stable link on the Web\footnote{\href{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/f8dddbcb-f08c-4217-99de-0b59d2c11bb1.xml}{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/[UUID].xml} where [UUID] is replaced with the process's unique identifier. For example, \texttt{f8dddbcb-f08c-4217-99de-0b59d2c11bb1} corresponds to production of 1,3-Butadiene, via catalytic dehydrogenation of butane, in Germany in 2015.}.  Similarly, impact assessment providers should publish their characterization data in a way that allows for automatic query and validation.  Both of these tasks can be accomplished independently through the efforts of members of the LCA community; however, the accuracy and validity of the data would be improved if the work were done by the data providers themselves.

Foreground studies are distinct from process inventory data sets, about which there is much ongoing work to establish file formats and standards.  To avoid the need for a similar enterprise for foreground publications, a minimal, inclusive standardization of information should be sought that makes the fewest possible prescriptions on its content.  The publication model presented here is meant to identify the minimal requirements that a study can be both reproduced and interpreted accurately: an \emph{entity map}, a list of references to entities used in the study; and \emph{tabular data} to fill in the model in Eq.~\ref{eqn:study}.  From such a simple prescription, a framework for distributed, platform-independent LCA computation can be imagined.

### 2017-03-31 manuscript/publication-discussion.tex
Move paper to git

\section{Discussion}

An LCA foreground study can be published concisely by distinguishing foreground components from background components, explicitly identifying all entities involved in the study, and then reporting the data values that make up the submatrices in Eq.~\ref{eqn:study}.  As the examples demonstrate, a structured publication can be created that allows a reader to critically review, validate, and accurately reproduce the product system model without requiring additional information.  The entity map provides the foundation for a linked-data publication by explicitly naming all externally-referenced entities involved in the foreground model.  Ultimately it should be possible to name entities using linked data identifiers or URIs, allowing unambiguous, automatic retrieval of metadata about the entities.

The input matrices ($A_f$, $A_d$, or $B_f$) or aggregation results ($\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, $\tilde{\mathbf{b}}_f$), together with the entity map, form a research object that can be easily generated and shared with colleagues.  By separating foreground and background, the research object satisfies the objective for attribution, because only the information included in the research object is attributable to the study author; whereas information that came from the background database is excluded from the publication.

This framework allows an author a great deal of flexibility in publishing their model.  Any of the components or intermediate results visualized in Figure~\ref{fig:foreground} may be selected for publication at any level of detail.  By publishing a set of linked data identifiers corresponding to the rows and entries of the matrices, the author ensures that the data user will be able to identify the foreground and determine the background processes used. Again because the background is excluded, the size of the research object is kept down, and licensing limitations associated with the distribution of proprietary data are avoided.

\subsection{Reproducing LCIA Results}

As shown in Eq.~\ref{eqn:lci}, the LCIA results for a foreground system are the sum of the foreground and background scores.  The foreground scores are computed from the foreground emissions, while the background scores are computed from LCI results included in the background database.  These two components require distinct approaches for study validation.

For foreground scores, the LCIA results depend on the information provided in $B_f$ combined with the characterization factors for the LCIA methods selected.  Because LCIA methods are maintained independently from LCI databases and LCA software providers, the redundant implementation of these methods by different providers and/or software makers is a significant source of irreproducibility \citep{Speck_2015,Herrmann_2015}.  In the context of a structured publication, the most transparent approach is for the publication to include the characterization factors used in the study, which allows a reader to inspect them and check the results.

For background scores, the results cannot be validated without access to the background database.  Again, the characterization factors used would need to be disclosed along with the background LCI results.  Because background databases tend to be updated infrequently, it is important for a background dependency reference to include information about which version was used.  Assuming datasets within a particular background database do not change, it would be possible to create an archive of unit impact scores for each background process and LCIA method supported by the database.  These values could then be independently validated.

In the context of structured publication, a reader of a structured publication could obtain the unit LCIA scores for each background dataset referenced, and use them to validate the background LCIA result. If the background database is subject to licensing restrictions, it may be necessary to require readers or data users to independently obtain background LCIA scores.  However, it may also be possible for the author to publish unit impact scores for background processes used in the model without violating licensing terms. In this case, it is much easier to reproduce published results, though the reader may still wish to check the author's published unit scores for accuracy.  If unit impact scores for background processes are included in the publication, then it is a simple matter for a reader to perform a contribution analysis and inspect the sensitivity of the results to foreground parameters.  This is exemplified in the ``organic potato'' publication provided in the supplementary materials.

\subsection{Avoiding Redundancy in LCA Computation}

Considering the discussion of LCIA score reproducibility, it is evident that relying on study authors to individually perform flow characterization, background LCI computation, and background LCIA computation generates a tremendous duplication of effort in the LCA community.  Although background LCI and LCIA results are static as long as the background database is not revised, the conventional approach to LCA requires study authors to individually re-compute background inventory results and LCIA scores for every study.  This has weighty implications:

\begin{itemize}
\item If the database is provided in unit process form, then the practitioner must have access to the entire database in order to compute any LCI result.  This means that every licensee must download the full database upon purchasing a license.
\item Different LCA software providers must independently ensure that their software can support every LCI database their users may wish to use.  This %leads to redundancy of implementation at the inventory level, which
  causes redundancy and can introduce errors and inconsistencies.
\item Database updates become inconvenient, because every update requires a complete replacement of the database. This inhibits the deployment of incremental updates, even when errors are found.
\item Because of the increased burden of data updates, software providers may be slow to apply updates from LCI providers.  Additionally, it is more likely that users will postpone updating their software, leading to the possibility of version mismatches between users or software systems.
\item Software providers must also independently implement LCIA methods, which introduces the same possibilities of errors, inconsistencies, and version mismatches.
\end{itemize}

In contrast, a computational approach based on foreground study publication separates the concerns of study authorship from software and database maintenance.  In principle, the product system models presented as examples in this paper could remain unchanged even if background databases and LCIA methodologies were completely revised.  A study publication could be easily updated simply by changing which entities are pointed to by the entity map.  This exercise is well defined and easily reviewed.  It would also allow for a straightforward comparison of the changes to study results after a database update, a task that is currently challenging to perform.

A further implication of the linked data approach suggested here is a shift away from stand-alone computation of background data results, and toward a system where authoritative data providers compute aggregation and characterization results as a service for their users.  In such a system, data users would supply a query in the form of a particular reference to a background process or elementary flow, and service providers would answer the query with LCI results, LCIA results, or characterization factors.  Answers could be provided on-demand to study authors, readers, reviewers, or downstream data users.

This approach, which could be termed ``aggregation by reference'' and ``characterization by reference,'' could dramatically reduce the complexity of authoring LCA studies, in principle allowing authors to construct whole product system models without procuring any data.  Before computation of results could be performed, authors would have to obtain access to the desired datasets.  It would also situate the data providers as the sole authoritative sources for results derived from their data.  This has the dual effects of giving readers increased confidence in the correctness of the results, and in ensuring that the data providers continue to occupy an indispensible role in study preparation and review.

Finally, it would allow data providers to directly control access to their intellectual property.  A given query for an aggregation result could be answered in different ways depending on the credentials of the user making the query, and depending on the context of the query.  For example, an authorized critical reviewer of a published study may be given free access to the requested dataset, whereas a member of the general public would be given only the impact score.

\endinput

%% First, the reimplementation introduces the possibility for errors and discrepancies.  Second, the labor of revising and updating characterization factors is multiplied because every implementation must be updated separately.  Third, different implementations may be out of sync with each other.  Finally, the reimplementations may themselves not be easily reviewed when studies using them are reviewed.

If a publication includes LCIA results, then a reader with access to the same background database used by the author can validate the scores by computing the unit impact scores for background datasets referenced in the publication.  It may also be possible for the author to publish unit impact scores for background processes without violating licensing terms, in which case the reader can check the scores for accuracy.

By selecting which input matrices , $\tilde{\mathbf{b}}_x$, or $\tilde{\mathbf{b}}$, as well as the LCIA result $\tilde{s}$) are published, the author can determine the level of transparency of the publication.  By reporting the matrices or vectors in sparse form as delimited files, the author can enable a data user to easily reconstruct the model, and if applicable, reproduce the result.

\endinput

Points to cover here:

* attribution-- foreground contains only data attributable to study author
* isolation of background-- presents opportunities to background data providers (licensing / agg on demand)
* characterization section can stay

* handling allocation?

- need to re-read after examples are done

The publication then becomes the authoritative statement of that assertion.

Particularly in LCA, licensing restrictions may prohibit the inclusion of data from background databases in a study.

A research object must contain all the information necessary to accomplish its goal.  However, the object must also be of manageable size and should exclude information that is not required or that can be found elsewhere.

The separation of foreground elements from the background database(s) achieves the attribution objective because it distinguishes the study author's contribution from data from other sources.

This model is not useful for publishing background systems (i.e. where the background depends on the output of the product system being modeled)

### 2017-03-31 manuscript/publication-example.tex
Move paper to git

\section{Examples}

The product system models in most LCAs can be described as foreground studies (Eq.~\ref{eqn:foreground}).  In fact, LCI databases contain product systems that can also be modeled as foreground studies because they describe products that are not required by the background.  In this section, two product systems selected from LCI databases are used to illustrate the concept of structured publication.  The systems selected because are complex enough to illustrate the premise but simple enough to review easily.

Each system is illustrated as a table that shows the foreground model, cutoff flows, background dependencies and foreground emissions included in the system.  The table omits numeric data from most of the table for clarity. Instead, a black square indicates the presence of a nonzero value.  Aggregation results $\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, and $\tilde{\mathbf{b}}_f$ are also reported.  The systems are also reproduced as structured publications in the supplementary materials, formatted as Excel spreadsheets.

\subsection{US LCI: Secondary Aluminum}

The US LCI database contains a small background system of 39 processes, as well as 395 foreground product systems that range in size from one to 83 foreground flows.
%The largest, ``paper, freesheet, coated, at mill'' makes reference to 34 of the 39 background processes.

The example product system  reports the production of secondary aluminum from automotive scrap (Figure~\ref{table:aluminum}).   The aluminum production process has direct requirements for two other foreground systems, including transportation services and quicklime production.  The quicklime production itself depends on limestone extraction, which is also part of the foreground.  The aluminum production requires two background systems, electricity and natural gas combustion, and reports 10 cutoff flows.  The main input to the process, ``Aluminum scrap, automotive,'' appears as a cutoff flow.  Overall, the four foreground processes require the background for only combustion (five fuels), transport (three modes) and grid electricity.  The same natural gas combustion and electricity models were used in all three non-transport foreground nodes.

The system also reports several direct emissions, mainly from the aluminum process.  Only one resource consumption (input flow) is reported -- the extraction of limestone. The direct emission modeling is limited in scope and includes a number of ``unspecified'' flows that may not be well characterized in many impact methods.

The structured publication of the secondary aluminum model includes a total of 47 data points about 36 entities.  The $A_f$, $A_d$, and $B_f$ matrices are reported in sparse format, along with the aggregation vectors. In the publication, LCIA scores are reported for several impact indicators drawn from the Ecoinvent reference LCIA implementation.\footnote{The Ecoinvent LCIA factors are not published, but are available from the Ecoinvent Centre. The name of the file used in the current study is \texttt{LCIA implementation v3.1 2014\_08\_13.xlsx}.}    The $E$ matrix is also partially reproduced: entries are only reported if they match flows that appear in the foreground.

\input{fig_uslci}

\subsection{Ecoinvent: Organic Potatoes}

\input{fig_ecoinvent}

Ecoinvent version 3 is provided in three different system models that reflect different linking strategies.  The example is drawn from the ``Allocation: Point of substitution'' or APOS model, which includes 11,420 processes that produce 12,966 product flows.  Of these, 10,282 are background flows and the rest are foreground flows.

The example system reports production of organic potatoes supplied to the global market (Table \ref{table:potato}).  The foreground includes nine nodes, of which six (nodes 3 through 8) form a strongly connected component associated with the production of potato seeds.  The Ecoinvent database's use of ``markets'' as mixer processes is evident in the table: node 0 is a mixer process that combines Swiss (``CH'') production (node 1) weighted at 2.4\%  with rest-of-World (``RoW'') production (node 2) weighted at 97.6\%.  That same market split (2.4 / 97.6) can also be seen in nodes 7 and 8.

Nodes 4 and 5, which make seeds ready ``for setting,'' are mixed by node 8. Although nodes 4 and 5 are geographically distinguished (CH vs RoW), both can be seen to consume potato seed from the global market.

Looking at the dependency and emission lists, the different ``signatures'' of different kinds of processes can be seen: nodes 1, 2, 3, and 6 are clearly agricultural processes that require irrigation, tillage, manure and so on.  Nodes 0 and 7 are visible as market processes, their only requirements being transport processes.  Nodes 4 and 5 each consume electricity and require use of a ``multi-storey building.''  Taken together, the CH-locality processes appear to use CH irrigation and electricity supply but are otherwise similar to their RoW counterparts.  The product model includes no cutoff flows.

The structured publication of the organic potato model reports the $A_f$, $A_d$, and $B_f$ matrices in full rather than sparse format to enable a more accessible visual review.  In addition, the $E$ matrix is reported in full, for all emissions found in the model, rather than only the emissions found in the foreground.  This enables the publication to be used to visually evaluate different implementations of the same impact assessment method in the context of the inventory model (Ecoinvent vs ILCD implementations of ReCiPe Midpoint marine eutrophication; see supplementary materials for details).

### 2017-03-31 manuscript/publication-foreground.tex
Move paper to git

\subsection{The Foreground}

\input{fig_fragments}

Generally, any directed graph can be represented in $A_f$, but some fundamental designs are common.  Some simple foreground models are shown in Figure~\ref{fig:fragments}.  The first (a) is a sequential model, in which each node requires one foreground input and generates one output. This model is equivalent to a ``gate to gate'' model.  Here the weights $k_i$ indicate the amount of the preceding reference flow that is required by the subsequent node.  Figure~\ref{fig:fragments}(b) shows an additive model, in which the outputs of several foreground nodes are added together, equivalent to a ``mixer'' or a horizontal average.  In this arrangement the weights %represent the relative weights of each input, and
should add up to a unit output of the reference node.

\subsection{Model fragments and foreground inputs and outputs}

Conventionally, the foreground flows are entirely inside the system boundary: they are produced by one foreground node and consumed by one or more others.  In the conventional case, the only product flow which crosses the implicit system boundary is the reference flow, which provides the functional unit of the study given in $\tilde{\mathbf{y}}$.  By implication, the difference $\tilde{\mathbf{x}} - \tilde{\mathbf{y}}$ represents the magnitudes of interior flows within the foreground.

However, these flows may not all necessarily be interior.  Most complex life cycle models involve nested model designs, in which low-level model components are encapsulated within higher-level models.  A foreground matrix that describes an incomplete life cycle model could be called a model ``fragment.''  In model fragments, some foreground flows represent product flows consumed within the fragment but not produced there, or vice versa. These flows represent inputs and outputs to the foreground model that are not further exchanged among the foregrond nodes, but neither are they terminated to a background process or emitted into the environment.  Thus input/output flows can be detected by identifying foreground flows that show up as row entries but have no column entries.

The input/output flows provide a mechanism to link separate foreground fragments together.  Two foreground models that both specify foreground flows with the same identity can be linked together simply by combining their entries in a common set of $A_f$, $A_d$, and $B_f$ matrices.  Similarly, large foregrounds can be easily partitioned into fragments simply by selectively including or excluding columns of $A_f$, $A_d$, and $B_f$ in different publications.
In isolation from other foreground models, input/output flows may be considered to be ``cutoff flows'' that must be terminated by a data user.

An example of foreground composed of fragments is illustrated in Figure 3(c).  Here, the nodes labeled 1--5 represent one fragment, which generates the foreground's canonical reference flow $\tilde{y}$. This fragment requires two interior flows from separate fragments ($y_0$ and $y_1$), and has two cutoff flows (4 and 5).  The reference $y_0$ is supplied by a second fragment, made up of nodes 6-8.  The reference $y_1$ is supplied by another fragment made of only one node (9). The reference flow $y_1$ is consumed in two different places by the other fragments.

\subsection{Foreground Dependencies and Foreground Emissions}

While the $A_f$ matrix establishes exchanges of foreground flows, these exhanges have no environmental impacts and only serve to compute the activity levels of the foreground nodes themselves.  The quantifiable impacts from these nodes arise either from their dependency on background processes or their direct environmental emissions.

In this context, background processes include any activities that are not modeled explicitly by the model author, but are drawn from an external data source.  These generally include flows such as the consumption of grid electricity or fuels, the procurement of raw materials, the use of transportation or waste disposal services, and other points of reliance on the industrial ecosystem.  The contents of this matrix are often kept secret because they can reveal sensitive information about the foreground processes being modeled.  While the model depicted in Figure 2 shows reliance on a single background LCI database, foreground nodes could easily draw on processes that belong to multiple data sources.

Foreground emissions include direct emissions from foreground processes that are not already included in background models.  Examples include process emissions from vents or into waterways or soil, explicitly modeled foreground combustion processes, or use phase emissions.  It is tempting to think of the $B_f$ matrix as including all direct foreground emissions.  This is inaccurate because many direct-emitting  processes, such as fuel combustion, are in fact modeled in the background and only referenced by the foreground, via $A_d$.  Many studies could have a zero $B_f$ matrix.

\subsection{Publishing the Foreground Model}

A structured publication could include any of these elements, bundled into a research object that meets the objectives laid out above.  This research object must include:
\begin{itemize}
\item \emph{tabular data} which enables the data user to construct the matrices and perform the computation;
\item an \emph{entity map}, which allows for precise interpretation of the elements of the study.
\end{itemize}
These two components are discussed below.

%%Moreover,

%%%% Two types of information visible in Fig. 2 are of interest to publish: input data, which are the matrices used for the computation, and aggregation results, which are the vectors with tildes.  Given

%%%% \subsubsection{Characterizing The Foreground}

%%% The $A_f$ matrix reports the interconnections (exchanges) between the nodes in a foreground model or portion of a model.  Each entry $a_{ij}$ in the matrix represents the exchange of a flow between foreground nodes $i$ and $j$.  The $i$th column describes the characteristics of the process occurring at the $i$th node.  Similarly, the $i$th row represents the demand (or alternative supply) for the $i$th reference product.  Thus an entry in position $a_{ij}$ indicates that the process in column $j$ requires $a_{ij}$ amount of the reference product $i$, per unit of reference product $j$ produced.  When the foreground technology matrix $I-A_f$ is constructed and inverted, the $i$th column of the inverse reports the activity levels of the foreground nodes that result from a unit demand for the $i$th node.

%%%% This node weight vector is used to compute the aggregate levels of dependencies and background emissions generated by the foreground.  However, i

\subsubsection{Tabular Data}

The objective for reproducibility of computation can be achieved with reference to the formulation in Equations~\ref{eqn:study}--\ref{eqn:lci}.  The publication contents can include any of the intermediate results, or all of them from end to end.  These components are all vectors or matrices, and it is easy to produce a machine-readable rendition of a vector or matrix.  Of innumerable possible formats, two broad subtypes deserve explicit mention: ``full'' representation and ``sparse'' representation.  A full representation of a matrix is a literal grid of numbers, such as a spreadsheet.  The spreadsheet can be equivalently represented as a delimited file, where each row is a single line of the file (e.g. separated by newlines), and the cells within the row are separated by a delimiter such as a comma.

The sparse representation assumes the matrix is largely zeros, and only reports the non-zero elements.  A sparse representation could also take the form of a delimited file, except each line would contain exactly three entries: the row index, column index, and value of the non-zero matrix entry.  Given that the matrices being represented are largely made up of zeros, the sparse approach is preferable to the full one.  The exception is $B_x$ or any column thereof, which is likely to contain no non-zero entries.  Publication of $B_x$ is a special case that will be discussed in a later section.

The data tabulation described here is, ultimately, optional: for reasons of privacy, an author may wish to conceal some or all data points, or include the data points but exclude the actual data values.  This is allowable if it is consistent with the author's goal, and it will accordingly limit the ability of a data user to reproduce the model.

\subsubsection{The Entity Map}

The matrix formulation also provides a clear statement of the components that make up the LCA study.  To be specific, the rows and columns of each vector or matrix correspond to the entities in the model.  The publication should include a ``mapping'' in which every row and column that contains a nonzero matrix entry is linked to an externally published reference that describes what the row or column represents.  This mapping from mathematical element to semantic meaning forms the ``linked data foundation'' of the publication.

%%% The objective for a ``linked data foundation'' can be achieved simply by naming every row and column that contains a nonzero matrix entry, and linking it to an externally published reference.  Doing so provides a ``mapping'' from mathematical element to semantic meaning that
%%%  list of foreground flows, background processes, elementary flows, and characterization quantities

The entity map must provide some minimal information about the entities in each of the four groups discussed above:
\begin{itemize}
\item \textbf{Foreground flows}: Because the foreground is specific to the model, the publication itself \emph{is} the basis for the semantic meaning of the foreground flows. Authors should simpy provide enough detail that the significance of each flow is clear to a data user.  Each foreground flow must specify its \emph{reference unit}, which is the amount of the flow that corresponds to a unit activity of the foreground node.  The data values reported elsewhere in the column must be normalized to this unit.
    % and does not necessarily relate to any other publication, there are no restrictions on how the individual flows should be named. T
    %It is also not necessary for the foreground flows to be mapped to anything outside the model. Rather,
\item \textbf{Background processes}: A properly identified background data set should correspond to a distinct aggregated unit process dataset from a background data provider.  The reference units of the background processes should already be part of the background data publication and do not need to be reproduced in the foreground publication except for clarity.  
\item \textbf{Elementary flows}: The best practices for identifying elementary flows are a matter of ongoing discussion and research.  At minimum, each elementary flow should be identified in terms of (1) its chemical composition, using a reference nomenclature, CAS number, chemical formula, or other concrete identifier; and (2) the environmental compartment with which it is being exchanged.  Geographic locale may also be a part of an elementary flow specification.  However, if the flow originates in the foreground, the locale may be inferred from the locale of the originating foreground node.
  \item \textbf{Characterization quantities}: The LCIA characterization method is already required to be unambiguously reported in current LCA practice.  If a structured publication includes results, it should also include a link to an authoritative reference for the method(s) computed.  
\end{itemize}

The entity map should include all of the foreground flows, background processes, elementary flows, and characterization quantities referred to in the publication.  Taken together, the entity map provides a working formal definition of the scope of the model.  Even in the absence of the numerical data that make up the input matrices or aggregation results, this \emph{functional scope} can be used as a basis to describe the contents of the model and formally state the system boundary of the foreground.

The entity map also provides a template for empirically-generated LCA studies: once the flows are enumerated, an observer can simply report the flow magnitudes observed over a reference period, and use it to construct an LCA study.

### 2017-03-31 manuscript/publication-forms.tex
Move paper to git

\section{Aggregation of Foreground Models}

\input{fig_aggregation}

This section describes what data points must be reported in order to generate a structured publication of a foreground model in various stages of aggregation.  In each case, full publication must include the two components already introduced: the entity map, used to and enumerate the matrix rows and columns; and the tabular data, used to populate the matrices.  For complete fidelity to the model, each non-zero entry in any matrix should be represented in the tabular data.   If results are included, the transparency of the publication can be improved by reporting the characterization vector as well.  Note that the reference flows of each foreground node are implicit in Eq.~6 and should not be replicated in the tabular data. Different aggregation forms are illustrated in Figure~\ref{fig:aggregation}.

\subsection{Unit Process Inventory}

A unit process makes up a single shared column of $A_f$, $A_d$, and $B_f$, as depicted in Figure~\ref{fig:aggregation}(a).  Publication formats for unit processes already exist, the most well-known being the ILCD and Ecospold XML formats.  Current LCA software is based on the principle of the unit process as the sole consensus format for structured publication, although current formats do not uniformly distinguish between foreground flows, dependencies, and emissions.  The entity map must include at least one foreground flow, which is the reference flow of the process, and corresponds to the column of $A_f$ to which it belongs.  Multi-functional processes can also be expressed, but a reference exchange must still be selected before the process can be entered into a foreground matrix.  Other co-products should in this case be listed as additional foreground flows in the entity map, and their exchange values included in the tabulated data; absent further specification, these flows would be interpreted as cutoff flows.

\subsection{Foreground}

See Figure~\ref{fig:aggregation}(b).  A foreground can be viewed as a linked set of unit processes.  In the mode of structured publication as a research object, the entries in the unit process inventories must be segregated into foreground exchanges (entries in $A_f$), background dependencies (entries in $A_d$), and emissions (entries in $B_f$).  A foreground or foreground fragment comprises these three complete matrices, all having a common set of columns.  A  publication in this vein would allow a data user to completely reproduce the study fragment.  A foreground fragment is also the smallest publication unit that can fully express an allocation treatment, because such treatment requires the involvement of multiple unit processes, either for the purposes of partitioning, co-product substitution, or system expansion.  The two examples presented above show foreground publications.

\subsection{Aggregated Foreground}

See Figure~\ref{fig:aggregation}(c). Any collection of foreground nodes can also be expressed in aggregated form by performing a row-wise weighted sum of the foreground columns.  This is equivalent to multiplying the $A_d$ and $B_f$ matrices by $\tilde{\mathbf{x}}$ to yield $\tilde{\mathbf{a}}_d$ and $\tilde{\mathbf{b}}_f$.
The foreground vector $\tilde{\mathbf{a}}_f$ is obtained by selecting all the elements of $\tilde{\mathbf{x}} - \tilde{\mathbf{y}}$ that represent input, output, or cutoff flows.  A publication of the aggregated foreground is identical to the publication of a unit process, except that the result was derived through aggregation.  As in the unit process case, the reference flow should appear in the entity map but the reference output should not be reported in the tabular data.

\subsection{Partial Background Aggregation}

See Figure~\ref{fig:aggregation}(d). This approach is useful when an author wishes not to disclose certain dependency relationships in the foreground but still wants to publish a complete, reproducible model.  The aggregated dependency vector is split into two parts that sum to the original:
\begin{equation}
 \tilde{\mathbf{a}}_d = \tilde{\mathbf{a}}_{d,priv} + \tilde{\mathbf{a}}_d'
\end{equation}
The disclosed dependencies $\tilde{\mathbf{a}}_d'$ are reported, and the private dependencies $\tilde{\mathbf{a}}_{d,priv}$ are replaced with a private impact score $\tilde{s}_{x,priv}$, computed from corresponding life cycle inventories from the background database.  Foreground emissions can still be reported separately, or they can be characterized and combined with the private score as appropriate.  This approach allows the author to limit the disclosure of confidential information; however, the reader will still learn how large of an impact can be attributed to the undisclosed portion.  It isimpossible to discern the makeup of the private dependency vector from the private impact score.

\subsection{Full Background Aggregation}

See Figure~\ref{fig:aggregation}(e). In this approach, all dependency information is concealed through replacement with an impact score derived from the background life cycle inventory.  In this publication, foreground emissions are still reported separately, along with any foreground inputs, outputs, and cutoffs.  For a marginal increase in transparency, the full background inventory can be reported as $\tilde{\mathbf{b}}_x$.  However, taking this approach significantly increases the size of the publication because the background inventory vector is full (not sparse).  Full background aggregation is most commonly practiced

\subsection{Full LCI}

See Figure~\ref{fig:aggregation}(f). Here, all foreground information is concealed and only the aggregated life cycle inventory vector $\tilde{\mathbf{b}}$ is reported.  The life cycle inventory $\tilde{\mathbf{b}}$ provides the most aggregated form of the study that can still be independently validated with an external characterization vector $\mathbf{e}$.

### 2017-03-31 manuscript/publication-intro.tex
Move paper to git

\section{Introduction}

Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver a quantitative measure of utility to a consumer.  The technique is well-established, widely practiced, and draws on an extensive body of standardization and scholarship.
%Performing an LCA first requires preparing an inventory of the amounts of environmental interventions associated with the product, such as resources extracted or emissions released.  The inventory is the input to an impact assessment computation, in which the inventory items are scaled by their relative significance according to impact categories of interest, and summed together.  %% 55 words
Increasingly, the results of LCA studies are appearing in environmental declarations, corporate sustainability reports, and product marketing information.  LCA is also gaining prominence as a tool for developing and evaluating environmental policy.  
%
At the core of an LCA is a model of a ``product system,'' which represents a collection of ``processes $\ldots$ performing one or more defined functions'', that describes the life cycle of a product \citep{iso14044}.  Product system models encode information about how products are manufactured, distributed, used by the consumer, and what happens to them after they are disposed.  A product system can be divided into a foreground, which denotes the portions of the life cycle whose operations are directly modeled by the study, and a background, which represents the global industrial system \citep{SETAC_inventory_1998}.  While the foreground is often modeled from direct observation or other primary data, background processes are typically drawn from a life cycle inventory (LCI) database prepared by a third party.
%A relatively small number of comprehensive LCI databases exist, owing to their vast data requirements and the complexity of preparing and maintaining them \citep{UNEP_2011}.  The product system model itself is often constructed in specialized software, which is used to compute and analyze impact assessment results.  $$ 45 words

Although there has been considerable effort to normalize and harmonize %%important methodological decisions around %% -4 +1 
LCI database design methodology and data interchange \citep{UNEP_2011, JRC_ILCD_ELCD_2013, Mila_e_Canals_2015, Ingwersen_JLCA_2015}, less attention has been paid to harmonizing product system models.  These models are often highly complex and include countless modeling decisions, approximations, and assumptions made by a study author \citep{Lloyd2007, reap2008_I}.  Consequently, while comparative evaluations can be made regarding alternative cases or scenarios within a single study, comparisons across studies are much more challenging \citep{Heath2012, Henriksson2014}.  Even when relatively simple and homogeneous systems are considered, wide variations can be found in the results of studies from different authors \citep{van_der_Harst_2013, Turconi_2013}.
%Some types of product systems may require coordination on key methodological issues before results can be compared \citep{Ahlgren_2015}, while others face more fundamental challenges that arise from the complexity of the supply chain and the lack of suitable inventory or proxy data \citep{Bull2014}. 43 words
The lack of transparency in reporting,
particularly regarding definitions of study scope and system boundary,
is a crucial factor challenging the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.
%In particular, this lack of transparency may manifest as a imprecision in the definitions of study scope and system boundary \citep{Laurent_2014, Turconi_2013}. %% -23 + 9
One consequence
%% of the wide variation in model construction, as well as the lack of consistent transparency in reporting, %% -18
is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.

As LCA gains prominence, particularly in the policy realm, these problems become more acute.  If LCA is to be used in policy, it will necessarily have the effect of recommending certain product systems, technologies, or approaches at the expense of others; yet in that event it is critically important that there be a consensus among stakeholders that the quantitative results are credible and well-supported \citep{Rainville_2015, McManus_2015}.  As noted above, this level of consensus is hard to realize, in part because of the conflicting implications inherent in different analytic modes.
%It is well known that allocation decisions can have dispositive effects on the outcome of a study; yet no consensus exists for specifying allocation methods (e.g. \cite{Hanes_2015, Pelletier_2014}), much less evaluating their significance. %% 33 words
Other factors can also render study findings unreliable or highly contingent.  Plevin et al.\ famously observed in the strongest terms that careless reporting of attributional LCA study results can distort the significance of findings in a policy context \citep{Plevin_2013}, a shortcoming that can be found in other modes as well \citep{Brandao_2014}.  In the scope of Environmental Product Declarations (EPDs), studies within an industry are supposed to be rendered comparable by adhering to a common product category rule (PCR) \citep{Fet_2006}.  However, this prescription is insufficient to ensure the comparability of different declarations, even if they use the same PCR \citep{Modahl_2012}.  Moreover, PCRs are often themselves not unique and can include wide variations in scope and boundary requirements \citep{Subramanian_2012}, leading to ambiguity.  %%PCRs themselves tend, like EPDs and other LCA studies, to be written qualitatively, to be unstructured, and to be open to considerable interpretation

A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the product system model.  In the scope of academic research, the transparency of techniques and reproducibility of results are paramount concerns \citep{Mesirov_2010}.  But reproducibility of a computation requires a distinction between the computing algorithm and the input data \citep{Buckheit_1995, Fomel_2009}; contemporary LCA lacks this distinction.  While the ostensible input to an LCIA computation is a product system model, this input cannot be separated from the computational environment used to author it.  Instead, models are distilled into written documents describing LCA results, be they academic papers, ISO reports, PCRs or EPDs.  These are uniformly \textit{unstructured}, meaning their interpretation must be secured by a human reading text and graphical images such as system boundary diagrams, but cannot be expressed to a machine.

The objective of \textit{structured} publication, on the other hand, is to enable the reader of a publication to automatically interpret its contents and, ideally, reproduce its results.  Such a publication is both machine-readable and easily human-interpretable.  In this paper, I propose a framework for the structured publication of product system models in LCA that achieves these dual requirements.  %% \textit{Publication}, for the purpose of this article, means disclosing sufficient detail to allow a reader to reproduce an LCA result, or all or part of the model that produced it.  I begin with a set of objectives for structured publication of a study model. Then I show how a simple analytic formulation of study structure can meet these objectives.   I provide examples of structured formulations of studies drawn from the US LCI and Ecoinvent databases.  %% 75 words

### 2017-03-31 manuscript/publication-method.tex
Move paper to git

\subsection{Mathematical Formulation of an LCA study}

At the heart of this article is a precise formulation of an LCA \emph{foreground study}, as distinct from (though likely dependent on) an LCI database, hybrid or input-output database, or software system.  The study formulation relies on the observation that LCA data sets can be ordered based on their dependency on one another \citep{Kuczenski_JLCA_2015}.  Background databases generally contain cyclic dependencies, and so it is not useful to publish unit-level descriptions of background processes unless they are all published as a group.
In contrast, foreground systems often %can be found to
depend on background systems, while the background systems do not %show a reciprocal dependency
depend on the foreground.  Because of this it is possible to describe foreground systems precisely %how foreground systems depend on the background database
without including the background.

We begin with the classic formulation of an LCA problem \citep{Heijungs2002}:
\begin{equation}
s = \mathbf{e} \times B \times A^{-1} \times \mathbf{y}
\end{equation}
where $A$ is the technology matrix, $B$ is the environmental intervention or emission matrix, $\mathbf{e}$ is a vector of characterization factors, $\mathbf{y}$ is the externally specified final demand, and $s$ is the numerical impact score or category indicator.   For simplicity, we will consider the computation of a single impact result.  However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.

\input{fig_lca}
It can be shown that, under conditions typical for LCA databases, the technology matrix can be represented equivalently in the input-output format, in which the technology matrix is actually the combination of separate supply and use tables, each of which has been normalized to a unit of output \citep{Suh_JIE_2010, Pauliuk_2015_framework}:
\begin{equation}
s = \mathbf{e} \times B \times \left(I - A'\right)^{-1} \times \mathbf{y}
\label{eqn:leontief}
\end{equation}
Here $A'$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values) per unit of a process's reference output.  The conditions under which this system is solvable, particularly with respect to different strategies for handling co-production, have been discussed extensively \citep{Majeau_Bettez_2014}.  Eq.~\ref{eqn:leontief} is visualized in Figure~\ref{fig:lca}.

The direct requirements matrix also describes a directed graph, in which each column is a node, and nonzero entries of $A'$ define edges between nodes.  This graph can be partially ordered based on the directions of the edges, so the foreground can be identified as the set of processes earliest in the ordering, on which the background does not depend.  An appropriately ordered $A'$ matrix can be written in block-triangular form:
\begin{equation}
A' = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A*
  \end{array}
\right];\;\;\;  B' = \left[\begin{array}{cc} B_f & B*   \end{array}\right]
\label{eqn:foreground}
\end{equation}

\input{fig_foreground}
In this case, the submatrix $A_f$ represents the foreground; $A*$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.  The ordered $B$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B*$ which includes background emissions. The constitutive characteristic of this formulation is that the background system does not depend on the foreground, allowing for the submatrix in the top right corner of Eq.~\ref{eqn:foreground} to be zero.  As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.  The following formulation is equivalent to Eq.~\ref{eqn:foreground}:
\begin{equation}
 A' = \left[\begin{array}{cc} 
A_f &  0 \\
A_d & 0 
   \end{array}\right];\;\;\;  B = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
\end{equation}
where $B_x = B* \times (I - A*)^{-1}$ is a background LCI database, which can be prepared in advance by a background database maintainer, as is currently common practice.  In this case, Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:
\begin{equation}
s = \mathbf{e} \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \mathbf{y}_f
\label{eqn:study}
\end{equation}

where $\mathbf{y_f}$ represents final demand of the foreground processes only.  Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}.  This formulation shows that given access to the background database $B_x$, the study results can be fully reproduced using only the submatrices $A_f$, $A_d$, and $B_f$, as well as the characterization vector $\mathbf{e}$.  %The foreground, which represents the activities and processes that are directly within the scope of the study, can be fully defined by its graph topology ($A_f$); its dependency on the background ($A_d$); and its direct emissions ($B_f$).

The foreground graph $A_f$ can be distilled into a vector of foreground activity levels for the foreground nodes, which determine the magnitude of dependencies and emissions for a particular output, case, or scenario.  With the functional unit for one study formulation given as $\tilde{\mathbf{y}}$, the foreground activity levels can be given as:
\begin{equation}
\tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
\label{eqn:inv}
\end{equation}
For a typical foreground study, a ``canonical'' $\tilde{\mathbf{y}}$ may be suggested in which the first foreground node produces the functional unit as output:
\begin{equation}
\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}
\end{equation}
Although the variable $\tilde{\mathbf{x}}$ can be computed through matrix inversion as expressed in Eq.~\ref{eqn:inv}, often matrix inversion is not required.  If the foreground does not include any feedback loops, then the activity levels of foreground can be determined simply by traversing the graph \citep{Bapat_LAA_2013}, which can be performed in linear computational time with respect to the number of foreground links.  In the canonical case the vector $\tilde{\mathbf{x}}$ is the first column of the matrix $(I-A_f)^{-1}$  and can be computed by traversal of the foreground tree beginning on the first node.

The node weights $\tilde{\mathbf{x}}$ can be used to describe an \emph{aggregated foreground}, which generates the same results as the fully expanded foreground:
\begin{equation}
\begin{array}{rl}
    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
\end{array}
\label{eqn:agg}
\end{equation}

These vectors summarize the contents of the foreground without disclosing its detailed structure.  The aggregated dependency vector $\tilde{\mathbf{a}}_d$ has the same dimension as the background database, and the aggregated emission vector $\tilde{\mathbf{b}}_f $has the same dimension as the elementary flow matrix.  The dependency vector can be transformed into an elementary flow vector by multiplying by $B_x$.  The result of the study $\tilde{s}$ is the sum of foreground and background impact scores:

\begin{equation}
\begin{array}{rl}
   \tilde{\mathbf{b}} & = \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x \\
   \tilde{s} &= \mathbf{e} \times \tilde{\mathbf{b}} \\
   & = \mathbf{e} \times \tilde{\mathbf{b}}_f + \mathbf{e} \times \tilde{\mathbf{b}}_x \\
   &= \tilde{s}_f + \tilde{s}_x
\end{array}
\label{eqn:lci}
\end{equation}

### 2017-03-31 manuscript/publication-models.tex
Move paper to git

\section{Constructing Foreground Publications}

According to the definition in the previous section, the foreground of an LCA study is defined through its relationship with the background: the foreground depends on the background, but the background does not depend on the foreground.  In this section, I assume a study background is drawn from one or several background databases, each of which have been independently published as a collection of aggregated LCI results for background processes.  I show how

A foreground can be thought of as the vertices of a directed graph.  Each node in the graph represents a process, and each edge in the graph represents a flow between processes.  When the

both a process and a flow of that process's reference product.  The Af matrix reports the interconnections (exchanges) between the nodes in a foreground model or portion of a model.  Each entry a_ij in the matrix represents the exchange of a flow between foreground nodes i and j.  The ith column describes the characteristics of the process occurring at the ith node.  Similarly, the ith row represents the demand (or alternative supply) for the ith reference product.  Thus an entry in position a_ij indicates that thee process in column j requires a_ij amount of the reference product i, per unit of reference product j produced.  When the foreground technology matrix I-Af is constructed and inverted, the ith column of the inverse reports the activity levels of the foreground nodes that result from a unit demand for the ith node.  For the canonical foreground, x~ is the first column of (I-Af)^{-1}.

### 2017-03-31 manuscript/publication-objective.tex
Move paper to git

\section{Requirements for Structured Publication of Study Models}

An LCA study result is ultimately an assertion that, for some input data known as the ``model,'' the delivery of a particuar reference flow %of products or services
is associated with a certain amount of environmental impact or impact potential.  Structured publication of a model by itself is simply documentation of the model's design.  On the other hand, structured publication of the model together with the result constitutes an assertion that the model and the result correspond.  The fundamental objective of structured publication is to enable a data user to validate that assertion.

As discussed in the introduction, a structured publication %of a study model
must achieve the dual requirements of human interpretability and machine readability.  %In the space of modern communications, t
This is often accomplished using linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.  The resource at the end of the link functions as a point of agreement: both study author and data user can follow it from anywhere on the Internet to obtain the same information.  The content pointed to by the hyperlink, along with context provided by \emph{other} references to the same resource, gives meaning to the information and allows it to be curated by the study authors and other members of the community \citep{Khan_2011}.  Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.  
Some preliminary descriptions of the entity types involved in LCA computation, and their relationships to one another, have recently been developed \citep{Ciroth_Srocka_2014, Janowicz_WOP_2015, Kuczenski_JCP_2016}.

However, the existence of semantic data resources is not enough by itself to establish the reproducibility of a model or result: it must also be clear how the information is used to generate the result.  A \emph{research object} \citep{Bechhofer_2013} is an aggregation of linked references, combined with a structure or system for converting the information into the scientific model for reproducing the result.  In order to develop a research object that can encapsulate the publication of an LCA study, it is necessary to identify the computation to be performed, and then supply the data necessary to perform it. The following objectives will be used to guide the this development:

\begin{enumerate}
\item \emph{Linked Data Foundation}.  Human interpretation of the publication should rely on the use of references to shared resources identified by URIs or hyperlinks.
\item \emph{Computation}.  The role of the publication in an LCA computation should be clear.  If the publication includes a result, it should be clear how the result was obtained.
\item \emph{Attribution}.  The publication should distinguish information attributable to the author from information attributable to other sources.
\item \emph{Completeness}.  The publication should contain all the parts of the model necessary to accomplish the goal.  
\item \emph{Conciseness}.  The publication should includes the smallest amount of information necessary to accomplish the goal.
\item \emph{Flexibility}.  The publication form should accommodate a variety of levels of disclosure and transparency.
\end{enumerate}

### 2017-03-31 manuscript/publication-si-recipe.tex
Move paper to git

\section{LCIA Case Study: Marine Eutrophication}

In the Organic Potatoes example presented in the text, the product system was characterized with respect to six different LCIA methods, all representing different implementations of the ReCiPe Marine Eutrophication midpoint indicator, measured in kg N-equivalents.  Ecoinvent's LCIA implementation is available by request from the Ecoinvent Centre; it has not changed since 2014. The ReCiPe method has been implemented according to the Individualist (I), Hierarchist (H), and Egalitarian (E) modes specified in the methodology; in addition, the (I) and (H) methods have been reimplemented in a way that omits long-term emissions.  The European Life Cycle Database's reference implementation of marine eutrophication was last updated in 2011.  The dataset cites the 2008 version of ReCiPe.

The tabular output of the structured publication provides a convenient method to compare the implementations. Inspection reveals that, as far as marine eutrophication  is concerned, there is no difference among the (I), (H), and (E) modes, which all have the same characterization factors.  Similarly, the (I) and (H) methods omitting long-term emissions are also the same.  Both sets of Ecoinvent indicators bear similarities and differences to the ILCD implementation, however.  The differences are shown in Table~\ref{table:recipe}.

Briefly, the implementations correspond on many significant flows but differ on several as well.
\begin{itemize}
\item While all methods characterize ammonia to air and ammonium ion to water, only ILCD characterizes ammonia to water and ammonium to air.
\item Where Ecoinvent characterizes ``nitrogen oxides'', ILCD separately characterizes nitrogen dioxide and nitrogen monoxide.  Since neither are present in the Ecoinvent data sets, the ILCD method will miss these flows entirely.
\item Ecoinvent includes characterizations for two different forms of elemental nitrogen, but ILCD is characterized with respect to a third flow, which is likely a synonym.  As a consequence, the ILCD implementation will not capture elemental nitrogen emissions from Ecoinvent datasets.
\item Cyanide emissions are only included in ``without long-term'', and do not appear in the ``long-term'' emissions nor in ILCD.
\item Nitrite is omitted from the ``long-term'' characterization but included in ``without long-term.''
\item The characterization for nitrogen oxides in ``without long-term'' is a factor of 10 smaller than it is in ``long term.''
\end{itemize}
This example demonstrates how independent reimplementation of LCIA methods can produce inaccurate and/or inconsistent results.

\begin{table}
  \footnotesize\sffamily
  \caption{ReCiPe Marine Eutrophication, Midpoint, kg N-equivalent}
  \input{lcia_table}
\end{table}

### 2017-03-31 manuscript/publication-si-spreadsheet.tex
Move paper to git

\section{Interpreting the Research Object}

The product systems are provided as ``research objects'' that take the form of Excel spreadsheets.  The spreadsheets are provided in the supplementary information accompanying the paper.  Each example is also illustrated with a table in the body of the paper.
The spreadsheets allow a data user to interpret, validate, and re-create the foreground study.  Each spreadsheet contains an entity map, a table of LCIA category scores, a set of data matrices, and a set of aggregation results.  The aggregation results can be computed from the data tables, and the LCIA results can be computed from the aggregation results.

\subsection{The Entity Map}

The ``EntityMap'' tab describes each entity in the model, broken into five sections: characterization quantities (for LCIA); foreground nodes; background dependencies; cutoff flows; and elementary flows.  Each entity is given an identifying key that is used to represent the entity in the data tables.  In addition to the key, each entity has four core properties:
\begin{itemize}
\item The \emph{origin} defines the data resource from which the entity was derived.  In an ideal case, the origin would be a URI of a namespace on the Semantic Web.  Since the research object is itself the data resource that defines the foreground, the origin for foreground nodes is simply ``Foreground.''
\item The \emph{identifier} is unique with respect to the origin.  The identifier should allow a data user to find the entity's dataset when consulting the origin's namespace. Ideally, the origin and the identifier together form a URI for the entity.   In principle a data user should be able to retrieve the correct entity using only the origin and the identifier fields.
\item Each entity has a \emph{reference unit} that defines how numeric values should be interpreted.
\item Each entity also has a \emph{name} for convenient interpretation of the entity.  
\end{itemize}

Each entity type (aside from characterization quantities) has additional descriptive fields:
\begin{itemize}
\item The \emph{flow direction} describes the direction of each flow with respect to the entity being defined.  Foreground flows, background dependencies, cutoff flows and elementary flows all have directions.
\item Foreground nodes also have a \emph{flow name} that reports the name of the reference flow emanating from the node.
\item Background dependencies also have a \emph{reference flow} that tells which of the process's reference outputs is being used.  In the case of multi-output background processes, this field is necessary to identify the appropriate allocation result.
  \item Cutoff and elementary flows also have a \emph{compartment} that describes which environmental compartment the flow is exchanged with. For cutoff flows, the compartment may be used to document some kind of classification of the flow.
\end{itemize}
Note that cutoff flows and elementary flows are both included in the foreground emission matrix, $B_f$, because they are mathematically equivalent.  In other words, ``emissions'' can be interpreted to include any flows that are exterior to the technology matrix, including resource extraction, environmental releases, and cutoffs.

\subsection{LCIA Scores}

The ``LciaScores'' tab has one data column for each LCIA method included in the entity map.  The rows of data include:
\begin{itemize}
\item \texttt{s\_tilde}, for $\tilde{s}$, the total LCIA result for the product system;
\item \texttt{sf\_tilde}, for $\tilde{s}_f = \mathbf{e}\times B_f \times \tilde{\mathbf{x}}$, the sum of characterized foreground emissions;
\item \texttt{sx\_tilde}, for $\tilde{s}_x = \mathbf{e}\times B_x \times A_d \times \tilde{\mathbf{x}}$, the sum of characterized background emissions;
  \item One row for each background dependency, reporting the unit impact score of the named dependency. These values can be validated by consulting the original LCI database provider.
\end{itemize}

\subsection{Data Sheets}

After the entity map and the LCIA results, the spreadsheet includes worksheets that report the contents of the $E$, $A_f$, $A_d$, and $B_f$ matrices, as well as the aggregation results $\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, and $\tilde{\mathbf{b}}_f$.  The aggregation results can be validated from the data tables using Eqs.~6 and~8 in the main paper.  The data sheets can be presented in sparse format, which contains three columns: a row key, a column key, and a data value, where the row and column keys each correspond to an entity on the entity map.  Alternately, the data sheets can be presented in full format, in which the top row contains a list of column keys, the first column contains a list of row keys, and the data region include nonzero values at the intersection of the corresponding row and column keys (zero valued cells are left blank.)

### 2017-03-31 manuscript/publication-conclusion.tex
Takeout thinkstep footnote

@@
 Providers of aggregated inventory data should make available static hyperlinks to their content, so that licensees and data users can make shared reference to the same entities, even if the data may not be accessible to the general public.
-Thinkstep already does this: every process inventory they publish has a stable link on the Web\footnote{\href{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/f8dddbcb-f08c-4217-99de-0b59d2c11bb1.xml}{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/[UUID].xml} where [UUID] is replaced with the process's unique identifier.
+%Thinkstep already does this: every process inventory they publish has a stable link on the Web\footnote{\href{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/f8dddbcb-f08c-4217-99de-0b59d2c11bb1.xml}{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/[UUID].xml} where [UUID] is replaced with the process's unique identifier.
 For example, \texttt{f8dddbcb-f08c-4217-99de-0b59d2c11bb1} corresponds to production of 1,3-Butadiene, via catalytic dehydrogenation of butane, in Germany in 2015.}.

### 2017-03-31 manuscript/publication-si-openaccess.tex
Jupyter notebook for USLCI

\section{Open Access to Paper Materials}
The software used to generate the examples published in this paper is available online.  Please consult the GitHub repository found at \url{http://github.com/bkuczenski/Publication-JIE}.  The python code used to produce the examples is found in the `python` directory, as a set of Jupyter notebooks.  Because the Ecoinvent database cannot be distributed due to licensing restrictions, only the US LCI example works ``out of the box.''

Please contact the author with any questions about the code, via the GitHub page or via email at \texttt{bkuczenski@ucsb.edu}.

### 2017-03-31 manuscript/publication-si-recipe.tex
tweaks

@@
 Ecoinvent's LCIA implementation is available by request from the Ecoinvent Centre; it has not changed since 2014.
-The ReCiPe method has been implemented according to the Individualist (I), Hierarchist (H), and Egalitarian (E) modes specified in the methodology; in addition, the (I) and (H) methods have been reimplemented in a way that omits long-term emissions.
+The ReCiPe method has been implemented according to the Individualist (I), Hierarchist (H), and Egalitarian (E) modes specified in the methodology; in addition, the (E) and (H) methods have been reimplemented in a way that omits long-term emissions.
 The European Life Cycle Database's reference implementation of marine eutrophication was last updated in 2011.
@@
 Inspection reveals that, as far as marine eutrophication  is concerned, there is no difference among the (I), (H), and (E) modes, which all have the same characterization factors.
-Similarly, the (I) and (H) methods omitting long-term emissions are also the same.
+Similarly, the (E) and (H) methods omitting long-term emissions are also the same.
 Both sets of Ecoinvent indicators bear similarities and differences to the ILCD implementation, however.

### 2017-04-01 manuscript/publication-si-recipe.tex
Final review of supporting information

@@
+\
 \section{LCIA Case Study: Marine Eutrophication}
+\renewcommand{\bottomfraction}{.8}
+\begin{table}[b]
+  \footnotesize\sffamily
+  \caption{ReCiPe Marine Eutrophication, Midpoint, kg N-equivalent}
+  \label{table:recipe}
+  \input{lcia_table}
+\end{table}
 
@@
 As a consequence, the ILCD implementation will not capture elemental nitrogen emissions from Ecoinvent datasets.
-\item Cyanide emissions are only included in ``without long-term'', and do not appear in the ``long-term'' emissions nor in ILCD.
-\item Nitrite is omitted from the ``long-term'' characterization but included in ``without long-term.''
+\item Cyanide and nitrite emissions are only included in ``without long-term'', and do not appear in the ``long-term'' emissions.  Cyanide emissions do not appear at all in ILCD.
 \item The characterization for nitrogen oxides in ``without long-term'' is a factor of 10 smaller than it is in ``long term.''
@@
 This example demonstrates how independent reimplementation of LCIA methods can produce inaccurate and/or inconsistent results.
-
-\begin{table}
-  \footnotesize\sffamily
-  \caption{ReCiPe Marine Eutrophication, Midpoint, kg N-equivalent}
-  \input{lcia_table}
-\end{table}

### 2017-04-01 manuscript/publication-si-spreadsheet.tex
Final review of supporting information

@@
 In the case of multi-output background processes, this field is necessary to identify the appropriate allocation result.
-  \item Cutoff and elementary flows also have a \emph{compartment} that describes which environmental compartment the flow is exchanged with.
-For cutoff flows, the compartment may be used to document some kind of classification of the flow.
+  \item Elementary flows also have a \emph{compartment} that describes which environmental compartment the flow is exchanged with. This is used to determine the correct characterization factor.
 \end{itemize}
 Note that cutoff flows and elementary flows are both included in the foreground emission matrix, $B_f$, because they are mathematically equivalent.
-In other words, ``emissions'' can be interpreted to include any flows that are exterior to the technology matrix, including resource extraction, environmental releases, and cutoffs.
+In other words, ``emissions'' can be interpreted to include any flows that are exterior to the technology matrix, including resource withdrawals, environmental releases, and cutoffs.
 

### 2017-04-01 manuscript/publication-abstract.tex
Final read through and submission (almost)

@@
 \begin{abstract}
-  Many of the challenges facing interpretation and knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of life cycle inventory models.
+  Many of the challenges facing %interpretation and
+  knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of life cycle inventory models.
 This article presents a framework for LCA study authors to publish inventory models in a structured format based on a mathematical formulation of a \emph{foreground study}.
-Using linked semantic data as a foundation, the publication framework offers a mechanism to provide both machine readability and unambiguous human interpretation of published studies.
-Implementing the framework requires only that the LCA community provides a way for study authors to make common reference to shared data sources, including background databases, elementary flows, and impact assessment methods, that are widely used in many studies.
+Using linked semantic data as a foundation, the publication framework %offers a mechanism to provide
+  provides for both machine readability and unambiguous human interpretation of published studies.
+To implement the framework, the LCA community should find ways to enable study authors and data users to make common reference to shared data sources, including background databases, elementary flows, and impact assessment methods, that are widely used in many studies.
 The framework contributes to ongoing efforts within industrial ecology to improve the reproducibility and verifiability of scholarly works, and if implemented, plots a course toward distributed, platform-independent computation and validation of LCA results.

### 2017-04-01 manuscript/publication-components.tex
Final read through and submission (almost)

@@
 
-The collection of foreground nodes and their reference flows expresses a de facto definition of the scope of the model, which is enclosed by an implicit system boundary.
+The foreground is enclosed by an implicit system boundary which includes all foreground nodes and their reference flows.
+Therefore, the list of foreground nodes and reference flows can act as a de facto definition of the effective scope of the model.
 %Most published studies disclose at least the structure of their foreground models by providing a graphical system diagram in the LCA report.

### 2017-04-01 manuscript/publication-conclusion.tex
Final read through and submission (almost)

@@
 The four entity classes found in foreground studies, namely foreground flows, background processes, elementary flows, and characterization quantities, must shift from being recorded locally to being stored on the Semantic Web, where their shared meaning can be curated.
-Fortunately, this is easy to do: authoritative sources simply need to publish hyperlinks to their content.
-Providers of aggregated inventory data should make available static hyperlinks to their content, so that licensees and data users can make shared reference to the same entities, even if the data may not be accessible to the general public.
+Fortunately, this is easy to do: authoritative sources simply need to maintain static hyperlinks to their content, so that licensees and data users can make shared reference to the same entities, even if the data may not be accessible to the general public.
 %Thinkstep already does this: every process inventory they publish has a stable link on the Web\footnote{\href{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/f8dddbcb-f08c-4217-99de-0b59d2c11bb1.xml}{http://gabi-documentation-2016.gabi-software.com/xml-data/processes/[UUID].xml} where [UUID] is replaced with the process's unique identifier.
@@
 Both of these tasks can be accomplished independently through the efforts of members of the LCA community; however, the accuracy and validity of the data would be improved if the work were done by the data providers themselves.
+A shared semantic framework for foreground flows would also support collaborative development of inventory data resources.
 

### 2017-04-01 manuscript/publication-discussion.tex
Final read through and submission (almost)

@@
 An LCA foreground study can be published concisely by distinguishing foreground components from background components, explicitly identifying all entities involved in the study, and then reporting the data values that make up the submatrices in Eq.~\ref{eqn:study}.
-As the examples demonstrate, a structured publication can be created that allows a reader to critically review, validate, and accurately reproduce the product system model without requiring additional information.
+As the examples demonstrate, a structured publication can be created that allows a reader to critically review, validate, and accurately reproduce the product system model without any additional information.
 The entity map provides the foundation for a linked-data publication by explicitly naming all externally-referenced entities involved in the foreground model.
@@
 
-The input matrices ($A_f$, $A_d$, or $B_f$) or aggregation results ($\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, $\tilde{\mathbf{b}}_f$), together with the entity map, form a research object that can be easily generated and shared with colleagues.
-By separating foreground and background, the research object satisfies the objective for attribution, because only the information included in the research object is attributable to the study author; whereas information that came from the background database is excluded from the publication.
+The input matrices ($A_f$, $A_d$, and $B_f$) or aggregation results ($\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, $\tilde{\mathbf{b}}_f$), together with the entity map, form a research object that can be easily generated and shared with colleagues.
+By separating foreground and background, the research object satisfies the objective for attribution, because only the information attributable to the study author is included in the research object; whereas information that came from the background database is excluded from the publication.
 
@@
 Again because the background is excluded, the size of the research object is kept down, and licensing limitations associated with the distribution of proprietary data are avoided.
+Use of the research object to selectively exclude private data is discussed in the supplementary materials.
 
-\subsection{Reproducing LCIA Results}
+\subsection{Validating LCIA Results}
 
@@
 Again, the characterization factors used would need to be disclosed along with the background LCI results.
-Because background databases tend to be updated infrequently, it is important for a background dependency reference to include information about which version was used.
+Because background databases are occasionally updated, it is important for a background dependency reference to include information about which version was used.
 Assuming datasets within a particular background database do not change, it would be possible to create an archive of unit impact scores for each background process and LCIA method supported by the database.
-These values could then be independently validated.
+These values could then be independently shared and validated.
 
-In the context of structured publication, a reader of a structured publication could obtain the unit LCIA scores for each background dataset referenced, and use them to validate the background LCIA result.
+In the context of structured publication, a reader of the publication could obtain the unit LCIA scores for each background dataset referenced, and use them to validate the background LCIA result.
 If the background database is subject to licensing restrictions, it may be necessary to require readers or data users to independently obtain background LCIA scores.
@@
 
+%%%% TODO TODO TODO TODO!!!!!!! save it for the revision
+\subsection{Interlinking Foreground Models}
+
+Current practice does not provide a way for study authors to easily provide their models to collaborators, or for  independently created models to be linked together.
+However, structured publication of foreground studies could allow independent authors to work together on complex models.
+Large foregrounds can be easily partitioned into fragments simply by selectively including or excluding columns of $A_f$, $A_d$, and $B_f$ in different publications, allowing for independent maintenance of different subsystems.
+Similarly, an isolated product system described in a structured format could easily be combined with other models of complementary systems to form a larger system.
+Independent foreground systems that both reference the same shared cutoff flows could be automatically linked without any intervention by the authors.
+
+In both cases, the interlinking of foreground models rests on the principle of shared reference to common entities.
+Linked semantic data provides an enabling technology, both allowing standalone foreground publications to be automatically linked to data resources and to one another, and by providing interpretive services such as dataset discovery, flow matching, unit conversion, and metadata retrieval.
+This represents an opportunity for further research and development.
+
 \endinput

### 2017-04-01 manuscript/publication-example.tex
Final read through and submission (almost)

@@
 
-The product system models in most LCAs can be described as foreground studies (Eq.~\ref{eqn:foreground}).
-In fact, LCI databases contain product systems that can also be modeled as foreground studies because they describe products that are not required by the background.
+LCI databases often contain product systems that can be modeled as foreground studies because they describe products that are not required by the background.
 In this section, two product systems selected from LCI databases are used to illustrate the concept of structured publication.
@@
 Each system is illustrated as a table that shows the foreground model, cutoff flows, background dependencies and foreground emissions included in the system.
+Aggregation results $\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, and $\tilde{\mathbf{b}}_f$ are also reported.
 The table omits numeric data from most of the table for clarity.
 Instead, a black square indicates the presence of a nonzero value.
-Aggregation results $\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, and $\tilde{\mathbf{b}}_f$ are also reported.
 The systems are also reproduced as structured publications in the supplementary materials, formatted as Excel spreadsheets.
@@
 \subsection{US LCI: Secondary Aluminum}
+
+\input{fig_uslci}
 
@@
 
-\input{fig_uslci}
-
 \subsection{Ecoinvent: Organic Potatoes}
@@
 Ecoinvent version 3 is provided in three different system models that reflect different linking strategies.
-The example is drawn from the ``Allocation: Point of substitution'' or APOS model, which includes 11,420 processes that produce 12,966 product flows.
+The example is drawn from the ``Allocation at the point of substitution'' or APOS model, which includes 11,420 processes that produce 12,966 product flows.
 Of these, 10,282 are background flows and the rest are foreground flows.
@@
 The example system reports production of organic potatoes supplied to the global market (Table \ref{table:potato}).
-The foreground includes nine nodes, of which six (nodes 3 through 8) form a strongly connected component associated with the production of potato seeds.
+The foreground includes nine nodes, of which six (nodes 3 through 8) form a cyclic dependency associated with the production of potato seeds.
 The Ecoinvent database's use of ``markets'' as mixer processes is evident in the table: node 0 is a mixer process that combines Swiss (``CH'') production (node 1) weighted at 2.4\%  with rest-of-World (``RoW'') production (node 2) weighted at 97.6\%.
@@
 
-The structured publication of the organic potato model reports the $A_f$, $A_d$, and $B_f$ matrices in full rather than sparse format to enable a more accessible visual review.
+The structured publication of the organic potato model reports the $A_f$, $A_d$, and $B_f$ matrices in full rather than sparse format to enable a more visually immediate review.
 In addition, the $E$ matrix is reported in full, for all emissions found in the model, rather than only the emissions found in the foreground.

### 2017-04-01 manuscript/publication-foreground.tex
Final read through and submission (almost)

@@
 
-\subsection{Model fragments and foreground inputs and outputs}
+\subsubsection{Model fragments and foreground inputs and outputs}
 
 Conventionally, the foreground flows are entirely inside the system boundary: they are produced by one foreground node and consumed by one or more others.
-In the conventional case, the only product flow which crosses the implicit system boundary is the reference flow, which provides the functional unit of the study given in $\tilde{\mathbf{y}}$.
+In this case, the only product flow which crosses the implicit system boundary is the reference flow, which provides the functional unit of the study given in $\tilde{\mathbf{y}}$.
 By implication, the difference $\tilde{\mathbf{x}} - \tilde{\mathbf{y}}$ represents the magnitudes of interior flows within the foreground.
 
-However, these flows may not all necessarily be interior.
+However, not all these flows are necessarily interior.
 Most complex life cycle models involve nested model designs, in which low-level model components are encapsulated within higher-level models.
 A foreground matrix that describes an incomplete life cycle model could be called a model ``fragment.''  In model fragments, some foreground flows represent product flows consumed within the fragment but not produced there, or vice versa.
-These flows represent inputs and outputs to the foreground model that are not further exchanged among the foregrond nodes, but neither are they terminated to a background process or emitted into the environment.
-Thus input/output flows can be detected by identifying foreground flows that show up as row entries but have no column entries.
-
-The input/output flows provide a mechanism to link separate foreground fragments together.
+These flows represent inputs and outputs to the foreground model that may be produced or consumed by other foreground systems.
+%are not further exchanged among the foregrond nodes, but neither are they terminated to a background process or emitted into the environment.  Thus
+Input/output flows can be detected by identifying foreground flows that show up as row entries but have no column entries in the $A_f$, $A_d$, or $B_f$ matrices.
+If there is no other foreground system that connects to an input/output flow, it may be considered to be a ``cutoff flow'' that is excluded from the model scope.
+%
+%%The input/output flows provide a mechanism to link separate foreground fragments together.
 Two foreground models that both specify foreground flows with the same identity can be linked together simply by combining their entries in a common set of $A_f$, $A_d$, and $B_f$ matrices.
-Similarly, large foregrounds can be easily partitioned into fragments simply by selectively including or excluding columns of $A_f$, $A_d$, and $B_f$ in different publications.
-In isolation from other foreground models, input/output flows may be considered to be ``cutoff flows'' that must be terminated by a data user.
+Similarly,
+%%In isolation from other foreground models, input/output flows may be considered to be ``cutoff flows'' that must be terminated by a data user.
 
@@
 
-\subsection{Foreground Dependencies and Foreground Emissions}
+\subsubsection{Foreground Dependencies and Foreground Emissions}
 
@@
 \end{itemize}
-These two components are discussed below.
+%These two components are discussed below.
 
@@
 Given that the matrices being represented are largely made up of zeros, the sparse approach is preferable to the full one.
-The exception is $B_x$ or any column thereof, which is likely to contain no non-zero entries.
+%The exception is $B_x$ or any column thereof, which is likely to contain no non-zero entries.
 Publication of $B_x$ is a special case that will be discussed in a later section.
@@
 The reference units of the background processes should already be part of the background data publication and do not need to be reproduced in the foreground publication except for clarity.
+\item \textbf{Cutoff flows}: Cutoff flows are explicit exclusions from the system boundary.
+They may represent inputs or outputs from the foreground that are produced by another foreground system, or they may represent dependencies for which no background process is available.
+Thus, they are semantically similar to foreground flows.
+They may or may not make reference to externally defined entities.
+However, complex foreground models can be published in parts by making shared reference to a set of cutoff flows.
 \item \textbf{Elementary flows}: The best practices for identifying elementary flows are a matter of ongoing discussion and research.
@@
 
-The entity map should include all of the foreground flows, background processes, elementary flows, and characterization quantities referred to in the publication.
-Taken together, the entity map provides a working formal definition of the scope of the model.
+The entity map should include all of the foreground flows, background processes, cutoff flows, elementary flows, and characterization quantities referred to in the publication.
+Taken together, the entities in the map provide a working formal definition of the scope of the model.
 Even in the absence of the numerical data that make up the input matrices or aggregation results, this \emph{functional scope} can be used as a basis to describe the contents of the model and formally state the system boundary of the foreground.

### 2017-04-01 manuscript/publication-intro.tex
Final read through and submission (almost)

@@
 
-Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver a quantitative measure of utility to a consumer.
-The technique is well-established, widely practiced, and draws on an extensive body of standardization and scholarship.
+Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver %a quantitative measure of utility
+them to a consumer.  The technique is well-established, widely practiced, and draws on an extensive body of standardization and scholarship.
 %Performing an LCA first requires preparing an inventory of the amounts of environmental interventions associated with the product, such as resources extracted or emissions released.
@@
 %
-At the core of an LCA is a model of a ``product system,'' which represents a collection of ``processes $\ldots$ performing one or more defined functions'', that describes the life cycle of a product \citep{iso14044}.
+At the core of an LCA study is a model of a ``product system,'' which represents a collection of ``processes $\ldots$ performing one or more defined functions'', that describes the life cycle of a product \citep{iso14044}.
 Product system models encode information about how products are manufactured, distributed, used by the consumer, and what happens to them after they are disposed.
@@
 %Some types of product systems may require coordination on key methodological issues before results can be compared \citep{Ahlgren_2015}, while others face more fundamental challenges that arise from the complexity of the supply chain and the lack of suitable inventory or proxy data \citep{Bull2014}.
-43 words
-The lack of transparency in reporting,
+The  ## 43 words
+A lack of transparency in reporting,
 particularly regarding definitions of study scope and system boundary,
-is a crucial factor challenging the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.
+is a crucial challenge to the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.
 %In particular, this lack of transparency may manifest as a imprecision in the definitions of study scope and system boundary \citep{Laurent_2014, Turconi_2013}. %% -23 + 9
@@
 %% 33 words
-Other factors can also render study findings unreliable or highly contingent.
+Other factors such as allocation strategy \citep{Pelletier_2014} can also render study findings unreliable or highly contingent.
 Plevin et al.\ famously observed in the strongest terms that careless reporting of attributional LCA study results can distort the significance of findings in a policy context \citep{Plevin_2013}, a shortcoming that can be found in other modes as well \citep{Brandao_2014}.

### 2017-04-01 manuscript/publication-method.tex
Final read through and submission (almost)

@@
 
-At the heart of this article is a precise formulation of an LCA \emph{foreground study}, as distinct from (though likely dependent on) an LCI database, hybrid or input-output database, or software system.
-The study formulation relies on the observation that LCA data sets can be ordered based on their dependency on one another \citep{Kuczenski_JLCA_2015}.
-Background databases generally contain cyclic dependencies, and so it is not useful to publish unit-level descriptions of background processes unless they are all published as a group.
-In contrast, foreground systems often %can be found to
-depend on background systems, while the background systems do not %show a reciprocal dependency
-depend on the foreground.  Because of this it is possible to describe foreground systems precisely %how foreground systems depend on the background database
-without including the background.
+Collections of LCA datasets can be partially ordered based on their dependency on one another \citep{Kuczenski_JLCA_2015}.
+This observation allows for a formal distinction between foreground systems and background systems.
+A foreground system has a dependency on an LCI database, hybrid or input-output database, or software system, whereas the background system does not have a reciprocal dependency on the foreground.
+Because of this, it is possible to describe foreground systems precisely %how foreground systems depend on the background database
+without including the background in the description.  This section develops a precise formulation of an LCA \emph{foreground study}, which is used as the computational basis for the research object.
 
-We begin with the classic formulation of an LCA problem \citep{Heijungs2002}:
+%% Background databases generally contain cyclic dependencies, and so it is not useful to publish unit-level descriptions of background processes unless they are all published as a group.
+%% In contrast, foreground systems often %can be found to
+%% depend on background systems, while the background systems do not %show a reciprocal dependency
+%% depend on the foreground.  Because of this
+
+We begin with the classic formulation of the LCA problem \citep{Heijungs2002}:
 \begin{equation}
@@
 
-\input{fig_lca}
-It can be shown that, under conditions typical for LCA databases, the technology matrix can be represented equivalently in the input-output format, in which the technology matrix is actually the combination of separate supply and use tables, each of which has been normalized to a unit of output \citep{Suh_JIE_2010, Pauliuk_2015_framework}:
+It can be shown that, under conditions typical for LCA databases, the technology matrix can be represented as an input-output database, which is made up of separate supply and use tables, each of which has been normalized to a unit of output \citep{Suh_JIE_2010, Pauliuk_2015_framework}:
 \begin{equation}
@@
 Eq.~\ref{eqn:leontief} is visualized in Figure~\ref{fig:lca}.
+\input{fig_lca}
 
@@
 \end{equation}
-where $B_x = B* \times (I - A*)^{-1}$ is a background LCI database, which can be prepared in advance by a background database maintainer, as is currently common practice.
+where $B_x = B* \times (I - A*)^{-1}$ is a background LCI database.
+It is currently common practice for $B_x$ to be prepared in advance by a background database maintainer.
 In this case, Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:
@@
 Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}.
-This formulation shows that given access to the background database $B_x$, the study results can be fully reproduced using only the submatrices $A_f$, $A_d$, and $B_f$, as well as the characterization vector $\mathbf{e}$.
+This formulation shows that given access to the background database $B_x$ and the characterization vector $\mathbf{e}$, the study results can be fully reproduced using only the submatrices $A_f$, $A_d$, and $B_f$.
 %The foreground, which represents the activities and processes that are directly within the scope of the study, can be fully defined by its graph topology ($A_f$); its dependency on the background ($A_d$); and its direct emissions ($B_f$).
 
-The foreground graph $A_f$ can be distilled into a vector of foreground activity levels for the foreground nodes, which determine the magnitude of dependencies and emissions for a particular output, case, or scenario.
+The foreground graph $A_f$ can be distilled into a vector of activity levels for the foreground nodes, which determine the magnitude of dependencies and emissions for a particular output.
 With the functional unit for one study formulation given as $\tilde{\mathbf{y}}$, the foreground activity levels can be given as:
@@
 \end{equation}
+Although the variable $\tilde{\mathbf{x}}$ can be computed through matrix inversion as expressed in Eq.~\ref{eqn:inv}, often matrix inversion is not required.
+If the foreground does not include any feedback loops, then the activity levels of foreground can be determined simply by traversing the graph \citep{Bapat_LAA_2013}, which can be performed in linear computational time with respect to the number of foreground links.
+
 For a typical foreground study, a ``canonical'' $\tilde{\mathbf{y}}$ may be suggested in which the first foreground node produces the functional unit as output:
@@
 \end{equation}
-Although the variable $\tilde{\mathbf{x}}$ can be computed through matrix inversion as expressed in Eq.~\ref{eqn:inv}, often matrix inversion is not required.
-If the foreground does not include any feedback loops, then the activity levels of foreground can be determined simply by traversing the graph \citep{Bapat_LAA_2013}, which can be performed in linear computational time with respect to the number of foreground links.
 In the canonical case the vector $\tilde{\mathbf{x}}$ is the first column of the matrix $(I-A_f)^{-1}$  and can be computed by traversal of the foreground tree beginning on the first node.

### 2017-04-01 manuscript/publication-objective.tex
Final read through and submission (almost)

@@
 
-However, the existence of semantic data resources is not enough by itself to establish the reproducibility of a model or result: it must also be clear how the information is used to generate the result.
+However, the existence of semantic data resources is not enough by itself to establish a reproducible result: it must also be clear how the information is used to generate the result.
 A \emph{research object} \citep{Bechhofer_2013} is an aggregation of linked references, combined with a structure or system for converting the information into the scientific model for reproducing the result.
@@
 \begin{enumerate}
-\item \emph{Linked Data Foundation}.  Human interpretation of the publication should rely on the use of references to shared resources identified by URIs or hyperlinks.
+\item \emph{Linked Data Foundation}.  Both human and machine interpretation of the publication should rely on the use of references to shared resources identified by URIs or hyperlinks.
 \item \emph{Computation}.  The role of the publication in an LCA computation should be clear.  If the publication includes a result, it should be clear how the result was obtained.

### 2017-08-08 revision/reviewer-response.txt
Reading R3, reviewing doc for big changes

==========
Tue Aug 08 10:15:39 -0700 2017

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
R1-

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
R2-

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
R3-

Why JIE and not JLCA? short reason- better community for 'scary new' ideas. but I can't just say that...

(1) Page 17 line 12: To my knowledge, this short sentence question the potential usefulness of the approach developed in the article. Data confidentiality in the proposed research object should be discussed more thoroughly, in particular with respect to the current critical review process in ISO LCA.

We take the author to refer to page 17 of the PDF proof, which is manuscript page 15.  Line 12 reads "...for reasons of privacy, an author may wish to conceal some or all data points, or include the data points but exclude the actual data values. This is allowable if it is consistent with the author’s goal, and it will accordingly limit the ability of a data user to reproduce the model."

Definitely. OK.

(2) Page 17 line 40: Similarly, data confidentiality might result in a very sparse entity map. Can such a research object be organized in order to safeguard confidentiality and allow reproducibility of the results?

### 2017-09-05 revision/reviewer-response.txt
Last month's work on reviewer comments

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-R2-
+R2- Pauliuk
+
+"... sound mathematical description and a guideline and examples of how to extract and format crucial LCA results ...  well suited for publication in JIE."
+
+* Right now, most LCA studies are close to useless when being compared to others because of complete lack of transparency.
+From that state we have a ladder of improvements to make, and that ladder should be described in detail and linked to the different aggregation steps.
+
+" ... we need to provide for unique labelling of data points across databases and computation platforms."
+
+" version clarity"
+
+" reference products"
+
+Difference between Heijungs and Leontief style matrices
 

### 2017-09-05 manuscript/agg-intro.tex
Today's work on rewriting the intro.

--- 2017-04-01 manuscript/publication-intro.tex
@@
 
-Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver %a quantitative measure of utility
+Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver
 them to a consumer.  The technique is well-established, widely practiced, and draws on an extensive body of standardization and scholarship.
-%Performing an LCA first requires preparing an inventory of the amounts of environmental interventions associated with the product, such as resources extracted or emissions released.
-The inventory is the input to an impact assessment computation, in which the inventory items are scaled by their relative significance according to impact categories of interest, and summed together.
-%% 55 words
 Increasingly, the results of LCA studies are appearing in environmental declarations, corporate sustainability reports, and product marketing information.
@@
 While the foreground is often modeled from direct observation or other primary data, background processes are typically drawn from a life cycle inventory (LCI) database prepared by a third party.
-%A relatively small number of comprehensive LCI databases exist, owing to their vast data requirements and the complexity of preparing and maintaining them \citep{UNEP_2011}.
-The product system model itself is often constructed in specialized software, which is used to compute and analyze impact assessment results.
-$$ 45 words
 
-Although there has been considerable effort to normalize and harmonize %%important methodological decisions around %% -4 +1 
+Although there has been considerable effort to normalize and harmonize 
 LCI database design methodology and data interchange \citep{UNEP_2011, JRC_ILCD_ELCD_2013, Mila_e_Canals_2015, Ingwersen_JLCA_2015}, less attention has been paid to harmonizing product system models.
@@
 Even when relatively simple and homogeneous systems are considered, wide variations can be found in the results of studies from different authors \citep{van_der_Harst_2013, Turconi_2013}.
-%Some types of product systems may require coordination on key methodological issues before results can be compared \citep{Ahlgren_2015}, while others face more fundamental challenges that arise from the complexity of the supply chain and the lack of suitable inventory or proxy data \citep{Bull2014}.
-The  ## 43 words
-A lack of transparency in reporting,
-particularly regarding definitions of study scope and system boundary,
+A lack of transparency in reporting, particularly regarding definitions of study scope and system boundary,
 is a crucial challenge to the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.
-%In particular, this lack of transparency may manifest as a imprecision in the definitions of study scope and system boundary \citep{Laurent_2014, Turconi_2013}. %% -23 + 9
-One consequence
-%% of the wide variation in model construction, as well as the lack of consistent transparency in reporting, %% -18
-is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.
+One consequence is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.
 
@@
 As noted above, this level of consensus is hard to realize, in part because of the conflicting implications inherent in different analytic modes.
-%It is well known that allocation decisions can have dispositive effects on the outcome of a study; yet no consensus exists for specifying allocation methods (e.g.
-\cite{Hanes_2015, Pelletier_2014}), much less evaluating their significance.
-%% 33 words
 Other factors such as allocation strategy \citep{Pelletier_2014} can also render study findings unreliable or highly contingent.
@@
 Moreover, PCRs are often themselves not unique and can include wide variations in scope and boundary requirements \citep{Subramanian_2012}, leading to ambiguity.
-%%PCRs themselves tend, like EPDs and other LCA studies, to be written qualitatively, to be unstructured, and to be open to considerable interpretation
+
+\endinput
+
+Some thoughts to put together and rearrange:
+
+ * The task of computing life cycle impact assessment (LCIA) results from a product system model is often described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.
+
+ * In the process of aggregating the models, much of the information about their construction is lost.
+This is partly by design, in order to protect the privacy of the input data \citep[Ch.
+3]{UNEP_2011}.
+
+% \citep{Kuczenski_2017}.
+
+ * However, that same feature is also a shortcoming when considered from the perspective of study transparency, replication, and extension.
+
+ * Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature (Suh 2005; Lenzen 2007), the computational representation of product system models, and the generation of aggregated results, remains steadfastly software-specific.
+
+ * When a study is reviewed, it is often the written documentation and \emph{not} the product system model itself that is the object of review.
+
+  - In current practice, it is the responsibility of the study author to ensure that the written report accurately describes the model.
+
+  - Because the model itself may not be available for review, it is impossible for a reviewer to directly evaluate the author's claim that the model is accurately represented.
+
+  - Because the representation of product system models is software-specific, it is also not generally possible to recreate a product system modeled in one software system in a separate software system.
+
+ * Documents such as ISO reports, PCRs, or EPDs are uniformly \textit{unstructured}, meaning their interpretation must be secured by a human reading text and graphical images such as system boundary diagrams, but cannot be expressed to a machine.
+
+ *
+
+\oldinput
 
@@
 In this paper, I propose a framework for the structured publication of product system models in LCA that achieves these dual requirements.
+
+\cruft
+
 %% \textit{Publication}, for the purpose of this article, means disclosing sufficient detail to allow a reader to reproduce an LCA result, or all or part of the model that produced it.

### 2017-09-18 manuscript/agg-intro.tex
previous work

@@
 
+What is the point of the new paper?
+
 Some thoughts to put together and rearrange:
+
+ * A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the product system model.
+
+ * Although the general structure for life cycle assessment can be formally stated, there is no current practice for formally describing the product system model and thus, no formal mechanisms for interpretation or review of published LCA studies.
+
+ * When a study is reviewed, it is often the written documentation and \emph{not} the product system model itself that is the object of review.
+
+  - In current practice, it is the responsibility of the study author to ensure that the written report accurately describes the model.
+
+  - Because the model itself may not be available for review, it is impossible for a reviewer to directly evaluate the author's claim that the model is accurately represented.
+
+or for distinguishing between the foreground and the background of a model.
 
@@
 
- * Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature (Suh 2005; Lenzen 2007), the computational representation of product system models, and the generation of aggregated results, remains steadfastly software-specific.
+ * Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature (Suh 2005; Lenzen 2007), the computational representation of product system models, and the generation of aggregated results, remains software-specific.
 
- * When a study is reviewed, it is often the written documentation and \emph{not} the product system model itself that is the object of review.
+ * Because of the nature of the computation, it is straightforward to provide varying degrees of
 
-  - In current practice, it is the responsibility of the study author to ensure that the written report accurately describes the model.
-
-  - Because the model itself may not be available for review, it is impossible for a reviewer to directly evaluate the author's claim that the model is accurately represented.
-
-  - Because the representation of product system models is software-specific, it is also not generally possible to recreate a product system modeled in one software system in a separate software system.
+The aim of this paper is to present a consistent and reproducible formulation of LCA foreground models, and to focus on the aggregation step
 

### 2017-09-18 manuscript/agg-intro.tex
Starting the objectives

@@
 is a crucial challenge to the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.
-One consequence is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.
+A further challenge to transparency can be found in the use of input data that are confidential or proprietary \citep{Kuczenski_2017}.
+One consequence of all these challenges is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.
 
@@
 
+An LCA study result is ultimately an assertion that for some constructed product model, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.
+But if the product model itself is ambiguously stated, then the significance of the result is ambiguous as well.
+For a computational model to have scientific significance, its results need to be both verifiable and replicable by other scientists \citep{Fomel_2009, Mesirov_2010}.
+But in the case of LCA, the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.
+When a study is reviewed, it is often the written report and \emph{not} the product system model itself that is the object of review.
+In the end, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the product model itself is not available.
+
+A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the product system model.
+Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the product system model and thus, no automatic mechanisms for interpretation or review of published LCA studies.
+
+The purpose of this paper is to advance a framework for the formal description of product system models that would provide robust transparency and reproducibility of computed results.
+
+The purpose of this paper is three-fold:
+\begin{enumerate}[label=\roman*.]
+\item to advance a fram
+\item to describe the computation of aggregated LCIA results using a product system model as input data;
+\item to show how the framework may be used to protect confidential information while still providing transparency and reproducibility during critical review.
+\end{enumerate}
+
 \endinput
+
+o advance a framework for the formal description of a subset of LCA studies that can be used to document, validate, and reproduce the computations that lead from model specification to LCIA results.
+The framework can also
+
+But reproducibility of a computation requires a distinction between the computing algorithm and the input data \citep{Buckheit_1995, }; contemporary LCA lacks this distinction.
 
@@
 
- * A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the product system model.
-
- * Although the general structure for life cycle assessment can be formally stated, there is no current practice for formally describing the product system model and thus, no formal mechanisms for interpretation or review of published LCA studies.
+ *
 
@@
 
- * The task of computing life cycle impact assessment (LCIA) results from a product system model is often described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.
+ *
 
@@
 
- * Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature (Suh 2005; Lenzen 2007), the computational representation of product system models, and the generation of aggregated results, remains software-specific.
+ *
 

### 2017-09-19 aggregation/agg-objective.tex
New folder for new paper

\section{Formalizing LCA Study Aggregation}

The task of computing life cycle impact assessment (LCIA) results from a product system model is often described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.  Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh_2005, Peters_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.

The practice of reporting the results of an LCA study implies the existence of two parties with different information: the study Author has full knowledge of the product system model; and the Reader only has knowledge of what the Author has disclosed.  Obviously, the level of detail of the disclosure will determine how much of the computation can be reproduced by the Reader.  In this context, the Reader will want to know the answers to various questions about the model, which we will call \emph{review objectives}.  At the same time the Author will be limited in what she can disclose, which we will call \emph{disclosure constraints}.  The purpose of the framework is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.

In the context of this paper, an ``LCA Study'' should be understood as any aggregation of data under consideration, from a single-operation unit process to a full ISO-style product LCA.  The framework should meet the following objectives:

\begin{itemize}
\item \textbf{Computability}. The result should be computed using only the disclosure, plus resources available to the reader.  The computation should follow the consensus understanding of process LCA computation.
\item \textbf{Completeness / Minimality}.  The computation should use all of the information provided in the disclosure.
\item \textbf{Reproducibility}.  The result obtained by the reader should match the result provided by the author.
\item \textbf{Authority}.  The disclosure should include only information for which the author is responsible.
\item \textbf{Autonomy}.  The disclosure should exclude information that is owned by a third party.
\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the author.

Let's say the author (whom we'll call Alice) wishes to make a disclosure to the reader (whom we'll call Rob) that would enable the reader to \emph{automatically} reproduce a given numerical result.

The framework advanced in this paper aims to specify the contents of a \emph{minimal disclosure} by the author that would enable the reader to  reproduce

A framework for formally describing the product system model must meet the following objectives:
\begin{itemize}
\item \textbf{Transparency}.  It must be possible for a reader to identify the contents of the model (i.e. the selection of data sets) and the structure of the model (i.e. the way the data sets are connected).
\item \textbf{Reproducibility}.  If the reader has possession of the same input data, then she may use

A framework to formally describe product system models and compute aggregation results

should meet the following objectives:

\begin{itemize}
\item It must be possible for a data to  reconstruct

known as ``foreground studies,'' in which the results depend on a background database but the background database does not depend on the result.

The framework can be applied to a common subclass of LCA studies, ``foreground studies''.

### 2017-09-19 aggregation/agg-basic-lca.tex
Mathiness

\section{Mathematical Formulation of LCA}

Because the objective of the framework is to validate a computation, we begin by providing a mathematical formulation of the LCA problem.  An LCA computation is performed over a collection of linked unit processes, each of which describes a single operation or reference product.  The task of LCA can be described in two steps: determining the required outputs of each process; and determining the environmental impacts of each process.  If the processes are assumed to be linear, the problem can be written in terms of matrix algebra.  There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Suh_2002, Pauliuk_2015}.  Although the two models are understood to be equivalent, the input-output approach is computationally more efficient because it lends itself to iterative solution \citep{Peters_2007}.  That formulation may be stated:
\begin{equation}
s = \mathbf{e}^T \cdot B \cdot \left(I - A\right)^{-1} \cdot \mathbf{y}
\label{eqn:leontief}
\end{equation}
Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values) per unit of a process's reference output; $B$ is the environmental intervention or emission matrix; and $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions with respect to a given impact assessment method.  In this model, $\mathbf{y}$ is the externally specified final demand, which is the product or service whose delivery is to be assessed, and $s$ is the numerical impact score or category indicator.

The term $(I-A)$ is equivalent to the normalized technology matrix, i.e. the difference between the supply and use tables in classical input-output analysis, $(V'-U)$ \citep{Suh_2010}.  The conditions under which this matrix has an inverse, particularly with respect to different strategies for handling co-production, have been discussed extensively \citep{Suh_2010, Majeau_Bettez_2014}.  Eq.~\ref{eqn:leontief} is visualized in Figure~\ref{fig:lca}. 
\input{fig_lca}

In classical LCA, the matrices $A$ (or $I-A$) and $B$ are called a life cycle inventory database (LCIDB), and the functional unit of a study is given by the final demand vector $\mathbf{y}$.  Intermediate results of this computation include:
\begin{itemize}
\item $\mathbf{x} = \left(I-A\right)^{-1}\cdot\mathbf{y}$, the activity level vector;
\item $\mathbf{b} = B\cdot\mathbf{x}$, the life cycle inventory;
\item $s = \mathbf{e}^T\cdot\mathbf{b}$, the life cycle impact category indicator, the study result.
\end{itemize}
For simplicity, we will consider the computation of a single impact result.  However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.

This formulation is equivalent to the implementations of all major LCA software systems, some of which provide the capacity to export the $A$ and $B$ matrices for a given study model.  However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are the same for all users.  Thus exporting $A$ and $B$ violates our authority and primacy objectives.  Additionally, replicating the entire LCIDB is cumbersome because of the large size of the matrices.   As a consequence, the matrix formulation in Eq.~\ref{eqn:leontief} is unsuitable as a disclosure framework.

### 2017-09-19 aggregation/agg-lca-study.tex
Mathiness

\section{Foreground Study}

In practice, most LCA studies contain foreground information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with foreground models.  In consideration of our disclosure objectives, a disclosure should be limited to the foreground of a study and exclude the background.  Because background databases are prepared independently of any study, their computational results do not depend on the studies in which they are used.  However, the foreground systems themselves cannot be computed without the background.  Thus the dependency relationships among processes in a life cycle model allow the processes to be sorted into foreground and background groups, and an $A$ matrix augmented with foreground information can be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.  An appropriately ordered $A$ matrix can be written in block-triangular form:
\begin{equation}
A = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A*
  \end{array}
\right];\;\;\;  B = \left[\begin{array}{cc} B_f & B*   \end{array}\right]
\label{eqn:foreground}
\end{equation}

\input{fig_foreground}
In this case, the submatrix $A_f$ represents the foreground; $A*$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.  The ordered $B$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B*$ which includes background emissions. The constitutive characteristic of this formulation is that the background system does not depend on the foreground, allowing for the submatrix in the top right corner of Eq.~\ref{eqn:foreground} to be zero.  As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.\footnote{It should be noted that this condition does not hold for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells.  The approach presented here applies only to LCA studies in which the background system is invariant with respect to the foreground, a class that includes the vast majority of LCA case studes.}

### 2017-09-19 aggregation/agg-objective.tex
Mathiness

@@
 The practice of reporting the results of an LCA study implies the existence of two parties with different information: the study Author has full knowledge of the product system model; and the Reader only has knowledge of what the Author has disclosed.
-Obviously, the level of detail of the disclosure will determine how much of the computation can be reproduced by the Reader.
+Obviously, the level of detail of the disclosure will determine how useful the computation will be to the Reader.
 In this context, the Reader will want to know the answers to various questions about the model, which we will call \emph{review objectives}.
@@
 In the context of this paper, an ``LCA Study'' should be understood as any aggregation of data under consideration, from a single-operation unit process to a full ISO-style product LCA.
-The framework should meet the following objectives:
+Furthermore, the aggregation can be computed at any level: a foreground (inventory) result; a background (accumulated life cycle inventory) result; or a computed LCIA score.
+These different levels are discussed in detail in the following sections.
 
+On the basis of transparency, the framework should meet the following objectives:
 \begin{itemize}
 \item \textbf{Computability}.
-The result should be computed using only the disclosure, plus resources available to the reader.
+The result should be computed using only the disclosure, plus resources available independently to the Reader.
 The computation should follow the consensus understanding of process LCA computation.
 \item \textbf{Completeness / Minimality}.  The computation should use all of the information provided in the disclosure.
-\item \textbf{Reproducibility}.  The result obtained by the reader should match the result provided by the author.
-\item \textbf{Authority}.  The disclosure should include only information for which the author is responsible.
-\item \textbf{Autonomy}.  The disclosure should exclude information that is owned by a third party.
-\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the author.
+\item \textbf{Reproducibility}.  The result obtained by the Reader should match the result provided by the Author.
+\end{itemize}
+One key objective of disclosure is to inspect the work of the Author.
+Consequently, the disclosure should be made up of the Author's own work product and should exclude, for instance, data that were drawn from a reference database or other independent source and used in an unmodified form.
+The purpose of this is twofold: to focus the efforts of the Reader on the work of the Author, and to enable the Reader to obtain and inspect the third-party data independently, thus ensuring its integrity.
+\begin{itemize}
+\item \textbf{Authority}.  The disclosure should include only information for which the Author is responsible.
+\item \textbf{Primacy}.  The disclosure should exclude information that is made available by a third party.
+\end{itemize}
+In addition, because of the ubiquity of confidential and sensitive information in LCA, any operational disclosure framework must also meet the following:
+\begin{itemize}
+\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the Author.
+\end{itemize}
+
+\endinput
 

### 2017-09-19 figures/aluminum-secondary.tex
Move figures folder

Aluminum, secondary, ingot, from automotive scrap, at plant [RNA]

{\scriptsize\sffamily
\begin{tabularx}{\textwidth}{|>{\hangindent=2ex}X|c@{~}c@{~}c@{~}c@{~}|c|}
\hline
\everypar{\hangindent1em \hangafter1}
(node) Foreground flow \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & \\ 
\hline
(0) Aluminum, secondary, ingot, from automotive scrap, at plant (kg) [RNA] & \refbox  &  &  &  & \\ 
(1) Aluminum recovery, transport, to plant (kg) [RNA] & 1.03 & \refbox  &  &  & \\ 
(2) Quicklime, at plant (kg) [RNA] & 2.35e-05 &  & \refbox  &  & \\ 
(3) Limestone, at mine (kg) [RNA] &  &  & 1.87 & \refbox  & \\ 
\hline
Foreground Node Weights $\tilde{x}$ &    1 & 1.03 & 2.35e-05 & 4.39e-05& \\ 
\hline
Input: CUTOFF Disposal, solid waste, unspecified, to sanitary landfill [CUTOFF Flows] & \dependency &  & \dependency &  & 0.0842\\ 
Input: CUTOFF Filter media, at plant [CUTOFF Flows] & \dependency &  &  &  & 4.79e-05\\ 
Input: CUTOFF Lube oil, at plant [CUTOFF Flows] & \dependency &  &  &  & 8.75e-07\\ 
Input: CUTOFF Treatment gases, unspecified, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.003\\ 
Input: CUTOFF Alloying additives, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.021\\ 
Input: CUTOFF Chemicals, unspecified, used for wastewater treatment [CUTOFF Flows] & \dependency &  &  &  & 0.008\\ 
Input: CUTOFF Grain refiners, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.0003\\ 
Input: CUTOFF Treatment salts, unspecified, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.009\\ 
Input: CUTOFF Packaging, unspecified, at plant [CUTOFF Flows] & \dependency &  &  &  & 1.67e-05\\ 
Input: CUTOFF Aluminum scrap, automotive [CUTOFF Flows] & \dependency &  &  &  &  1.03\\ 
\hline
\hline
Background Dependencies \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & $\tilde{a_d}$\\ 
\hline
Diesel, combusted in industrial boiler (l) [RNA] &  &  & \dependency & \dependency & 4.79e-08\\ 
Transport, train, diesel powered (t*km) [RNA] &  & \dependency & \dependency &  & 0.0415\\ 
Electricity, at grid, US, 2000 (kWh) [RNA] & \dependency &  & \dependency & \dependency & 0.668\\ 
Transport, combination truck, diesel powered (t*km) [RNA] &  & \dependency & \dependency &  & 0.374\\ 
Natural gas, combusted in industrial boiler (m3) [RNA] & \dependency &  & \dependency & \dependency & 0.223\\ 
Bituminous coal, combusted in industrial boiler (kg) [RNA] &  &  & \dependency & \dependency & 4.04e-06\\ 
Gasoline, combusted in equipment (l) [RNA] &  &  &  & \dependency & 2.25e-09\\ 
Liquefied petroleum gas, combusted in industrial boiler (l) [RNA] &  &  & \dependency &  & 7.57e-10\\ 
Transport, barge, average fuel mix (t*km) [RNA] &  &  & \dependency &  & 5.66e-07\\ 
\hline
Foreground Emissions \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & $\tilde{b_f}$\\ 
\hline
Output: Lead [air, unspecified] & \dependency &  &  &  & 2.16e-07\\ 
Output: Suspended solids, unspecified [water, unspecified] & \dependency &  &  &  & 2.56e-05\\ 
Output: Particulates, unspecified [air, unspecified] & \dependency &  & \dependency & \dependency & 1.78e-07\\ 
Output: NMVOC, non-methane volatile organic compounds [air, unspecified] & \dependency &  &  &  & 4.7e-05\\ 
Output: BOD5, Biological Oxygen Demand [water, unspecified] & \dependency &  &  &  & 1.35e-09\\ 
Output: COD, Chemical Oxygen Demand [water, unspecified] & \dependency &  &  &  & 8.3e-07\\ 
Output: Sulfur dioxide  [air, unspecified] &  &  & \dependency &  & 3.52e-09\\ 
Output: Dissolved solids [water, unspecified] & \dependency &  &  &  & 2.38e-08\\ 
Output: carbon dioxide [air, unspecified] &  &  & \dependency &  & 1.8e-05\\ 
Input: Limestone [resource, ground-] &  &  &  & \dependency & 4.39e-05\\ 
Output: Heavy metals, unspecified [water, unspecified] & \dependency &  &  &  & 2.25e-05\\ 
Output: Organic substances, unspecified [water, unspecified] & \dependency &  &  &  & 1.29e-06\\ 
Output: Acids, unspecified [air, unspecified] & \dependency &  &  &  & 4.31e-05\\ 
\hline
\end{tabularx}
}

### 2017-09-19 aggregation/agg-JIE.tex
Further progress on math

\documentclass[12pt]{article}

\input{formatting}

\usepackage{tabularx}
\usepackage[round]{natbib}
\bibliographystyle{authordate_bk}

%\thispagestyle{empty}

\input{psstyle}
\input{draw-matrices}

\title{Structured Representation of Aggregated Product System Models in Life Cycle Assessment}
\author{Brandon Kuczenski\\
Institute for Social, Behavioral, and Economic Research\\
University of California, Santa Barbara}
\date{\mydate\today}

%\usepackage{lineno}
\let\showgrid\relax
\let\oldtimes\times
\let\times\cdot

\def\mainpaper{

  \maketitle
  %\emph{For submission to the Journal of Industrial Ecology; February 2017}

  \input{agg-abstract}

  \doublespacing

  \input{agg-intro}

  \input{publication-objective}

  \input{publication-method}

  \input{publication-components}

  \input{publication-foreground}

  \input{publication-example}

  \input{publication-discussion}

  \input{publication-conclusion}

  \input{publication-acknowledgments}

  \singlespacing
  \bibliography{jabref}
}

\def\supporting{
  \renewcommand{\thefigure}{S\arabic{figure}}
  \renewcommand{\thetable}{S\arabic{table}}

  \maketitle

  \begin{center}
    \Large Supporting Information
  \end{center}

  \doublespacing

  \input{publication-si-spreadsheet}

  \input{publication-forms}

  \input{publication-si-recipe}

  \input{publication-si-openaccess}

  \singlespacing
  %\bibliography{jabref}
}

\def\unused{
  \input{publication-background}
}

\begin{document}

%\mainpaper
\supporting

%% %\unused

\end{document}

### 2017-09-19 aggregation/agg-lca-study.tex
Further progress on math

@@
 As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.\footnote{It should be noted that this condition does not hold for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells.
-The approach presented here applies only to LCA studies in which the background system is invariant with respect to the foreground, a class that includes the vast majority of LCA case studes.}
+The approach presented here applies only to LCA studies in which the background system is invariant with respect to the foreground, a class that includes the vast majority of LCA case studes.}  The following formulation is equivalent to Eq.~\ref{eqn:foreground}:
+\begin{equation}
+ A' = \left[\begin{array}{cc} 
+A_f &  0 \\
+A_d & 0 
+   \end{array}\right];\;\;\;  B = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
+\label{eqn:fg_bg}
+\end{equation}
+where $B_x = B* \times (I - A*)^{-1}$ is a background LCI database.
+The proof of the equivalency of Eq.s\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary information to \cite{Kuczenski_JLCA_2015}.
+It is currently common practice for $B_x$ to be prepared in advance by a background database maintainer.
+
+[partition y into yf; 0]
+
+needswork to respond to Suh...
+
+In this case, Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:
+\begin{equation}
+s = \mathbf{e} \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \mathbf{y}_f
+\label{eqn:study}
+\end{equation}
+where $\mathbf{y_f}$ represents final demand of the foreground processes only.
+This formulation shows that given access to the background database $B_x$ and the characterization vector $\mathbf{e}$, the study results can be fully reproduced using only the submatrices $A_f$, $A_d$, and $B_f$.
+
+Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}.

### 2017-09-20 aggregation/agg-lca-study.tex
math revisions: I think I got it.

--- 2017-09-19 aggregation/agg-lca-study.tex
@@
 In practice, most LCA studies contain foreground information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with foreground models.
-In consideration of our disclosure objectives, a disclosure should be limited to the foreground of a study and exclude the background.
-Because background databases are prepared independently of any study, their computational results do not depend on the studies in which they are used.
+In consideration of our objectives, a disclosure should be limited to the foreground of a study and exclude the background.
+The simplest conceivable foreground is simply what is given as the final demand, in other words the list of background processes invoked by the study and their activity levels.
+For clarity, I introduce a ``canonical'' final demand vector in which the first entry is 1 and all successive entries are 0:
+\begin{equation}
+\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}
+\label{eqn:canonical_y}
+\end{equation}
+The $A$ and $B$ matrices can be augmented with the $\mathbf{y]$ vector and a $\mathbf{0}$ vector respectively, to generate the same results using the canonical $\tilde{\mathbf{y}}$:
+\begin{eqnarray}
+s &= \mathbf{e}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
+ &= \mathbf{e}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde\mathbf{y}\\
+\tilde{A} &= \left[\begin{array}{cc}
+ 0 & 0 \\
+ \mathbf{y} & A
+  \end{array}\right]\\
+\tilde{B} &= \left[\begin{array}{cc} \mathbf{0} & B\right]  \end{array}
+\label{eqn:canonical}
+\end{eqnarray}
+This is intuitively true because the canonical $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.
+It can easily be shown that the first equality in Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.
+\textit{author's note: if need be, a proof for this equality can be provided in the electronic supplementary information.}
+
+Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database.
+To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.
+Because background databases are prepared independently of any study, their computational results do not vary across all the studies in which they are used.
 However, the foreground systems themselves cannot be computed without the background.
-Thus the dependency relationships among processes in a life cycle model allow the processes to be sorted into foreground and background groups, and an $A$ matrix augmented with foreground information can be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.
-An appropriately ordered $A$ matrix can be written in block-triangular form:
+Put another way, the foreground depends on the background but the background does not depend on the foreground.
+This condition allows the processes in an LCIDB to be sorted into foreground and background groups.
+The processes an LCIDB containing both a foreground and a background can then be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.
+
+Thus a foreground study can be written in block-triangular form, using the canonical functional unit $\tilde{\mathbf{y}}$, as follows:
 \begin{equation}
-A = \left[\begin{array}{cc}
+\tilde{A} = \left[\begin{array}{cc}
 A_f & 0 \\
-A_d &  A*
+A_d &  A
   \end{array}
-\right];\;\;\;  B = \left[\begin{array}{cc} B_f & B*   \end{array}\right]
+\right];\;\;\;  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right]
 \label{eqn:foreground}
@@
 \input{fig_foreground}
-In this case, the submatrix $A_f$ represents the foreground; $A*$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.
-The ordered $B$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B*$ which includes background emissions.
-The constitutive characteristic of this formulation is that the background system does not depend on the foreground, allowing for the submatrix in the top right corner of Eq.~\ref{eqn:foreground} to be zero.
+In this case, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.
+The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions.
+The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.
 As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.\footnote{It should be noted that this condition does not hold for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells.
-The approach presented here applies only to LCA studies in which the background system is invariant with respect to the foreground, a class that includes the vast majority of LCA case studes.}
+The approach presented here applies only to LCA studies in which the background system is invariant with respect to the foreground.
+This class includes the vast majority of LCA case studes, and very likely includes all case studies prepared using commercial LCA software.}
+
+It is currently common practice for background database maintainers to pre-compute the life cycle inventory results for their databases.
+Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, Eq.s~\ref{eqn:foreground} can be rewritten as:
+\begin{equation}
+ \tilde{A} = \left[\begin{array}{cc} 
+A_f &  0 \\
+A_d & 0 
+   \end{array}\right];\;\;\;  \tilde{B} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
+\label{eqn:fg_bg}
+\end{equation}
+The proof of the equivalency of Eq.s\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary information to \cite{Kuczenski_JLCA_2015}.
+In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining there pertains to the study foreground.
+Furthermore, the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported.
+
+Using the ordered matrices, Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:\footnote{For a demonstration of this equivalence, the reader is again directed to \cite{Kuczenski_JLCA_2015}.}
+\begin{equation}
+s = \mathbf{e}^T \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \tilde{\mathbf{y}}
+\label{eqn:study}
+\end{equation}
+
+This formulation shows that given access to the background database $B_x$ and the characterization vector $\mathbf{e}$, the study results can be fully reproduced using only the submatrices $A_f$, $A_d$, and $B_f$.
+Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$, having been subsumed into $B_x$, represented by an identity matrix $I$.
+
+An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
+\begin{equation}
+\tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
+\label{eqn:inv}
+\end{equation}
+This vector can be used to describe an \emph{aggregated foreground}, which generates the same results as the fully expanded foreground:
+\begin{equation}
+\begin{array}{rl}
+    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
+    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
+    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
+\end{array}
+\label{eqn:agg}
+\end{equation}
+
+These vectors summarize the contents of the foreground without disclosing its detailed structure.
+The aggregated dependency vector $\tilde{\mathbf{a}}_d$ has the same dimension as the background database, and the aggregated emission vector $\tilde{\mathbf{b}}_f $has the same dimension as the elementary flow matrix.
+The dependency vector can be transformed into an elementary flow vector by multiplying by $B_x$.
+The result of the study $s$ is the sum of foreground and background impact scores:
+
+\begin{equation}
+\begin{array}{rl}
+   \tilde{\mathbf{b}} & = \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x \\
+   s &= \mathbf{e}^T \times \tilde{\mathbf{b}} \\
+   & = \mathbf{e}^T \times \tilde{\mathbf{b}}_f + \mathbf{e}^T \times \tilde{\mathbf{b}}_x \\
+   &= \tilde{s}_f + \tilde{s}_x
+\end{array}
+\label{eqn:lci}
+\end{equation}

### 2017-09-25 aggregation/agg-lca-study.tex
Morning work on agg math

@@
 In practice, most LCA studies contain foreground information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with foreground models.
-In consideration of our objectives, a disclosure should be limited to the foreground of a study and exclude the background.
-The simplest conceivable foreground is simply what is given as the final demand, in other words the list of background processes invoked by the study and their activity levels.
-For clarity, I introduce a ``canonical'' final demand vector in which the first entry is 1 and all successive entries are 0:
-\begin{equation}
-\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}
-\label{eqn:canonical_y}
-\end{equation}
-The $A$ and $B$ matrices can be augmented with the $\mathbf{y]$ vector and a $\mathbf{0}$ vector respectively, to generate the same results using the canonical $\tilde{\mathbf{y}}$:
+To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.
+Because background databases are prepared independently of any study, their computational results do not vary across the studies in which they are used.
+However, the foreground systems themselves cannot be computed without the background.
+Put another way, the foreground depends on the background but the background does not depend on the foreground.
+This condition allows the processes in an LCIDB to be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.
+
+The simplest conceivable foreground is simply what is given as the final demand $\mathbf{y}$, in other words the list of background processes invoked by the study and their activity levels.
+For an LCA computation as expressed in Eq.~\ref{eqn:leontief}, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
 \begin{eqnarray}
@@
 \tilde{B} &= \left[\begin{array}{cc} \mathbf{0} & B\right]  \end{array}
+\tilde{\mathbf{y}} &=  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}\\
 \label{eqn:canonical}
 \end{eqnarray}
-This is intuitively true because the canonical $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.
 It can easily be shown that the first equality in Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.
-\textit{author's note: if need be, a proof for this equality can be provided in the electronic supplementary information.}
+\textit{author's note: a proof for this equality is provided in the electronic supplementary information.}  This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.
 
-Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database.
-To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.
-Because background databases are prepared independently of any study, their computational results do not vary across all the studies in which they are used.
-However, the foreground systems themselves cannot be computed without the background.
-Put another way, the foreground depends on the background but the background does not depend on the foreground.
-This condition allows the processes in an LCIDB to be sorted into foreground and background groups.
-The processes an LCIDB containing both a foreground and a background can then be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.
+Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.
+For clarity in this paper, I will call the augmented $\tilde{A}$ and $\tilde{B}$ matrices an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the \emph{canonical} functional unit.
 
-Thus a foreground study can be written in block-triangular form, using the canonical functional unit $\tilde{\mathbf{y}}$, as follows:
-\begin{equation}
-\tilde{A} = \left[\begin{array}{cc}
+In a canonical LCA foreground study, the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground.
+An LCA foreground study can be written in block-triangular form:
+\begin{eqnarray}
+\tilde{A} &= \left[\begin{array}{cc}
 A_f & 0 \\
@@
   \end{array}
-\right];\;\;\;  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right]
+\right]\\
+  \tilde{B} &= \left[\begin{array}{cc} B_f & B   \end{array}\right]
 \label{eqn:foreground}
-\end{equation}
+\end{eqnarray}
 

### 2017-10-17 aggregation/agg-JIE.tex
Morning work on agg math
foreground work

@@
 \input{psstyle}
-\input{draw-matrices}
+\input{../manuscript/draw-matrices}
 
-\title{Structured Representation of Aggregated Product System Models in Life Cycle Assessment}
+\title{Aggregated Product System Models in Life Cycle Assessment: Transparency versus Privacy}
 \author{Brandon Kuczenski\\
@@
   \maketitle
-  %\emph{For submission to the Journal of Industrial Ecology; February 2017}
+  \emph{For submission to the Journal of Industrial Ecology; October 2017}
 
@@
 
+  {\red http://pubs.acs.org/doi/pdf/10.1021/acs.est.5b01558}
+
   \input{agg-intro}
 
-  \input{publication-objective}
+  \input{agg-objective}
 
-  \input{publication-method}
+  \input{agg-basic-lca}
 
-  \input{publication-components}
+  \input{agg-lca-study}
 
-  \input{publication-foreground}
+  %% \input{publication-foreground}
 
-  \input{publication-example}
+  %% \input{publication-example}
 
-  \input{publication-discussion}
+  %% \input{publication-discussion}
 
-  \input{publication-conclusion}
+  %% \input{publication-conclusion}
 
-  \input{publication-acknowledgments}
+  %% \input{publication-acknowledgments}
 
@@
 
-%\mainpaper
-\supporting
+\mainpaper
+%\supporting
 

### 2017-10-17 aggregation/agg-abstract.tex

--- 2017-04-01 manuscript/publication-abstract.tex
@@
   knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of life cycle inventory models.
-This article presents a framework for LCA study authors to publish inventory models in a structured format based on a mathematical formulation of a \emph{foreground study}.
-Using linked semantic data as a foundation, the publication framework %offers a mechanism to provide
-  provides for both machine readability and unambiguous human interpretation of published studies.
-To implement the framework, the LCA community should find ways to enable study authors and data users to make common reference to shared data sources, including background databases, elementary flows, and impact assessment methods, that are widely used in many studies.
-The framework contributes to ongoing efforts within industrial ecology to improve the reproducibility and verifiability of scholarly works, and if implemented, plots a course toward distributed, platform-independent computation and validation of LCA results.
+This article presents a framework for computing aggregation results in LCA $\ldots$
 \end{abstract}

### 2017-10-17 aggregation/agg-basic-agg.tex

\subsection{The Foreground}

The foreground of an LCA study is a collection of nodes that represent points in the inventory model where flows are exchanged.  The foreground is modelled as a weighted directed graph in which there is a 1:1 correspondence between nodes and product flows.

The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.  These nodes typically represent unit processes, but they can also indicate exchanges across an implicit system boundary that encloses the nodes.  If a column in $A_f$ is empty, then that is an indication that the corresponding product flow is excluded from the foreground.  In order for this flow to be accounted for in the model, it must either be ``terminated'' in the dependency matrix $A_d$, or represented in the corresponding column of $B_f$ as an emission.  Flows whose entire column consists of 0s throughout $A_f$, $A_d$, and $B_f$ are ``cut offs'' that exit the model with zero burdens.

\input{fig_fragments}

Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e. any allocated supply-and-use table, but some fundamental designs are common.  Some simple foreground models are shown in Figure~\ref{fig:fragments}.  The first (a) is a sequential model, in which each node requires one foreground input and generates one output. This model is equivalent to a ``gate to gate'' model.  Here the weights $k_i$ indicate the amount of the preceding reference flow that is required by the subsequent node.  Figure~\ref{fig:fragments}(b) shows an additive model, in which the outputs of several foreground nodes are added together, equivalent to a ``mixer'' or a horizontal average.  In this arrangement the weights %represent the relative weights of each input, and
should add up to a unit output of the reference node.

add a cyclic model

\subsubsection{Model fragments and subsystem boundaries}

The nodes that make up the foreground of an LCA study describe an implicit system boundary

Most complex

\section{Preparing and Reviewing Aggregation Results}

### 2017-10-19 aggregation/agg-disclosure.tex
Work on foreground + disclosure

\subsection{Disclosure of Study Design}

The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.  Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$, are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.  These three submatrices together are referred to as the product system model, or PSM.

This set of three submatrices can be used to represent the product system model; thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.  Recognizing that all three matrices are likely to be sparse (that is, that most of the entries are zero), the most efficient disclosure would be to only report non-zero entries.  A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.  Thus, a formal representation of the PSM can be stated in six parts:
\begin{enumerate}[label=\alph*.]
\item\label{itm:fg} An ordered list of foreground nodes (rows/columns of $A_f$);
\item\label{itm:bg} An ordered list of background flows, making reference to a particular background database (rows of $A_d$, mapping to columns of $B_x$);
\item\label{itm:em} An ordered list of exchanges with environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
\item A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
\item a list of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
\item a list of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
\end{enumerate}

For a disclosure that includes the computation of an LCIA indicator or indicators, a seventh part is required:
\begin{enumerate}[resume]
\item the identity of the LCIA indicator or indicators computed (identity of $\mathbf{e}$).
\end{enumerate}

A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the paper):
\begin{itemize}
  \item The result is \textbf{computable} using Eq.~\ref{eqn:study};
  \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references to $B_x$ and $\mathbf{e}$;
  \item The disclosure includes only information for which the author can claim \textbf{authority};
  \item The disclosure includes only information for which the study is the \textbf{primary} source.
\end{itemize}
As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows.  If LCIA indicator results are included, the author must also unambiguously identify the method computed.  Implications of this requirement are discussed below.

For each row in $A_d$ containing a non-zero entry, the author must unambiguously identify the version of the database used, as well as the exact process and reference flow used from the background database; the dimension (reference unit) of each reference flow must be specified; and the sign of the numeric entry in $A_d$ must be consistent with the implementation of the process in the background database.  For the reader's part, the reader must be able to obtain either the column of $B_x$, or the scalar value that

Similarly, for each row in $B_f$ containing a non-zero entry, the author must unambiguously identify the substance being exchanged with the environment, the compartment or context into which it is being exchanged, and

\subsection{Foreground Aggregation}

Often the contents of the foreground

An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
\begin{equation}
\tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
\label{eqn:inv}
\end{equation}
This vector can be used to describe an \emph{aggregated foreground}, which generates the same results as the fully expanded foreground:
\begin{equation}
\begin{array}{rl}
    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
\end{array}
\label{eqn:agg}
\end{equation}

These vectors summarize the contents of the foreground without disclosing its detailed structure.  The aggregated dependency vector $\tilde{\mathbf{a}}_d$ has the same dimension as the background database, and the aggregated emission vector $\tilde{\mathbf{b}}_f $has the same dimension as the elementary flow matrix.  The dependency vector can be transformed into an elementary flow vector by multiplying by $B_x$.  The result of the study $s$ is the sum of foreground and background impact scores:

\begin{equation}
\begin{array}{rl}
   \tilde{\mathbf{b}} & = \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x \\
   s &= \mathbf{e}^T \times \tilde{\mathbf{b}} \\
   & = \mathbf{e}^T \times \tilde{\mathbf{b}}_f + \mathbf{e}^T \times \tilde{\mathbf{b}}_x \\
   &= \tilde{s}_f + \tilde{s}_x
\end{array}
\label{eqn:lci}
\end{equation}

\subsubsection{Model fragments and subsystem boundaries}

The nodes that make up the foreground of an LCA study describe an implicit system boundary

Most complex

\section{Preparing and Reviewing Aggregation Results}

### 2017-10-19 aggregation/agg-lca-study.tex
New foreground graph
Work on foreground + disclosure

@@
-\section{Foreground Study}
+\subsection{Foreground Study}
 
@@
 For an LCA computation as expressed in Eq.~\ref{eqn:leontief}, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
-\begin{eqnarray}
-s &= \mathbf{e}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
- &= \mathbf{e}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde\mathbf{y}\\
-\tilde{A} &= \left[\begin{array}{cc}
+\begin{equation}
+  \begin{array}{cc}
+s =& \mathbf{e}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
+ =& \mathbf{e}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
+  \end{array}
+  \label{eqn:canonical}
+\end{equation}
+where $\tilde{A} = \left[\begin{smallmatrix}
  0 & 0 \\
  \mathbf{y} & A
-  \end{array}\right]\\
-\tilde{B} &= \left[\begin{array}{cc} \mathbf{0} & B\right]  \end{array}
-\tilde{\mathbf{y}} &=  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}\\
-\label{eqn:canonical}
-\end{eqnarray}
-It can easily be shown that the first equality in Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.
+  \end{smallmatrix}\right]$, $\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
+It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.
 \textit{author's note: a proof for this equality is provided in the electronic supplementary information.}  This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.
@@
 \end{equation}
-
-This formulation shows that given access to the background database $B_x$ and the characterization vector $\mathbf{e}$, the study results can be fully reproduced using only the submatrices $A_f$, $A_d$, and $B_f$.
 Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$, having been subsumed into $B_x$, represented by an identity matrix $I$.
-
-An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
-\begin{equation}
-\tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
-\label{eqn:inv}
-\end{equation}
-This vector can be used to describe an \emph{aggregated foreground}, which generates the same results as the fully expanded foreground:
-\begin{equation}
-\begin{array}{rl}
-    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
-    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
-    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
-\end{array}
-\label{eqn:agg}
-\end{equation}
-
-These vectors summarize the contents of the foreground without disclosing its detailed structure.
-The aggregated dependency vector $\tilde{\mathbf{a}}_d$ has the same dimension as the background database, and the aggregated emission vector $\tilde{\mathbf{b}}_f $has the same dimension as the elementary flow matrix.
-The dependency vector can be transformed into an elementary flow vector by multiplying by $B_x$.
-The result of the study $s$ is the sum of foreground and background impact scores:
-
-\begin{equation}
-\begin{array}{rl}
-   \tilde{\mathbf{b}} & = \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x \\
-   s &= \mathbf{e}^T \times \tilde{\mathbf{b}} \\
-   & = \mathbf{e}^T \times \tilde{\mathbf{b}}_f + \mathbf{e}^T \times \tilde{\mathbf{b}}_x \\
-   &= \tilde{s}_f + \tilde{s}_x
-\end{array}
-\label{eqn:lci}
-\end{equation}

### 2017-10-19 aggregation/agg-JIE.tex
New foreground graph
Work on foreground + disclosure
Work on study review

@@
 
-\title{Aggregated Product System Models in Life Cycle Assessment: Transparency versus Privacy}
+\title{Aggregated Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy}
 \author{Brandon Kuczenski\\
@@
 %\usepackage{lineno}
-\let\showgrid\relax
+%\let\showgrid\relax
 \let\oldtimes\times
@@
 
-  \input{agg-intro}
+  \input{agg-intro}  % motivations
 
-  \input{agg-objective}
+\section{Formalizing LCA Studies}
 
-  \input{agg-basic-lca}
+  \input{agg-objective}  % Authors, Readers, review objectives, disclosure constraints
 
-  \input{agg-lca-study}
+  \input{agg-basic-lca}  % basic Leontief model
+
+  \input{agg-lca-study}  % foreground JLCA model
+
+  \input{agg-disclosure} % disclosure requirements satisfied
+
+\section{Reviewing Product System Models}
+
+Objectives in critical review vary with the study and the reviewer, but they generally include consistency with the ISO 14044 standard, the validity of methods, appropriateness of data used, correctness of interpretations, and transparency of reporting \citep[Section 6.1]{iso14044}.
+Although the object of critical review is typically a written report, the use of a structured disclosure of study design introduces the possibility that reviewers can inspect the model itself and even verify the computations and reported results.
+This section discusses how a disclosure in the format described above can enable robust review of a product system model.
+Note that, for purposes of critical review, the objective for data privacy can be omitted because the reviewer may be granted privileged access to confidential data.
+
+  \input{agg-foreground} % Af and graph/matrix equivalency
+
+  \input{agg-deps}       % Ad and Bf
+
+\section{Protecting Privacy through Aggregation}
+
+  \input{agg-agg}        % notation for full aggregation
+
+%  \input{agg-pagg}       % notation for partial aggregation
+
+%  \input{agg-forms}      % the elusive Figure 4
+
+\section{Discussion}
+
+%  \input{agg-review}     % review objectives
+
+%  \input{agg-privacy}    % disclosure constraints
+
+%  \input{agg-semantic}   % the need for stable references
+
+\section{Conclusion}
 

### 2017-10-19 aggregation/agg-agg.tex
Work on study review

--- 2017-10-19 aggregation/agg-disclosure.tex
@@
-\subsection{Disclosure of Study Design}
-
-The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.
-Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$, are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.
-These three submatrices together are referred to as the product system model, or PSM.
-
-This set of three submatrices can be used to represent the product system model; thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.
-Recognizing that all three matrices are likely to be sparse (that is, that most of the entries are zero), the most efficient disclosure would be to only report non-zero entries.
-A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.
-Thus, a formal representation of the PSM can be stated in six parts:
-\begin{enumerate}[label=\alph*.]
-\item\label{itm:fg} An ordered list of foreground nodes (rows/columns of $A_f$);
-\item\label{itm:bg} An ordered list of background flows, making reference to a particular background database (rows of $A_d$, mapping to columns of $B_x$);
-\item\label{itm:em} An ordered list of exchanges with environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
-\item A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
-\item a list of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
-\item a list of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
-\end{enumerate}
-
-For a disclosure that includes the computation of an LCIA indicator or indicators, a seventh part is required:
-\begin{enumerate}[resume]
-\item the identity of the LCIA indicator or indicators computed (identity of $\mathbf{e}$).
-\end{enumerate}
-
-A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the paper):
-\begin{itemize}
-  \item The result is \textbf{computable} using Eq.~\ref{eqn:study};
-  \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
-  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references to $B_x$ and $\mathbf{e}$;
-  \item The disclosure includes only information for which the author can claim \textbf{authority};
-  \item The disclosure includes only information for which the study is the \textbf{primary} source.
-\end{itemize}
-As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows.
-If LCIA indicator results are included, the author must also unambiguously identify the method computed.
-Implications of this requirement are discussed below.
-
-For each row in $A_d$ containing a non-zero entry, the author must unambiguously identify the version of the database used, as well as the exact process and reference flow used from the background database; the dimension (reference unit) of each reference flow must be specified; and the sign of the numeric entry in $A_d$ must be consistent with the implementation of the process in the background database.
-For the reader's part, the reader must be able to obtain either the column of $B_x$, or the scalar value that
-
-Similarly, for each row in $B_f$ containing a non-zero entry, the author must unambiguously identify the substance being exchanged with the environment, the compartment or context into which it is being exchanged, and
-
 \subsection{Foreground Aggregation}

### 2017-10-19 aggregation/agg-deps.tex
Work on study review

\subsection{Dependencies, Emissions, and Cutoffs}

The foreground matrix only describes the structure of the PSM; it does not provide any information about inventory or impacts.  The inventory requirements and environmental burdens associated with the model are all described in the dependency matrix $A_d$ and the foreground emission matrix $B_f$.  The dependency matrix shows how foreground nodes depend on activities provided by processes in a background database; the emission matrix shows direct exchanges with the environment.  In both cases, the respective columns of $A_d$ and $B_f$ correspond with columns in $A_f$ and denote dependencies and/or emissions of the corresponding nodes.  Review of the dependencies of foreground nodes must include review of the selection of background processes to which they are connected (i.e. the identities of the columns of $B_x$ referred to in disclosure item~\ref{itm:bg}, the list of background flows). Likewise, the identities of the emissions (entries in disclosure item~\ref{itm:em}) must also be reviewed.

Interpretation of the contents of these matrices thus hinges on the contents of $A_f$.  If a column in $A_f$ is nonempty, then the corresponding node has exchanges with other nodes in the foreground.  If columns of $A_d$ and/or $B_f$ are empty, then the node is simply a pass-through process that serves to connect different foreground elements.  However, if $A_d$ and/or $B_f$ have nonzero entries in that column, then that indicates some transformation process or emission is occurring at that point in the foreground.

On the other hand, if a column in $A_f$ is empty, then that is an indication that the corresponding product flow is crossing the foreground system boundary.  In order for this flow to contribute any impacts, it must either be ``terminated'' in the dependency matrix, or represented as an emission.

Flows whose entire column consists of 0s throughout $A_f$, $A_d$, and $B_f$ are ``cut-offs'' that exit the model with zero burdens.  Specifying cut-off criteria and reporting of cut-off flows is an important part of system boundary definition.  Thus, reviewing cut-off flows (including evaluating their significance) is a crucial part of review that is facilitated by having a structured disclosure of the model.

### 2017-10-19 aggregation/agg-disclosure.tex
Work on study review

@@
 
-This set of three submatrices can be used to represent the product system model; thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.
+This set of three submatrices can be used to represent the PSM; thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.
 Recognizing that all three matrices are likely to be sparse (that is, that most of the entries are zero), the most efficient disclosure would be to only report non-zero entries.
 A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.
-Thus, a formal representation of the PSM can be stated in six parts:
-\begin{enumerate}[label=\alph*.]
+Thus, a formal disclosure of the PSM can be stated in six parts:
+\begin{enumerate}[label={\em d-\roman*}., ref={\em d-\roman*}]
 \item\label{itm:fg} An ordered list of foreground nodes (rows/columns of $A_f$);
@@
 \item\label{itm:em} An ordered list of exchanges with environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
-\item A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
-\item a list of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
-\item a list of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
+\item\label{itm:af} A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
+\item\label{itm:ad} a list of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
+\item\label{itm:bf} a list of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
 \end{enumerate}
 
-For a disclosure that includes the computation of an LCIA indicator or indicators, a seventh part is required:
-\begin{enumerate}[resume]
-\item the identity of the LCIA indicator or indicators computed (identity of $\mathbf{e}$).
-\end{enumerate}
-
-A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the paper):
+A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):
 \begin{itemize}
@@
 \end{itemize}
-As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows.
-If LCIA indicator results are included, the author must also unambiguously identify the method computed.
-Implications of this requirement are discussed below.
-
-For each row in $A_d$ containing a non-zero entry, the author must unambiguously identify the version of the database used, as well as the exact process and reference flow used from the background database; the dimension (reference unit) of each reference flow must be specified; and the sign of the numeric entry in $A_d$ must be consistent with the implementation of the process in the background database.
-For the reader's part, the reader must be able to obtain either the column of $B_x$, or the scalar value that
-
-Similarly, for each row in $B_f$ containing a non-zero entry, the author must unambiguously identify the substance being exchanged with the environment, the compartment or context into which it is being exchanged, and
-
-\subsection{Foreground Aggregation}
-
-Often the contents of the foreground
-
-An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
-\begin{equation}
-\tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
-\label{eqn:inv}
-\end{equation}
-This vector can be used to describe an \emph{aggregated foreground}, which generates the same results as the fully expanded foreground:
-\begin{equation}
-\begin{array}{rl}
-    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
-    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
-    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
-\end{array}
-\label{eqn:agg}
-\end{equation}
-
-These vectors summarize the contents of the foreground without disclosing its detailed structure.
-The aggregated dependency vector $\tilde{\mathbf{a}}_d$ has the same dimension as the background database, and the aggregated emission vector $\tilde{\mathbf{b}}_f $has the same dimension as the elementary flow matrix.
-The dependency vector can be transformed into an elementary flow vector by multiplying by $B_x$.
-The result of the study $s$ is the sum of foreground and background impact scores:
-
-\begin{equation}
-\begin{array}{rl}
-   \tilde{\mathbf{b}} & = \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x \\
-   s &= \mathbf{e}^T \times \tilde{\mathbf{b}} \\
-   & = \mathbf{e}^T \times \tilde{\mathbf{b}}_f + \mathbf{e}^T \times \tilde{\mathbf{b}}_x \\
-   &= \tilde{s}_f + \tilde{s}_x
-\end{array}
-\label{eqn:lci}
-\end{equation}
-
-\subsubsection{Model fragments and subsystem boundaries}
-
-The nodes that make up the foreground of an LCA study describe an implicit system boundary
-
-Most complex
-
-\section{Preparing and Reviewing Aggregation Results}
+As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{e}$.
+These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.
+However, the the challenges of operationalizing this requirement are tremendously complex.
+Some implications of the requirement for shared agreement are considered in the Discussion.

### 2017-10-19 aggregation/agg-foreground.tex
Work on foreground + disclosure
Work on study review

--- 2017-10-17 aggregation/agg-basic-agg.tex
@@
-\subsection{The Foreground}
+\subsection{Interpreting the Foreground}
 
 The foreground of an LCA study is a collection of nodes that represent points in the inventory model where flows are exchanged.
+The foreground matrix $A_f$ describes in precise terms what components are included in the product system and how they connect to one another.
 The foreground is modelled as a weighted directed graph in which there is a 1:1 correspondence between nodes and product flows.
+The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.
+These nodes typically represent unit processes, but they can also indicate exchanges across an implicit system boundary that encloses the foreground.
 
-The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.
-These nodes typically represent unit processes, but they can also indicate exchanges across an implicit system boundary that encloses the nodes.
-If a column in $A_f$ is empty, then that is an indication that the corresponding product flow is excluded from the foreground.
-In order for this flow to be accounted for in the model, it must either be ``terminated'' in the dependency matrix $A_d$, or represented in the corresponding column of $B_f$ as an emission.
-Flows whose entire column consists of 0s throughout $A_f$, $A_d$, and $B_f$ are ``cut offs'' that exit the model with zero burdens.
+A reviewer with access to items \ref{itm:fg} and \ref{itm:af} in the disclosure would be able to construct $A_f$ and automatically create a process-flow diagram, annotated with information about each node.
+Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e.
+any allocated supply-and-use table, but some fundamental designs are common.
+Some simple foreground models are shown in Figure~\ref{fig:fragments}.
 
@@
 
-Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e.
-any allocated supply-and-use table, but some fundamental designs are common.
-Some simple foreground models are shown in Figure~\ref{fig:fragments}.
 The first (a) is a sequential model, in which each node requires one foreground input and generates one output.
@@
 should add up to a unit output of the reference node.
+Finally, Figure~\ref{fig:fragments}(c) shows a foreground model with a cyclic dependency, where some of the reference output is consumed by another foreground node.
 
-add a cyclic model
-
-\subsubsection{Model fragments and subsystem boundaries}
-
-The nodes that make up the foreground of an LCA study describe an implicit system boundary
-
-Most complex
-
-\section{Preparing and Reviewing Aggregation Results}
+A typical PSM may contain multiple fragmentary elements that are interconnected.
+An example of foreground composed of several fragments is illustrated in Figure~\ref{fig:fragments}(d).
+Here, the nodes labeled 1--5 represent one fragment, which generates the foreground's canonical reference flow $\tilde{y}$.
+This fragment requires two interior flows from separate fragments ($y_0$ and $y_1$), and has two unconnected flows (4 and 5).
+The reference $y_0$ is supplied by a second fragment, made up of nodes 6-8.
+The reference $y_1$ is supplied by another fragment made of only one node (9).
+The reference flow $y_1$ is consumed in two different places by the other fragments.

### 2017-10-19 aggregation/agg-semantic.tex
Work on study review

For each row in $A_d$ containing a non-zero entry, the author must unambiguously identify the version of the database used, as well as the exact process and reference flow used from the background database; the dimension (reference quantity or unit) of each reference flow must be specified; and the sign of the numeric entry in $A_d$ must be consistent with the implementation of the process in the background database.

Similarly, for each row in $B_f$ containing a non-zero entry, the author must unambiguously identify the substance being exchanged with the environment, the compartment or context into which it is being exchanged, and the reference quantity or unit associated with the flow.  Sign consistency must also be assured.  If LCIA indicator results are included, the author must also unambiguously identify the method computed (identity of $\mathbf{e}$).

For the reader's part, the reader must be able to produce an $\mathbf{e}$ vector that matches the one used by the author and identify which entries correspond with which

either the column of $B_x$, or the scalar value that corresponds to the dot product of this column with $\mathbf{e}$

### 2017-10-20 aggregation/agg-agg.tex
move basic agg back into the study definition

@@
 Often the contents of the foreground
-
-An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
-\begin{equation}
-\tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
-\label{eqn:inv}
-\end{equation}
-This vector can be used to describe an \emph{aggregated foreground}, which generates the same results as the fully expanded foreground:
-\begin{equation}
-\begin{array}{rl}
-    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
-    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
-    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
-\end{array}
-\label{eqn:agg}
-\end{equation}
 
@@
 \end{equation}
-
-\subsubsection{Model fragments and subsystem boundaries}
-
-The nodes that make up the foreground of an LCA study describe an implicit system boundary
-
-Most complex
-
-\section{Preparing and Reviewing Aggregation Results}

### 2017-10-20 aggregation/agg-disclosure.tex
move basic agg back into the study definition

@@
 Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$, are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.
-These three submatrices together are referred to as the product system model, or PSM.
+These three submatrices together can be considered a complete representation of the PSM.
 
-This set of three submatrices can be used to represent the PSM; thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.
+Thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.
 Recognizing that all three matrices are likely to be sparse (that is, that most of the entries are zero), the most efficient disclosure would be to only report non-zero entries.
@@
   \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
-  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references to $B_x$ and $\mathbf{e}$;
+  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references in \ref{itm:bg} and \ref{itm:em};
   \item The disclosure includes only information for which the author can claim \textbf{authority};

### 2017-10-20 aggregation/agg-intro.tex
move basic agg back into the study definition

--- 2017-09-18 manuscript/agg-intro.tex
@@
 At the core of an LCA study is a model of a ``product system,'' which represents a collection of ``processes $\ldots$ performing one or more defined functions'', that describes the life cycle of a product \citep{iso14044}.
-Product system models encode information about how products are manufactured, distributed, used by the consumer, and what happens to them after they are disposed.
+The product system model, or PSM, encodes information about how a product is manufactured, distributed, used by the consumer, and what happens to it after it is disposed.
 A product system can be divided into a foreground, which denotes the portions of the life cycle whose operations are directly modeled by the study, and a background, which represents the global industrial system \citep{SETAC_inventory_1998}.
@@
 Although there has been considerable effort to normalize and harmonize 
-LCI database design methodology and data interchange \citep{UNEP_2011, JRC_ILCD_ELCD_2013, Mila_e_Canals_2015, Ingwersen_JLCA_2015}, less attention has been paid to harmonizing product system models.
+LCI database design methodology and data interchange \citep{UNEP_2011, JRC_ILCD_ELCD_2013, Mila_e_Canals_2015, Ingwersen_JLCA_2015}, less attention has been paid to harmonizing PSMs.
 These models are often highly complex and include countless modeling decisions, approximations, and assumptions made by a study author \citep{Lloyd2007, reap2008_I}.
@@
 
-An LCA study result is ultimately an assertion that for some constructed product model, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.
-But if the product model itself is ambiguously stated, then the significance of the result is ambiguous as well.
+An LCA study result is ultimately an assertion that for some constructed PSM, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.
+But if the PSM itself is ambiguously stated, then the significance of the result is ambiguous as well.
 For a computational model to have scientific significance, its results need to be both verifiable and replicable by other scientists \citep{Fomel_2009, Mesirov_2010}.
 But in the case of LCA, the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.
-When a study is reviewed, it is often the written report and \emph{not} the product system model itself that is the object of review.
-In the end, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the product model itself is not available.
+When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.
+In the end, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.
 
-A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the product system model.
-Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the product system model and thus, no automatic mechanisms for interpretation or review of published LCA studies.
+A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.
+Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the PSM and thus, no automatic mechanisms for interpretation or review of published LCA studies.
 

### 2017-10-20 aggregation/agg-lca-study.tex
Work on study review
move basic agg back into the study definition

@@
 where $\tilde{A} = \left[\begin{smallmatrix}
- 0 & 0 \\
+ 0 & \mathbf{0} \\
  \mathbf{y} & A
@@
 Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.
-For clarity in this paper, I will call the augmented $\tilde{A}$ and $\tilde{B}$ matrices an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the \emph{canonical} functional unit.
-
-In a canonical LCA foreground study, the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground.
-An LCA foreground study can be written in block-triangular form:
-\begin{eqnarray}
-\tilde{A} &= \left[\begin{array}{cc}
+An LCA study augmented with foreground information can be written in block-triangular form:
+\begin{equation}
+\tilde{A} = \left[\begin{array}{cc}
 A_f & 0 \\
@@
   \end{array}
-\right]\\
-  \tilde{B} &= \left[\begin{array}{cc} B_f & B   \end{array}\right]
+\right];\;\;\;
+  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right]
 \label{eqn:foreground}
-\end{eqnarray}
-
-\input{fig_foreground}
-In this case, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.
+\end{equation}
+Here, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.
 The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions.
 The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.
-As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.\footnote{It should be noted that this condition does not hold for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells.
-The approach presented here applies only to LCA studies in which the background system is invariant with respect to the foreground.
-This class includes the vast majority of LCA case studes, and very likely includes all case studies prepared using commercial LCA software.}
+As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.
+\input{fig_foreground}
+
+For clarity, I will call the augmented $\tilde{A}$ and $\tilde{B}$, in which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be a \emph{canonical LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.
+This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.
+However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2016}).
 
@@
 \end{equation}
-The proof of the equivalency of Eq.s\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary information to \cite{Kuczenski_JLCA_2015}.
+The proof of the equivalency of Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary information to \cite{Kuczenski_JLCA_2015}.
 In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining there pertains to the study foreground.
@@
 Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$, having been subsumed into $B_x$, represented by an identity matrix $I$.
+
+The activity levels of the foreground nodes are represented by $\tilde{\mathbf{x}}$, which is the result of $\tilde{\mathbf{y}}$ selecting the first column of $(I-A_f)^{-1}$.
+%% An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
+%% \begin{equation}
+%% \tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
+%% \label{eqn:inv}
+%% \end{equation}
+This vector can be used to describe an \emph{aggregated foreground}:
+\begin{equation}
+\begin{array}{rl}
+    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
+    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
+    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
+\end{array}
+\label{eqn:agg}
+\end{equation}
+The study result can also be computed from the aggregated foreground:
+\begin{equation}
+\begin{array}{rl}
+s & = \mathbf{e}^T \times( B_f + B_x \times A_d) \times \tilde{\mathbf{x}} \\
+  & = \mathbf{e}^T \times( \tilde{\mathbf{b}}_f + B_x \times\tilde{\mathbf{a}}_d) \\
+  & = \mathbf{e}^T \times( \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x) \\
+\end{array}
+\end{equation}

### 2017-10-21 aggregation/agg-agg.tex
basic aggregation; figure 4

@@
-\subsection{Foreground Aggregation}
+\subsection{Forms of Aggregation}
 
-Often the contents of the foreground
+Figure~\ref{fig:foreground} illustrates the LCA computation in various successive aggregation steps, from full matrices to vectors to a scalar value.
+The level of aggregation provided in a disclosure will constrain what review objectives can be met by the reader of a study.
+Different aggregation forms are illustrated in Figure~\ref{fig:aggregation}.
 
-These vectors summarize the contents of the foreground without disclosing its detailed structure.
-The aggregated dependency vector $\tilde{\mathbf{a}}_d$ has the same dimension as the background database, and the aggregated emission vector $\tilde{\mathbf{b}}_f $has the same dimension as the elementary flow matrix.
-The dependency vector can be transformed into an elementary flow vector by multiplying by $B_x$.
-The result of the study $s$ is the sum of foreground and background impact scores:
+\subsubsection{Unit Process Inventory}
 
+A unit process makes up a single shared column of $A_f$, $A_d$, and $B_f$, as depicted in Figure~\ref{fig:aggregation}(a).
+Documentation of unit processes is specified in ISO 14048, and several compliant formats exist, the most well-known being the ILCD and Ecospold XML formats.
+ISO 14048 specifies a distinction between intermediate and elementary flows, but does not include the concept of a model foreground.
+In a matrix representation, each unit process must have exactly one reference flow, which corresponds to its row and column in $A_f$.
+
+\input{fig_aggregation}
+
+\subsubsection{Complete Foreground}
+
+See Figure~\ref{fig:aggregation}(b).
+A foreground can be viewed as a linked set of unit processes.
+The collection of processes themselves make up the foreground of the study, and so exchanges among them are recorded in $A_f$, and exchanges outside the foreground are represented in $A_d$ or $B_f$.
+A complete foreground represents the full disclosure of a model, and a reader can replicate and extend it independently.
+Any co-production treatment, such as an allocation or system expansion, can only be fully expressed as a foreground model, because it intrinsically involves a collection of linked unit processes.
+
+\subsubsection{Aggregated Foreground}
+
+See Figure~\ref{fig:aggregation}(c).
+Any collection of foreground nodes can also be expressed in aggregated form by performing the computations in Eq.~\ref{eqn:agg}.
+A publication of the aggregated foreground is identical to the publication of a unit process, except that the result was derived through aggregation.
+As in the unit process case, the aggregated foreground has exactly one reference flow, which is $\tilde{\mathbf{y}}$.
+The aggregated foreground still reports explicit links between the model and a background database.
+
+\subsubsection{Partial Background Aggregation}
+
+See Figure~\ref{fig:aggregation}(d).  Using this approach, the author can replace portions of the
+The aggregated dependency vector is split into two parts that sum to the original:
 \begin{equation}
-\begin{array}{rl}
-   \tilde{\mathbf{b}} & = \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x \\
-   s &= \mathbf{e}^T \times \tilde{\mathbf{b}} \\
-   & = \mathbf{e}^T \times \tilde{\mathbf{b}}_f + \mathbf{e}^T \times \tilde{\mathbf{b}}_x \\
-   &= \tilde{s}_f + \tilde{s}_x
-\end{array}
-\label{eqn:lci}
+ \tilde{\mathbf{a}}_d = \tilde{\mathbf{a}}_{d,priv} + \tilde{\mathbf{a}}_d'
 \end{equation}
+The disclosed dependencies $\tilde{\mathbf{a}}_d'$ are reported, and the private dependencies $\tilde{\mathbf{a}}_{d,priv}$ are replaced with an aggregated background inventory derived from the background database, $\tilde{\mathbf{b}}_{x,agg}$.
+This approach can be used when a reader lacks access to the background inventory sets referred to in $\tilde{\mathbf{a}}_{d,priv}$, or when the author wishes not to disclose the private dependencies.
+Foreground emissions $\tilde{\mathbf{b}}_f$ can be reported separately from the aggregation result, but often in current practice they are not distinguished from the background flows.
+
+\subsubsection{Full Background Aggregation and LCI}
+
+See Figure~\ref{fig:aggregation}(e) and (f).
+At this level of aggregation, the entire dependency vector is replaced with an aggregated life cycle inventory vector $\tilde{\mathbf{b}}_x$ derived from the background LCIDB.
+The reader no longer requires access to any background database to perform the computation, but all dependency information is concealed.
+Again, foreground emissions are often not distinguished from the aggregation result.
+The life cycle inventory $\tilde{\mathbf{b}}$ provides the most aggregated form of the study that can still be independently validated with an external characterization vector $\mathbf{e}$.

### 2017-10-21 aggregation/agg-basic-lca.tex
Work on foreground + disclosure
basic aggregation; figure 4

@@
-\section{Mathematical Formulation of LCA}
+\subsection{Mathematical Formulation of LCA}
 
@@
 An LCA computation is performed over a collection of linked unit processes, each of which describes a single operation or reference product.
-The task of LCA can be described in two steps: determining the required outputs of each process; and determining the environmental impacts of each process.
+The task of LCA can be described in two steps: making an inventory of environmental interventions; and determining their environmental impacts.
 If the processes are assumed to be linear, the problem can be written in terms of matrix algebra.
 There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Suh_2002, Pauliuk_2015}.
-Although the two models are understood to be equivalent, the input-output approach is computationally more efficient because it lends itself to iterative solution \citep{Peters_2007}.
+Although the two models are understood to be equivalent, the input-output approach is computationally more efficient in some applications because it lends itself to iterative solution \citep{Peters_2007}.
 That formulation may be stated:
@@
 In this model, $\mathbf{y}$ is the externally specified final demand, which is the product or service whose delivery is to be assessed, and $s$ is the numerical impact score or category indicator.
+For simplicity, we will consider the computation of a single impact result.
+However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.
 
@@
 \end{itemize}
-For simplicity, we will consider the computation of a single impact result.
-However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.
 

### 2017-10-21 aggregation/agg-lca-study.tex
basic aggregation; figure 4

@@
 
+\input{fig_foreground}
 Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.
@@
 As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.
-\input{fig_foreground}
 

### 2017-10-21 aggregation/agg-objective.tex
Morning work on agg math
Work on foreground + disclosure
basic aggregation; figure 4

@@
-\section{Formalizing LCA Study Aggregation}
+\subsection{Objectives}
 
-The task of computing life cycle impact assessment (LCIA) results from a product system model is often described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.
+The task of computing life cycle impact assessment (LCIA) results from a product system model can be described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.
 Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh_2005, Peters_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.
 
-The practice of reporting the results of an LCA study implies the existence of two parties with different information: the study Author has full knowledge of the product system model; and the Reader only has knowledge of what the Author has disclosed.
-Obviously, the level of detail of the disclosure will determine how useful the computation will be to the Reader.
-In this context, the Reader will want to know the answers to various questions about the model, which we will call \emph{review objectives}.
-At the same time the Author will be limited in what she can disclose, which we will call \emph{disclosure constraints}.
+In the context of this paper, an ``LCA Study'' should be understood as any aggregation of data under consideration, from a single-operation unit process to a full ISO-style product LCA.
+The practice of reporting the results of an LCA study implies the existence of two parties with different information: the study author has full knowledge of the product system model; and the Reader only has knowledge of what the author has disclosed.
+Obviously, the level of detail of the disclosure will determine how useful it is to the Reader.
+In this context, the reader will want to know the answers to various questions about the model, which we will call \emph{review objectives}.
+At the same time the author may be limited in what she can disclose, which we will call \emph{disclosure constraints}.
 The purpose of the framework is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.
 
-In the context of this paper, an ``LCA Study'' should be understood as any aggregation of data under consideration, from a single-operation unit process to a full ISO-style product LCA.
-Furthermore, the aggregation can be computed at any level: a foreground (inventory) result; a background (accumulated life cycle inventory) result; or a computed LCIA score.
+%Furthermore, the aggregation can be computed at any level: a foreground (inventory) result; a background (accumulated life cycle inventory) result; or a computed LCIA score.
 These different levels are discussed in detail in the following sections.

### 2017-10-21 aggregation/agg-agg.tex
fig labels

@@
 
-A unit process makes up a single shared column of $A_f$, $A_d$, and $B_f$, as depicted in Figure~\ref{fig:aggregation}(a).
+A unit process makes up a single shared column of $A_f$, $A_d$, and $B_f$, as depicted in Figure~\ref{fig:aggregation}a.
 Documentation of unit processes is specified in ISO 14048, and several compliant formats exist, the most well-known being the ILCD and Ecospold XML formats.
@@
 
-See Figure~\ref{fig:aggregation}(b).
+See Figure~\ref{fig:aggregation}b.
 A foreground can be viewed as a linked set of unit processes.
@@
 
-See Figure~\ref{fig:aggregation}(c).
+See Figure~\ref{fig:aggregation}c.
 Any collection of foreground nodes can also be expressed in aggregated form by performing the computations in Eq.~\ref{eqn:agg}.
 A publication of the aggregated foreground is identical to the publication of a unit process, except that the result was derived through aggregation.
-As in the unit process case, the aggregated foreground has exactly one reference flow, which is $\tilde{\mathbf{y}}$.
-The aggregated foreground still reports explicit links between the model and a background database.
+As in the unit process case, the aggregated foreground has exactly one reference flow, the canonical functional unit $\tilde{\mathbf{y}}$.
+The aggregated foreground still reports explicit links between the model and a background database and can thus be used to review data set selection.
 
@@
 
-See Figure~\ref{fig:aggregation}(d).  Using this approach, the author can replace portions of the
+See Figure~\ref{fig:aggregation}d.  Using this approach, the author can replace portions of the
 The aggregated dependency vector is split into two parts that sum to the original:
@@
 
-See Figure~\ref{fig:aggregation}(e) and (f).
+See Figure~\ref{fig:aggregation}e and f.
 At this level of aggregation, the entire dependency vector is replaced with an aggregated life cycle inventory vector $\tilde{\mathbf{b}}_x$ derived from the background LCIDB.
 The reader no longer requires access to any background database to perform the computation, but all dependency information is concealed.
-Again, foreground emissions are often not distinguished from the aggregation result.
+Again, foreground emissions are often not distinguished from the aggregation result (Figure~\ref{fig:aggregation}f).
 The life cycle inventory $\tilde{\mathbf{b}}$ provides the most aggregated form of the study that can still be independently validated with an external characterization vector $\mathbf{e}$.

### 2017-10-23 aggregation/proofs.tex
proofs. ugh.

\section{Mathematical Relations}

\subsection{Final Demand Augmentation}

This is a proof of Equation~2 in the manuscript.

Given a life cycle inventory database (LCIDB) containing input-output matrix $A$ and emission matrix $B$, and a final demand vector $\mathbf{y}$, construct an augmented inventory database
$\tilde{A} = \left[\begin{smallmatrix} 0 & \mathbf{0} \\ \mathbf{y} & A \end{smallmatrix}\right]$,
$\tilde{B} = [ \mathbf{0},\, B ]$, and canonical functional unit $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.  Show that
\begin{equation} B\times\left(I-A\right)^{-1}\times\mathbf{y} = \tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
\end{equation}

We begin by constructing $(I-\tilde{A})^{-1}$.  Because the matrix is block triangular, we only need to determine the value of $\mathbf{q}$ in the matrix below:
\begin{eqnarray}
  I &=&  (I-\tilde{A})^{-1}\times(I-\tilde{A})\\
  \left[\begin{array}{cc} 1 & \mathbf{0} \\
    \mathbf{0} & I
  \end{array}\right] &=&
  \left[\begin{array}{cc} 1 & \mathbf{0} \\
      \mathbf{q} & (I-A)^{-1}
  \end{array}\right]\times
  \left[\begin{array}{cc} 1 & \mathbf{0} \\
      -\mathbf{y} & (I-A)
        \end{array}\right]
\end{eqnarray}
It can be seen that $\mathbf{q}-(I-A)^{-1}\times\mathbf{y}=0$ is a necessary condition to satisfy the equality, and thus $\mathbf{q}=(I-A)^{-1}\times\mathbf{y}$.

We know that $\tilde{\mathbf{y}}$ selects the first column of its argument, so:
\begin{equation}
  (I-\tilde{A})^{-1}\times\tilde{\mathbf{y}} = \left[\begin{array}{c}
      1\\
      \mathbf{q}
    \end{array}\right]
\end{equation}

And therefore

\begin{eqnarray}
  \tilde{B}\times(I-\tilde{A})^{-1}\times\tilde{\mathbf{y}} &=& \left[\begin{array}{cc}\mathbf{0}&B\end{array}\right]\times\left[\begin{array}{c}
      1\\
      \mathbf{q}
    \end{array}\right]\\
   &=&B\times\mathbf{q}\\
   &=&B\times(I-A)^{-1}\times\mathbf{y}
\end{eqnarray}
thus completing the proof.

\subsection{``Flattening'' the background database}

Given a canonical LCA foreground study, show that the following formulations are equivalent:
\begin{eqnarray}
\tilde{A} = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A
  \end{array}
\right];&
  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right] \\
 \tilde{A}_{flat} = \left[\begin{array}{cc} 
A_f &  0 \\
A_d & 0 
   \end{array}\right];&  \tilde{B}_{flat} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
\end{eqnarray}
where $B_x = B\times(I-A)^{-1}$.

It is sufficient to show that $\tilde{B}\times(I-\tilde{A})^{-1} = \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}$.

Using the shorthand $Z=I-A$, we define the following identity:
\begin{equation}
  I* = \left[\begin{array}{cc}
      I & 0 \\
      0 & Z
    \end{array}\right]^{-1} \times
  \left[\begin{array}{cc}
      I & 0 \\
      0 & Z
    \end{array}\right]
\end{equation}

We introduce the identity in between the two terms of the LCIDB:
\begin{align}
  \tilde{B}\times(I-\tilde{A})^{-1} &=
  \left[\begin{array}{cc}B_f & B\end{array}\right]\times I*  \times
  \left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I-A\end{array}\right]^{-1}\\
      &=  \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\times
      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right] \times
      \left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]^{-1}
\end{align}
Using the associative property of matrix multiplication, it can be shown that $M\times N^{-1} = (N\times M^{-1})^{-1}$ for any invertible $M$ and $N$ of the same rank, so:
\begin{eqnarray}
  \tilde{B}\times(I-\tilde{A})^{-1}
  &=& \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\times
\left(
\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]\times
    \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\right)^{-1}\\
  &=&   \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
      \left[\begin{array}{cc}I & 0 \\ 0 & Z^{-1}\end{array}\right]\times
\left(
\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]\times
\left[\begin{array}{cc}I & 0 \\ 0 & Z^{-1}\end{array}\right]\right)^{-1}\\
 &=& \left[\begin{array}{cc}B_f & B\times Z^{-1}\end{array}\right]\times
\left(\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I\end{array}\right]\right)^{-1}
\end{eqnarray}
But these terms are equivalent to the flattened LCIDB:
\begin{eqnarray}
  \tilde{B}\times(I-\tilde{A})^{-1} &=&
  \left[\begin{array}{cc}B_f & B\times Z^{-1}\end{array}\right]\times
\left(\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I\end{array}\right]\right)^{-1} \\
&=& \left[\begin{array}{cc}B_f & B\times (I-A)^{-1}\end{array}\right]\times
\left(I-\left[\begin{array}{cc}A_f & 0 \\ A_d & 0\end{array}\right]\right)^{-1}\\
&=& \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}
\end{eqnarray}
thus completing the proof.

### 2017-10-23 aggregation/agg-agg.tex
Private disclosure work

@@
 
-\subsubsection{Complete Foreground}
+\subsubsection{Foreground or Foreground Fragment}
 
@@
 The collection of processes themselves make up the foreground of the study, and so exchanges among them are recorded in $A_f$, and exchanges outside the foreground are represented in $A_d$ or $B_f$.
-A complete foreground represents the full disclosure of a model, and a reader can replicate and extend it independently.
-Any co-production treatment, such as an allocation or system expansion, can only be fully expressed as a foreground model, because it intrinsically involves a collection of linked unit processes.
+A complete foreground represents
+the full disclosure of a model, and a reader can replicate and extend it independently.
+A fragment of the PSM can also be published as a complete foreground.
+Any co-production treatment, such as an allocation or system expansion, can only be fully expressed as a foreground fragment, because it intrinsically involves a collection of linked unit processes.
 

### 2017-10-23 aggregation/agg-pagg.tex
Private disclosure work

\subsection{Aggregation and Privacy}

\input{fig_private}

### 2017-10-23 aggregation/agg-objective.tex
More tweaks

@@
 Obviously, the level of detail of the disclosure will determine how useful it is to the Reader.
-In this context, the reader will want to know the answers to various questions about the model, which we will call \emph{review objectives}.
-At the same time the author may be limited in what she can disclose, which we will call \emph{disclosure constraints}.
+In this context, the reader will want to know the answers to various review objectives.
+At the same time the author may have constraints that limit what she can disclose.
 The purpose of the framework is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.

### 2017-10-24 aggregation/agg-JIE.tex
basic aggregation; figure 4
More tweaks
private aggregation

@@
 \bibliographystyle{authordate_bk}
+
+\renewcommand{\topfraction}{0.7}
 
@@
 %\usepackage{lineno}
-%\let\showgrid\relax
+\let\showgrid\relax
 \let\oldtimes\times
@@
 
-\section{Formalizing LCA Studies}
+\section{Disclosing the Product System Model}
 
@@
 Note that, for purposes of critical review, the objective for data privacy can be omitted because the reviewer may be granted privileged access to confidential data.
+We return to the data privacy requirement in the next section.
 
@@
 
-\section{Protecting Privacy through Aggregation}
+\section{Foreground Aggregation}
 
-  \input{agg-agg}        % notation for full aggregation
+When an LCA study result is computed, the foreground and background matrices are aggregated into a scalar indicator score (or a small vector of a few scores).
+The formulation in Eq.~\ref{eqn:study} and the illustration in Fig.~\ref{fig:foreground} provides a framework for reporting results at varying levels of aggregation.
+In this section, I present several different aggregation computations that provide equivalent results for an LCA study.
+After that, I show how a foreground containing private data can be partitioned into public and private portions, each of which can be comprehensively checked by a reviewer having privileged access, and subsequently disclosed in a manner that protects the private data.
 
-%  \input{agg-pagg}       % notation for partial aggregation
+  \input{agg-agg}         % the elusive Figure 4
 
-%  \input{agg-forms}      % the elusive Figure 4
+  \input{agg-pagg}       % notation for partial aggregation
 
@@
 
-%  \input{agg-review}     % review objectives
+ %% \input{agg-review}     % review objectives
 
-%  \input{agg-privacy}    % disclosure constraints
+ %% \input{agg-privacy}    % disclosure constraints
 
-%  \input{agg-semantic}   % the need for stable references
+ %% \input{agg-semantic}   % the need for stable references
 
@@
 
-  %% \input{publication-foreground}
+%  \input{publication-foreground}
 
-  %% \input{publication-example}
+%  \input{publication-example}
 
-  %% \input{publication-discussion}
+%  \input{publication-discussion}
 
-  %% \input{publication-conclusion}
+%  \input{publication-conclusion}
 
-  %% \input{publication-acknowledgments}
+%  \input{publication-acknowledgments}
 
@@
 
-  \doublespacing
+%  \doublespacing
 
-  \input{publication-si-spreadsheet}
+  \input{proofs}
 
-  \input{publication-forms}
+%  \input{publication-si-spreadsheet}
 
-  \input{publication-si-recipe}
+%  \input{publication-forms}
 
-  \input{publication-si-openaccess}
+%  \input{publication-si-recipe}
+
+%  \input{publication-si-openaccess}
 

### 2017-10-24 aggregation/agg-lca-study.tex
private aggregation

@@
 \end{equation}
-The proof of the equivalency of Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary information to \cite{Kuczenski_JLCA_2015}.
+The proof of the equivalency of Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary materials.
 In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining there pertains to the study foreground.
@@
 
-Using the ordered matrices, Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:\footnote{For a demonstration of this equivalence, the reader is again directed to \cite{Kuczenski_JLCA_2015}.}
+Using the ordered matrices, Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:
 \begin{equation}
-s = \mathbf{e}^T \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \tilde{\mathbf{y}}
+s = \mathbf{e}^T \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \tilde{\mathbf{y}}_f
 \label{eqn:study}
 \end{equation}
+where $\tilde{\mathbf{y}}_f$ is a canonical functional unit vector having the same rank as $A_f$. Eq.~\ref{eqn:study} is derived in the supplementary materials.
 Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$, having been subsumed into $B_x$, represented by an identity matrix $I$.
 
-The activity levels of the foreground nodes are represented by $\tilde{\mathbf{x}}$, which is the result of $\tilde{\mathbf{y}}$ selecting the first column of $(I-A_f)^{-1}$.
+The activity levels of the foreground nodes are represented by $\tilde{\mathbf{x}}$, which is the result of $\tilde{\mathbf{y}}_f$ selecting the first column of $(I-A_f)^{-1}$.
 %% An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:

### 2017-10-24 aggregation/agg-pagg.tex
private aggregation

@@
-\subsection{Aggregation and Privacy}
+\subsection{Reviewable Private Aggregation}
 
 \input{fig_private}
+
+Often, a study author may wish to make a partial disclosure of a PSM that retains the privacy of confidential data, while still permitting comprehensive critical review of the complete model.
+In this case it is necessary to grant a reviewer privileged access to the private data in order to meet the review objectives.
+The study formulation in Eq.~\ref{eqn:study} may be used to partition a foreground into public and private portions in order to comply with any applicable constraints on what may be disclosed.
+The approach is illustrated in Figure~\ref{fig:private}.
+
+The first step is to identify the constraints on the disclosure.
+In the framework presented here, constraints may include the locations or values of any subset of nonzero entries in the $A_d$ and $B_f$ matrices, as well as any subset of nodes in $A_f$.
+Fig.~\ref{fig:private}a shows a PSM in which disclosure constraints are indicated by shaded regions of the matrices.
+In this case the nodes represented by the three rightmost columns of $A_f$, as well as one entire row of $A_d$, are to remain secret.
+
+The $A_d$ and $B_f$ matrices are then partitioned into two components that sum to the original:
+\begin{equation}\begin{aligned}
+  A_d =\; & A_{d,pub} + A_{d,priv} \\
+  B_f =\; & B_{f,pub} + B_{f,priv}
+  \label{eqn:partition}
+  \end{aligned}
+  \end{equation}
+This is illustrated in Fig.~\ref{fig:private}b.
+Substituted back into Eq.~\ref{eqn:study}, the private portions are aggregated into a private life cycle inventory vector $\mathbf{b}_{priv}$, while the public portions remain disaggregated:
+\begin{align}
+  s =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
+  =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\\
+  \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}
+\end{align}
+Any foreground nodes that are completely contained within the private partition can be omitted from the disclosed foreground.
+The aggregation result can then be included as a separate foreground node with a unit weight, and the aggregated inventory vector included as the corresponding column of $B_f$.
+This is illustrated in Fig.~\ref{fig:private}c.
+
+The reviewer must be granted access to both the public and private partitions in order to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly; however, it is not thought possible for a reader to discern any information about $A_{d,priv}$ or $B_{f,priv}$ solely from $\mathbf{b}_{priv}$.
+An advantage to this approach is that the reader can evaluate easily what fraction of the overall impact score is accounted for by the public versus the private portions of the model.
+A ``disclosure completeness'' metric $\varphi$ can be defined as the fraction of the overall score which is accounted for by the disclosed portion of the PSM:
+\begin{equation}
+  \varphi =  \frac{\mathbf{e}^T\cdot(\mathbf{b} - \mathbf{b}_{priv})}{\mathbf{e}^T\cdot\mathbf{b}} = 1 - \frac{\mathbf{e}^T\cdot\mathbf{b}_{priv}}{\mathbf{e}^T\cdot\mathbf{b}}
+  \label{eqn:metric}
+\end{equation}
+where $\mathbf{b}$ is the complete life cycle inventory.
+The value of this metric can provide an indication of the level of transparency of the disclosure with respect to a given impact category indicator.
+The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values.

### 2017-10-24 aggregation/proofs.tex
private aggregation

@@
   \tilde{B}\times(I-\tilde{A})^{-1} &=&
-  \left[\begin{array}{cc}B_f & B\times Z^{-1}\end{array}\right]\times
+  \left[\begin{array}{cc}B_f & B\times (I-A)^{-1}\end{array}\right]\times
 \left(\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I\end{array}\right]\right)^{-1} \\
-&=& \left[\begin{array}{cc}B_f & B\times (I-A)^{-1}\end{array}\right]\times
+&=& \left[\begin{array}{cc}B_f & B_x\end{array}\right]\times
 \left(I-\left[\begin{array}{cc}A_f & 0 \\ A_d & 0\end{array}\right]\right)^{-1}\\
@@
 thus completing the proof.
+
+\subsection{The Study Foreground Equation}
+
+Equation 5 in the manuscript is derived from Equation 4.
+
+We begin by computing $(I-\tilde{A})^{-1}$, again taking advantage of the matrix's block triangularity.  Find $Q$ such that:
+\begin{equation}
+  \left[\begin{array}{cc} (I-A_f) & 0 \\ -A_d & I\end{array}\right]\times
+  \left[\begin{array}{cc} (I-A_f)^{-1} & 0 \\ Q & I\end{array}\right] =
+  \left[\begin{array}{cc} I & 0 \\ 0 & I\end{array}\right]
+\end{equation}
+
+It can be seen that $Q=A_d\times(I-A_f)^{-1}$ is a necessary condition to satisfy the equality, and thus:
+\begin{equation}
+  (I-\tilde{A})^{-1} = \left[\begin{array}{cc} (I-A_f)^{-1} & 0 \\ A_d\times(I-A_f)^{-1} & I\end{array}\right]
+\end{equation}.
+
+Substituting this into the LCA system equation:
+\begin{align}
+  s =\;& \mathbf{e}^T\times \tilde{B}\times(I-\tilde{A})^{-1}\tilde{\mathbf{y}}\\
+  =\;& \mathbf{e}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
+  \left[\begin{array}{cc} (I-A_f)^{-1} & 0 \\ A_d\times(I-A_f)^{-1} & I \end{array}\right]
+  \times\left[\begin{array}{c} \tilde{\mathbf{y}}_f \\ \mathbf{0}\end{array}\right]
+\end{align}
+where $\tilde{\mathbf{y}}_f$ is a canonical functional unit having the same dimension as $A_f$.  Thus:
+\begin{align}
+s =\;& \mathbf{e}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
+ \left[\begin{array}{c} (I-A_f)^{-1}\times\tilde{\mathbf{y}}_f \\ A_d\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f\end{array}\right]\\
+ =\;& \mathbf{e}^T\times \left(B_f\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f
+ + B_x\times A_d\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f\right)\\
+ =\;& \mathbf{e}^T\times \left(B_f + B_x\times A_d\right)\times(I-A_f)^{-1}\times \tilde{\mathbf{y}}_f\\
+ =\;& \mathbf{e}^T\times \left(B_f + B_x\times A_d\right)\times \tilde{\mathbf{x}}
+\end{align}
+thus completing the derivation.

### 2017-10-24 aggregation/agg-JIE.tex
morning tweaks

@@
 
-\title{Aggregated Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy}
+\title{Disclosure of Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy}
 \author{Brandon Kuczenski\\
@@
 Objectives in critical review vary with the study and the reviewer, but they generally include consistency with the ISO 14044 standard, the validity of methods, appropriateness of data used, correctness of interpretations, and transparency of reporting \citep[Section 6.1]{iso14044}.
-Although the object of critical review is typically a written report, the use of a structured disclosure of study design introduces the possibility that reviewers can inspect the model itself and even verify the computations and reported results.
+Although the object of critical review is typically a written report, the use of a structured disclosure of a PSM introduces the possibility that reviewers can inspect the model itself and even verify the computations and reported results.
 This section discusses how a disclosure in the format described above can enable robust review of a product system model.

### 2017-10-24 aggregation/agg-lca-study.tex
morning tweaks

@@
 
-For clarity, I will call the augmented $\tilde{A}$ and $\tilde{B}$, in which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be a \emph{canonical LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.
+For clarity, I will call the augmented $\tilde{A}$ and $\tilde{B}$, in which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.
 This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.
@@
 \begin{equation}
- \tilde{A} = \left[\begin{array}{cc} 
+ \tilde{A}_{flat} = \left[\begin{array}{cc} 
 A_f &  0 \\
 A_d & 0 
-   \end{array}\right];\;\;\;  \tilde{B} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
+   \end{array}\right];\;\;\;  \tilde{B}_{flat} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
 \label{eqn:fg_bg}
@@
 The proof of the equivalency of Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary materials.
-In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining there pertains to the study foreground.
+In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining in $\tilde{A}_{flat}$ pertains to the study foreground.
 Furthermore, the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported.

### 2017-10-24 aggregation/agg-review.tex
morning tweaks

\subsection{Critical Review of PSMs}

When a study result is published in an aggregated form, it is the responsibility of the critical reviewer to validate the aggregation result.

### 2017-10-24 aggregation/proofs.tex
morning tweaks

@@
 
-We begin by computing $(I-\tilde{A})^{-1}$, again taking advantage of the matrix's block triangularity.  Find $Q$ such that:
+We begin by computing $(I-\tilde{A}_{flat})^{-1}$, again taking advantage of the matrix's block triangularity.  Find $Q$ such that:
 \begin{equation}
@@
 \begin{align}
-  s =\;& \mathbf{e}^T\times \tilde{B}\times(I-\tilde{A})^{-1}\tilde{\mathbf{y}}\\
+  s =\;& \mathbf{e}^T\times \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}\tilde{\mathbf{y}}\\
   =\;& \mathbf{e}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times

### 2017-10-25 aggregation/agg-intro.tex
paltry work today

@@
 However, this prescription is insufficient to ensure the comparability of different declarations, even if they use the same PCR \citep{Modahl_2012}.
-Moreover, PCRs are often themselves not unique and can include wide variations in scope and boundary requirements \citep{Subramanian_2012}, leading to ambiguity.
+In addition, PCRs are often themselves not unique and can include wide variations in scope and boundary requirements \citep{Subramanian_2012}, leading to ambiguity.
 
@@
 When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.
-In the end, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.
+ISO 14071, which provides guidelines on critical review, emphasizes that review of the PSM or individual datasets is optional \citep{iso14071}.
 
@@
 Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the PSM and thus, no automatic mechanisms for interpretation or review of published LCA studies.
+Moreover, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.
+The purpose of this paper is to advance a framework for the formal description of product system models that would provide robust transparency and reproducibility of computed results.
 
-The purpose of this paper is to advance a framework for the formal description of product system models that would provide robust transparency and reproducibility of computed results.
+\endinput
 
@@
 \end{enumerate}
-
-\endinput
 

### 2017-10-25 aggregation/agg-pagg.tex
paltry work today

@@
   s =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
-  =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\\
+  =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}\\
   \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}
@@
 This is illustrated in Fig.~\ref{fig:private}c.
-
-The reviewer must be granted access to both the public and private partitions in order to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly; however, it is not thought possible for a reader to discern any information about $A_{d,priv}$ or $B_{f,priv}$ solely from $\mathbf{b}_{priv}$.
-An advantage to this approach is that the reader can evaluate easily what fraction of the overall impact score is accounted for by the public versus the private portions of the model.
-A ``disclosure completeness'' metric $\varphi$ can be defined as the fraction of the overall score which is accounted for by the disclosed portion of the PSM:
-\begin{equation}
-  \varphi =  \frac{\mathbf{e}^T\cdot(\mathbf{b} - \mathbf{b}_{priv})}{\mathbf{e}^T\cdot\mathbf{b}} = 1 - \frac{\mathbf{e}^T\cdot\mathbf{b}_{priv}}{\mathbf{e}^T\cdot\mathbf{b}}
-  \label{eqn:metric}
-\end{equation}
-where $\mathbf{b}$ is the complete life cycle inventory.
-The value of this metric can provide an indication of the level of transparency of the disclosure with respect to a given impact category indicator.
-The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values.
+The

### 2017-10-25 aggregation/agg-review.tex
paltry work today

@@
-\subsection{Critical Review of PSMs}
+\subsection{Documentation and Critical Review of PSMs}
 
-When a study result is published in an aggregated form, it is the responsibility of the critical reviewer to validate the aggregation result.
+The product system model (PSM) represents the product of an analyst's labor to perform an LCA.
+It contains all study-specific information about what processes are modeled, how those processes exchange flows with one another, and what background datasets have been selected.
+
+In current practice, this model is then documented as a written report, and the report can be both critically reviewed and distributed to data users, but this process introduces significant interpretive challenges.
+It also results in the model being frozen to a particular configuration or set of configurations from which the results are generated.
+
+When a study result is published in an aggregated form, often it is beyond the capacity of the critical reviewer to validate the aggregation result, even if the reviewer has privileged access to confidential materials used in preparing the report.
+
+This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.
+A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.
+Once equipped with the PSM, the reviewer or data user can potentially go much further, adding or removing elements from the model, altering dataset selections, and applying other impact assessment methods.
+
+If a study containing confidential information is partitioned into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated without including private data in the disclosure.
+The reviewer must be granted access to both the public and private partitions in order to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly; however, it is not thought possible for a reader to discern any information about $A_{d,priv}$ or $B_{f,priv}$ solely from $\mathbf{b}_{priv}$.
+
+A further advantage to this approach is that the reader can evaluate easily what fraction of the overall impact score is accounted for by the public versus the private portions of the model.
+A ``disclosure completeness'' metric $\varphi$ can be defined as the fraction of the overall score which is accounted for by the disclosed portion of the PSM:
+\begin{equation}
+  \varphi =  \frac{\mathbf{e}^T\cdot(\mathbf{b} - \mathbf{b}_{priv})}{\mathbf{e}^T\cdot\mathbf{b}} = 1 - \frac{\mathbf{e}^T\cdot\mathbf{b}_{priv}}{\mathbf{e}^T\cdot\mathbf{b}}
+  \label{eqn:metric}
+\end{equation}
+where $\mathbf{b}$ is the complete life cycle inventory.
+The value of this metric can provide an indication of the level of transparency of the disclosure with respect to a given impact category indicator.
+The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values.

### 2017-10-30 aggregation/agg-discussion.tex
minor discussion work

\subsection{Meeting the Disclosure Objectives}

\subsubsection{Transparency}

The objective for transparency was identified as requiring computability, completeness, and reproducibility.  These requirements can be met by a disclosure that clearly describes the identities of each row or column of $A_f$, $A_d$, and $B_f$ containing non-zero entries, and the locations and values of those entries.  In principle, a reader with this information would have the capability to construct the augmented LCIDB and perform the computation in Eq.~\ref{eqn:study}.  In actuality, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity in stating the identities of the rows and columns of those matrices.

The foreground nodes, which make up the columns of $A_f$, can be chosen freely by the study author according to the objectives of the study.  Large studies may contain hundreds of foreground nodes, and the nodes can correspond to physical activities, logical operations, unit conversions, accumulation or distribution points, or any other aspect of model construction that can be reflected in a process-flow diagram.  The only requirement on their disclosure is that the identity of each node's reference flow, including its unit of measure, is clearly stated.  In formalizing the study as a normalized direct requirements matrix, the activity level of each foreground node necessarily equals the magnitude of the total reference flow emanating from the node.

For each row in $A_d$ containing a non-zero entry, the author must unambiguously identify the exact dataset used, including the version of the database, as well as the exact process and reference flow selected; the dimension (reference quantity or unit) of each reference flow must be specified; and the sign of the numeric entry in $A_d$ must be consistent with the implementation of the process in the background database.  Similarly, for each row in $B_f$ containing a non-zero entry, the author must unambiguously identify the substance being exchanged with the environment, the compartment or context into which it is being exchanged, and the reference quantity or unit associated with the flow.  Sign consistency must also be assured.  If LCIA indicator results are included, the author must also unambiguously identify the method computed (identity of $\mathbf{e}$).

\subsubsection{Authority and Primacy}

### 2017-10-30 aggregation/agg-semantic.tex
morning tweaks
paltry work today
minor discussion work

@@
-For each row in $A_d$ containing a non-zero entry, the author must unambiguously identify the version of the database used, as well as the exact process and reference flow used from the background database; the dimension (reference quantity or unit) of each reference flow must be specified; and the sign of the numeric entry in $A_d$ must be consistent with the implementation of the process in the background database.
-
-Similarly, for each row in $B_f$ containing a non-zero entry, the author must unambiguously identify the substance being exchanged with the environment, the compartment or context into which it is being exchanged, and the reference quantity or unit associated with the flow.
-Sign consistency must also be assured.
-If LCIA indicator results are included, the author must also unambiguously identify the method computed (identity of $\mathbf{e}$).
-
-For the reader's part, the reader must be able to produce an $\mathbf{e}$ vector that matches the one used by the author and identify which entries correspond with which
+\subsection{Avoiding Ambiguous Disclosures}
 
 either the column of $B_x$, or the scalar value that corresponds to the dot product of this column with $\mathbf{e}$
+
+Multiple background databases
+
+\subsection{Stable Semantic References}

### 2017-11-01 aggregation/agg-disclosure.tex
private aggregation

@@
-\subsection{Disclosure of Study Design}
+\subsection{Contents of a Disclosure}
 
@@
 These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.
-However, the the challenges of operationalizing this requirement are tremendously complex.
-Some implications of the requirement for shared agreement are considered in the Discussion.
+However, the the challenges of operationalizing this requirement are complex.
+Some implications on the information required for a successful disclosure are considered in the Discussion.

### 2017-11-01 aggregation/agg-objective.tex

@@
 In the context of this paper, an ``LCA Study'' should be understood as any aggregation of data under consideration, from a single-operation unit process to a full ISO-style product LCA.
-The practice of reporting the results of an LCA study implies the existence of two parties with different information: the study author has full knowledge of the product system model; and the Reader only has knowledge of what the author has disclosed.
-Obviously, the level of detail of the disclosure will determine how useful it is to the Reader.
+The practice of reporting the results of an LCA study implies the existence of two parties with different information: the study author has full knowledge of the product system model; and the reader only has knowledge of what the author has disclosed.
+Obviously, the level of detail of the disclosure will determine how useful it is to the reader.
 In this context, the reader will want to know the answers to various review objectives.
@@
 \item \textbf{Computability}.
-The result should be computed using only the disclosure, plus resources available independently to the Reader.
+The result should be computed using only the disclosure, plus resources available independently to the reader.
 The computation should follow the consensus understanding of process LCA computation.
 \item \textbf{Completeness / Minimality}.  The computation should use all of the information provided in the disclosure.
-\item \textbf{Reproducibility}.  The result obtained by the Reader should match the result provided by the Author.
+\item \textbf{Reproducibility}.  The result obtained by the reader should match the result provided by the author.
 \end{itemize}
-One key objective of disclosure is to inspect the work of the Author.
-Consequently, the disclosure should be made up of the Author's own work product and should exclude, for instance, data that were drawn from a reference database or other independent source and used in an unmodified form.
-The purpose of this is twofold: to focus the efforts of the Reader on the work of the Author, and to enable the Reader to obtain and inspect the third-party data independently, thus ensuring its integrity.
+One key objective of disclosure is to inspect the work of the author.
+Consequently, the disclosure should be made up of the author's own work product and should exclude, for instance, data that were drawn from a reference database or other independent source and used in an unmodified form.
+The purpose of this is twofold: to focus the efforts of the reader on the work of the author, and to enable the reader to obtain and inspect the third-party data independently, thus ensuring its integrity.
 \begin{itemize}
-\item \textbf{Authority}.  The disclosure should include only information for which the Author is responsible.
+\item \textbf{Authority}.  The disclosure should include only information for which the author is responsible.
 \item \textbf{Primacy}.  The disclosure should exclude information that is made available by a third party.
@@
 \begin{itemize}
-\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the Author.
+\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the author.
 \end{itemize}

### 2017-11-01 aggregation/agg-review.tex

@@
 In current practice, this model is then documented as a written report, and the report can be both critically reviewed and distributed to data users, but this process introduces significant interpretive challenges.
-It also results in the model being frozen to a particular configuration or set of configurations from which the results are generated.
+It also results in the model being frozen to a particular configuration or set of configurations at the time the results are generated.
+If the review objectives include sensitivity of the model to parameter variations and/or alternative scenarios, then the review is limited to only the parameters and scenarios considered by the author.
 

### 2017-11-01 aggregation/agg-review.tex
comparative review?

@@
 
-In current practice, this model is then documented as a written report, and the report can be both critically reviewed and distributed to data users, but this process introduces significant interpretive challenges.
+The current practice of documenting the model with a written report sharply limits the ability of readers to interpret and reuse the data.
 It also results in the model being frozen to a particular configuration or set of configurations at the time the results are generated.
-If the review objectives include sensitivity of the model to parameter variations and/or alternative scenarios, then the review is limited to only the parameters and scenarios considered by the author.
+Any review objectives involving sensitivity of the model to parameter variations and/or alternative scenarios are limited to the selections considered by the author.
 
@@
 Once equipped with the PSM, the reviewer or data user can potentially go much further, adding or removing elements from the model, altering dataset selections, and applying other impact assessment methods.
+
+\subsection{Comparative Review: Formalizing Scope and System Boundary}
+
+The disclosure of a PSM contains six elements, including three lists and three tables of sparse matrix data.
+While the numeric data are required to critically evaluate and reproduce the study results, substantial information about the study design can be found in the list of background dependencies (\ref{itm:bg})and emissions (\ref{itm:em}).
+These lists report all the distinct sources of environmental impacts in the model.
+The cutoff flows, which are foreground nodes that have no nonzero entries in $A_f$, $A_d$, or $B_f$, report on the other hand what aspects were explicitly excluded from the model.
+
+Together, these lists can be considered a functional definition of the study scope that can be used to conduct a qualitative comparison of multiple studies by different authors.
+
+\subsection{Privacy and Partial Aggregation}
+
+Partial aggregation is already used in practice to conceal private data, but the
 

### 2017-11-02 aggregation/agg-JIE.tex
paltry work today
comparative review?
Finish discussion?

@@
 
- %% \input{agg-review}     % review objectives
+The current practice of documenting a PSM with a written report sharply limits the ability of readers to interpret and reuse the data.
+It also results in the model being frozen to a particular configuration or set of configurations used to generate the published results.
+Any review objectives involving sensitivity of the model to parameter variations and/or alternative scenarios are limited to the selections considered by the author.
 
- %% \input{agg-privacy}    % disclosure constraints
+When a study result is published in an aggregated form, often it is beyond the capacity of the critical reviewer to validate the aggregation result, even if the reviewer has privileged access to confidential materials used in preparing the report.
 
- %% \input{agg-semantic}   % the need for stable references
+This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.
+A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.
+Once equipped with the PSM, the reviewer or data user can potentially go much further.
+The graph structure encoded in the foreground matrices can be automatically rendered as a process-flow diagram, supplementing or supplanting the hand-crafted system diagrams that currently accompany ISO reports.
+Reviewers or data users who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, and applying other impact assessment methods.
+Moreover, because the disclosure of the PSM includes only references to background data without including the data themselves, it is trivial for item~\ref{itm:bg} to include references to several background databases without adding any additional requirements.
+
+  \input{agg-review}     % review objectives and disclosure constraints
+
+  \input{agg-semantic}   % the need for stable references
 
@@
 
+  \input{agg-requirements}
+
+  \input{../publication/publication-example}
+
 %  \input{publication-si-spreadsheet}

### 2017-11-02 aggregation/agg-review.tex
Finish discussion?

@@
-\subsection{Documentation and Critical Review of PSMs}
+%\subsection{Documentation and Critical Review of PSMs}
 
-The product system model (PSM) represents the product of an analyst's labor to perform an LCA.
+%%The product system model (PSM) represents the product of an analyst's labor to perform an LCA.
 It contains all study-specific information about what processes are modeled, how those processes exchange flows with one another, and what background datasets have been selected.
-
-The current practice of documenting the model with a written report sharply limits the ability of readers to interpret and reuse the data.
-It also results in the model being frozen to a particular configuration or set of configurations at the time the results are generated.
-Any review objectives involving sensitivity of the model to parameter variations and/or alternative scenarios are limited to the selections considered by the author.
-
-When a study result is published in an aggregated form, often it is beyond the capacity of the critical reviewer to validate the aggregation result, even if the reviewer has privileged access to confidential materials used in preparing the report.
-
-This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.
-A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.
-Once equipped with the PSM, the reviewer or data user can potentially go much further, adding or removing elements from the model, altering dataset selections, and applying other impact assessment methods.
 
@@
 
-The disclosure of a PSM contains six elements, including three lists and three tables of sparse matrix data.
-While the numeric data are required to critically evaluate and reproduce the study results, substantial information about the study design can be found in the list of background dependencies (\ref{itm:bg})and emissions (\ref{itm:em}).
+The disclosure of a PSM contains six parts, including three lists and three tables of sparse matrix data.
+While the numeric data are required to critically evaluate and reproduce the study results, substantial information about the study design can be found in the list of background dependencies (\ref{itm:bg}) and emissions (\ref{itm:em}).
 These lists report all the distinct sources of environmental impacts in the model.
@@
 Together, these lists can be considered a functional definition of the study scope that can be used to conduct a qualitative comparison of multiple studies by different authors.
+Background database choice, dataset selection, and version information can all be made available in disclosure item~\ref{itm:bg}, thereby streamlining what is presently a highly interpretive and labor-intensive process of extracting this information from written text.
+Similarly, the set of elementary flows explicitly modeled in the study foreground (item~\ref{itm:em}) expose whether any study-specific modeling of direct emissions was performed and what flows were included.
 
@@
 
-Partial aggregation is already used in practice to conceal private data, but the
-
-If a study containing confidential information is partitioned into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated without including private data in the disclosure.
-The reviewer must be granted access to both the public and private partitions in order to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly; however, it is not thought possible for a reader to discern any information about $A_{d,priv}$ or $B_{f,priv}$ solely from $\mathbf{b}_{priv}$.
+Partial aggregation is already used in practice to conceal private data, but current practices used to generate aggregated results from input model configurations do not provide a way for a reviewer to validate the aggregation.
+In contrast, if a study containing confidential information is partitioned into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated directly.
+To achieve this, the reviewer must be granted access to both the public and private partitions in order to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly.
+However, and the private data can still be withheld from the disclosure.
 
@@
 The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values.
+
+\subsection{Attacks on Privacy}
+
+It is not thought possible for a reader to discern any information about a disaggregated system solely from an aggregated disclosure such as $\mathbf{b}_{priv}$ \citep[Ch.
+3]{UNEP_2011}, although that assumption has never been rigorously tested.
+Compressed sensing is a signal processing technique that seeks solutions to an underdetermined linear system, in which there are more unknowns than equations, based on the assumption that the solution is sparse \citep{Donoho_2006}.
+A form of linear optimization called ``basis pursuit'' can be used to try to detect the signal.
+
+In the LCA case, if the private LCI result $\mathbf{b}_{priv}$ includes inventory data from a background database but no foreground emissions, then it can be written as:
+\begin{equation}
+  \mathbf{b}_{priv} = B_x\times \mathbf{a}_{d,priv}
+\end{equation}
+Typically this equation is highly underdetermined (In ecoinvent, for instance, $B_x$ has roughly 1,800 rows and over 12,000 columns).
+If the private input $\mathbf{a}_{d,priv}$ is highly sparse, and the identity of $B_x$ is known, then $\mathbf{a}_{d,priv}$ may be vulnerable to a basis pursuit attack.
+Further research is required to evaluate whether this attack can be implemented in a real situation, and how it may be defended against.
+%10.1109/TIT.2006.871582

### 2017-11-02 aggregation/agg-semantic.tex
Finish discussion?

@@
-\subsection{Avoiding Ambiguous Disclosures}
+\subsection{Stable Semantic References to Study Elements}
 
-either the column of $B_x$, or the scalar value that corresponds to the dot product of this column with $\mathbf{e}$
+In actuality, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity
+in identifying the background datasets and emissions to which those matrix entries correspond.
+It is necessary for the entries in disclosure items~\ref{itm:bg} and \ref{itm:em} to be unambiguous and easy to interpret.
+Entries in \ref{itm:bg}, which represent rows in $A_d$, must clearly identify the originating database, version, and configuration, along with the precise dataset, that dataset's reference flow, and the dimension (reference quantity or unit) of that flow.
+In addition, the sign of the numeric entry in $A_d$ must be consistent with the implementation of the background dataset.
+Similar requirements also hold for entries in \ref{itm:em}, which represent rows in $B_f$.
+Current research has revealed widespread challenges to finding consistency on the identities of elementary flows \citep{Edelen_2017}.
 
-Multiple background databases
+In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of those references.
+This is often accomplished using linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.
+The resource at the end of the link functions as a point of agreement: both study author and data user can follow it from anywhere on the Internet to obtain the same information.
+The content pointed to by the hyperlink, along with context provided by \emph{other} references to the same resource, gives meaning to the information and allows it to be curated by the study authors and other members of the community \citep{Khan_2011}.
+% Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.
+Some preliminary descriptions of the entity types involved in LCA computation, and their relationships to one another, have recently been developed \citep{Ciroth_Srocka_2014, Janowicz_WOP_2015, Kuczenski_JCP_2016}.
 
-\subsection{Stable Semantic References}
+Before the goal of easily reproducible foreground models can be realized, the linked data foundation must be laid.
+Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that distinguish database, version, configuration, dataset, and product flow to which the reference refers.
+These references should take the form of URIs that can be accessed using any browser and can provide a reader with documentary information describing the dataset, the quantitative dimension of the reference flow, and a mechanism for the reader to obtain access to the data.
+Similarly, LCIA method providers should ensure that their methods make use of stable semantic references to elementary flows and contexts.

### 2017-11-03 aggregation/agg-JIE.tex
Some morning work

@@
 
-  \input{agg-foreground} % Af and graph/matrix equivalency
+\subsection{Interpreting the Foreground}
+
+The foreground of an LCA study is a collection of nodes that represent points in the inventory model where flows are exchanged.
+The foreground matrix $A_f$ describes in precise terms what components are included in the product system and how they connect to one another.
+The foreground is modelled as a weighted directed graph in which there is a 1:1 correspondence between nodes and product flows.
+The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.
+These nodes typically represent unit processes, but they can also indicate exchanges across an implicit system boundary that encloses the foreground.
+Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e.
+any allocated supply-and-use table, but some fundamental designs are common.
+Some simple foreground models are shown in the supplementary materials.
 
@@
 
-  \input{../publication/publication-example}
+  \input{agg-foreground} % Af and graph/matrix equivalency
+
+  \input{../manuscript/publication-example}
 

### 2017-11-03 aggregation/agg-foreground.tex
Private disclosure work
Some morning work

@@
-\subsection{Interpreting the Foreground}
+\section{Foreground Configurations}
 
-The foreground of an LCA study is a collection of nodes that represent points in the inventory model where flows are exchanged.
-The foreground matrix $A_f$ describes in precise terms what components are included in the product system and how they connect to one another.
-The foreground is modelled as a weighted directed graph in which there is a 1:1 correspondence between nodes and product flows.
-The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.
-These nodes typically represent unit processes, but they can also indicate exchanges across an implicit system boundary that encloses the foreground.
-
+Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.
+Most product system models would include many of these elements mixed together.
 A reviewer with access to items \ref{itm:fg} and \ref{itm:af} in the disclosure would be able to construct $A_f$ and automatically create a process-flow diagram, annotated with information about each node.
-Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e.
-any allocated supply-and-use table, but some fundamental designs are common.
-Some simple foreground models are shown in Figure~\ref{fig:fragments}.
 
@@
 
-A typical PSM may contain multiple fragmentary elements that are interconnected.
+A typical PSM may contain multiple modules or fragments that are interconnected.
 An example of foreground composed of several fragments is illustrated in Figure~\ref{fig:fragments}(d).

### 2017-11-03 aggregation/agg-pagg.tex
Some morning work

@@
 This is illustrated in Fig.~\ref{fig:private}c.
-The
+
+The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values.
+This would enable a reader to understand \emph{that} a certain background process or emission was included in the model, but not to know \emph{how much} was included.

### 2017-11-03 aggregation/agg-basic-lca.tex
afternoon work- cites; semantics; editing

@@
 If the processes are assumed to be linear, the problem can be written in terms of matrix algebra.
-There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Suh_2002, Pauliuk_2015}.
-Although the two models are understood to be equivalent, the input-output approach is computationally more efficient in some applications because it lends itself to iterative solution \citep{Peters_2007}.
+There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Heijungs2002, Pauliuk_2015_framework}.
+Although the two models are understood to be equivalent, the input-output approach is computationally more efficient in some applications because it lends itself to iterative solution \citep{Peters_JLCA_2007}.
 That formulation may be stated:
@@
 The term $(I-A)$ is equivalent to the normalized technology matrix, i.e.
-the difference between the supply and use tables in classical input-output analysis, $(V'-U)$ \citep{Suh_2010}.
-The conditions under which this matrix has an inverse, particularly with respect to different strategies for handling co-production, have been discussed extensively \citep{Suh_2010, Majeau_Bettez_2014}.
+the difference between the supply and use tables in classical input-output analysis, $(V'-U)$ \citep{Suh_JIE_2010}.
+The conditions under which this matrix has an inverse, particularly with respect to different strategies for handling co-production, have been discussed extensively \citep{Majeau_Bettez_2014}.
 Eq.~\ref{eqn:leontief} is visualized in Figure~\ref{fig:lca}.

### 2017-11-03 aggregation/agg-lca-study.tex
afternoon work- cites; semantics; editing

@@
 This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.
-However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2016}).
+However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2015}).
 

### 2017-11-03 aggregation/agg-objective.tex
afternoon work- cites; semantics; editing

@@
 The task of computing life cycle impact assessment (LCIA) results from a product system model can be described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.
-Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh_2005, Peters_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.
+Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh2005a, Peters_JLCA_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.
+In the context of this paper, an ``LCA Study'' should be understood as any collection of inventory data under consideration, from a single-operation unit process to a full ISO-style product LCA.
 
-In the context of this paper, an ``LCA Study'' should be understood as any aggregation of data under consideration, from a single-operation unit process to a full ISO-style product LCA.
-The practice of reporting the results of an LCA study implies the existence of two parties with different information: the study author has full knowledge of the product system model; and the reader only has knowledge of what the author has disclosed.
+Reporting the results of an LCA study implies the existence of two parties with different information: the study author has full knowledge of the product system model; and the reader only has knowledge of what the author has disclosed.
 Obviously, the level of detail of the disclosure will determine how useful it is to the reader.
 In this context, the reader will want to know the answers to various review objectives.
-At the same time the author may have constraints that limit what she can disclose.
-The purpose of the framework is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.
+At the same time the author may have constraints that limit what can be disclosed.
+The main goal of the framework is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.
 

### 2017-11-03 aggregation/agg-pagg.tex
afternoon work- cites; semantics; editing

@@
 The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values.
-This would enable a reader to understand \emph{that} a certain background process or emission was included in the model, but not to know \emph{how much} was included.
+This would enable a reader to understand \emph{that} a certain background process or emission was included in the model, but not to know \emph{how much} was included, which may jointly satisfy review objectives and disclosure constraints in some cases.

### 2017-11-03 aggregation/agg-review.tex
Some morning work
afternoon work- cites; semantics; editing

@@
 Together, these lists can be considered a functional definition of the study scope that can be used to conduct a qualitative comparison of multiple studies by different authors.
-Background database choice, dataset selection, and version information can all be made available in disclosure item~\ref{itm:bg}, thereby streamlining what is presently a highly interpretive and labor-intensive process of extracting this information from written text.
+Background database choice, dataset selection, and version information is all made available in disclosure item~\ref{itm:bg}, thereby streamlining what is presently an interpretive and study-specific process of extracting this information from written text.
 Similarly, the set of elementary flows explicitly modeled in the study foreground (item~\ref{itm:em}) expose whether any study-specific modeling of direct emissions was performed and what flows were included.
@@
 Partial aggregation is already used in practice to conceal private data, but current practices used to generate aggregated results from input model configurations do not provide a way for a reviewer to validate the aggregation.
-In contrast, if a study containing confidential information is partitioned into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated directly.
+If a study containing confidential information is split into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated directly by the reviewer.
 To achieve this, the reviewer must be granted access to both the public and private partitions in order to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly.
-However, and the private data can still be withheld from the disclosure.
+However, the private information used in the computation can still be withheld from the disclosure.
 
@@
 The value of this metric can provide an indication of the level of transparency of the disclosure with respect to a given impact category indicator.
-The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values.
 

### 2017-11-03 aggregation/agg-semantic.tex
afternoon work- cites; semantics; editing

@@
 
-In actuality, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity
+In practice, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity
 in identifying the background datasets and emissions to which those matrix entries correspond.
+In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of those references.
+In particular, the database configuration, dataset version, reference flow and dimension (quantitative unit of measure) must be unambiguous and easy to interpret.
+
+Agreement about the meaning of concepts is often accomplished using linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.
+The resource at the end of the link functions as a point of agreement: both study author and data user can follow it from anywhere on the Internet to obtain the same information.
+The content pointed to by the hyperlink, along with context provided by \emph{other} references to the same content, gives meaning to the information and allows it to be curated by the study authors and other members of the community \citep{Khan_2011}.
+Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.
+Some preliminary descriptions of the entity types involved in LCA computation, and their relationships to one another, have recently been developed \citep{Ciroth_Srocka_2014, Janowicz_WOP_2015, Kuczenski_JCP_2016}.
+
+Before the goal of easily reproducible foreground models can be realized, the linked data foundation must be laid.
+Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that denote a particular activity in a particular database version and configuration.
+These references should take the form of URIs that can be accessed using any Internet browser and can provide both author and reader with documentary information describing the referred dataset, the available reference flows and their dimensions, and a mechanism for the reader to obtain access to the exchange data or LCI/LCIA computations for the purposes of model validation.
+A dependency reference in disclosure item~\ref{itm:bg} should include both an activity and a reference flow.
+
+Similarly, LCIA method providers should ensure that their methods make use of stable semantic references to elementary flows and contexts.
+Current research has revealed widespread challenges to finding consistency on the identities of elementary flows despite the community's longstanding awareness of the problem, and several mutually inconsistent reference flow sets exist \citep{Edelen_2017}.
+The conventional understanding of a ``flow'' as comprising a substance and a context together (e.g.
+methane, emissions to air) causes a combinatorial increase in the number of flows in a database, and also multiplies the potential points of disagreement across sources.
+I reiterate Edelen et al.'s (ibid.) suggestion to explicitly regard substances (which can easily include non-material ``flowables'' such as land occupation and transformation) and contexts as independent semantic classes.
+In this case, an elementary flow entry in the disclosure item~\ref{itm:em} would require both a flowable and a context.
+
+\endinput
+
 It is necessary for the entries in disclosure items~\ref{itm:bg} and \ref{itm:em} to be unambiguous and easy to interpret.
@@
 Similar requirements also hold for entries in \ref{itm:em}, which represent rows in $B_f$.
-Current research has revealed widespread challenges to finding consistency on the identities of elementary flows \citep{Edelen_2017}.
-
-In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of those references.
-This is often accomplished using linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.
-The resource at the end of the link functions as a point of agreement: both study author and data user can follow it from anywhere on the Internet to obtain the same information.
-The content pointed to by the hyperlink, along with context provided by \emph{other} references to the same resource, gives meaning to the information and allows it to be curated by the study authors and other members of the community \citep{Khan_2011}.
-% Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.
-Some preliminary descriptions of the entity types involved in LCA computation, and their relationships to one another, have recently been developed \citep{Ciroth_Srocka_2014, Janowicz_WOP_2015, Kuczenski_JCP_2016}.
-
-Before the goal of easily reproducible foreground models can be realized, the linked data foundation must be laid.
-Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that distinguish database, version, configuration, dataset, and product flow to which the reference refers.
-These references should take the form of URIs that can be accessed using any browser and can provide a reader with documentary information describing the dataset, the quantitative dimension of the reference flow, and a mechanism for the reader to obtain access to the data.
-Similarly, LCIA method providers should ensure that their methods make use of stable semantic references to elementary flows and contexts.

### 2017-11-04 aggregation/agg-JIE.tex
Late afternoon read through

@@
   \maketitle
-  \emph{For submission to the Journal of Industrial Ecology; October 2017}
+  \emph{For submission to the Journal of Industrial Ecology; November 2017}
 
@@
 
-  {\red http://pubs.acs.org/doi/pdf/10.1021/acs.est.5b01558}
+%  {\red http://pubs.acs.org/doi/pdf/10.1021/acs.est.5b01558}
 
@@
 Reviewers or data users who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, and applying other impact assessment methods.
-Moreover, because the disclosure of the PSM includes only references to background data without including the data themselves, it is trivial for item~\ref{itm:bg} to include references to several background databases without adding any additional requirements.
 

### 2017-11-04 aggregation/agg-abstract.tex
Late afternoon read through

@@
 \begin{abstract}
-  Many of the challenges facing %interpretation and
-  knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of life cycle inventory models.
-This article presents a framework for computing aggregation results in LCA $\ldots$
+  Many of the challenges facing knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of the product system models used to perform LCA computations.
+  This article presents a framework for distinguishing the portions of an LCA study that are attributable to the design of the model from
+
+  isolating the
+
+  computing aggregation results in LCA $\ldots$
 \end{abstract}

### 2017-11-04 aggregation/agg-basic-lca.tex
Late afternoon read through

@@
 
-Because the objective of the framework is to validate a computation, we begin by providing a mathematical formulation of the LCA problem.
+We begin by providing a mathematical formulation of the LCA problem.
 An LCA computation is performed over a collection of linked unit processes, each of which describes a single operation or reference product.
-The task of LCA can be described in two steps: making an inventory of environmental interventions; and determining their environmental impacts.
-If the processes are assumed to be linear, the problem can be written in terms of matrix algebra.
+The task of procedure can be described in two steps: making an inventory of environmental interventions; and determining their environmental impacts.
+If the processes are assumed to be linear and time-invariant, the problem can be written in terms of matrix algebra.
 There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Heijungs2002, Pauliuk_2015_framework}.
@@
 \end{equation}
-Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values) per unit of a process's reference output; $B$ is the environmental intervention or emission matrix; and $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions with respect to a given impact assessment method.
+Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values)%\footnote{Whether negative values are present or allowed depends on the particular construct or co-production method being used.
+\cite{Majeau_Bettez_2014}}
+per unit of a process's reference output; $B$ is the environmental intervention or emission matrix; and $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions with respect to a given impact assessment method.
 In this model, $\mathbf{y}$ is the externally specified final demand, which is the product or service whose delivery is to be assessed, and $s$ is the numerical impact score or category indicator.
@@
 
-This formulation is equivalent to the implementations of all major LCA software systems, some of which provide the capacity to export the $A$ and $B$ matrices for a given study model.
-However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are the same for all users.
-Thus exporting $A$ and $B$ violates our authority and primacy objectives.
+This formulation implemented in all major LCA software systems, some of which provide the capacity to export the $A$ and $B$ matrices for a given study model.
+However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are not usually modified by study authors.
+Thus exporting $A$ and $B$ does not meet our authority and primacy objectives.
 Additionally, replicating the entire LCIDB is cumbersome because of the large size of the matrices.

### 2017-11-04 aggregation/agg-disclosure.tex
Late afternoon read through

@@
 The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.
-Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$, are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.
+Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$ (or their product $\mathbf{e}\times B_x$) are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.
 These three submatrices together can be considered a complete representation of the PSM.
@@
 These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.
-However, the the challenges of operationalizing this requirement are complex.
-Some implications on the information required for a successful disclosure are considered in the Discussion.
+However, the the challenges of operationalizing this requirement are complex (see Discussion and Supplementary materials).
+%Some implications on the information required for a successful disclosure are considered in the Discussion.

### 2017-11-04 aggregation/agg-intro.tex
Finish discussion?
afternoon work- cites; semantics; editing
Late afternoon read through

@@
 When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.
-ISO 14071, which provides guidelines on critical review, emphasizes that review of the PSM or individual datasets is optional \citep{iso14071}.
+\cite{iso14071}, which provides guidelines on critical review, emphasizes that review of the PSM or individual datasets is optional.
 
-A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.
+%A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.
 Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the PSM and thus, no automatic mechanisms for interpretation or review of published LCA studies.

### 2017-11-04 aggregation/agg-lca-study.tex
Late afternoon read through

@@
 To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.
-Because background databases are prepared independently of any study, their computational results do not vary across the studies in which they are used.
+Background databases are prepared independently of any study and do not typically vary across the studies in which they are used.
 However, the foreground systems themselves cannot be computed without the background.
@@
 As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.
+However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2015}).
 
-For clarity, I will call the augmented $\tilde{A}$ and $\tilde{B}$, in which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.
+%For clarity, I will call thein which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be an
+
+The augmented $\tilde{A}$ and $\tilde{B}$, makes up an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.
 This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.
-However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2015}).
 

### 2017-11-04 aggregation/agg-objective.tex
Late afternoon read through

@@
 Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh2005a, Peters_JLCA_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.
-In the context of this paper, an ``LCA Study'' should be understood as any collection of inventory data under consideration, from a single-operation unit process to a full ISO-style product LCA.
+In the context of this paper, an ``LCA Study'' should be understood as any collection of inventory data under consideration as a group, from a single-operation unit process to a full ISO-style product LCA.
 
@@
 The main goal of the framework is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.
-
-%Furthermore, the aggregation can be computed at any level: a foreground (inventory) result; a background (accumulated life cycle inventory) result; or a computed LCIA score.
-These different levels are discussed in detail in the following sections.
 
@@
 \begin{itemize}
-\item \textbf{Authority}.  The disclosure should include only information for which the author is responsible.
-\item \textbf{Primacy}.  The disclosure should exclude information that is made available by a third party.
+\item \textbf{Authority}.  The contents of the disclosure should authentically represent the outcomes of the author's labor in constructing the PSM.
+\item \textbf{Primacy}.
+The disclosure should exclude information that is made available by a third party.
+(It must, however, still include previously-published data that were re-implemented by the author in the study.)
 \end{itemize}
@@
 \endinput
+
+%Furthermore, the aggregation can be computed at any level: a foreground (inventory) result; a background (accumulated life cycle inventory) result; or a computed LCIA score.
+These different levels are discussed in detail in the following sections.
 

### 2017-11-04 aggregation/agg-pagg.tex
Late afternoon read through

--- 2017-10-25 aggregation/agg-pagg.tex
@@
 This is illustrated in Fig.~\ref{fig:private}c.
-The
+
+The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values, or disclosing a range that includes the actual value.
+This would enable a reader to understand \emph{that} a certain background process or emission was included in the model (and checked by the critical reviewer) without knowing how much, which may jointly satisfy review objectives and disclosure constraints in some cases.

### 2017-11-04 aggregation/agg-review.tex
Late afternoon read through

@@
 Similarly, the set of elementary flows explicitly modeled in the study foreground (item~\ref{itm:em}) expose whether any study-specific modeling of direct emissions was performed and what flows were included.
+
+Moreover, because the disclosure of the PSM includes only references to background data without including the data themselves, it is trivial for item~\ref{itm:bg} to include references to a variety of background databases without adding any additional requirements.
+For instance, a study that makes reference to a set of EPDs published by one firm, while using a background LCI database from another firm, can express both dependencies in an analogous way, even if the two sources are published independently.
 

### 2017-11-07 aggregation/agg-review.tex
minor fixes

@@
 
-Partial aggregation is already used in practice to conceal private data, but current practices used to generate aggregated results from input model configurations do not provide a way for a reviewer to validate the aggregation.
+Partial aggregation is already used in practice to conceal private data, but current methods do not provide a way for a reviewer to validate the result.
 If a study containing confidential information is split into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated directly by the reviewer.

### 2017-11-07 aggregation/agg-semantic.tex
minor fixes

@@
 The content pointed to by the hyperlink, along with context provided by \emph{other} references to the same content, gives meaning to the information and allows it to be curated by the study authors and other members of the community \citep{Khan_2011}.
-Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.
+%Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.
 Some preliminary descriptions of the entity types involved in LCA computation, and their relationships to one another, have recently been developed \citep{Ciroth_Srocka_2014, Janowicz_WOP_2015, Kuczenski_JCP_2016}.
 
-Before the goal of easily reproducible foreground models can be realized, the linked data foundation must be laid.
+Before the goal of easily reproducible foreground models can be realized, this linked data foundation must be laid.
 Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that denote a particular activity in a particular database version and configuration.

### 2017-11-10 revision/reviewer-response.txt
reviewer response

@@
 Tue Aug 08 10:15:39 -0700 2017
+
+Many thanks to the anonymous reviewers for their close review of my manuscript.
+Based in part on their recommendations I have significantly revised the manuscript.
+Every section after the introduction has been either reworked or newly created.
+I have narrowed the scope to concentrate on structured representation of product system models, and to exclude many aspects of publishing those models that were present in the first draft.
+With this narrowing of scope I have been able to add new focus areas on private or confidential information in the model and on implications for critical review, both of which were highlighed by Reviewer #3.
+I have also thoroughly revised the mathematical development to be clearer and more precise in accordance with the suggestions from reviewer #1, which were highly appropriate, and I hope that the reviewer can now make it to the end of the text.
+
+Below I respond to the reviewers' comments individually.  Page and line numbers refer to the revised manuscript.
 
@@
 
+In general, the mathematics have been cleaned up extensively, and I adopted almost all of the reviewer's suggestions.
+The notation should be self-consistent throughout the manuscript (though not, unfortunately, consistent with Heijungs and Suh) and all symbols are properly introduced and defined.
+The supplementary information also includes proofs or derivations for all equations for which this is necessary.
+Finally, I have attempted to introduce the "canonical" LCA foreground study (p.
+X, l xx) in a way that is both easy to understand and useful for the development.
+The comments were very helpful (and in some cases, such as in observing my misuse of the cross product notation, a bit embarrassing).
+I believe the mathematical development has been greatly strengthened in the revision, and I hope the changes make the manuscript more intelligible to the reviewer and ultimately to readers.
+
+1.
+Dependency and ordering.
+Page 5, line 54 states that foreground processes are dependent on background processes but not the other way around.
+Hance the processes can be partially ordered.
+The validity of these statements depend on the meaning of the word "dependent"....
+
+Thank you for this observation.
+In the revised manuscript I have emphasized that "dependent on x" means "x is required to perform the LCA computation."  In this sense, the foreground of the study depends on the background but not vice-versa.
+This is explained at p.X, l XX.
+
+... Another example is the so-called integrated hybrid IOA-LCA, where all quadrants contain non-zero cells.
+
+I added discussion of integrated hybrid models to the section on LCA foreground studies, p. X, l XX.
+
+2. Most equations contain a cross (x) for matrix and vector multiplication. Please do not
+
+Thank you for pointing out this error. I have corrected the notation.
+
+3. Why do you start using "classic formulation ... Heijungs & Suh", but use different symbols (y for f, s for h), which are moreover confusing (because s means something different to Heijungs & Suh).
+
+I have removed the reference to Heijungs and Suh's notation.
+
+4. The notation in eq (1) has y as a column vector and e as a row vector. It would be better to adopt one convention (please make it column vectors), and write a row vector with a transpose symbol...
+
+I have done as the reviewer suggests, both in the text and in the figures.
+
+5. What is the reason of rewriting eq (1) into eq (2)? As I see it, it only complicates, because every appearance of "A^-1" now reads "(I-A')^-1"....
+
+Reviewer #2 also commented on this confusing aspect.
+I have adopted the input-output notation exclusively in the revision, because it is more convenient to represent the ordering in terms of a normalized adjacency graph than a technology matrix.
+The complication to which the reviewer objects is no longer present, because I no longer introduce the "classical" technology matrix formulation at all.
+
+Comments 6-11 are either not applicable in the revision because of notation changes and editing.
+
+Comments 12-13: proofs have been provided for both these relations, which are between equations 3 and 4, and equations 1 and 5 respectively in the new manuscript.
+
+Comments 14-17 and 19 have also been accounted for through improved notation.
+
+In these passages, I have used the tilde modifier to represent any LCI database that is made up of a background database augmented with foreground content and appropriately ordered so that the foreground is to the left / above the background.
+The "canonical" functional unit is now the functional unit that selects a unit output of the first column of the technology matrix.
+The x-tilde, y-tilde, and y_f vectors are defined explicitly on p.
+X l xx, p.X l xx, and p.X l xx respectively.
+
+Comment 18 ... states that the foreground needs no matrix inversion, because you can "traverse the graph". Can you specify concrete formulas...
+
+The passage has been removed.
+However, the equivalency is quite interesting.
+There is a useful treatment in Bapat and Ghorbani, 2014 in Linear Algebra and Applications (DOI: 10.1016/j.laa.2013.03.002), hereafter BG14.
+It is not easy to make the analogy because technology matrices, though derivable from bipartite graphs, are not themselves bipartite.
+The bipartite graph in BG14 theorem 1 essentially can be imagined to correspond to an LCI database in supply/use form, and the adjacency matrix A in BG14 theorem 1 is the normalized technology matrix that is derived from the supply and use matrices.
+
+The traversal process cannot be stated easily as concrete formulas because it is essentially algorithmic (the proof in BG14 is by induction).
+But the algorithm is the same as what is used to iteratively compute the matrix inverse from the direct requirements matrix.
+Sequential multiplications of a unit demand vector by the technology matrix represent sequential steps in a traversal of the graph.
+Every nonzero value in the resulting multiplication will be the product of the exchange values that have been traversed, i.e.
+the node weight of the node that has been arrived at in the traversal.
+If this matrix is sparse and acyclic (lower-triangular), then the traversal will end after a finite number of multiplications that cannot exceed the rank of the matrix.
+The sum of these vectors will report the accumulated weights of each node that was traveled through during the traversal, in the same way that the total requirements matrix can be derived by an infinite sum of powers of the direct requirements matrix.
+But the infinite sum is not required if the graph has no cycles.
+
+Ultimately, the requirement that the foreground has no cycles is overly limiting and not necessary, so I removed it.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-R2- Pauliuk
+R2- Pauliuk (he disclosed his identity to me)
 
 "... sound mathematical description and a guideline and examples of how to extract and format crucial LCA results ...  well suited for publication in JIE."
+
+Thank you for your constructive review.
+In my revision I have steered away from the practical implications of sharing models, which are complex and include challenges that I was unable to address in the prior draft, and focused more on the mathematical framework that would allow these challenges to be addressed in the future.
+Thus the scope of the manuscript has been reduced.
+One advantage of this is that I could focus more on aspects that relate to the quality of the disclosure-- namely, in discerning how varying levels of transparency can be achieved through different partial aggregation strategies.
+This also allowed me to strongly consider the implications for critical review, as requested by Reviewer #3.
+
++ The level of detail at which the results are shared determines the types of analysis that others can perform.
+When only total background emissions are shared a contribution or structural path analysis is rendered impossible.
+Next to figure S1 (which is great and should be moved to the main paper), there should be a table listing the types of analysis that can be performed depending on the level of detail at which the results are reported.
+
+I did move the supplemental figure to the main text; however, I did not explicitly list "the types of analysis that can be performed" because I could not imagine making such a list comprehensive.
+However, in the revision I introduce the concepts of 'review objectives' and 'disclosure constraints' on p.
+X l xx.
+While I do not explicitly mention structural path analysis, this form of analysis is really not applicable to foreground studies because the 'structural paths' are explicit in the patterns of nonzero entries in the foreground.
+SPA essentially provides a method for contribution analysis that is important in relatively dense and cyclically dependent input-output networks (as discussed in Peters 2007; DOI: 10.1016/j.ecolmodel.2006.07.041) but largely superfluous in foreground networks of the type described in the manuscript.
 
@@
 
-" ... we need to provide for unique labelling of data points across databases and computation platforms."
+I agree with this assessment.
+In the revision I emphasize the creation of a "disclosure" in which a portion of the model (up to the entire model) is explicitly reported, while other portions can be held back.
+The idea is that this, if adopted, will advance the quality of reporting of LCAs and "move us up the ladder".
 
-" version clarity"
++ I am missing emphasis on the classification of processes, products, and emissions.
+Reporting the full names of the data objects (processes, products, elementary flows) is important, but ...
+we need to provide for unique labelling of data points across databases and computation platforms."
 
-" reference products"
+"... version clarity ..."
 
-Difference between Heijungs and Leontief style matrices
+"... reference products ..."
+
+I agree, the problem of reaching agreement on the meaning of descriptors is quite challenging.
+The prior version of the manuscript depended on this problem being solved; in the revision, the manuscript is able to elucidate this requirement as an implication of the work without necessarily providing a solution.
+The discussion section entitled "stable semantic references to study elements" provides an explicit statement of these requirements, and they are also addressed in the Supplementary materials.
+
++ Throughout the manuscript there is a slight lack of clearness regarding the distinction between flows and coefficients, the difference between the Heijungs and Suh A-matrix and the Leontief-type A-matrix, and the question of which results can be reproduced from which aggregation level or results.
+More clarity will greatly help the readers less familiar with the computational structure of LCS to grasp the ideas presented here.
+
+I have tried to take this advice to heart in the revision, particularly with the introduction of "Reviewable Private Aggregation" and the new Figure 4.
+
++ Just a general feeling I had when reading the manuscript: There is a lot of generic description on the importance of metadata....
+but I miss the discussion of the real problems ...
+matching concerns ...
+to link flows ...
+automatically to the right background data....
+"
+
+Again, I have tried to take this advice to heart, but to some extent I have reduced the scope of the manuscript to exclude some of these concerns.
+Again the section on "Stable semantic references" is meant to be a call to the community to consider precisely these problems and develop collective solutions.
+
++ The expected consequences for data providers, data users, database providers, and software developers should be mapped out in more detail. Where are the bottlenecks in setting up this process?
+
+In the revised manuscript, these implications are no longer in scope.
+The idea of structured disclosure of product system models is directed at modellers and report authors; it will ultimately bear on software makers but that is a matter to be addressed later.
 

### 2017-11-15 aggregation/agg-JIE.tex
Abstract and conclusion

@@
 
-The current practice of documenting a PSM with a written report sharply limits the ability of readers to interpret and reuse the data.
-It also results in the model being frozen to a particular configuration or set of configurations used to generate the published results.
-Any review objectives involving sensitivity of the model to parameter variations and/or alternative scenarios are limited to the selections considered by the author.
-
-When a study result is published in an aggregated form, often it is beyond the capacity of the critical reviewer to validate the aggregation result, even if the reviewer has privileged access to confidential materials used in preparing the report.
-
-This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.
-A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.
-Once equipped with the PSM, the reviewer or data user can potentially go much further.
-The graph structure encoded in the foreground matrices can be automatically rendered as a process-flow diagram, supplementing or supplanting the hand-crafted system diagrams that currently accompany ISO reports.
-Reviewers or data users who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, and applying other impact assessment methods.
-
   \input{agg-review}     % review objectives and disclosure constraints
@@
 
-\section{Conclusion}
+  \input{agg-conclusion}
 

### 2017-11-15 aggregation/agg-abstract.tex
Abstract and conclusion

@@
   Many of the challenges facing knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of the product system models used to perform LCA computations.
-  This article presents a framework for distinguishing the portions of an LCA study that are attributable to the design of the model from
+  This article presents a framework for formally disclosing the foreground of an LCA study in a way that permits the computations to be inspected, verified, and reproduced by a reader, provided that the reader has access to the same life cycle inventory and impact characterization resources as the author.
+  The framework can also be used to partition a study into public and private portions, allowing both portions to be critically reviewed but omitting the private information from the disclosure.
+  A disclosure is made up of six components, including three lists of entities in the model and three sparse matrices describing their interconnections.
+The entity lists make reference to previously-published resources, including background inventory databases and characterized elementary flows, and the disclosure framework requires both author and reader to agree on the meaning of each of these references.
+  The framework contributes to ongoing efforts within and beyond industrial ecology to improve the reproducibility and verifiability of scholarly works, and if implemented, plots a course toward distributed, platform-independent computation and validation of LCA results.
+\end{abstract}
 
-  isolating the
+\endinput
 
-  computing aggregation results in LCA $\ldots$
-\end{abstract}
+The author's
+
+disclosing the portions of an LCA study that embody the work of the author

### 2017-11-15 aggregation/agg-conclusion.tex
Abstract and conclusion

\section{Conclusion}

The current practice of documenting a PSM with a written report sharply limits the ability of readers to interpret and reuse the data.  It also results in the model being frozen to a particular configuration or set of configurations used to generate the published results.  Any review objectives involving sensitivity of the model to parameter variations and/or alternative scenarios are limited to the selections considered by the author.
When a study result is published in an aggregated form, often it is beyond the capacity of the critical reviewer to validate the aggregation result, even if the reviewer has privileged access to confidential materials used in preparing the report.

This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.  A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.  Once equipped with the PSM, the reviewer or data user can potentially go much further.  Reviewers or data users who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, and applying other impact assessment methods. Portions of the model which have been pre-aggregated can remain private in the disclosure. Further research is required to determine whether or under what conditions aggregation results can be reverse-engineered.

If data providers also make available stable semantic references to background data sets and elementary flow characterizations, then it becomes possible for a stand-alone description of a PSM to be used to reproduce an LCA result.  The same description can be extended or incorporated into subsequent studies by other authors. From this prescription, a framework for distributed, platform-independent LCA computation can be imagined.

### 2017-11-15 aggregation/agg-lca-study.tex
reviewer response
Abstract and conclusion

@@
 
-The simplest conceivable foreground is simply what is given as the final demand $\mathbf{y}$, in other words the list of background processes invoked by the study and their activity levels.
+The simplest conceivable foreground is simply the final demand $\mathbf{y}$, in other words the list of background processes invoked by the study and their activity levels.
 For an LCA computation as expressed in Eq.~\ref{eqn:leontief}, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
@@
 \end{equation}
-The proof of the equivalency of Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary materials.
+The proof of the equivalency of results derived from Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary materials.
 In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining in $\tilde{A}_{flat}$ pertains to the study foreground.

### 2017-11-15 aggregation/agg-review.tex
Abstract and conclusion

@@
 The disclosure of a PSM contains six parts, including three lists and three tables of sparse matrix data.
+The graph structure encoded in the foreground matrices can be automatically rendered as a process-flow diagram, supplementing or supplanting the hand-crafted system diagrams that currently accompany ISO reports.
 While the numeric data are required to critically evaluate and reproduce the study results, substantial information about the study design can be found in the list of background dependencies (\ref{itm:bg}) and emissions (\ref{itm:em}).
@@
 Typically this equation is highly underdetermined (In ecoinvent, for instance, $B_x$ has roughly 1,800 rows and over 12,000 columns).
-If the private input $\mathbf{a}_{d,priv}$ is highly sparse, and the identity of $B_x$ is known, then $\mathbf{a}_{d,priv}$ may be vulnerable to a basis pursuit attack.
+If the private input $\mathbf{a}_{d,priv}$ contains only a few nonzero elements, and the identity of $B_x$ is known, then $\mathbf{a}_{d,priv}$ may be vulnerable to a basis pursuit attack.
 Further research is required to evaluate whether this attack can be implemented in a real situation, and how it may be defended against.

### 2017-11-15 revision/reviewer-response.txt
Pauliuk comments

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-R2- Pauliuk (he disclosed his identity to me)
+R2-
 
@@
 
+Minor comments:
+
++ Please also refer to the allocation problem (co-production) and to the crediting problem (material or waste enter different product systems), and cite some relevant literature here.
+
+Factors such as co-production strategy (Finnveden_1999, Pelletier_2014) can also render study findings unreliable or highly contingent.
+
++ P6L13ff: “An LCA study result is ultimately an assertion that, for some input data known as the “model,”…”  1) That describes Attributional LCA, please specify.
+2) please rephrase the start of this sentence, as it is unclear how (input) data can be a model.
+
+The comment has been updated to read: "An LCA study result is ultimately an assertion that for some constructed PSM, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact."  The applicability of the statement is not limited to attributional LCA.
+
++ P8L31: Please also cite DOI 10.1111/jiec.12142 (Majeau-Bettez et al. 2014) here for the link between supply and use tables and input-output models in LCA.
+
+This is now done.
+
++ P8L40: “generated non-reference outputs (negative values)” Only for some constructs.
+In the case of partitioning allocation there are no negative inflows as input requirements are split so that only single output process descriptions result.
+
+Noted in a footnote.
+
++ Equations 3 and 4 are not equivalent. Equ. 4 cannot be used for a structural path analysis, for example, as it contains aggregate information only. Please be more specific.
+
+Updated to read "...equivalency of results derived from Eq.s 3 and 4..."
+
++ P11L22: “The foreground graph Af can be distilled into a vector of…” please use more scientific language here.
+A_f is not a graph but a description of a graph, and distilling reminds me of other things than science…
+
+The passage has been removed.
+
++ Figure S1 provides a very good summary of the different aggregation steps; consider moving it to the main paper.
+
+Thank you for the compliment and the suggestion; this has been done in the revision.
+
++ There are two A-matrices in LCA: One is the Suh and Heijungs matrix of flows and the other one is the matrix of technical coefficients.
+Please distinguish better between the two, e.g., on P12L45, where you write about flows in the A_f matrix, but it should read ‘coefficients in the A_f matrix’.
+
+In the revision, only the Leontief matrix is used. Other passages have been updated or removed.
+
++ P13L31: Only aggregated results can be reproduced from the aggregated foreground. Please make this limitation explicit.
+
+Noted: "The aggregated result can also be computed from the aggregated foreground..." (p. X l x)
+
++ P31L41: “Generally, any directed graph can be represented in Af ,…” That is not correct.
+Af can only contain bipartite graphs with a 1:1 correspondence between industries and nodes, that means, an industrial inventory (SUT) after allocation.
+Cf.
+DOI 10.1111/jiec.12306 for the argument.
+
+Noted: "Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e. any allocated supply-and-use table..."
+
++ Figures 4 and 5: Units are missing.
+Is the rightmost column the row sum of the detailed exchange table?
+Please provide more explanations in the Table caption.
+The symbols A_f, B_f, etc.
+that are frequently used in the text should be displayed here as well.
+
+These exhibits have been moved into the supplementary materials.
+NEED MORE FINISH HERE
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-11-15 aggregation/agg-basic-lca.tex
Pauliuk comments

@@
 \end{equation}
-Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values)%\footnote{Whether negative values are present or allowed depends on the particular construct or co-production method being used.
+Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values)\footnote{Whether negative values are present or allowed depends on the particular construct or co-production method being used.
 \cite{Majeau_Bettez_2014}}

### 2017-11-15 aggregation/agg-intro.tex
Pauliuk comments

@@
 These models are often highly complex and include countless modeling decisions, approximations, and assumptions made by a study author \citep{Lloyd2007, reap2008_I}.
+Factors such as co-production strategy \citep{Finnveden_1999, Pelletier_2014} can also render study findings unreliable or highly contingent.
 Consequently, while comparative evaluations can be made regarding alternative cases or scenarios within a single study, comparisons across studies are much more challenging \citep{Heath2012, Henriksson2014}.
@@
 is a crucial challenge to the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.
-A further challenge to transparency can be found in the use of input data that are confidential or proprietary \citep{Kuczenski_2017}.
+A further challenge to transparency can be found in the use of input data that are confidential or proprietary \citep{Kuczenski_ESD_2017}.
 One consequence of all these challenges is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.
@@
 As noted above, this level of consensus is hard to realize, in part because of the conflicting implications inherent in different analytic modes.
-Other factors such as allocation strategy \citep{Pelletier_2014} can also render study findings unreliable or highly contingent.
 Plevin et al.\ famously observed in the strongest terms that careless reporting of attributional LCA study results can distort the significance of findings in a policy context \citep{Plevin_2013}, a shortcoming that can be found in other modes as well \citep{Brandao_2014}.

### 2017-11-16 aggregation/agg-foreground.tex
Supplementary stuffs

@@
 \section{Foreground Configurations}
-
-Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.
-Most product system models would include many of these elements mixed together.
-A reviewer with access to items \ref{itm:fg} and \ref{itm:af} in the disclosure would be able to construct $A_f$ and automatically create a process-flow diagram, annotated with information about each node.
 
@@
 
+Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.  Most product system models would include many of these elements mixed together.  
 The first (a) is a sequential model, in which each node requires one foreground input and generates one output.
@@
 The reference flow $y_1$ is consumed in two different places by the other fragments.
+
+As noted, a reviewer with access to items \ref{itm:fg} and \ref{itm:af} in the disclosure would be able to construct $A_f$ and automatically create a process-flow diagram, annotated with information about each node.

### 2017-11-16 aggregation/agg-requirements.tex
comparative review?
Late afternoon read through
Supplementary stuffs

--- 2017-10-30 aggregation/agg-discussion.tex
@@
 \subsubsection{Authority and Primacy}
+
+It may be observed that many LCA studies make use of data sources that have been previously published but that have not been included in any LCI reference or background database.
+As long as these data sources can be integrated seamlessly into an LCA computation, it is not necessary to reproduce them in a disclosure.
+However, in most applications, previously published data must be re-implemented by the author in the LCA software context, and often this re-implementation requires a re-interpretation of the data source as a unit process inventory, where one exchange is recognized as a reference flow and other exchange values are reported in proportionality to the reference flow.
+
+Because of these conditions, in the vast majority of cases the LCA study disclosure must include the author's reimplementations in order to achieve both the aims for authority and primacy.
+In so doing, the disclosure enables a critical reviewer to evaluate whether the author's implementations are generally correct and complete.
+
+In the future, the inclusion of external data automatically can be accomplished in the same way that reference LCI data could conceivably be included automatically: by the data providers making their information available  using a stable semantic reference to a specialized Web-based application programming interface (API).
+This would have the benefits of enabling downstream users to access the information without having to re-implement it, thus reducing the size of the disclosure necessary to describe the PSM and simplifying the task of the modeler.

### 2017-11-16 figures/aluminum-secondary.tex
Supplementary stuffs

@@
 {\scriptsize\sffamily
-\begin{tabularx}{\textwidth}{|>{\hangindent=2ex}X|c@{~}c@{~}c@{~}c@{~}|c|}
+\begin{tabularx}{\textwidth}{|>{\hangindent=3ex}X|c@{~}c@{~}c@{~}c@{~}|c|}
 \hline
-\everypar{\hangindent1em \hangafter1}
-(node) Foreground flow \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & \\ 
+(node) Foreground flows -- $A_f$ \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & \\ 
 \hline
-(0) Aluminum, secondary, ingot, from automotive scrap, at plant (kg) [RNA] & \refbox  &  &  &  & \\ 
-(1) Aluminum recovery, transport, to plant (kg) [RNA] & 1.03 & \refbox  &  &  & \\ 
-(2) Quicklime, at plant (kg) [RNA] & 2.35e-05 &  & \refbox  &  & \\ 
-(3) Limestone, at mine (kg) [RNA] &  &  & 1.87 & \refbox  & \\ 
+(0) Aluminum, secondary, ingot, from automotive scrap, at plant [RNA] (kg) & \refbox  &  &  &  & \\ 
+(1) Aluminum recovery, transport, to plant [RNA] (kg) & 1.03 & \refbox  &  &  & \\ 
+(2) Quicklime, at plant [RNA] (kg) & 2.35e-05 &  & \refbox  &  & \\ 
+(3) Limestone, at mine [RNA] (kg) &  &  & 1.87 & \refbox  & \\ 
 \hline
@@
 \hline
-Input: CUTOFF Disposal, solid waste, unspecified, to sanitary landfill [CUTOFF Flows] & \dependency &  & \dependency &  & 0.0842\\ 
-Input: CUTOFF Filter media, at plant [CUTOFF Flows] & \dependency &  &  &  & 4.79e-05\\ 
-Input: CUTOFF Lube oil, at plant [CUTOFF Flows] & \dependency &  &  &  & 8.75e-07\\ 
-Input: CUTOFF Treatment gases, unspecified, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.003\\ 
-Input: CUTOFF Alloying additives, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.021\\ 
-Input: CUTOFF Chemicals, unspecified, used for wastewater treatment [CUTOFF Flows] & \dependency &  &  &  & 0.008\\ 
-Input: CUTOFF Grain refiners, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.0003\\ 
-Input: CUTOFF Treatment salts, unspecified, at plant [CUTOFF Flows] & \dependency &  &  &  & 0.009\\ 
-Input: CUTOFF Packaging, unspecified, at plant [CUTOFF Flows] & \dependency &  &  &  & 1.67e-05\\ 
-Input: CUTOFF Aluminum scrap, automotive [CUTOFF Flows] & \dependency &  &  &  &  1.03\\ 
+Input: CUTOFF Disposal, solid waste, unspecified, to sanitary landfill [CUTOFF Flows] (kg) & \dependency &  & \dependency &  & 0.0842\\ 
+Input: CUTOFF Filter media, at plant [CUTOFF Flows] (kg) & \dependency &  &  &  & 4.79e-05\\ 
+Input: CUTOFF Lube oil, at plant [CUTOFF Flows] (kg) & \dependency &  &  &  & 8.75e-07\\ 
+Input: CUTOFF Treatment gases, unspecified, at plant [CUTOFF Flows] (kg) & \dependency &  &  &  & 0.003\\ 
+Input: CUTOFF Alloying additives, at plant [CUTOFF Flows] (kg) & \dependency &  &  &  & 0.021\\ 
+Input: CUTOFF Chemicals, unspecified, used for wastewater treatment [CUTOFF Flows] (kg) & \dependency &  &  &  & 0.008\\ 
+Input: CUTOFF Grain refiners, at plant [CUTOFF Flows] (kg) & \dependency &  &  &  & 0.0003\\ 
+Input: CUTOFF Treatment salts, unspecified, at plant [CUTOFF Flows] (kg) & \dependency &  &  &  & 0.009\\ 
+Input: CUTOFF Packaging, unspecified, at plant [CUTOFF Flows] (kg) & \dependency &  &  &  & 1.67e-05\\ 
+Input: CUTOFF Aluminum scrap, automotive [CUTOFF Flows] (kg) & \dependency &  &  &  &  1.03\\ 
 \hline
 \hline
-Background Dependencies \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & $\tilde{a_d}$\\ 
+Background Dependencies -- $A_d$ \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & $\tilde{a_d}$\\ 
 \hline
-Diesel, combusted in industrial boiler (l) [RNA] &  &  & \dependency & \dependency & 4.79e-08\\ 
-Transport, train, diesel powered (t*km) [RNA] &  & \dependency & \dependency &  & 0.0415\\ 
-Electricity, at grid, US, 2000 (kWh) [RNA] & \dependency &  & \dependency & \dependency & 0.668\\ 
-Transport, combination truck, diesel powered (t*km) [RNA] &  & \dependency & \dependency &  & 0.374\\ 
-Natural gas, combusted in industrial boiler (m3) [RNA] & \dependency &  & \dependency & \dependency & 0.223\\ 
-Bituminous coal, combusted in industrial boiler (kg) [RNA] &  &  & \dependency & \dependency & 4.04e-06\\ 
-Gasoline, combusted in equipment (l) [RNA] &  &  &  & \dependency & 2.25e-09\\ 
-Liquefied petroleum gas, combusted in industrial boiler (l) [RNA] &  &  & \dependency &  & 7.57e-10\\ 
-Transport, barge, average fuel mix (t*km) [RNA] &  &  & \dependency &  & 5.66e-07\\ 
+Transport, barge, average fuel mix [RNA] (t*km) &  &  & \dependency &  & 5.66e-07\\ 
+Liquefied petroleum gas, combusted in industrial boiler [RNA] (l) &  &  & \dependency &  & 7.57e-10\\ 
+Transport, combination truck, diesel powered [RNA] (t*km) &  & \dependency & \dependency &  & 0.374\\ 
+Diesel, combusted in industrial boiler [RNA] (l) &  &  & \dependency & \dependency & 4.79e-08\\ 
+Natural gas, combusted in industrial boiler [RNA] (m3) & \dependency &  & \dependency & \dependency & 0.223\\ 
+Transport, train, diesel powered [RNA] (t*km) &  & \dependency & \dependency &  & 0.0415\\ 
+Electricity, at grid, US, 2000 [RNA] (kWh) & \dependency &  & \dependency & \dependency & 0.668\\ 
+Bituminous coal, combusted in industrial boiler [RNA] (kg) &  &  & \dependency & \dependency & 4.04e-06\\ 
+Gasoline, combusted in equipment [RNA] (l) &  &  &  & \dependency & 2.25e-09\\ 
 \hline
-Foreground Emissions \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & $\tilde{b_f}$\\ 
+Foreground Emissions -- $B_f$ \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & $\tilde{b_f}$\\ 
 \hline
-Output: Lead [air, unspecified] & \dependency &  &  &  & 2.16e-07\\ 
-Output: Suspended solids, unspecified [water, unspecified] & \dependency &  &  &  & 2.56e-05\\ 
-Output: Particulates, unspecified [air, unspecified] & \dependency &  & \dependency & \dependency & 1.78e-07\\ 
-Output: NMVOC, non-methane volatile organic compounds [air, unspecified] & \dependency &  &  &  & 4.7e-05\\ 
-Output: BOD5, Biological Oxygen Demand [water, unspecified] & \dependency &  &  &  & 1.35e-09\\ 
-Output: COD, Chemical Oxygen Demand [water, unspecified] & \dependency &  &  &  & 8.3e-07\\ 
-Output: Sulfur dioxide  [air, unspecified] &  &  & \dependency &  & 3.52e-09\\ 
-Output: Dissolved solids [water, unspecified] & \dependency &  &  &  & 2.38e-08\\ 
-Output: carbon dioxide [air, unspecified] &  &  & \dependency &  & 1.8e-05\\ 
-Input: Limestone [resource, ground-] &  &  &  & \dependency & 4.39e-05\\ 
-Output: Heavy metals, unspecified [water, unspecified] & \dependency &  &  &  & 2.25e-05\\ 
-Output: Organic substances, unspecified [water, unspecified] & \dependency &  &  &  & 1.29e-06\\ 
-Output: Acids, unspecified [air, unspecified] & \dependency &  &  &  & 4.31e-05\\ 
+Output: Lead [air, unspecified] (kg) & \dependency &  &  &  & 2.16e-07\\ 
+Output: Suspended solids, unspecified [water, unspecified] (kg) & \dependency &  &  &  & 2.56e-05\\ 
+Output: Particulates, unspecified [air, unspecified] (kg) & \dependency &  & \dependency & \dependency & 1.78e-07\\ 
+Output: NMVOC, non-methane volatile organic compounds [air, unspecified] (kg) & \dependency &  &  &  & 4.7e-05\\ 
+Output: BOD5, Biological Oxygen Demand [water, unspecified] (kg) & \dependency &  &  &  & 1.35e-09\\ 
+Output: COD, Chemical Oxygen Demand [water, unspecified] (kg) & \dependency &  &  &  & 8.3e-07\\ 
+Output: Sulfur dioxide  [air, unspecified] (kg) &  &  & \dependency &  & 3.52e-09\\ 
+Output: Dissolved solids [water, unspecified] (kg) & \dependency &  &  &  & 2.38e-08\\ 
+Output: carbon dioxide [air, unspecified] (kg) &  &  & \dependency &  & 1.8e-05\\ 
+Input: Limestone [resource, ground-] (kg) &  &  &  & \dependency & 4.39e-05\\ 
+Output: Heavy metals, unspecified [water, unspecified] (kg) & \dependency &  &  &  & 2.25e-05\\ 
+Output: Organic substances, unspecified [water, unspecified] (kg) & \dependency &  &  &  & 1.29e-06\\ 
+Output: Acids, unspecified [air, unspecified] (kg) & \dependency &  &  &  & 4.31e-05\\ 
 \hline

### 2017-11-16 figures/potato-organic.tex
Supplementary stuffs

market for potato, organic [GLO]

{\scriptsize\sffamily
\begin{tabularx}{\textwidth}{|>{\hangindent=3ex}X|c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}|c|}
\hline
(node) Foreground flows -- $A_f$ \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & \\ 
\hline
(0) potato, organic [GLO] (kg) & \refbox  &  &  &  &  &  &  &  &  & \\ 
(1) potato, organic [CH] (kg) & 0.024 & \refbox  &  &  &  &  &  &  &  & \\ 
(2) potato, organic [RoW] (kg) & 0.976 &  & \refbox  &  &  &  &  &  &  & \\ 
(3) potato seed, organic, at farm [CH] (kg) &  &  &  & \refbox  & 0.024 &  &  &  &  & \\ 
(4) potato seed, organic, at farm [GLO] (kg) &  &  &  &  & \refbox  &    1 &    1 &  &  & \\ 
(5) potato seed, organic, for setting [RoW] (kg) &  &  &  &  &  & \refbox  &  &  & 0.976 & \\ 
(6) potato seed, organic, for setting [CH] (kg) &  &  &  &  &  &  & \refbox  &  & 0.024 & \\ 
(7) potato seed, organic, at farm [RoW] (kg) &  &  &  &  & 0.976 &  &  & \refbox  &  & \\ 
(8) potato seed, organic, for setting [GLO] (kg) &  & 0.11 & 0.11 & 0.16 &  &  &  & 0.16 & \refbox  & \\ 
\hline
Foreground Node Weights $\tilde{x}$ &    1 & 0.024 & 0.976 & 0.00315 & 0.131 & 0.128 & 0.00315 & 0.128 & 0.131& \\ 
\hline
\hline
Background Dependencies -- $A_d$ \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & $\tilde{a_d}$\\ 
\hline
building, multi-storey [GLO] (m3) &  &  &  &  &  & \dependency & \dependency &  &  & 5.24e-06\\ 
electricity, low voltage [GLO] (kWh) &  &  &  &  &  & \dependency &  &  &  & 0.00933\\ 
potato haulm cutting [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 5.04e-05\\ 
tillage, hoeing and earthing-up, potatoes [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.000101\\ 
transport, tractor and trailer, agricultural [GLO] (metric ton*km) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.00113\\ 
transport, freight, light commercial vehicle [GLO] (metric ton*km) & \dependency &  &  &  & \dependency &  &  &  &  & 0.0258\\ 
tillage, ploughing [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 5.04e-05\\ 
green manure, organic, until March [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 5.04e-05\\ 
transport, freight, lorry, unspecified [GLO] (metric ton*km) & \dependency &  &  &  & \dependency &  &  &  &  & 0.404\\ 
transport, freight train [GLO] (metric ton*km) & \dependency &  &  &  & \dependency &  &  &  &  & 0.138\\ 
transport, freight, inland waterways, barge [GLO] (metric ton*km) & \dependency &  &  &  & \dependency &  &  &  &  & 0.0857\\ 
copper oxide [GLO] (kg) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.000107\\ 
tillage, harrowing, by spring tine harrow [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 5.04e-05\\ 
potato planting [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 5.04e-05\\ 
solid manure loading and spreading, by hydraulic loader and spreader [GLO] (kg) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.717\\ 
potato grading [GLO] (kg) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  &  1.13\\ 
tillage, harrowing, by rotary harrow [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 5.04e-05\\ 
harvesting, by complete harvester, potatoes [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 5.04e-05\\ 
transport, freight, sea, transoceanic ship [GLO] (metric ton*km) & \dependency &  &  &  & \dependency &  &  &  &  & 0.525\\ 
liquid manure spreading, by vacuum tanker [GLO] (m3) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.000565\\ 
electricity, low voltage [CH] (kWh) &  &  &  &  &  &  & \dependency &  &  & 0.00023\\ 
irrigation [GLO] (m3) &  &  & \dependency &  &  &  &  & \dependency &  & 0.0173\\ 
tillage, currying, by weeder [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.000101\\ 
application of plant protection product, by field sprayer [GLO] (ha) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.000222\\ 
irrigation [CH] (m3) &  & \dependency &  & \dependency &  &  &  &  &  & 0.000426\\ 
\hline
Foreground Emissions -- $B_f$ \rule[-3pt]{0pt}{12pt} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & $\tilde{b_f}$\\ 
\hline
Input: Occupation, construction site [natural resource, land] (m2*year) &  &  &  &  &  & \dependency & \dependency &  &  & 2.1e-06\\ 
Input: Transformation, from unspecified [natural resource, land] (m2) &  &  &  &  &  & \dependency & \dependency &  &  & 1.05e-06\\ 
Input: Transformation, to industrial area [natural resource, land] (m2) &  &  &  &  &  & \dependency & \dependency &  &  & 1.05e-06\\ 
Input: Energy, gross calorific value, in biomass [natural resource, biotic] (MJ) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  &  3.87\\ 
Output: Phosphate [water, ground-] (kg) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 6.25e-06\\ 
Output: Cadmium, ion [water, ground-] (kg) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 1.14e-09\\ 
Output: Zinc, ion [water, ground-] (kg) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 1.35e-06\\ 
Output: Nitrate [water, ground-] (kg) &  & \dependency & \dependency & \dependency &  &  &  & \dependency &  & 0.00696\\ 
$\ldots$ (31 rows omitted) &  &  &  &  &  &  &  &  & \\ 
\hline
\end{tabularx}
}

### 2017-11-16 aggregation/publication-example.tex
preparing for submission

--- 2017-04-01 manuscript/publication-example.tex
@@
 Instead, a black square indicates the presence of a nonzero value.
-The systems are also reproduced as structured publications in the supplementary materials, formatted as Excel spreadsheets.
+%The systems are also reproduced as structured publications in the supplementary materials, formatted as Excel spreadsheets.
 
@@
 
-The structured publication of the secondary aluminum model includes a total of 47 data points about 36 entities.
+%The structured publication of the secondary aluminum model includes a total of 47 data points about 36 entities.
 The $A_f$, $A_d$, and $B_f$ matrices are reported in sparse format, along with the aggregation vectors.
@@
 
-The structured publication of the organic potato model reports the $A_f$, $A_d$, and $B_f$ matrices in full rather than sparse format to enable a more visually immediate review.
+%The structured publication of the organic potato model reports the $A_f$, $A_d$, and $B_f$ matrices in full rather than sparse format to enable a more visually immediate review.
 In addition, the $E$ matrix is reported in full, for all emissions found in the model, rather than only the emissions found in the foreground.

### 2017-11-16 revision/reviewer-response.txt
R#3

@@
 
-Noted: "Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e. any allocated supply-and-use table..."
+Noted: "Generally, the foreground matrix A_f can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e. any allocated supply-and-use table..."
 
@@
 
-Why JIE and not JLCA? short reason- better community for 'scary new' ideas. but I can't just say that...
++  The objective of the article is clear and the case for it well made.
+Despites minor mistakes the article is also well written.
+However, the article focuses on a very specific aspect of LCA and the results might be more suitable for the Int.
+J LCA.
 
-(1) Page 17 line 12: To my knowledge, this short sentence question the potential usefulness of the approach developed in the article.
+Thank you for your positive review.
+I have considered submission to JLCA but I feel the JIE readership and editorial staff is more aligned with the paper's objectives of advancing disclosure and transparency at the present time.
+Additionally I was not happy with the formatting of the mathematics obtained in Kuczenski (2015), "Partial ordering of life cycle inventory databases", which I published in JLCA.
+
++ That said I found the paper interesting and I have some general comments which focus mostly on the discussion. ....
++ The added value of the proposed structured publication of results from a machine reading perspective is obvious.
+However its value added for the review process, comparison across models, benchmarking, etc.
+in the context of private data must be addressed from the outset.
+Otherwise the relevance and practicality of the proposed structure seems unclear.
+
+Your commentary on the discussion was the most influential regarding the path I took with the revision.
+I strongly agree with your argument that overriding concerns about confidentiality and critical review both affect the usefulness of the proposed framework.
+Therefore in the revision I focused less on "publication" of the results and introduced both privacy protection and critical review as fundamental concerns.
+
++ The mathematical formulation is straightforward and shows a minimal set of data which should be included in a “research object”.
+
+Please note that in the revision I have used the term 'disclosure' to refer to the "research object" of the initial manuscript.
+
++ 1)      Page 17 line 12: To my knowledge, this short sentence question the potential usefulness of the approach developed in the article.
 Data confidentiality in the proposed research object should be discussed more thoroughly, in particular with respect to the current critical review process in ISO LCA.
 
-We take the author to refer to page 17 of the PDF proof, which is manuscript page 15.
-Line 12 reads "...for reasons of privacy, an author may wish to conceal some or all data points, or include the data points but exclude the actual data values.
-This is allowable if it is consistent with the author’s goal, and it will accordingly limit the ability of a data user to reproduce the model."
+The referenced sentence was, "This is allowable if it is consistent with the author’s goal, and it will accordingly limit the ability of a data user to reproduce the model."
 
-Definitely. OK.
+In the revision, the perspective taken is best summarized on p. X l. XX, "the reviewer must be granted access to
+both the public and private partitions in order to confirm that the partitioning is valid according
+to Eq. 9, and that b priv was computed correctly. However, the private information used in the
+computation can still be withheld from the disclosure."  Here the reviewer, rather than the data user, is the primary audience, which is more consistent with current practice.
+The "disclosure completeness" metric (Eq.
+13) also provides a measure of how much of the model's impacts are disclosed, which at least provides some intelligence to the general reader about the degree of transparency.
+If this metric is close to (or equal to) 0, then no significant information is disclosed.
+THis may not be desirable, but at least it is clear.
 
-(2) Page 17 line 40: Similarly, data confidentiality might result in a very sparse entity map.
++ 2)      Page 17 line 40: Similarly, data confidentiality might result in a very sparse entity map.
 Can such a research object be organized in order to safeguard confidentiality and allow reproducibility of the results?
+
+In the revision, this question is somewhat out of scope.  Again, I agree with you and the focus is more on what the author chooses to disclose, for better or worse.
+
++ 3)      Page 24 line 26: This is where it gets interesting, especially when the readers move to the supplementary information and pages 44 and 45, sections 2.4 and 2.5.
+Here the proposed research object is compared to current practice and the aggregation of foreground results for the sake of confidentiality.
+
+Thank you for these comments.
+The material you referenced on "pages 44 and 45" was of course in the supplementary materials in the first version, and has now been moved to the main paper.
+Additionally, "section 2.4", which dealt with partial background aggregation, has been expanded into a new main body section entitled "Reviewable Private Aggregation."  I look forward with great interest to hear your thoughts on the revised version.

### 2017-11-16 aggregation/agg-lca-study.tex
Pauliuk comments
Rev01 Submitted!

@@
 The augmented $\tilde{A}$ and $\tilde{B}$, makes up an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.
-This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.
+This formulation can be applied to the vast majority of LCA case studies, and very likely to all case studies prepared using commercial LCA software.
 
@@
 \end{equation}
-The study result can also be computed from the aggregated foreground:
+The aggregated result can also be computed from the aggregated foreground:
 \begin{equation}

### 2017-11-20 aggregation/publication-example.tex
Correct column numbers

@@
 Ecoinvent version 3 is provided in three different system models that reflect different linking strategies.
-The example is drawn from the ``Allocation at the point of substitution'' or APOS model, which includes 11,420 processes that produce 12,966 product flows.
+The example is drawn from the ``Allocation at the point of substitution'' or APOS model of Ecoinvent 3.2, which includes 11,420 processes that produce 12,966 product flows.
 Of these, 10,282 are background flows and the rest are foreground flows.
 
-The example system reports production of organic potatoes supplied to the global market (Table \ref{table:potato}).
-The foreground includes nine nodes, of which six (nodes 3 through 8) form a cyclic dependency associated with the production of potato seeds.
+The example system reports production of organic potatoes supplied to the global market (Figure \ref{table:potato}).
+The foreground includes nine nodes, of which six (nodes 3 through 8) contain cyclic dependencies associated with the production of potato seeds.
 The Ecoinvent database's use of ``markets'' as mixer processes is evident in the table: node 0 is a mixer process that combines Swiss (``CH'') production (node 1) weighted at 2.4\%  with rest-of-World (``RoW'') production (node 2) weighted at 97.6\%.
-That same market split (2.4 / 97.6) can also be seen in nodes 7 and 8.
+That same market split (2.4 / 97.6) can also be seen in nodes 4 and 8.
 
-Nodes 4 and 5, which make seeds ready ``for setting,'' are mixed by node 8.
-Although nodes 4 and 5 are geographically distinguished (CH vs RoW), both can be seen to consume potato seed from the global market.
+Nodes 5 and 6, which make seeds ready ``for setting,'' are mixed by node 8.
+Although nodes 5 and 6 are geographically distinguished (CH vs RoW), both can be seen to consume potato seed from the global market.
 
-Looking at the dependency and emission lists, the different ``signatures'' of different kinds of processes can be seen: nodes 1, 2, 3, and 6 are clearly agricultural processes that require irrigation, tillage, manure and so on.
-Nodes 0 and 7 are visible as market processes, their only requirements being transport processes.
-Nodes 4 and 5 each consume electricity and require use of a ``multi-storey building.''  Taken together, the CH-locality processes appear to use CH irrigation and electricity supply but are otherwise similar to their RoW counterparts.
+Looking at the dependency and emission lists, the different ``signatures'' of different kinds of processes can be seen: nodes 1, 2, 3, and 7 are clearly agricultural processes that require irrigation, tillage, manure and so on.
+Nodes 0 and 4 are visible as market processes, their only requirements being transport processes.
+Nodes 5 and 6 each consume electricity and require use of a ``multi-storey building.''  Taken together, the CH-locality processes appear to use CH irrigation and electricity supply but are otherwise similar to their RoW counterparts.
 The product model includes no cutoff flows.

### 2018-02-02 aggregation/agg-foreground.tex
Multioutput case in SI

@@
-\section{Foreground Configurations}
+\section{Preparing the Foreground Matrix}
+
+The development in the main text assumes the product system model is already described as a symmetric collection of single-output processes (activities) and flows (commodities) related by coefficients.
+In concrete terms, this means that:
+\begin{itemize}
+\item each node in the foreground (column shared by $A_f$, $A_d$, and $B_f$) corresponds to a particular flow in the model's foreground (row of $A_f$);
+\item a unit activity of each node in the foreground corresponds to a unit output of each flow;
+\item the coefficients in $A_f$, $A_d$, and $B_f$ are ratios expressing the quantity of the flow (row), in its native unit, exchanged per unit of the reference flow (column). 
+\end{itemize}
+
+In short: each foreground column has both an intrinsic reference flow and a unit of measure, and the use of this flow by other processes must be made with regard to this measure.
+The activity level of each foreground node necessarily equals the magnitude of the total reference flow emanating from the node.
+For background dependencies and emissions (rows of $A_d$ and $B_f$, respectively), the modeler must take care to use the units established by the references, but in the foreground case the modeler must ensure to be self-consistent in preparing the $A_f$ matrix.
+
+\subsection{Single-Output Foreground Configurations}
 
@@
 
-Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.  Most product system models would include many of these elements mixed together.  
+Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.
+Most product system models would include many of these elements mixed together.
+All the activities in these examples are single-output.
 The first (a) is a sequential model, in which each node requires one foreground input and generates one output.

### 2018-02-02 aggregation/agg-multioutput.tex
Multioutput case in SI

\subsection{Multi-Output Foreground Configurations}

The treatment of multi-functional activities in LCA is a topic of constant discussion among practitioners and theorists.  While the question of the correct, appropriate, or ideal method of apportioning impacts among co-products cannot be resolved in general, the utility of a disclosure framework must be to \textit{express} transparently how co-production was resolved within a given foreground model.  Majeau-Bettez et al. (2014) recently provided a general framework (referred to hereafter as MWS14) for describing the treatment of co-products, and I adopt that framework here to show how the technique used to handle co-production can be disclosed transparently for foreground processes.

\subsubsection{Example System}

In all strategies, an abbreviated example system drawn from Ecoinvent, reporting chlor-alkali electrolysis using a membrane cell, is used.  The un-allocated system is shown in Table~\ref{tbl:chlor-alkali}.

\begin{table}
  \begin{center}
  \caption{Chlor-Alkali Electrolysis, membrane cell [RER] (Ecoinvent 3.2; excerpt)}
  \label{tbl:chlor-alkali}
  \begin{tabular}{rccl}
    \toprule
    \bf Flow & \bf Unit & \bf Direction & \bf Value \\
    \midrule
    \multicolumn{4}{l}{Reference:} \\
    Chlorine, gaseous & kg & Output & 1 \\
    Hydrogen, liquid & kg & Output & 0.028 \\
    Sodium hydroxide & kg dry & Output & 1.13 \\
    \midrule
    \multicolumn{4}{l}{Requirements:} \\
    Electricity, medium voltage & kWh & Input & 2.97 \\
    Sodium chloride, powder & kg & Input & 1.75 \\
    Chemical factory, organics & unit & Input & 4e-10 \\
    Sludge, NaCl electrolysis & kg & Output & 0.0153 \\[1ex]
    \multicolumn{4}{l}{Emissions:} \\
    Chloride [16887-00-6], surface water & kg & Output & 0.015 \\
    Carbon dioxide [124-38-9], urban air & kg & Output & 0.0031 \\
    \bottomrule
  \end{tabular}
  \end{center}
\end{table}

The objective is to disclose explicitly how the multi-output system was resolved to single-output activities for the purposes of review, reproduction, or modification.  Therefore, the raw (un-allocated) process inventory is included in all cases.  Examples are provided for partitioning allocation, classical system expansion, and consequential system expansion.

\subsubsection{Partitioning Allocation}

According to MWS14, Partitioning allocation (PA) is a ``production-balanced model that artificially splits the requirements of an activity across its different outputs.''  The exact manner of the partitioning need not be specified, as long as the parts sum up to the whole.  In other words, for a process with $n$ outputs the modeler has $n-1$ degrees of freedom in specifying the allocated inventories, with the requirement that the $n$th inventory equals the remainder.

In the disclosure framework, any manner of production-balanced partitioning can be represented by including the raw, un-allocated inventory as one foreground column, choosing any output as the reference.  The other $n-1$ outputs are then included as arbitrarily-specified allocated inventories in additional columns, for a total of $n$ columns.  Each column must be normalized to a unit reference.  The reference flow not explicitly specified will simply be computed as the remainder when the allocated inventories are subtracted from the un-allocated inventory.

\begin{table}
  \begin{center}
  \caption{Chlor-Alkali Electrolysis as a foreground model, partitioning allocation}
  \label{tbl:partition}
  \footnotesize\sffamily
  \begin{tabular}{l|ccc}
    \midrule
    \bf $A_f$ & 0 & 1 & 2 \\
    \midrule
    \textbf{Chlorine, gaseous (kg)} & $\ast$ & & \\
    Hydrogen, liquid (kg) & -0.028 & $\ast$ & \\
    Sodium hydroxide (kg dry) & -1.13 & & $\ast$ \\
    \midrule
    \bf $A_d$ & 0 & 1 & 2 \\
    \midrule
    Electricity, medium voltage (kWh) & 2.97 & 1.38 & 1.38 \\
    Sodium chloride, powder (kg) & 1.75 & 0.811 & 0.811 \\
    Chemical factory, organics (unit) & 4e-10 & 1.9e-10 & 1.9e-10 \\
    Sludge, NaCl electrolysis (kg) & 0.0153 & 0.00709 & 0.00709 \\
    \midrule
    \bf $B_f$ & 0 & 1 & 2 \\
    \midrule
    Chloride, to surface water (kg) & 0.015 & 0.00695 & 0.00695 \\
    Carbon dioxide, to urban air (kg) & 0.0031 & 0.00144 & 0.00144 \\
    \midrule
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tbl:partition} illustrates this for the chlor-alkali example, for conventional allocation by mass.  The un-altered inventory is shown in column $0$, corresponding to the reference product of gaseous chlorine.  Co-products are shown as additional foreground flows, with negative signs indicating that the flows are outputs.  The reference for each column is indicated by an asterisk.

The inventories themselves are shown as entries in the $A_d$ and $B_f$ matrices. Because all co-products are already reported in mass, the allocated inventories for the two co-products are both the same and represent a scaling-down of the un-allocated inventory by the total mass of outputs (about 2.16 kg).

Using the foreground in Table~\ref{tbl:partition}, LCA computation using the canonical functional unit will automatically compute the self-consistent single-output result for 1 kg of chlorine gas because the allocated inventories for hydrogen and sodium hydroxide will be subtracted from the total.  The foreground activity levels for the system are:

\begin{equation}
  \tilde{\mathbf{x}} = [1,\,-0.028,\,-1.13]^T
\end{equation}

with the resulting aggregated dependency and emission vectors being, again, the same as the normalized and allocated inventories:

\begin{eqnarray}
  \tilde{\mathbf{a}}_d = A_d\times \tilde{\mathbf{x}} = & [1.38,\, 0.811,\,\num{1.9e-10},\,0.00709]^T \\
  \tilde{\mathbf{b}}_f = B_f\times \tilde{\mathbf{x}} = & [0.00695,\, 0.00144]^T
\end{eqnarray}

At the same time, LCA computation using a functional unit of one of the co-products will simply make use of the allocated inventory.  In this example, all processes are identical on a mass basis, but any other partitioning allocation can be automatically rendered self-consistent through this approach.  A reviewer would be able to inspect the foreground model to determine whether the allocation was performed as described in the report.

\subsubsection{``Classical'' System Expansion}

Although the term `system expansion' is often used to refer to product substitution and/or alternate activity allocation, these modeling techniques are distinct and should be referred to with different names, as noted in GWS14.  In the original definition of system expansion, the foreground model is simply enlarged to include all co-products.  Representation of such a system is trivial and is shown in Table~\ref{tbl:cse}.

\begin{table}[h]
  \begin{center}
  \caption{Chlor-Alkali Electrolysis as a foreground model, classical system expansion}
  \label{tbl:cse}
  \footnotesize\sffamily
  \begin{tabular}{l|ccc}
    \midrule
    \bf $A_f$ & 0 & 1 & 2 \\
    \midrule
    \textbf{Chlorine, gaseous (kg)} & $\ast$ & & \\
    Hydrogen, liquid (kg) & -0.028 & $\ast$ & \\
    Sodium hydroxide (kg dry) & -1.13 & & $\ast$ \\
    \midrule
    \bf $A_d$ & 0 & 1 & 2 \\
    \midrule
    Electricity, medium voltage (kWh) & 2.97 &  &  \\
    Sodium chloride, powder (kg) & 1.75 &  & \\
    Chemical factory, organics (unit) & 4e-10 &  &  \\
    Sludge, NaCl electrolysis (kg) & 0.0153 &  &  \\
    \midrule
    \bf $B_f$ & 0 & 1 & 2 \\
    \midrule
    Chloride, to surface water (kg) & 0.015 &  &  \\
    Carbon dioxide, to urban air (kg) & 0.0031 &  &  \\
    \midrule
  \end{tabular}
  \end{center}
\end{table}

In the classical system expansion case, the canonical functional unit, production of 1 kg of chlorine gas, will result in two cutoff flows corresponding to outputs of liquid hydrogen and sodium hydroxide.  Equivalently, an expanded functional unit of 1 kg chlorine, plus 0.028 kg hydrogen and 1.13 kg of sodium hydroxide, will result in the same activity of the un-allocated process, but will generate no cutoff flows at net.

\subsubsection{Consequential System Expansion}

The foreground disclosure framework is also well suited to describe consequential system expansion through product substitution or ``alternate activity allocation,'' which are the terms used in MWS14.  An example illustrating these methods is shown in Table~\ref{tbl:exp}

\begin{table}[t]
  \begin{center}
  \caption{Chlor-Alkali Electrolysis as a foreground model, consequential system expansion.  An inventory for the electrolysis of water (shown in blue) is provided as an alternate activity to produce hydrogen, while the sodium hydroxide product is assumed to substitute for for magnesium hydroxide (shown in red).}
  \label{tbl:exp}
  \footnotesize\sffamily
  \psset{unit=0.5cm}
  \begin{tabular}{l|cccc}
    \midrule
    \bf $A_f$ & 0 & 1 & 2 & \blue\textit{3} \\
    \midrule
    \textbf{Chlorine, gaseous (kg)} & $\ast$ & & &\\
    Hydrogen, liquid (kg) & -0.028 & $\ast$ & &\\
    Sodium hydroxide (kg dry) & -1.13 & & $\ast$ & \\
    \blue\textit{Hydrogen, electrolysis (kg)} & & \blue\textit{1} & & $\ast$ \\
    \midrule
    \bf $A_d$ & 0 & 1 & 2 & \blue\textit{3}\\
    \midrule
    Electricity, medium voltage (kWh) & 2.97 &  & & \blue\textit{50} \\
    Sodium chloride, powder (kg) & 1.75 &  & & \\
    Chemical factory, organics (unit) & 4e-10 &  & &  \\
    Sludge, NaCl electrolysis (kg) & 0.0153 &  & & \\
    \blue\textit{Water, distilled} (kg) & & & & \blue\textit{9} \\
    \red\textit{Magnesium hydroxide (kg dry)} & & & \red\textit{0.8} & \\
    \midrule
    \bf $B_f$ & 0 & 1 & 2 & \blue\textit{3}\\
    \midrule
    Chloride, to surface water (kg) & 0.015 &  & & \\
    Carbon dioxide, to urban air (kg) & 0.0031 &  & & \\
    \midrule
  \end{tabular}
  \end{center}
\end{table}

Here an alternate activity, namely electrolysis of water, is specified for liquid hydrogen production and is specified at index 3 in the foreground  (shown in blue).  The inventory converts 9 kg of water to 1 kg of hydrogen using 50 kWh of electricity.  The hydrogen produced from the multi-output process is then (somewhat dubiously) taken to displace hydrogen produced through electrolysis by introducing a unity coefficient in row 3, column 1 of $A_f$.

Meanwhile, sodium hydroxide from the process is assumed to substitute the production of magnesium hydroxide through the introduction of a new row and a new nonzero coefficient in column 2 of $A_d$ (shown in red). Here the match is inexact, and because magnesium hydroxide is more effective in some applications, 1 kg of sodium hydroxide is taken to displace only 0.8 kg of magnesium hydroxide.

### 2018-02-02 aggregation/agg-basic-lca.tex

@@
 
-We begin by providing a mathematical formulation of the LCA problem.
-An LCA computation is performed over a collection of linked unit processes, each of which describes a single operation or reference product.
-The task of procedure can be described in two steps: making an inventory of environmental interventions; and determining their environmental impacts.
-If the processes are assumed to be linear and time-invariant, the problem can be written in terms of matrix algebra.
-There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Heijungs2002, Pauliuk_2015_framework}.
-Although the two models are understood to be equivalent, the input-output approach is computationally more efficient in some applications because it lends itself to iterative solution \citep{Peters_JLCA_2007}.
-That formulation may be stated:
+In the most general terms, the LCA problem concerns the interaction of a collection of activities and commodities described by a supply-and-use inventory, in which $U$ is a table of the use of each commodity by each activity (inputs) and $V$ is a table of the supply of each commodity by each activity (outputs).
+Such an account can always be transformed into a \textit{symmetric} table of inputs and outputs on a commodity-by-commodity or industry-by-industry basis \citep{Eurostat_2008}.
+Doing so requires the application of various assumptions and transformations to the data, the details of which are not in scope of this paper, but which have recently been reviewed in depth \citep{Suh_JIE_2010, Majeau_Bettez_2014}.
+
+A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products).
+The table of coefficients, commonly known as a \textit{direct requirements matrix}, can be used to compute the output from all activities that is necessary to produce a given commodity.
+That vector of outputs can then be combined with a second table of coefficients, known as an \textit{emission matrix}, that reports exchanges with the environment associated with the production of each commodity, to determine the cumulative environmental emissions associated with the production of the commodity over its life cycle.
+Those emissions are individually \textit{characterized} with respect to an environmental impact category of interest, and the characterized emissions summed to report the \textit{impact score} or category indicator, which is the result of the study.
+The formulation may be summarized:
 \begin{equation}
@@
 \end{equation}
-Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values)\footnote{Whether negative values are present or allowed depends on the particular construct or co-production method being used.
-\cite{Majeau_Bettez_2014}}
-per unit of a process's reference output; $B$ is the environmental intervention or emission matrix; and $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions with respect to a given impact assessment method.
-In this model, $\mathbf{y}$ is the externally specified final demand, which is the product or service whose delivery is to be assessed, and $s$ is the numerical impact score or category indicator.
+Here $A$ represents the direct requirements; $B$ is the emission matrix;  $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions; $\mathbf{y}$ is the externally specified final demand; and $s$ is the numerical impact score or category indicator.
 For simplicity, we will consider the computation of a single impact result.
 However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.
-
-The term $(I-A)$ is equivalent to the normalized technology matrix, i.e.
-the difference between the supply and use tables in classical input-output analysis, $(V'-U)$ \citep{Suh_JIE_2010}.
-The conditions under which this matrix has an inverse, particularly with respect to different strategies for handling co-production, have been discussed extensively \citep{Majeau_Bettez_2014}.
 Eq.~\ref{eqn:leontief} is visualized in Figure~\ref{fig:lca}.
@@
 
-In classical LCA, the matrices $A$ (or $I-A$) and $B$ are called a life cycle inventory database (LCIDB), and the functional unit of a study is given by the final demand vector $\mathbf{y}$.
-Intermediate results of this computation include:
+The term $(I-A)$ is equivalent to a normalized \textit{technology matrix}, and in certain cases can be computed from the difference between the supply and use tables, $(V'-U)$, with equivalent results \citep{Suh_JIE_2010}.
+The coefficient matrices $A$ (or $I-A$) and $B$ together are called a \textit{life cycle inventory database} (LCIDB), and the functional unit of a study is given by the final demand vector $\mathbf{y}$, which is the product or service whose delivery is to be assessed.\footnote{%
+Differences in convention have led to both the direct requirements matrix $A$, and the technology matrix, analogous to $(I-A)$, being denoted by the same symbol in different contexts.
+If the data contained in the LCIDB is sufficiently rich, either approach may be implemented with equivalent computational results.
+The direct requirement formulation has the advantage of permitting iterative solution, which can dramatically reduce the computational resources required \citep{Peters_JLCA_2007}.}  Intermediate results of this computation include:
 \begin{itemize}

### 2018-02-02 aggregation/agg-multioutput.tex
tweeeeeeak

@@
 The inventory converts 9 kg of water to 1 kg of hydrogen using 50 kWh of electricity.
-The hydrogen produced from the multi-output process is then (somewhat dubiously) taken to displace hydrogen produced through electrolysis by introducing a unity coefficient in row 3, column 1 of $A_f$.
+The hydrogen produced from the multi-output process is then (somewhat dubiously) taken to displace hydrogen produced through the alternate activity by introducing a unity coefficient in row 3, column 1 of $A_f$.
 

### 2018-02-05 aggregation/agg-intro.tex
Streamlined Intro

@@
 Plevin et al.\ famously observed in the strongest terms that careless reporting of attributional LCA study results can distort the significance of findings in a policy context \citep{Plevin_2013}, a shortcoming that can be found in other modes as well \citep{Brandao_2014}.
-In the scope of Environmental Product Declarations (EPDs), studies within an industry are supposed to be rendered comparable by adhering to a common product category rule (PCR) \citep{Fet_2006}.
-However, this prescription is insufficient to ensure the comparability of different declarations, even if they use the same PCR \citep{Modahl_2012}.
-In addition, PCRs are often themselves not unique and can include wide variations in scope and boundary requirements \citep{Subramanian_2012}, leading to ambiguity.
 
@@
 But if the PSM itself is ambiguously stated, then the significance of the result is ambiguous as well.
-For a computational model to have scientific significance, its results need to be both verifiable and replicable by other scientists \citep{Fomel_2009, Mesirov_2010}.
-But in the case of LCA, the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.
+Often the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.
 When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.
 \cite{iso14071}, which provides guidelines on critical review, emphasizes that review of the PSM or individual datasets is optional.
-
 %A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.
@@
 Moreover, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.
-The purpose of this paper is to advance a framework for the formal description of product system models that would provide robust transparency and reproducibility of computed results.
+
+Because of the diversity of software and techniques for performing LCA, it is important to develop methods for communicating the structure of the PSM that is not dependent on any particular software or database \cite{Kuczenski_JLCA_2018}.
+The purpose of this paper is to advance a framework for the formal description of PSM that would allow for sharing, revision, and critical review of models while also enabling authors to protect the privacy of confidential information.
+This objective is also in line with the broader imperative throughout industrial ecology to improve transparency and reproducibility of computed results \cite{Hertwich_JIE_2018}.
 

### 2018-02-05 aggregation/agg-multioutput.tex
Streamlined Intro

@@
-\subsection{Multi-Output Foreground Configurations}
+\subsection{Multi-Output Foreground Processes}
 
@@
   \begin{center}
-  \caption{Chlor-Alkali Electrolysis, membrane cell [RER] (Ecoinvent 3.2; excerpt)}
+  \caption{Example: Chlor-alkali electrolysis, membrane cell [RER] (Ecoinvent 3.2; excerpt)}
   \label{tbl:chlor-alkali}
@@
   \begin{center}
-  \caption{Chlor-Alkali Electrolysis as a foreground model, partitioning allocation}
+  \caption{Chlor-alkali electrolysis as a foreground model, partitioning allocation}
   \label{tbl:partition}
@@
   \begin{center}
-  \caption{Chlor-Alkali Electrolysis as a foreground model, classical system expansion}
+  \caption{Chlor-alkali electrolysis as a foreground model, classical system expansion}
   \label{tbl:cse}
@@
   \begin{center}
-  \caption{Chlor-Alkali Electrolysis as a foreground model, consequential system expansion.
+  \caption{Chlor-alkali electrolysis as a foreground model, consequential system expansion.
 An inventory for the electrolysis of water (shown in blue) is provided as an alternate activity to produce hydrogen, while the sodium hydroxide product is assumed to substitute for for magnesium hydroxide (shown in red).}

### 2018-02-06 aggregation/agg-basic-lca.tex
tweak IO discussion

@@
 
-In the most general terms, the LCA problem concerns the interaction of a collection of activities and commodities described by a supply-and-use inventory, in which $U$ is a table of the use of each commodity by each activity (inputs) and $V$ is a table of the supply of each commodity by each activity (outputs).
+In the most general terms, the LCA problem concerns the interaction of a collection of activities and commodities described by a supply-and-use inventory, in which $U$ is a table of the use of each commodity by each activity (inputs) and $V$ is a table of the supply of each commodity by each activity (outputs) \citep{Pauliuk_framework_2015}.
 Such an account can always be transformed into a \textit{symmetric} table of inputs and outputs on a commodity-by-commodity or industry-by-industry basis \citep{Eurostat_2008}.
@@
 
-A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products).
+A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products), equivalent to a ``unit process'' in ISO LCA, except that the reference flow is implicit in the column's index.
 The table of coefficients, commonly known as a \textit{direct requirements matrix}, can be used to compute the output from all activities that is necessary to produce a given commodity.
@@
 
-This formulation implemented in all major LCA software systems, some of which provide the capacity to export the $A$ and $B$ matrices for a given study model.
+Some LCA software systems have the capacity to export the $A$ and $B$ matrices for a given study model.
 However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are not usually modified by study authors.

### 2018-02-06 aggregation/agg-agg.tex
afternoon work- cites; semantics; editing
Disclosure card

@@
 the full disclosure of a model, and a reader can replicate and extend it independently.
-A fragment of the PSM can also be published as a complete foreground.
+%A {\red fragment} subset of the columns of $A_f$ may also be published  of the PSM can also be published as a complete foreground. %
 Any co-production treatment, such as an allocation or system expansion, can only be fully expressed as a foreground fragment, because it intrinsically involves a collection of linked unit processes.
@@
 
-See Figure~\ref{fig:aggregation}d.  Using this approach, the author can replace portions of the
-The aggregated dependency vector is split into two parts that sum to the original:
+See Figure~\ref{fig:aggregation}d.  Using this approach, the
+ aggregated dependency vector is split into two parts that sum to the original:
 \begin{equation}

### 2018-02-06 aggregation/agg-disclosure.tex
Rev01 Submitted!
Disclosure card

@@
 \begin{enumerate}[label={\em d-\roman*}., ref={\em d-\roman*}]
-\item\label{itm:fg} An ordered list of foreground nodes (rows/columns of $A_f$);
-\item\label{itm:bg} An ordered list of background flows, making reference to a particular background database (rows of $A_d$, mapping to columns of $B_x$);
-\item\label{itm:em} An ordered list of exchanges with environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
+\item\label{itm:fg} An ordered list of foreground nodes, beginning with the functional unit (rows/columns of $A_f$);
+\item\label{itm:bg} An ordered list of background flows, each making reference to a particular background dataset (rows of $A_d$, mapping to columns of $B_x$);
+\item\label{itm:em} An ordered list of exchanges with the environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
 \item\label{itm:af} A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
@@
 
+In both items~\ref{itm:bg} and~\ref{itm:em}, only rows containing non-zero entries need to be mentioned, since the index of each row is common information.
 A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):
@@
 These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.
-However, the the challenges of operationalizing this requirement are complex (see Discussion and Supplementary materials).
+However, the challenges of operationalizing this requirement are complex (see Discussion and Supplementary materials).
 %Some implications on the information required for a successful disclosure are considered in the Discussion.

### 2018-02-07 aggregation/agg-basic-lca.tex
Rewrite foreground study

@@
 
-In the most general terms, the LCA problem concerns the interaction of a collection of activities and commodities described by a supply-and-use inventory, in which $U$ is a table of the use of each commodity by each activity (inputs) and $V$ is a table of the supply of each commodity by each activity (outputs) \citep{Pauliuk_framework_2015}.
+In the most general terms, the LCA problem concerns the interaction of a collection of activities and commodities described by a supply-and-use inventory, in which $U$ is a table of the use of each commodity by each activity (inputs) and $V$ is a table of the supply of each commodity by each activity (outputs) \citep{Pauliuk_2015_framework}.
 Such an account can always be transformed into a \textit{symmetric} table of inputs and outputs on a commodity-by-commodity or industry-by-industry basis \citep{Eurostat_2008}.

### 2018-02-07 aggregation/agg-intro.tex
Rewrite foreground study

@@
 
-Because of the diversity of software and techniques for performing LCA, it is important to develop methods for communicating the structure of the PSM that is not dependent on any particular software or database \cite{Kuczenski_JLCA_2018}.
+Because of the diversity of software and techniques for performing LCA, it is important to develop methods for communicating the structure of the PSM that is not dependent on any particular software or database \citep{Kuczenski_JLCA_2018}.
 The purpose of this paper is to advance a framework for the formal description of PSM that would allow for sharing, revision, and critical review of models while also enabling authors to protect the privacy of confidential information.
-This objective is also in line with the broader imperative throughout industrial ecology to improve transparency and reproducibility of computed results \cite{Hertwich_JIE_2018}.
+This objective is also in line with the broader imperative throughout industrial ecology to improve transparency and reproducibility of computed results \citep{Hertwich_JIE_2018}.
 

### 2018-02-07 aggregation/agg-lca-study.tex
Rewrite foreground study

--- 2017-11-15 aggregation/agg-lca-study.tex
@@
 
-In practice, most LCA studies contain foreground information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with foreground models.
+In practice, most LCA studies contain information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with their own modeling information, known as the study foreground.
 To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.
 Background databases are prepared independently of any study and do not typically vary across the studies in which they are used.
-However, the foreground systems themselves cannot be computed without the background.
-Put another way, the foreground depends on the background but the background does not depend on the foreground.
+However, the results of a study with foreground content cannot generally be computed without knowledge of the background.
+%Put another way, the foreground results depend on the background but the background results do not depend on the foreground.
 This condition allows the processes in an LCIDB to be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.
 
-The simplest conceivable foreground is simply the final demand $\mathbf{y}$, in other words the list of background processes invoked by the study and their activity levels.
-For an LCA computation as expressed in Eq.~\ref{eqn:leontief}, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
+To understand how a foreground study is constructed on top of a background database, imagine a simple product system that can be fully described only making reference to processes that exist in a background database.
+For instance, the use of a computer may be modeled using a reference process for producing and disposing the computer, and a reference process for electricity production.
+Assuming those reference processes are available in a background database, the product system can be represented entirely by specifying a final demand vector $\mathbf{y}$ whose nonzero values correspond to computer production, computer disposal, and electricity.
+The LCIA results can be computed using Eq.~\ref{eqn:leontief}.
+However, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
 \begin{equation}
@@
 \end{equation}
-where $\tilde{A} = \left[\begin{smallmatrix}
- 0 & \mathbf{0} \\
- \mathbf{y} & A
-  \end{smallmatrix}\right]$, $\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
+where
+$\tilde{A} = \left[\begin{smallmatrix} 0 & \mathbf{0}^T \\ \mathbf{y} & A  \end{smallmatrix}\right]$,
+$\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
 It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.
@@
 
-\input{fig_foreground}
-Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.
-An LCA study augmented with foreground information can be written in block-triangular form:
+This same process can be repeated to build a more elaborate foreground.
+For instance, a model of private vehicle travel may make use of reference processes for vehicle production and decommissioning, road maintenance, fuel retail, and a model of direct emissions from fuel combustion.
+This model could occupy an additional column of $\tilde{A}$, with the coefficients for fuel combustion emissions being contained in the corresponding column of $\tilde{B}$.
+A process for brewing coffee could be created in a third column that requires coffee beans, water, and electricity.
+Finally, a fourth column could be added that invokes the first three columns in proportion to describe the activity of traveling to a coffeeshop to work\footnote{No prizes for guessing where this passage was written.}.
+
+In general, a product system model may be constructed through the successive augmentation of a background database with foreground content.
+As long as the background database is not altered to depend on the newly created foreground, such a study can be written in block triangular form:
 \begin{equation}
@@
 \end{equation}
-Here, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.
+Here, the submatrix $A_f$ represents the foreground; $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.
 The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions.
 The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.
-As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.
-However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2015}).
+The augmented $\tilde{A}$ and $\tilde{B}$ together make up an \emph{LCA foreground study}.
+This formulation can be applied to the vast majority of LCA case studies, and very likely to all case studies prepared using commercial LCA software.\footnote{It does not apply to cases in which the background database must be altered to adjust for double counting, such as in so-called ``integrated hybrid'' models in which all quadrants contain non-zero cells \citep{Suh2004}.
+These models are more complex than the models typically built by LCA practitioners, and in conventional LCA practice it is inadvisable to alter background databases to adapt them to study conditions.
+Instead, background processes can be replicated in the foreground, where they can be adapted as needed (cf.
+\cite{Bourgault_JLCA_2013}), without affecting the integrity of the background database.}
 
-%For clarity, I will call thein which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be an
+Without loss of generality, it is possible to construct the study such that the \textit{first} column of $\tilde{A}$ delivers the functional unit of the study.
+This allows us to designate the vector $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ as the \textit{canonical functional unit} of a foreground study, which allows us to introduce convenient notation later on.
 
-The augmented $\tilde{A}$ and $\tilde{B}$, makes up an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.
-This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.
-
+\input{fig_foreground}
 It is currently common practice for background database maintainers to pre-compute the life cycle inventory results for their databases.
-Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, Eq.s~\ref{eqn:foreground} can be rewritten as:
+In cases where the background database is used without modification, including the contents of $A$ in $\tilde{A}$ becomes unnecessary.
+Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, $\tilde{A}$ can be ``flattened'' so that it no longer includes inventories for background processes \citep{Kuczenski_JLCA_2015}.
+Eq.s~\ref{eqn:foreground} can be recast as:
 \begin{equation}
@@
 The proof of the equivalency of results derived from Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary materials.
-In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining in $\tilde{A}_{flat}$ pertains to the study foreground.
-Furthermore, the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported.
+In this form, the only information remaining in $\tilde{A}_{flat}$ pertains to the study foreground.
 
@@
 \end{equation}
-where $\tilde{\mathbf{y}}_f$ is a canonical functional unit vector having the same rank as $A_f$. Eq.~\ref{eqn:study} is derived in the supplementary materials.
+where $\tilde{\mathbf{y}}_f$ is a canonical functional unit vector whose length is the same as the rank of $A_f$. Eq.~\ref{eqn:study} is derived in the supplementary materials.
 Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$, having been subsumed into $B_x$, represented by an identity matrix $I$.
@@
 The activity levels of the foreground nodes are represented by $\tilde{\mathbf{x}}$, which is the result of $\tilde{\mathbf{y}}_f$ selecting the first column of $(I-A_f)^{-1}$.
-%% An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
-%% \begin{equation}
-%% \tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
-%% \label{eqn:inv}
-%% \end{equation}
-This vector can be used to describe an \emph{aggregated foreground}:
+This vector can be used to describe an \emph{aggregated foreground} in coefficient form, which delivers the same results as the foreground study given in Eq.~\ref{eqn:foreground}, but without disclosing any details of the construction of the foreground:
 \begin{equation}
@@
 \end{equation}
-The study result can also be computed from the aggregated foreground:
+Here $\tilde{\mathbf{a}}_d$ reports the aggregated foreground's dependency on background processes in background flows per canonical functional unit;  $\tilde{\mathbf{b}}_f$ reports foreground emissions per functional unit; and $\tilde{\mathbf{b}}_x$ reports the aggregated background life cycle emissions per functional unit.
+
+The aggregated result can also be computed from the aggregated foreground:
 \begin{equation}

### 2018-02-07 aggregation/agg-lca-study.tex
Further tweaking foreground derivation and proofs

@@
 
-To understand how a foreground study is constructed on top of a background database, imagine a simple product system that can be fully described only making reference to processes that exist in a background database.
-For instance, the use of a computer may be modeled using a reference process for producing and disposing the computer, and a reference process for electricity production.
+To understand how a foreground study is constructed on top of a background database, imagine a simple product system that can be fully described by only making reference to processes that exist in a background database.
+For instance, the use of a computer may be modeled using reference processes for producing and disposing the computer, and a reference process for electricity consumption.
 Assuming those reference processes are available in a background database, the product system can be represented entirely by specifying a final demand vector $\mathbf{y}$ whose nonzero values correspond to computer production, computer disposal, and electricity.
@@
 $\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
-It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.
-\textit{author's note: a proof for this equality is provided in the electronic supplementary information.}  This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.
+It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$  (A proof is provided in the electronic supplementary information).
+This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.
 
-This same process can be repeated to build a more elaborate foreground.
+This same workflow can be repeated to build a more elaborate foreground.
 For instance, a model of private vehicle travel may make use of reference processes for vehicle production and decommissioning, road maintenance, fuel retail, and a model of direct emissions from fuel combustion.
@@
 In general, a product system model may be constructed through the successive augmentation of a background database with foreground content.
-As long as the background database is not altered to depend on the newly created foreground, such a study can be written in block triangular form:
+As long as the background database is not altered to depend on the newly created foreground, such a study can be written in block triangular form \citep{Kuczenski_JLCA_2015}:
 \begin{equation}
@@
 \end{equation}
-Here, the submatrix $A_f$ represents the foreground; $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.
-The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions.
-The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.
+Here, the submatrix $A_f$ represents the foreground; $A$ represents the background; the rectangular matrix $A_d$ represents the dependency of the foreground on the background; and the top right submatrix is zero.
+The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$, which includes background emissions.
+%The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.
 The augmented $\tilde{A}$ and $\tilde{B}$ together make up an \emph{LCA foreground study}.
 This formulation can be applied to the vast majority of LCA case studies, and very likely to all case studies prepared using commercial LCA software.\footnote{It does not apply to cases in which the background database must be altered to adjust for double counting, such as in so-called ``integrated hybrid'' models in which all quadrants contain non-zero cells \citep{Suh2004}.
-These models are more complex than the models typically built by LCA practitioners, and in conventional LCA practice it is inadvisable to alter background databases to adapt them to study conditions.
+These models are more complex than the models typically built by LCA practitioners, and require more sophisticated computation.
+In conventional LCA practice it is inadvisable to alter background databases to adapt them to study conditions.
 Instead, background processes can be replicated in the foreground, where they can be adapted as needed (cf.
@@
 Without loss of generality, it is possible to construct the study such that the \textit{first} column of $\tilde{A}$ delivers the functional unit of the study.
-This allows us to designate the vector $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ as the \textit{canonical functional unit} of a foreground study, which allows us to introduce convenient notation later on.
+We thereby designate the vector $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ as the \textit{canonical functional unit} of a foreground study, which allows us to introduce convenient notation later on.
 
@@
 In cases where the background database is used without modification, including the contents of $A$ in $\tilde{A}$ becomes unnecessary.
-Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, $\tilde{A}$ can be ``flattened'' so that it no longer includes inventories for background processes \citep{Kuczenski_JLCA_2015}.
-Eq.s~\ref{eqn:foreground} can be recast as:
-\begin{equation}
- \tilde{A}_{flat} = \left[\begin{array}{cc} 
-A_f &  0 \\
-A_d & 0 
-   \end{array}\right];\;\;\;  \tilde{B}_{flat} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
-\label{eqn:fg_bg}
-\end{equation}
-The proof of the equivalency of results derived from Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary materials.
+Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, %% $\tilde{A}$ can be ``flattened'' so that it no longer includes inventories for background processes.
+%% Eq.s~\ref{eqn:foreground} can be recast as:
+%% \begin{equation}
+%%  \tilde{A}_{flat} = \left[\begin{array}{cc} 
+%% A_f &  0 \\
+%% A_d & 0 
+%%    \end{array}\right];\;\;\;  \tilde{B}_{flat} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
+%% \label{eqn:fg_bg}
+%% \end{equation}
+%% The proof of the equivalency of results derived from Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary materials.
 In this form, the only information remaining in $\tilde{A}_{flat}$ pertains to the study foreground.
-
-Using the ordered matrices, Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:
+%
+%% Using the ordered matrices, 
+Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:
 \begin{equation}
@@
 where $\tilde{\mathbf{y}}_f$ is a canonical functional unit vector whose length is the same as the rank of $A_f$. Eq.~\ref{eqn:study} is derived in the supplementary materials.
-Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$, having been subsumed into $B_x$, represented by an identity matrix $I$.
+Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$ subsumed into $B_x$ and replaced by an identity matrix $I$.
 

### 2018-02-07 aggregation/proofs.tex
Further tweaking foreground derivation and proofs

@@
 
-\subsection{``Flattening'' the background database}
-
-Given a canonical LCA foreground study, show that the following formulations are equivalent:
-\begin{eqnarray}
-\tilde{A} = \left[\begin{array}{cc}
-A_f & 0 \\
-A_d &  A
-  \end{array}
-\right];&
-  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right] \\
- \tilde{A}_{flat} = \left[\begin{array}{cc} 
-A_f &  0 \\
-A_d & 0 
-   \end{array}\right];&  \tilde{B}_{flat} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
-\end{eqnarray}
-where $B_x = B\times(I-A)^{-1}$.
-
-It is sufficient to show that $\tilde{B}\times(I-\tilde{A})^{-1} = \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}$.
-
-Using the shorthand $Z=I-A$, we define the following identity:
-\begin{equation}
-  I* = \left[\begin{array}{cc}
-      I & 0 \\
-      0 & Z
-    \end{array}\right]^{-1} \times
-  \left[\begin{array}{cc}
-      I & 0 \\
-      0 & Z
-    \end{array}\right]
-\end{equation}
-
-We introduce the identity in between the two terms of the LCIDB:
-\begin{align}
-  \tilde{B}\times(I-\tilde{A})^{-1} &=
-  \left[\begin{array}{cc}B_f & B\end{array}\right]\times I*  \times
-  \left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I-A\end{array}\right]^{-1}\\
-      &=  \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
-      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\times
-      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right] \times
-      \left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]^{-1}
-\end{align}
-Using the associative property of matrix multiplication, it can be shown that $M\times N^{-1} = (N\times M^{-1})^{-1}$ for any invertible $M$ and $N$ of the same rank, so:
-\begin{eqnarray}
-  \tilde{B}\times(I-\tilde{A})^{-1}
-  &=& \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
-      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\times
-\left(
-\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]\times
-    \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\right)^{-1}\\
-  &=&   \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
-      \left[\begin{array}{cc}I & 0 \\ 0 & Z^{-1}\end{array}\right]\times
-\left(
-\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]\times
-\left[\begin{array}{cc}I & 0 \\ 0 & Z^{-1}\end{array}\right]\right)^{-1}\\
- &=& \left[\begin{array}{cc}B_f & B\times Z^{-1}\end{array}\right]\times
-\left(\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I\end{array}\right]\right)^{-1}
-\end{eqnarray}
-But these terms are equivalent to the flattened LCIDB:
-\begin{eqnarray}
-  \tilde{B}\times(I-\tilde{A})^{-1} &=&
-  \left[\begin{array}{cc}B_f & B\times (I-A)^{-1}\end{array}\right]\times
-\left(\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I\end{array}\right]\right)^{-1} \\
-&=& \left[\begin{array}{cc}B_f & B_x\end{array}\right]\times
-\left(I-\left[\begin{array}{cc}A_f & 0 \\ A_d & 0\end{array}\right]\right)^{-1}\\
-&=& \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}
-\end{eqnarray}
-thus completing the proof.
-
 \subsection{The Study Foreground Equation}
 
-Equation 5 in the manuscript is derived from Equation 4.
+This section contains the derivation of Equation~\ref{eqn:study} in the manuscript, in two steps.
+
+\input{proof-flattening}
+
+\subsubsection{Separating Background from Foreground}
+
+Equation~\ref{eqn:study} in the manuscript is derived from Equation~\ref{eqn:si-flat} above.  It was also derived in Kuczenski (2015).
 

### 2018-02-07 aggregation/agg-disclosure.tex
Fig4 caption; disclosure + foreground work

@@
 
+\input{fig_disclosure}
 Thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.
@@
 A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.
-Thus, a formal disclosure of the PSM can be stated in six parts:
+Thus, a formal disclosure of the PSM can be stated in six parts, including three lists of entities and three sets of data values:
 \begin{enumerate}[label={\em d-\roman*}., ref={\em d-\roman*}]
@@
 \item\label{itm:em} An ordered list of exchanges with the environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
-\item\label{itm:af} A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
-\item\label{itm:ad} a list of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
-\item\label{itm:bf} a list of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
+\item\label{itm:af} A set of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
+\item\label{itm:ad} a set of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
+\item\label{itm:bf} a set of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
 \end{enumerate}
+In both items~\ref{itm:bg} and~\ref{itm:em}, only rows containing non-zero entries need to be mentioned, since the index of each row is common information.
+The contents of a disclosure are represented graphically in Fig.~\ref{fig:disclosure}., and two example disclosures drawn from reference databases can be found in the supplementary materials.
 
-In both items~\ref{itm:bg} and~\ref{itm:em}, only rows containing non-zero entries need to be mentioned, since the index of each row is common information.
 A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):

### 2018-02-07 aggregation/agg-lca-study.tex
Fig4 caption; disclosure + foreground work

@@
 The LCIA results can be computed using Eq.~\ref{eqn:leontief}.
-However, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
+
+The same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
 \begin{equation}
@@
 %% Using the ordered matrices, 
-Eq.~\ref{eqn:leontief} can be written so that the computationally costly foreground and background inversions can be performed separately:
+Eq.~\ref{eqn:leontief} can be written so that the foreground matrix inversion is  performed separately from the computationally costly background computation:
 \begin{equation}

### 2018-02-07 aggregation/agg-JIE.tex
Pauliuk comments
Headers for ERI submission
Supplementary stuffs
Multioutput case in SI
Streamlined Intro
Rewrite foreground study
Fig4 caption; disclosure + foreground work
Finishing body revision

@@
 
+%\lhead{\sffamily Disclosure of product system models}
+%\rhead{\sffamily Kuczenski -- 2017}
+
 \usepackage{tabularx}
 \usepackage[round]{natbib}
+\usepackage{siunitx}
+
+\makeatletter
+\let\oldsection\section
+\renewcommand\section{\@startsection {section}{1}{\z@}%
+                                   {-3.5ex \@plus -1ex \@minus -.2ex}%
+                                   {2.3ex \@plus.2ex}%
+                                   {\normalfont\Large\bfseries <Level 1> }*}
+\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
+                                     {-3.25ex\@plus -1ex \@minus -.2ex}%
+                                     {1.5ex \@plus .2ex}%
+                                     {\normalfont\large\bfseries <Level 2> }*}
+\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
+                                     {-3.25ex\@plus -1ex \@minus -.2ex}%
+                                     {1.5ex \@plus .2ex}%
+                                     {\normalfont\normalsize\bfseries <Level 3> }*}
+
+\makeatother
+
+\usepackage{endnotes}
+\let\footnote=\endnote
+
 \bibliographystyle{authordate_bk}
@@
 \renewcommand{\topfraction}{0.7}
+\renewcommand{\bottomfraction}{0.7}
 
@@
 
-%\usepackage{lineno}
+\usepackage{lineno}
 \let\showgrid\relax
@@
   \maketitle
-  \emph{For submission to the Journal of Industrial Ecology; November 2017}
+  \emph{Submitted after revision to the Journal of Industrial Ecology; February 2018}
 
@@
   \doublespacing
+  %\linenumbers
 
@@
 Objectives in critical review vary with the study and the reviewer, but they generally include consistency with the ISO 14044 standard, the validity of methods, appropriateness of data used, correctness of interpretations, and transparency of reporting \citep[Section 6.1]{iso14044}.
-Although the object of critical review is typically a written report, the use of a structured disclosure of a PSM introduces the possibility that reviewers can inspect the model itself and even verify the computations and reported results.
+Although the object of critical review is typically a written report, this level of review is not sufficient to ensure the correctness of quantitative results.
+The use of a structured disclosure of a PSM introduces the possibility that reviewers can inspect the model itself and even verify the computations and reported results.
 This section discusses how a disclosure in the format described above can enable robust review of a product system model.
+%\footnote{Although ISO 14071 makes explicit that reviewers are not required to review the system model, it is the opinion of both the author and at least one anonymous reviewer that the status quo is insufficient.}
 Note that, for purposes of critical review, the objective for data privacy can be omitted because the reviewer may be granted privileged access to confidential data.
@@
 
-\subsection{Interpreting the Foreground}
-
-The foreground of an LCA study is a collection of nodes that represent points in the inventory model where flows are exchanged.
-The foreground matrix $A_f$ describes in precise terms what components are included in the product system and how they connect to one another.
-The foreground is modelled as a weighted directed graph in which there is a 1:1 correspondence between nodes and product flows.
-The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.
-These nodes typically represent unit processes, but they can also indicate exchanges across an implicit system boundary that encloses the foreground.
-Generally, the foreground matrix $A_f$ can represent any process-flow model with a 1:1 correspondence between processes and flows, i.e.
-any allocated supply-and-use table, but some fundamental designs are common.
-Some simple foreground models are shown in the supplementary materials.
+\input{agg-foreground}
 
@@
 
+%  \input{agg-community}
+
   \input{agg-conclusion}
@@
 
+\let\section\oldsection
+    \section*{Acknowledgments}
+
+Many thanks to the anonymous reviewers for significant improvement of the manuscript. This work was partially supported by the US National Science Foundation (CCF-1442966).
+
  \singlespacing
+
+ \theendnotes
+
  \bibliography{jabref}
@@
 \def\supporting{
+  \setcounter{figure}{0}
+  \setcounter{table}{0}
   \renewcommand{\thefigure}{S\arabic{figure}}
@@
 
-  \input{agg-foreground} % Af and graph/matrix equivalency
+  \input{agg-preparing} % Af and graph/matrix equivalency
 
-  \input{../manuscript/publication-example}
+  \input{agg-multioutput}
+
+  \input{publication-example}
 

### 2018-02-07 aggregation/agg-agg.tex
Fig4 caption; disclosure + foreground work
Finishing body revision

@@
 
-\subsubsection{Foreground or Foreground Fragment}
+\subsubsection{Complete Foreground}
 
@@
 the full disclosure of a model, and a reader can replicate and extend it independently.
-%A {\red fragment} subset of the columns of $A_f$ may also be published  of the PSM can also be published as a complete foreground. %
-Any co-production treatment, such as an allocation or system expansion, can only be fully expressed as a foreground fragment, because it intrinsically involves a collection of linked unit processes.
+%An excerpt from a larger study may also be published as a complete foreground. %
+Any modeling technique used to treat a multi-functional process, such as an allocation or system expansion, can only be fully expressed as a foreground with at least as many columns as outputs.
+For examples of multi-output processes represented as foreground models, please see the supporting information.
+The example shown in Figure~\ref{fig:aggregation}b shows five nodes, including one cut-off flow
 
@@
 Any collection of foreground nodes can also be expressed in aggregated form by performing the computations in Eq.~\ref{eqn:agg}.
-A publication of the aggregated foreground is identical to the publication of a unit process, except that the result was derived through aggregation.
+A disclosure of an aggregated foreground resembles the description of a unit process.
 As in the unit process case, the aggregated foreground has exactly one reference flow, the canonical functional unit $\tilde{\mathbf{y}}$.
 The aggregated foreground still reports explicit links between the model and a background database and can thus be used to review data set selection.
+
+An aggregated foreground with cut-off flows resembles a multi-output process.
+After aggregation, foreground flows that correspond to cut-offs must still be reported in a disclosure, while foreground flows interior to the model can be omitted.
+This is done with the introduction of a new vector, $\tilde{\mathbf{a}}_f$, which includes the reference flow plus any entries from $\tilde{\mathbf{x}}$ that correspond to cut-off flows.
 

### 2018-02-07 aggregation/agg-conclusion.tex
Finishing body revision

@@
 This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.
+The disclosure framework also allows modelers to express precisely how multi-functionality is handled through allocation or system expansion.
 A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.
 Once equipped with the PSM, the reviewer or data user can potentially go much further.
-Reviewers or data users who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, and applying other impact assessment methods.
+Practitioners who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, applying other impact assessment methods, and considering alternative allocation methods.
 Portions of the model which have been pre-aggregated can remain private in the disclosure.

### 2018-02-07 aggregation/agg-deps.tex
Disclosure card
Finishing body revision

@@
-\subsection{Dependencies, Emissions, and Cutoffs}
+\subsection{Dependencies and Emissions}
 
@@
 In both cases, the respective columns of $A_d$ and $B_f$ correspond with columns in $A_f$ and denote dependencies and/or emissions of the corresponding nodes.
+If columns of $A_d$ and/or $B_f$ are empty, then the node is simply a pass-through process that serves to connect different foreground elements.
+However, if $A_d$ and/or $B_f$ have nonzero entries in that column, then that indicates some transformation process or emission is occurring at that point in the foreground.
+
 Review of the dependencies of foreground nodes must include review of the selection of background processes to which they are connected (i.e.
@@
 Likewise, the identities of the emissions (entries in disclosure item~\ref{itm:em}) must also be reviewed.
-
-Interpretation of the contents of these matrices thus hinges on the contents of $A_f$.
-If a column in $A_f$ is nonempty, then the corresponding node has exchanges with other nodes in the foreground.
-If columns of $A_d$ and/or $B_f$ are empty, then the node is simply a pass-through process that serves to connect different foreground elements.
-However, if $A_d$ and/or $B_f$ have nonzero entries in that column, then that indicates some transformation process or emission is occurring at that point in the foreground.
-
-On the other hand, if a column in $A_f$ is empty, then that is an indication that the corresponding product flow is crossing the foreground system boundary.
-In order for this flow to contribute any impacts, it must either be ``terminated'' in the dependency matrix, or represented as an emission.
-
-Flows whose entire column consists of 0s throughout $A_f$, $A_d$, and $B_f$ are ``cut-offs'' that exit the model with zero burdens.
-Specifying cut-off criteria and reporting of cut-off flows is an important part of system boundary definition.
-Thus, reviewing cut-off flows (including evaluating their significance) is a crucial part of review that is facilitated by having a structured disclosure of the model.

### 2018-02-07 aggregation/agg-disclosure.tex
Finishing body revision

@@
 \end{enumerate}
+Each entity included in items \ref{itm:fg}--\ref{itm:em} describes a flow of some kind, and its unit of measure must be reported in the disclosure.
 In both items~\ref{itm:bg} and~\ref{itm:em}, only rows containing non-zero entries need to be mentioned, since the index of each row is common information.
-The contents of a disclosure are represented graphically in Fig.~\ref{fig:disclosure}., and two example disclosures drawn from reference databases can be found in the supplementary materials.
+The contents of a disclosure are represented graphically in Fig.~\ref{fig:disclosure}., and two example disclosures drawn from reference databases can be found in the supporting information .
 
@@
   \item The disclosure includes only information for which the author can claim \textbf{authority};
-  \item The disclosure includes only information for which the study is the \textbf{primary} source.  
+  \item The disclosure includes all information specific to the study, or for which the study is the \textbf{primary} source, including re-implementations of already-published data. 
 \end{itemize}
@@
 These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.
-However, the challenges of operationalizing this requirement are complex (see Discussion and Supplementary materials).
+However, the challenges of operationalizing this requirement are complex (see the discussion and supporting information).
 %Some implications on the information required for a successful disclosure are considered in the Discussion.

### 2018-02-07 aggregation/agg-foreground.tex
Finishing body revision

--- 2017-11-03 aggregation/agg-foreground.tex
@@
-\section{Foreground Configurations}
+\subsection{The Foreground System Boundary and Cut-off Flows}
 
-Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.
-Most product system models would include many of these elements mixed together.
-A reviewer with access to items \ref{itm:fg} and \ref{itm:af} in the disclosure would be able to construct $A_f$ and automatically create a process-flow diagram, annotated with information about each node.
+The foreground of an LCA study is a collection of nodes that represent points in the inventory model where flows are exchanged.
+The foreground matrix $A_f$ describes in precise terms what components are included in the product system and how they connect to one another.
+The foreground is modelled as a weighted directed graph in which there is a 1:1 correspondence between nodes and product flows.
+The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.
+Some simple foregrounds are described in the supporting information.
 
-\input{fig_fragments}
+Generally, the direct requirements matrix can represent any process-flow model with a 1:1 correspondence between processes and flows.
+Normally this is understood as a requirement to include only allocated single-output processes.
+However, the foreground disclosure framework is also well suited to documenting how multi-output processes are transformed into single-output processes through either partitioning allocation or different forms of system expansion.
+Different approaches for creating foregrounds to model multi-output process are presented in the supporting information.
 
-The first (a) is a sequential model, in which each node requires one foreground input and generates one output.
-This model is equivalent to a ``gate to gate'' model.
-Here the weights $k_i$ indicate the amount of the preceding reference flow that is required by the subsequent node.
-Figure~\ref{fig:fragments}(b) shows an additive model, in which the outputs of several foreground nodes are added together, equivalent to a ``mixer'' or a horizontal average.
-In this arrangement the weights %represent the relative weights of each input, and
-should add up to a unit output of the reference node.
-Finally, Figure~\ref{fig:fragments}(c) shows a foreground model with a cyclic dependency, where some of the reference output is consumed by another foreground node.
+An implicit ``foreground system boundary'' can be imagined to contain all the nodes in $A_f$.
+Non-zero values in this matrix represent exchanges inside this boundary.
+If a column in $A_f$ is nonempty, then the corresponding node has exchanges with other nodes in the foreground.
+On the other hand, if a column in $A_f$ is empty, then that is an indication that the corresponding product flow is crossing the foreground system boundary.
+In order for this flow to contribute any impacts, it must either be linked to a background process in the dependency matrix, or represented as an emission.
 
-A typical PSM may contain multiple modules or fragments that are interconnected.
-An example of foreground composed of several fragments is illustrated in Figure~\ref{fig:fragments}(d).
-Here, the nodes labeled 1--5 represent one fragment, which generates the foreground's canonical reference flow $\tilde{y}$.
-This fragment requires two interior flows from separate fragments ($y_0$ and $y_1$), and has two unconnected flows (4 and 5).
-The reference $y_0$ is supplied by a second fragment, made up of nodes 6-8.
-The reference $y_1$ is supplied by another fragment made of only one node (9).
-The reference flow $y_1$ is consumed in two different places by the other fragments.
+A flow whose entire column consists of 0s throughout $A_f$, $A_d$, and $B_f$ is a ``cut-off'' that exits the model with zero burdens.
+Specifying cut-off criteria and reporting of cut-off flows is an important part of system boundary definition.
+Thus, reviewing cut-off flows (including evaluating their significance) is a crucial part of review that is facilitated by having a structured disclosure of the model.

### 2018-02-07 aggregation/agg-lca-study.tex
Finishing body revision

@@
 $\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
-It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$  (A proof is provided in the electronic supplementary information).
+It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$  (A proof is provided in the electronic supporting information).
 This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.
@@
 \end{equation}
-where $\tilde{\mathbf{y}}_f$ is a canonical functional unit vector whose length is the same as the rank of $A_f$. Eq.~\ref{eqn:study} is derived in the supplementary materials.
+where $\tilde{\mathbf{y}}_f$ is a canonical functional unit vector whose length is the same as the rank of $A_f$. Eq.~\ref{eqn:study} is derived in the supporting information.
 Eq.~\ref{eqn:study} is visualized in Figure~\ref{fig:foreground}, with the background technology matrix $A$ subsumed into $B_x$ and replaced by an identity matrix $I$.

### 2018-02-07 aggregation/agg-objective.tex
Further tweaking foreground derivation and proofs
Finishing body revision

@@
 Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh2005a, Peters_JLCA_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.
-In the context of this paper, an ``LCA Study'' should be understood as any collection of inventory data under consideration as a group, from a single-operation unit process to a full ISO-style product LCA.
+%In the context of this paper, an ``LCA Study'' should be understood as any collection of inventory data under consideration as a group, from a single-operation unit process to a full ISO-style product LCA.
 
@@
 At the same time the author may have constraints that limit what can be disclosed.
-The main goal of the framework is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.
+The main goal of the framework presented here is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.
 
-On the basis of transparency, the framework should meet the following objectives:
+On the basis of transparency, the framework for disclosure of the PSM should meet the following objectives:
 \begin{itemize}

### 2018-02-07 aggregation/agg-preparing.tex
Finishing body revision

\section{Preparing the Foreground Matrix}

The development in the main text assumes the product system model is already described as a symmetric collection of single-output processes (activities) and flows (commodities) related by coefficients.  In concrete terms, this means that:
\begin{itemize}
\item each node in the foreground (column shared by $A_f$, $A_d$, and $B_f$) corresponds to a particular flow in the model's foreground (row of $A_f$);
\item a unit activity of each node in the foreground corresponds to a unit output of each flow;
\item the coefficients in $A_f$, $A_d$, and $B_f$ are ratios expressing the quantity of the flow (row), in its native unit, exchanged per unit of the reference flow (column). 
\end{itemize}

In short: each foreground column has both an intrinsic reference flow and a unit of measure, and the use of this flow by other processes must be made with regard to this measure.  The activity level of each foreground node necessarily equals the magnitude of the total reference flow emanating from the node.  For background dependencies and emissions (rows of $A_d$ and $B_f$, respectively), the modeler must take care to use the units established by the references, but in the foreground case the modeler must ensure to be self-consistent in preparing the $A_f$ matrix.

\subsection{Single-Output Foreground Processes}

\input{fig_fragments}

Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.  Most product system models would include many of these elements mixed together.  All the activities in these examples are single-output.
The first (a) is a sequential model, in which each node requires one foreground input and generates one output. This model is equivalent to a ``gate to gate'' model.  Here the weights $k_i$ indicate the amount of the preceding reference flow that is required by the subsequent node.  Figure~\ref{fig:fragments}(b) shows an additive model, in which the outputs of several foreground nodes are added together, equivalent to a ``mixer'' or a horizontal average.  In this arrangement the weights %represent the relative weights of each input, and
should add up to a unit output of the reference node.  Figure~\ref{fig:fragments}(c) shows a foreground model with a cyclic dependency, where some of the reference output is consumed by another foreground node.

A typical PSM may contain multiple modules or fragments that are interconnected.  An example of foreground composed of several fragments is illustrated in Figure~\ref{fig:fragments}(d).  Here, the nodes labeled 1--5 represent one fragment, which generates the foreground's canonical reference flow $\tilde{y}$. This fragment requires two interior flows from separate fragments ($y_0$ and $y_1$), and has two unconnected flows (4 and 5).  The reference $y_0$ is supplied by a second fragment, made up of nodes 6-8.  The reference $y_1$ is supplied by another fragment made of only one node (9). The reference flow $y_1$ is consumed in two different places by the other fragments.

As noted, a reviewer with access to items \ref{itm:fg} and \ref{itm:af} in the disclosure would be able to construct $A_f$ and automatically create a process-flow diagram, annotated with information about each node.

### 2018-02-07 aggregation/agg-review.tex
Finishing body revision

@@
 
-\subsection{Comparative Review: Formalizing Scope and System Boundary}
+\subsection{Comparative Review: Formalizing Model Scope}
 
@@
 These lists report all the distinct sources of environmental impacts in the model.
-The cutoff flows, which are foreground nodes that have no nonzero entries in $A_f$, $A_d$, or $B_f$, report on the other hand what aspects were explicitly excluded from the model.
+The cut-off flows, which are foreground nodes that have no nonzero entries in $A_f$, $A_d$, or $B_f$, report on the other hand what aspects were explicitly excluded from the model.
 

### 2018-02-07 aggregation/agg-semantic.tex
Finishing body revision

@@
 Similarly, LCIA method providers should ensure that their methods make use of stable semantic references to elementary flows and contexts.
-Current research has revealed widespread challenges to finding consistency on the identities of elementary flows despite the community's longstanding awareness of the problem, and several mutually inconsistent reference flow sets exist \citep{Edelen_2017}.
+Current research has revealed widespread challenges to finding consistency on the identities of elementary flows despite the community's longstanding awareness of the problem \citep{Speck_2015,Herrmann_2015}, and several mutually inconsistent reference flow sets exist \citep{Edelen_2017}.
 The conventional understanding of a ``flow'' as comprising a substance and a context together (e.g.

### 2018-02-07 aggregation/agg-agg.tex
Change $ to $

@@
 Again, foreground emissions are often not distinguished from the aggregation result (Figure~\ref{fig:aggregation}f).
-The life cycle inventory $\tilde{\mathbf{b}}$ provides the most aggregated form of the study that can still be independently validated with an external characterization vector $\mathbf{e}$.
+The life cycle inventory $\tilde{\mathbf{b}}$ provides the most aggregated form of the study that can still be independently validated with an external characterization vector $\mathbf{c}$.

### 2018-02-07 aggregation/agg-basic-lca.tex
Change $ to $

@@
 \begin{equation}
-s = \mathbf{e}^T \cdot B \cdot \left(I - A\right)^{-1} \cdot \mathbf{y}
+s = \mathbf{c}^T \cdot B \cdot \left(I - A\right)^{-1} \cdot \mathbf{y}
 \label{eqn:leontief}
 \end{equation}
-Here $A$ represents the direct requirements; $B$ is the emission matrix;  $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions; $\mathbf{y}$ is the externally specified final demand; and $s$ is the numerical impact score or category indicator.
+Here $A$ represents the direct requirements; $B$ is the emission matrix;  $\mathbf{c}$ is a column vector of characterization factors for the environmental emissions; $\mathbf{y}$ is the externally specified final demand; and $s$ is the numerical impact score or category indicator.
 For simplicity, we will consider the computation of a single impact result.
-However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.
+However, it is straightforward to imagine the more typical case, in which $C$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.
 Eq.~\ref{eqn:leontief} is visualized in Figure~\ref{fig:lca}.
@@
 \item $\mathbf{b} = B\cdot\mathbf{x}$, the life cycle inventory;
-\item $s = \mathbf{e}^T\cdot\mathbf{b}$, the life cycle impact category indicator, the study result.
+\item $s = \mathbf{c}^T\cdot\mathbf{b}$, the life cycle impact category indicator, the study result.
 \end{itemize}

### 2018-02-07 aggregation/agg-disclosure.tex
Change $ to $

@@
 The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.
-Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$ (or their product $\mathbf{e}\times B_x$) are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.
+Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{c}$ (or their product $\mathbf{c}\times B_x$) are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.
 These three submatrices together can be considered a complete representation of the PSM.
@@
 \item\label{itm:bg} An ordered list of background flows, each making reference to a particular background dataset (rows of $A_d$, mapping to columns of $B_x$);
-\item\label{itm:em} An ordered list of exchanges with the environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
+\item\label{itm:em} An ordered list of exchanges with the environment (rows of $B_f$, mapping to entries in $\mathbf{c}$);
 \item\label{itm:af} A set of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
@@
 \end{itemize}
-As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{e}$.
+As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{c}$.
 These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.

### 2018-02-07 aggregation/agg-lca-study.tex
Change $ to $

@@
   \begin{array}{cc}
-s =& \mathbf{e}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
- =& \mathbf{e}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
+s =& \mathbf{c}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
+ =& \mathbf{c}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
   \end{array}
@@
 A process for brewing coffee could be created in a third column that requires coffee beans, water, and electricity.
-Finally, a fourth column could be added that invokes the first three columns in proportion to describe the activity of traveling to a coffeeshop to work\footnote{No prizes for guessing where this passage was written.}.
+Finally, a fourth column could be added that invokes the first three columns in proportion to describe the activity of traveling to a coffeeshop to work.%\footnote{No prizes for guessing where this passage was written.}.
 
@@
 \begin{equation}
-s = \mathbf{e}^T \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \tilde{\mathbf{y}}_f
+s = \mathbf{c}^T \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \tilde{\mathbf{y}}_f
 \label{eqn:study}
@@
 \begin{array}{rl}
-s & = \mathbf{e}^T \times( B_f + B_x \times A_d) \times \tilde{\mathbf{x}} \\
-  & = \mathbf{e}^T \times( \tilde{\mathbf{b}}_f + B_x \times\tilde{\mathbf{a}}_d) \\
-  & = \mathbf{e}^T \times( \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x) \\
+s & = \mathbf{c}^T \times( B_f + B_x \times A_d) \times \tilde{\mathbf{x}} \\
+  & = \mathbf{c}^T \times( \tilde{\mathbf{b}}_f + B_x \times\tilde{\mathbf{a}}_d) \\
+  & = \mathbf{c}^T \times( \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x) \\
 \end{array}

### 2018-02-07 aggregation/agg-pagg.tex
Change $ to $

@@
 \begin{align}
-  s =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
-  =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}\\
+  s =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
+  =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}\\
   \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}

### 2018-02-07 aggregation/proofs.tex
Change $ to $

@@
 \begin{align}
-  s =\;& \mathbf{e}^T\times \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}\tilde{\mathbf{y}}\\
-  =\;& \mathbf{e}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
+  s =\;& \mathbf{c}^T\times \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}\tilde{\mathbf{y}}\\
+  =\;& \mathbf{c}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
   \left[\begin{array}{cc} (I-A_f)^{-1} & 0 \\ A_d\times(I-A_f)^{-1} & I \end{array}\right]
@@
 \begin{align}
-s =\;& \mathbf{e}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
+s =\;& \mathbf{c}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
  \left[\begin{array}{c} (I-A_f)^{-1}\times\tilde{\mathbf{y}}_f \\ A_d\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f\end{array}\right]\\
- =\;& \mathbf{e}^T\times \left(B_f\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f
+ =\;& \mathbf{c}^T\times \left(B_f\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f
  + B_x\times A_d\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f\right)\\
- =\;& \mathbf{e}^T\times \left(B_f + B_x\times A_d\right)\times(I-A_f)^{-1}\times \tilde{\mathbf{y}}_f\\
- =\;& \mathbf{e}^T\times \left(B_f + B_x\times A_d\right)\times \tilde{\mathbf{x}}
+ =\;& \mathbf{c}^T\times \left(B_f + B_x\times A_d\right)\times(I-A_f)^{-1}\times \tilde{\mathbf{y}}_f\\
+ =\;& \mathbf{c}^T\times \left(B_f + B_x\times A_d\right)\times \tilde{\mathbf{x}}
 \end{align}

### 2018-02-07 aggregation/agg-agg.tex
Starting the reviewer response

@@
 The reader no longer requires access to any background database to perform the computation, but all dependency information is concealed.
-Again, foreground emissions are often not distinguished from the aggregation result (Figure~\ref{fig:aggregation}f).
+As mentioned above, foreground emissions are often not distinguished from the aggregation result; if foreground and background emissions are added together, the disclosure is as shown in Figure~\ref{fig:aggregation}f.
 The life cycle inventory $\tilde{\mathbf{b}}$ provides the most aggregated form of the study that can still be independently validated with an external characterization vector $\mathbf{c}$.

### 2018-02-07 aggregation/agg-review.tex
Change $ to $
Starting the reviewer response

@@
 Together, these lists can be considered a functional definition of the study scope that can be used to conduct a qualitative comparison of multiple studies by different authors.
-Background database choice, dataset selection, and version information is all made available in disclosure item~\ref{itm:bg}, thereby streamlining what is presently an interpretive and study-specific process of extracting this information from written text.
+Background database choice, dataset selection, and version information can all be made available in disclosure item~\ref{itm:bg}, thereby streamlining what is presently an interpretive and study-specific process of extracting this information from written text.
 Similarly, the set of elementary flows explicitly modeled in the study foreground (item~\ref{itm:em}) expose whether any study-specific modeling of direct emissions was performed and what flows were included.
@@
 If a study containing confidential information is split into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated directly by the reviewer.
-To achieve this, the reviewer must be granted access to both the public and private partitions in order to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly.
+To achieve this, the reviewer must be able to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly.
+This requires that the reviewer be granted access to both the public and private partitions.
 However, the private information used in the computation can still be withheld from the disclosure.
@@
 \begin{equation}
-  \varphi =  \frac{\mathbf{e}^T\cdot(\mathbf{b} - \mathbf{b}_{priv})}{\mathbf{e}^T\cdot\mathbf{b}} = 1 - \frac{\mathbf{e}^T\cdot\mathbf{b}_{priv}}{\mathbf{e}^T\cdot\mathbf{b}}
+  \varphi =  \frac{\mathbf{c}^T\cdot(\mathbf{b} - \mathbf{b}_{priv})}{\mathbf{c}^T\cdot\mathbf{b}} = 1 - \frac{\mathbf{c}^T\cdot\mathbf{b}_{priv}}{\mathbf{c}^T\cdot\mathbf{b}}
   \label{eqn:metric}

### 2018-02-07 revision/rev2-response.tex
Starting the reviewer response

\documentclass[11pt,letterpaper]{article}
\input{formatting}

\title{Disclosure of Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy}
\author{Brandon Kuczenski\\
Institute for Social, Behavioral, and Economic Research\\
University of California, Santa Barbara}
\date{\mydate\today}

\begin{document}
\maketitle

\section*{Response to Reviewers}

Thank you again to the reviewers for their robust reading of my revised manuscript.  I appreciate that all reviewers found the article to be engaging and important.  Their comments helped me to clarify the presentation of the methodology and to improve the connection to current scholarship, notably Majeau-Bettez et al (2014).  I have taken all of the reviewers' recommendations to heart.  Below I respond to each individually.

\subsection*{Reviewer \#1}

\emph{In the response to the reviewers I read: "I have adopted the input-output notation exclusively in the revision". That is true. There are no longer formulas with A^-1, only with the IO-style (I-A)^-1. Unfortunately this has weakened the relevance of the paper quite a bit. Let me explain. There are two ways of doing LCA: process-based and IO-based.  A major example of a process-based background database is Ecoinvent, a major example of an IO-based background database is Exiobase. The issue which this paper addresses (proprietary licensed data in the background) is the case for Ecoinvent, but not for Exiobase. So the paper makes a wrong choice in explaining how to protect non-proprietary data from IO but not explaining how to protect proprietary data from process-based DBs. I have no clue why you are making this choice, except one debatable statement that the IO-approach "is computationally more efficient" (which I personally doubt, but moreover is irrelevant for the present paper). As a matter of fact, most software for LCA (so page 7 line 49 is definitely wrong) is process-based, and so are the standards from ISO (in which the "unit process" assumes a core role). I strongly urge to phrase the paper entirely in process-based style, using A^-1 instead of (I-A)^-1. Editorially speaking it is a small change, but it increases the relevance of the paper tremendously.}

I thank the reviewer for his/her deep insights into the paper, which have led to a complete rewrite of the section of the manuscript dealing with the core LCA computation.  I now tie this strongly to the supply-and-use framework. (p.5-7 in the revision).

I note that the reviewer inaccurately conflates the \textit{contents} of an inventory database with their mathematical \textit{representation.}  In fact, both Ecoinvent and Exiobase are best represented using the supply-and-use framework.  It is only when preparing the LCA computation for matrix inversion that the supply and use values are placed into coefficient form.  In Ecoinvent, each unit process is described in terms of distinct products (outputGroup) and consumption (inputGroup) separately.  The un-allocated datasets are not suitable for computation, and the linked ecoinvent system models are stored as collections of normalized, single-output EcoSpold v2 files, which are trivial to compile in either technology-matrix or input-output form.  I have emphasized this in the revision.  The point of the article was to demonstrate how foreground models may be built on top of, and subsequently separated from, a background database, and the development applies equally well to a foreground model based on either Ecoinvent or Exiobase.

Regarding the computational efficiency of iteration vs inversion, I agree that the performance aspects are out of scope for the current paper.  However, the reviewer should note that in fact the Leontief inverse is the standard.  SimaPro, the leading worldwide software system for scientific LCA, converted to using an iterative Leontief inverse for Version 8, as disclosed in a whitepaper published in 2014 (https://www.pre-sustainability.com/news/new-calculation-engine-simapro-8), realizing significant speed and memory usage improvements.  OpenLCA, the open-source alternative, also reported a dramatic reduction in memory usage and computation time around the release of version 1.4 in 2014, coincident with a revised algorithm "optimized for huge databases" (http://www.openlca.org/openlca-beta-1-4-released-optimised-for-huge-databases/) which is highly suggestive of a switch to an iterative approach.  GaBi, the more industrially oriented alternative, does not use a matrix formulation at all, but has always computed activity levels using iteration at the plan level.

\emph{page 8 line 32 "The simplest conceivable foreground is simply the final demand y, in other words the list of background processes invoked by the study and their activity levels." I have no idea what this sentence means. The foreground is a collection of processes, but the final demand is not a process, so how can the final demand be a foreground? And as to the second part: what is the relation between the foreground and the list of background processes?}

The problematic sentence has been removed in revision, and the whole section has been revised to illustrate exactly how foreground models may be built.

* On page 8 line 48 I see a matrix ~A with a 0 in the top-left corner. Page 9 generalizes this, but suddenly has a matrix (A_f) at that place. This shift from page 8 to page 9 should be justified, because this is a major change.
* The same matrix ~A on page 8 line 48 contains a bold 0, which suggests a column of zeros, but actually is a row of zeros. Adding the subscript T would help here.
* Page 10 line 12 says "The augmented ~A and ~B, makes up an LCA foreground study". I am wondering if something went wrong. Two grammatical errors (the spurious comma and the singular verb) suggest an editorial mistake. But the meaning is also unclear: ~A and ~B contain foreground and background, so why is this a "foreground study"? What is a foreground study anyhow?
* Page 10 line 12 also speaks of "the canonical functional unit". Can you define the term canonical?
* Page 10 line 27 rewrites ~A as ~A_flat. What is the relation between these two symbols? Of course you may define a new symbol ~A_flat, but I even don't know what you intend do with ~A_flat: I see no equation in which it is used (like for instance ~A is used in eq (2)).
* Page 10 line 37 "the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported". Of course my critique depends on the missing definition of "canonical", but even then, do I understand that you write that case studies do no longer need to report a functional unit? That would be a revolutionary statement, so please expand or revise.

### 2018-02-08 aggregation/agg-disclosure.tex
Reviewer response

@@
 \end{enumerate}
-Each entity included in items \ref{itm:fg}--\ref{itm:em} describes a flow of some kind, and its unit of measure must be reported in the disclosure.
+Each entity included in items \ref{itm:fg} through \ref{itm:em} describes a flow of some kind, and its unit of measure must be reported in the disclosure.
 In both items~\ref{itm:bg} and~\ref{itm:em}, only rows containing non-zero entries need to be mentioned, since the index of each row is common information.

### 2018-02-08 aggregation/agg-preparing.tex
Reviewer response

@@
 
-In short: each foreground column has both an intrinsic reference flow and a unit of measure, and the use of this flow by other processes must be made with regard to this measure.
+Each foreground column has both an intrinsic reference flow and a unit of measure, and the use of this flow by other processes must be made with regard to this measure.
 The activity level of each foreground node necessarily equals the magnitude of the total reference flow emanating from the node.

### 2018-02-08 revision/rev2-response.tex
Reviewer response

@@
 That is true.
-There are no longer formulas with A^-1, only with the IO-style (I-A)^-1.
+There are no longer formulas with A\^{}-1, only with the IO-style (I-A)\^{}-1.
 Unfortunately this has weakened the relevance of the paper quite a bit.
@@
 As a matter of fact, most software for LCA (so page 7 line 49 is definitely wrong) is process-based, and so are the standards from ISO (in which the "unit process" assumes a core role).
-I strongly urge to phrase the paper entirely in process-based style, using A^-1 instead of (I-A)^-1.
+I strongly urge to phrase the paper entirely in process-based style, using A\^{}-1 instead of (I-A)\^{}-1.
 Editorially speaking it is a small change, but it increases the relevance of the paper tremendously.}
@@
 
-\emph{page 8 line 32 "The simplest conceivable foreground is simply the final demand y, in other words the list of background processes invoked by the study and their activity levels." I have no idea what this sentence means.
+\emph{* page 8 line 32 "The simplest conceivable foreground is simply the final demand y, in other words the list of background processes invoked by the study and their activity levels." I have no idea what this sentence means.
 The foreground is a collection of processes, but the final demand is not a process, so how can the final demand be a foreground?
@@
 
-The problematic sentence has been removed in revision, and the whole section has been revised to illustrate exactly how foreground models may be built.
+\emph{* On page 8 line 48 I see a matrix ~A with a 0 in the top-left corner.
+Page 9 generalizes this, but suddenly has a matrix (A\_f) at that place.
+This shift from page 8 to page 9 should be justified, because this is a major change.}
 
-* On page 8 line 48 I see a matrix ~A with a 0 in the top-left corner.
-Page 9 generalizes this, but suddenly has a matrix (A_f) at that place.
-This shift from page 8 to page 9 should be justified, because this is a major change.
-* The same matrix ~A on page 8 line 48 contains a bold 0, which suggests a column of zeros, but actually is a row of zeros. Adding the subscript T would help here.
-* Page 10 line 12 says "The augmented ~A and ~B, makes up an LCA foreground study".
+The passage introducing the foreground study has been revised extensively, and the shift from a final demand vector to a model for a process, and from a collection of processes to a complete foreground, has been re-written in a narrative format to illustrate the workflow (p.9).
+Hopefully the revision is clear.
+
+\emph{* The same matrix ~A on page 8 line 48 contains a bold 0, which suggests a column of zeros, but actually is a row of zeros. Adding the subscript T would help here.}
+
+Thank you for the observation.  This has been fixed.
+
+\emph{* Page 10 line 12 says "The augmented ~A and ~B, makes up an LCA foreground study".
 I am wondering if something went wrong.
@@
 But the meaning is also unclear: ~A and ~B contain foreground and background, so why is this a "foreground study"?
-What is a foreground study anyhow?
-* Page 10 line 12 also speaks of "the canonical functional unit". Can you define the term canonical?
-* Page 10 line 27 rewrites ~A as ~A_flat.
-What is the relation between these two symbols?
-Of course you may define a new symbol ~A_flat, but I even don't know what you intend do with ~A_flat: I see no equation in which it is used (like for instance ~A is used in eq (2)).
-* Page 10 line 37 "the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported".
+What is a foreground study anyhow?}
+
+The grammatical error has been corrected.
+A foreground study is defined on p.9 as the two augmented $\tilde{A}$ and $\tilde{B}$ matrices, containing foreground information described with reference to a background database.
+
+\emph{* Page 10 line 12 also speaks of "the canonical functional unit". Can you define the term canonical?}
+
+Some puzzles are best left as an exercise for the reader.
+
+\emph{* Page 10 line 37 "the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported".
 Of course my critique depends on the missing definition of "canonical", but even then, do I understand that you write that case studies do no longer need to report a functional unit?
 That would be a revolutionary statement, so please expand or revise.
+}
+
+The \textit{canonical functional unit} is now defined at the top of p.10 as a unit of the reference flow of the process described by the \textit{first} column of the $\tilde{A}$ matrix.
+It is introduced for the purposes of simplifying notation later on.
+While the use of a canonical functional unit means that the functional unit of a study no longer needs to be explicitly reported, it does still need to be defined as part of the disclosure.
+This is mentioned on p.12 in disclosure item \emph{d-i}.
+
+\emph{  * Page 10 line 27 rewrites ~A as ~A\_flat.
+What is the relation between these two symbols?
+Of course you may define a new symbol ~A\_flat, but I even don't know what you intend do with ~A\_flat: I see no equation in which it is used (like for instance ~A is used in eq (2)).}
+
+The equation formerly known as Eq.~4 is now removed from the manuscript.
+$A_{flat}$ still makes an appearance in the supporting information, but it is only an intermediate step toward deriving the main foreground study equation (old Eq.~5; new Eq.~4).
+
+\subsection*{Reviewer \#2}
+
+\emph{The resubmitted version prepared by the author represents a major improvement compared to the previous version in terms of clarity and conveyance of the message.
+I had a good time while reading the manuscript.
+  The derivation of the canonical LCA computation equation, the simplifications and data arrangements, the listing of the six data components d-i to d-vi, and the data privacy arrangement are convincing and clear.}
+
+Thank you for the positive comments.
+
+\emph{+ I have only one major issue with the framework presented: It starts with the A-matrix, which already includes a major modelling step: the conversion of raw unallocated processes data, including production volumes, etc.
+to 1:1 product-to-industry coefficients.
+What about the documentation of the raw, unallocated process data, which are the most valuable?
+  Including this part of the data work into the disclosure framework is crucial to its eventual success in my (and many others’) view.
+It may be beyond the scope of the current paper, but I still think the paper here must position itself.
+I therefore}
+
+\emph{+ request to comment on the issue in the manuscript, e.g., in the discussion, and}
+
+\emph{+ suggest to spend a couple of pages in the SI describing the issue.}
+
+I agree with the reviewer that omitting the conversion of unallocated processes to single-output processes leaves out an important modeling step.
+I first revised the introduction of Eq.1 by making strong ties to the supply-and-use framework of input-output analysis (beginning on p.5), as mentioned for the first reviewer.
+I then undertook a close review of Majeau-Bettez et al (2014) upon the reviewer's recommendation, and found that the article's taxonomy of allocation strategies could be captured within the foreground disclosure framework.
+
+In order to not clutter the text too much, I limit intensive discussion about this topic to the supporting information, where I demonstrate how an example multi-output process may be converted into single-output processes by means of the framework.
+I take care to note that each strategy preserves the raw, un-allocated process data in the disclosure.
+I draw attention to this fact a few times in the main text, first on p.14 bottom; again on p.17-18; and again in the conclusion.
+
+Hopefully this work is convincing for the reviewer.
+I was quite happy to come to the determination that the multi-functionality problem fits well into the scope of the current manuscript, and I feel the potential impact of the paper has been greatly improved because of this revision.
+
+\emph{Detailed comments}
+
+\emph{P4: “In the context of this paper, an “LCA Study” should be understood as …” is not very clear, please rephrase.}
+
+The problematic sentence has been removed.
+
+\emph{P5: “On the basis of transparency, the framework… “ -> write ‘product system model disclosure framework” or something else that is more explicit.}
+
+I accept the reviewer's suggestion.
+
+\emph{P6: “The task of procedure can be described in two steps: making …” Please link these steps to the ISO standard.}
+
+The sentence has been removed in the revision.
+
+\emph{Equation 1: I recommend using c instead of e, since e is often used as summation vector, i.e., a vector that contains only 1 as elements.}
+
+This has been done throughout the text, figures, and supporting information.
+
+\emph{Equation1 and in general: Please be consistent throughout the entire manuscript which variables contain flows (at scale) and which contain coefficients.
+Any jargon in that regard must be avoided, e.g., referring to B as ‘elementary flow matrix’, where, in reality, it is a normalized (per unit) elementary flow matrix.}
+
+Assuming you mean ``normalized (per unit) coefficient matrix,'' I have taken care to emphasize that all matrices in the disclosure framework are coefficient matrices.
+Notably p.6 middle; p.7; and p.11 top.
+No variables in the text refer to flows at scale.
+
+\emph{Another jargon phrase to be clarified:
+  “the foreground systems themselves cannot be computed without the background.” (P8)}
+
+Revision: ``However, the results of a study with foreground content cannot generally be computed without knowledge of the background.''
+
+\emph{P9: I think that at latest at this stage, many readers will wonder how the system you propose applies to cases where process descriptions in the background matrix are altered (also related to the comment of one of the reviewers regarding the zero-ness of the upper right corner of $\tilde{A}$.
+I think you should anticipate that question early on to keep your readers on board.
+In my opinion, ...) ...}
+
+This is discussed in Note \#2, referenced on p.9 bottom.
+
+\emph{Equation 4 is not a re-written version of equation 3! Please rephrase.}
+
+The equation formerly known as Eq.~4 has been removed from the manuscript.
+
+\emph{P10: I have never heard of the rank of a vector. Do you simply mean its length?}
+
+Revision: ``vector whose length is the same as the rank of $A_ f$.''
+
+\emph{P11: The term ‘aggregated foreground’ needs more introduction/explanation.
+At least the name, dimension, meaning, and unit (coefficient or flow??) of each of the newly introduced variables should be listed.}
+
+I have added text around Eq.~5 (formerly Eq.~6) to clarify these points.
+
+\emph{It would be nice to see Equation 6 as part of a larger ‘master equation’, for example, adding a bit more detail at the bottom of Figure 2.}
+
+Eq.~4 (formerly Eq.~5) is the ``master equation'' for the foreground model.  I have added it in a new row to the botttom of Fig.2.  Hopefully it made the figure more helpful than confusing.
+
+\emph{P10, “where ˜yf is a canonical functional unit vector …”, and P11: “The foreground matrix Af, dependency matrix Ad, and foreground emission matrix Bf now contain all study-specific information.”
+  That means that you are using the ‘canonical’ y vector all throughout your system, and that fact should be make more clear not that readers think that you forgot about y.
+Does ‘canonical’ here mean that yf is always [1,0,0,0, …., 0]?
+If so, that should be made clear on P10.}
+
+I have improved the declaration of the canonical functional unit on p. 10.
+
+\emph{P12: “The disclosure includes only information for which the study is the primary source.” To me that does not often seem to be the case, as also in the foreground, data from the literature and sometimes even recycled data from ecoinvent itself are used, e.g., when background inventories are copied into the foreground and modified there.
+  I think that the statement on the foreground system description as ‘primary source’ should be toned down/modified.}
+
+Revision: ``The disclosure includes all information specific to the study, or for which the study is the primary source, including re-implementations of already-published data.''
+
+\emph{P13: “Objectives in critical review vary… ” I recommend sharpening this paragraph.
+Reviewers should be held responsible for checking the model correctness and data/model/system structure as well, and a canonical PSM structure like the one proposed here would greatly facilitate the completion of this task.
+What is common now is clearly insufficient.}
+
+At your urging, I added the following to p.14: ``Although the object of critical review is typically a written report, this level of review is not sufficient to ensure the correctness of quantitative results.''  This is a statement of sentiment, but I do feel that it is accurate.
+
+\emph{P13: “These nodes typically represent unit processes, but they can also indicate exchanges across an implicit system boundary that encloses the foreground.” I don’t understand the second part.
+What is an implicit system boundary here?
+How can A-matrix columns describe unit processes and exchanges at the same time?
+I thought the latter is coded in Bf?
+Clarification needed.}
+
+I now explicitly introduce the notion of a ``foreground system boundary'' on p.15 and use it to define cut-off flows in a robust way.
+
+\emph{P14: “… it must either be “terminated” in the dependency matrix… ” What is the dependency matrix here? Ad? Please specify.}
+
+I removed the term ``terminated'' since it is of my own coinage; and yes, the dependency matrix is defined as $A_d$ in Eq.3 and so referenced on p.11 in ``Contents of a disclosure.''
+
+\emph{P15: “In a matrix representation, each unit process must have exactly one reference flow, which
+  corresponds to its row and column in Af.” That is a crucial LCA-matrix feature and should be described a bit more in detail: $\ldots$ In short: Each A-matrix column refers to the making 1 reference product AND unit, and exactly this reference product and unit is then used by other processes.}
+
+These points are made with greater clarity in the supporting information.
+
+\emph{In general, when writing about the classifications d-i, d-ii, and d-iii the importance of units should be stressed at some point in the manuscript.}
+
+This is now emphasized on p.13: ``Each entity included in items \ref{itm:fg} through \ref{itm:em} describes a flow of some kind, and its unit of measure must be reported in the disclosure.''  It is also discussed in the supporting information.
+
+\emph{Figure 3: A legend must be provided, explaining the meaning of the blue boxes, the dashed-line boxes, and the star.
+The variable $\tilde{a}_f$ is new to me and I don’t understand its shape.
+Notation in Fig.
+2, 3, and equation six must be checked for consistency.
+Else the figure is very informative and clear!}
+
+I added a legend and also a definition for the formerly un-defined $\tilde{\mathbf{a}}_f$ on p.18.  I also rendered the vectors in bold face for consistency with the text.
+
+\emph{P16: “A fragment of the PSM can also be published as a complete foreground.” I don’t understand, please explain.}
+
+The sentence has been removed, and the word ``fragment'' no longer appears in the manuscript.  Along with ``termination'' of foreground flows, this is rather a concession for me!
+
+\emph{P16: “Any co-production treatment, such as an allocation or system expansion, can only be fully expressed as a foreground fragment, because it intrinsically involves a collection of linked unit processes.” I don’t understand, please explain.}
+
+Revision: ``Any modeling technique used to treat a multi-functional process, such as an allocationor system expansion, can only be fully expressed as a foreground with at least as many columns as
+outputs. For examples of multi-output processes represented as foreground models, please see the
+supporting information.''
+
+\emph{P17: “A publication of the aggregated foreground is identical to the publication of a unit process, except that the result was derived through aggregation” I don’t understand, please explain.}
+
+Revision: ``A disclosure of an aggregated foreground resembles the description of a unit process.''
+
+\emph{Section 5.4. Consider citing DOI 10.1111/jiec.12245.}
+
+This has been added.
+
+\emph{I suggest to include another brief section 5.5., containing a comment on the role of the ISIE (it’s their journal!) or the SETAC group in facilitating the establishment of science-wide classifications.
+What about the responsibility of the database and LCIA-method providers?}
+
+At the moment I neglected to include a new section because the paper is already quite long.  But maybe I will change my mind in the next 24 hours.
+
+\subsection*{Reviewer \#3}
+
+\emph{First of all, the new title is much clearer and to the point.}
+
+\emph{The scope of the new version of the article is also clearer and more consistent with the
+objectives. I am satisfied with the responses to my comments on the earlier version as well as
+that of the other reviewers.}
+
+Thank you for your positive comments.
+
+\emph{One additional comment in the form of a suggestion is the following. From my point of view, the
+core result in this article is presented in section 2.4 with the list of elements from the PSM which
+should be disclosed:}
+
+\emph{1. This list could be represented graphically, similar to a “disclosure card” such as a
+  simplified versions of Figures S2 or S3.}
+
+New Figure 3 has been added to obtain this objective.  I hope you like it!
+
+\emph{2. A reference to the examples derived in the annexes would make the demonstration
+  more concrete (there were several in the previous version of the article?).}
+
+I have added a mention of the examples to the section formerly identified as 2.4.
+
+\emph{This would help in emphasizing the next steps, especially if the results draw on discussions
+within the SETAC working group, and possibly answer the question of whether this could
+become part of future LCA standards.}
+
+If I decide to do this, it will be included in the section that Reviewer \#2 wanted me to add at the end of the discussion.
+
+\end{document}

### 2018-02-08 aggregation/agg-multioutput.tex
Late nite work

@@
 Each column must be normalized to a unit reference.
-The reference flow not explicitly specified will simply be computed as the remainder when the allocated inventories are subtracted from the un-allocated inventory.
+The reference flow not explicitly allocated will simply be computed as the remainder when the allocated inventories are subtracted from the un-allocated inventory.
 
@@
 At the same time, LCA computation using a functional unit of one of the co-products will simply make use of the allocated inventory.
-In this example, all processes are identical on a mass basis, but any other partitioning allocation can be automatically rendered self-consistent through this approach.
+In this example, all processes are identical on a mass basis, but any other partitioning allocation can also be automatically rendered self-consistent through this approach.
 A reviewer would be able to inspect the foreground model to determine whether the allocation was performed as described in the report.
@@
 Although the term `system expansion' is often used to refer to product substitution and/or alternate activity allocation, these modeling techniques are distinct and should be referred to with different names, as noted in GWS14.
-In the original definition of system expansion, the foreground model is simply enlarged to include all co-products.
+In the original definition of system expansion, the functional unit is simply enlarged to include all co-products.
 Representation of such a system is trivial and is shown in Table~\ref{tbl:cse}.
@@
 
-In the classical system expansion case, the canonical functional unit, production of 1 kg of chlorine gas, will result in two cutoff flows corresponding to outputs of liquid hydrogen and sodium hydroxide.
-Equivalently, an expanded functional unit of 1 kg chlorine, plus 0.028 kg hydrogen and 1.13 kg of sodium hydroxide, will result in the same activity of the un-allocated process, but will generate no cutoff flows at net.
+In the classical system expansion case, applying the expanded functional unit of 1 kg chlorine, plus 0.028 kg hydrogen and 1.13 kg of sodium hydroxide, will result in a unit activity of the reference process, without any cut-off flows..
+Equivalently, the canonical functional unit, production of 1 kg of chlorine gas, will result in the same process activity level, and will also generate the co-products as cutoff flows.
 
@@
 
-Here an alternate activity, namely electrolysis of water, is specified for liquid hydrogen production and is specified at index 3 in the foreground  (shown in blue).
+Here an alternate activity, namely electrolysis of water, is included for liquid hydrogen production and is specified at index 3 in the foreground  (shown in blue).
 The inventory converts 9 kg of water to 1 kg of hydrogen using 50 kWh of electricity.
-The hydrogen produced from the multi-output process is then (somewhat dubiously) taken to displace hydrogen produced through the alternate activity by introducing a unity coefficient in row 3, column 1 of $A_f$.
+The hydrogen produced from the multi-output reference process is (somewhat dubiously) taken to displace hydrogen produced through the alternate activity by introducing a unity coefficient in row 3, column 1 of $A_f$.
 
-Meanwhile, sodium hydroxide from the process is assumed to substitute the production of magnesium hydroxide through the introduction of a new row and a new nonzero coefficient in column 2 of $A_d$ (shown in red).
+Meanwhile, sodium hydroxide from the reference process is modeled to substitute the production of magnesium hydroxide through the introduction of a new row and a new nonzero coefficient in column 2 of $A_d$ (shown in red).
 Here the match is inexact, and because magnesium hydroxide is more effective in some applications, 1 kg of sodium hydroxide is taken to displace only 0.8 kg of magnesium hydroxide.

### 2018-02-08 aggregation/agg-requirements.tex
Multioutput case in SI
Late nite work

@@
-\subsection{Meeting the Disclosure Objectives}
+\section{Meeting the Disclosure Objectives}
 
-\subsubsection{Transparency}
+\subsection{Transparency}
 
@@
 The only requirement on their disclosure is that the identity of each node's reference flow, including its unit of measure, is clearly stated.
-In formalizing the study as a normalized direct requirements matrix, the activity level of each foreground node necessarily equals the magnitude of the total reference flow emanating from the node.
 
@@
 Sign consistency must also be assured.
-If LCIA indicator results are included, the author must also unambiguously identify the method computed (identity of $\mathbf{e}$).
+If LCIA indicator results are included, the author must also unambiguously identify the method computed (identity of $\mathbf{c}$).
 
-\subsubsection{Authority and Primacy}
+\subsection{Authority and Primacy}
 

### 2018-02-08 revision/rev2-response.tex
response minor edits

@@
 Their comments helped me to clarify the presentation of the methodology and to improve the connection to current scholarship, notably Majeau-Bettez et al (2014).
+
 I have taken all of the reviewers' recommendations to heart.
-Below I respond to each individually.
+The major changes to the manuscript were in the mathematical introduction sections-- I have rewritten ``Mathematical Formulation of LCA'' to centralize the supply-and-use framework, and to emphasize that the matrices developed in the text are coefficient matrices derived from supply and use tables.
+I have also rewritten ``Foreground Study'' to better illustrate how a foreground study is constructed with reference to a background database, and to better define the canonical functional unit.
+I added a new figure at the request of Reviewer \#3, and a new discussion subsection dealing with the role of the community.
+I also made an assortment of clarifying changes throughout the document.
+
+Below I respond to each comment individually.
 
@@
 SimaPro, the leading worldwide software system for scientific LCA, converted to using an iterative Leontief inverse for Version 8, as disclosed in a whitepaper published in 2014 (https://www.pre-sustainability.com/news/new-calculation-engine-simapro-8), realizing significant speed and memory usage improvements.
-OpenLCA, the open-source alternative, also reported a dramatic reduction in memory usage and computation time around the release of version 1.4 in 2014, coincident with a revised algorithm "optimized for huge databases" (http://www.openlca.org/openlca-beta-1-4-released-optimised-for-huge-databases/) which is highly suggestive of a switch to an iterative approach.
-GaBi, the more industrially oriented alternative, does not use a matrix formulation at all, but has always computed activity levels using iteration at the plan level.
+OpenLCA, the open-source alternative, also reported a dramatic reduction in memory usage and computation time around the release of version 1.4 in 2014, coincident with a revised algorithm "optimized for huge databases" (http://www.openlca.org/openlca-beta-1-4-released-optimised-for-huge-databases/).
+This is highly suggestive of a switch to an iterative approach.
+GaBi, the more industrially oriented software, does not use a matrix formulation at all, but has always computed activity levels using iteration at the plan level.
 

### 2018-02-09 aggregation/agg-community.tex
Community first scratch

\subsection{Advancing LCA Practice}

\endinput

The need for a shareable product system model not dependent on any single software system was identified as a key objective for advancing the state of the art in LCA \citep{Kuczenski_JLCA_2018}.  The framework introduced in this article contributes to several milestones in that document, namely the ability to identify what datasets are used in a model and how they are linked together.  The framework can also be used to document how allocation was performed.  \emph{This does not really answer the question}

What is the role of the community?
 - need to band data providers together under a common framework
   = need stble semantic references
   = need provider-independent community-oriented definitions for flows, quantities, and contexts
   = need open source tools

Milestones:

Questions from the reviewers:

R2: **I suggest to include another brief section 5.5., containing a comment on the role of the ISIE (it’s their journal!) or the SETAC group in facilitating the establishment of science-wide classifications. What about the responsibility of the database and LCIA-method providers?

R3: This would help in emphasizing the next steps, especially if the results draw on discussions within the SETAC working group, and possibly answer the question of whether this could become part of future LCA standards.

Community is responsible for: consensus definitions of flows, quantities, and contexts
Mainly, the community is responsible for disclosing product system models according to this definition
software makers need to enable this capability

``How do we use it?''

It is my contention that any model can be represented within the disclosure framework
so practitioners who are interested in facilitating sharing and/or reuse of their models should implement them in such a framework.
Critical reviewers should consider ISO reports with a thought to how the model would be represented in the disclosure framework, were it to be disclosed
it will obviously take time for software makers to create new functionalities for generating disclosures
EPDs can be accompanied by disclosures
PCRs can be written as PSM specifications
WEB SERVICES can enable direct computation of LCA results from PSM documents (how? entity list includes a mapping to a web service instead of to a static document; queries can be asked of the web service to determine quantitative aspects)

Data providers are responsible for: stable semantic references to data sets
LCIA method providers are responsible for: using community-defined flows instead of their own flows

What is all this good for?

 - we want to better

### 2018-02-10 aggregation/agg-community.tex
tweak wording. this is really close.

@@
-\subsection{Advancing LCA Practice}
+\subsection{Recommendations for Advancing LCA Practice}
+
+The most transformative aspect of the disclosure framework is its ability to describe the productive output of LCA modeling in a way that transcends differences in software environments and databases.
+%In current practice, there is no way to separate the PSM from the computational environment used to create it; however, the formalism allows a study to be put into terms that are software independent.
+%The construction and and computation of Eq.~\ref{eqn:study} is equivalent to the operation of all LCA software that employs static background databases.
+A product system model described in the disclosure framework can %thus
+be implemented in any software system, though the manner of doing so, as well as availability of advanced features for analysis and visualization, varies widely.
+
+To advance LCA practice, the community should embrace the notion of a software-independent PSM description and consider how to best achieve it.
+The SETAC North America roadmap module for PSM description and revision \citep{Kuczenski_JLCA_2018} includes milestones toward this goal.
+The disclosure framework presented here is offered as a candidate for Milestone 2.1, ``a minimal description of a PSM.''  The community should give this proposal a robust critical evaluation, and researchers should consider how to conceptualize their own models in terms of the framework.
+Other milestones and cross-cutting issues are of crucial interest:
+\begin{enumerate}[label=(\alph*)]
+\item LCI database providers must provide their users with unambiguous, static, resolvable references to background datasets (Milestone 1.1) so that their use in a model may be documented precisely (Milestone 3.1);
+\item The transformation of multi-functional processes to single-function processes should be described using the framework (Milestones 1.2 and 1.4);
+\item Software makers should enable their users to automatically generate lists of foreground, background, and emission flows a model foreground (disclosure items \ref{itm:fg} through \ref{itm:em}) and to export foreground matrices (disclosure items \ref{itm:af} through \ref{itm:bf}) from a model in a concise format (Milestones 2.2 and 3.4);
+\item During ISO critical review, practitioners and reviewers should strive to come to agreement on the formal structure and contents of the model being reviewed;
+\item The LCA research community, including interested parties within the International Society for Industrial Ecology, the UNEP/SETAC Life Cycle Initiative, and other organizations, should collectively pursue consistent, shared definitions of quantities of measure, flows, and contexts used in LCIA.
+%These definitions should be published and curated using Semantic Web techniques and practices.
+LCIA method developers should immediately begin using these consensus definitions instead of their own internal lists when publishing flow characterization data, so that software providers can interpret the publications correctly.
+%A characterization of a flow defined privately, into a context defined privately, cannot be interpreted meaningfully by the public.
+\end{enumerate}
+As these milestones are pursued and achieved, it will be possible for a standalone description of a PSM to be computed and extended by another party, regardless of the software used to construct the model originally.
+This will tremendously improve the capacity for critical review and reuse of LCA models.
 
 \endinput
+
+Reproducing a study described in the disclosure framework is equivalent to constructing and computing Eq.~\ref{eqn:study}, which I assert is universally found in LCA computations using static background databases.
+
+A PSM fully defined in concise textual terms can be shared and distributed without
+
+The disclosure framework forproduct system models can advance LCA practice by transcending differences in modeling across databases and software systems.
+
+can advance LCA practice by
+
+\endinput
+
+* Role of ISIE or SETAC group
+
+- unclear
+it seems like we would need buy-in before going any farther.
+What can people do?
+
+* Responsibility of database and LCIA method providers
+
+* Next steps-- part of standards?
+
+* Potential for distributed LCA computation
 
@@
  - we want to better
+
+ ---------
+
+ most important innovation here is separating bg from fg data
+ if in fact bg data are static, and become referenceable via the Internet,
+
+ Distributed LCA

### 2018-02-10 aggregation/agg-pagg.tex
tweak wording. this is really close.

@@
 \begin{align}
+  \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}\\
   s =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
-  =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}\\
-  \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}
+  =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}
 \end{align}

### 2018-02-10 aggregation/agg-semantic.tex
tweak wording. this is really close.

--- 2017-11-07 aggregation/agg-semantic.tex
@@
 
-Similarly, LCIA method providers should ensure that their methods make use of stable semantic references to elementary flows and contexts.
-Current research has revealed widespread challenges to finding consistency on the identities of elementary flows despite the community's longstanding awareness of the problem, and several mutually inconsistent reference flow sets exist \citep{Edelen_2017}.
+Similarly, %LCIA method providers should ensure that their methods make use of stable semantic references to elementary flows and contexts.
+elementary flows and contexts should be given stable references that are held in commomn across data providers.
+Current research has revealed widespread challenges to finding consistency on the identities of elementary flows despite the community's longstanding awareness of the problem \citep{Speck_2015,Herrmann_2015}, and several mutually inconsistent reference flow sets now exist \citep{Edelen_2017}.
 The conventional understanding of a ``flow'' as comprising a substance and a context together (e.g.

### 2018-02-10 aggregation/agg-community.tex
Revise Semantics bit to cite FAIR

@@
 %These definitions should be published and curated using Semantic Web techniques and practices.
-LCIA method developers should immediately begin using these consensus definitions instead of their own internal lists when publishing flow characterization data, so that software providers can interpret the publications correctly.
+LCIA method developers should immediately begin using these consensus definitions instead of their own internal lists when publishing flow characterization data, so that software makers can interpret the publications correctly.
 %A characterization of a flow defined privately, into a context defined privately, cannot be interpreted meaningfully by the public.

### 2018-02-10 aggregation/agg-semantic.tex
Revise Semantics bit to cite FAIR

@@
 
-In practice, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity
-in identifying the background datasets and emissions to which those matrix entries correspond.
-In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of those references.
-In particular, the database configuration, dataset version, reference flow and dimension (quantitative unit of measure) must be unambiguous and easy to interpret.
+In practice, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity in identifying the background datasets and emissions to which those matrix entries correspond.
+In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of the references contained in items~\ref{itm:bg} and \ref{itm:em}.
+The FAIR guidelines, which were designed to help scientists organize data for Findability, Accessibility, Interoperability, and Reusability \citep{Wilkinson_2016}, provide valuable perspective on this goal.
 
-Agreement about the meaning of concepts is often accomplished using linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.
-The resource at the end of the link functions as a point of agreement: both study author and data user can follow it from anywhere on the Internet to obtain the same information.
-The content pointed to by the hyperlink, along with context provided by \emph{other} references to the same content, gives meaning to the information and allows it to be curated by the study authors and other members of the community \citep{Khan_2011}.
+Central to many of the FAIR guiding principles is the use of linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.
+This link serves as a unique identifier and a resolvable reference to metadata, and the linked content can be  curated by the study authors and other members of the community \citep{Khan_2011}.
+
 %Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.
@@
 Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that denote a particular activity in a particular database version and configuration.
-These references should take the form of URIs that can be accessed using any Internet browser and can provide both author and reader with documentary information describing the referred dataset, the available reference flows and their dimensions, and a mechanism for the reader to obtain access to the exchange data or LCI/LCIA computations for the purposes of model validation.
+These references should take the form of URIs that can be accessed using any Internet browser and can provide both author and reader with documentary information describing the referred dataset, including version information; database configuration; the available reference flows and their dimensions (quantitative units of measure); and a mechanism for the reader to obtain access to the exchange data or LCI/LCIA computations for the purposes of model validation.
 A dependency reference in disclosure item~\ref{itm:bg} should include both an activity and a reference flow.

### 2018-02-12 aggregation/agg-community.tex
Tiny tweaks. I should really put this down for awhile.

@@
 \item LCI database providers must provide their users with unambiguous, static, resolvable references to background datasets (Milestone 1.1) so that their use in a model may be documented precisely (Milestone 3.1);
-\item The transformation of multi-functional processes to single-function processes should be described using the framework (Milestones 1.2 and 1.4);
+\item The transformation of multi-functional processes to single-function processes should be described using the framework (Milestones 1.2 and 1.4; see supporting information);
 \item Software makers should enable their users to automatically generate lists of foreground, background, and emission flows a model foreground (disclosure items \ref{itm:fg} through \ref{itm:em}) and to export foreground matrices (disclosure items \ref{itm:af} through \ref{itm:bf}) from a model in a concise format (Milestones 2.2 and 3.4);

### 2018-02-12 aggregation/agg-pagg.tex
Tiny tweaks. I should really put this down for awhile.

@@
   \end{equation}
-This is illustrated in Fig.~\ref{fig:private}b.
-Substituted back into Eq.~\ref{eqn:study}, the private portions are aggregated into a private life cycle inventory vector $\mathbf{b}_{priv}$, while the public portions remain disaggregated:
+This is illustrated in Fig.~\ref{fig:private}b.  Substituted back into Eq.~\ref{eqn:study}, the private portions are aggregated into a private life cycle inventory vector $\mathbf{b}_{priv}$:
+\begin{equation}
+  \mathbf{b}_{priv} = (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}
+\end{equation}
+while the public portions remain disaggregated:
 \begin{align}
-  \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}\\
   s =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\

### 2018-02-12 aggregation/agg-semantic.tex
Tiny tweaks. I should really put this down for awhile.

@@
 In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of the references contained in items~\ref{itm:bg} and \ref{itm:em}.
-The FAIR guidelines, which were designed to help scientists organize data for Findability, Accessibility, Interoperability, and Reusability \citep{Wilkinson_2016}, provide valuable perspective on this goal.
+The FAIR guiding principles, which were designed to help scientists organize data for Findability, Accessibility, Interoperability, and Reusability \citep{Wilkinson_2016}, provide valuable perspective on this goal.
 
-Central to many of the FAIR guiding principles is the use of linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.
+Central to many of the FAIR guidelines is the use of linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.
 This link serves as a unique identifier and a resolvable reference to metadata, and the linked content can be  curated by the study authors and other members of the community \citep{Khan_2011}.

### 2018-02-13 revision/rev2-response.tex
reviewer response

@@
 I also made an assortment of clarifying changes throughout the document.
+
+Finally, I added a section entitled ``Recommendations to Advance LCA Practice'' at the end of the discussion, to respond to reviewers \#2 and \#3's suggestions about the role of the community and next steps.
+This section has added about 400 words to the paper.
 
@@
 
-At the moment I neglected to include a new section because the paper is already quite long.  But maybe I will change my mind in the next 24 hours.
+As noted in the introduction, I added a new section at the end of the Discussion entitled ``Recommendations to advance LCA practice,'' which addresses this.
 
@@
 
-If I decide to do this, it will be included in the section that Reviewer \#2 wanted me to add at the end of the discussion.
+As noted in the introduction, I added a new section at the end of the Discussion entitled ``Recommendations to advance LCA practice,'' which addresses this.
 

### 2018-02-13 aggregation/agg-disclosure.tex
Bugs

@@
 A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.
-Thus, a formal disclosure of the PSM can be stated in six parts, including three lists of entities and three sets of data values:
+A formal disclosure of the PSM can be stated in six parts, including three lists of entities and three sets of data values:
 \begin{enumerate}[label={\em d-\roman*}., ref={\em d-\roman*}]
@@
 In both items~\ref{itm:bg} and~\ref{itm:em}, only rows containing non-zero entries need to be mentioned, since the index of each row is common information.
-The contents of a disclosure are represented graphically in Fig.~\ref{fig:disclosure}., and two example disclosures drawn from reference databases can be found in the supporting information .
+The contents of a disclosure are represented graphically in Fig.~\ref{fig:disclosure}, and two example disclosures drawn from reference databases can be found in the supporting information.
 

### 2018-04-23 final-revision/final-response.tex
Final revision comments and response

\documentclass[11pt,letterpaper]{article}
\input{formatting}

\title{Disclosure of Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy}
\author{Brandon Kuczenski\\
Institute for Social, Behavioral, and Economic Research\\
University of California, Santa Barbara}
\date{\mydate\today}

\begin{document}
\maketitle

\def\rquote#1{\quote{\textit{#1}}}

\section*{Response to Reviewers}

Many thanks again to the editor and reviewers for providing me with constructive feedback to improve the submission.  In this revision I have added a thorough discussion regarding the equivalency of process-based versus input-output-based LCA computations to the Supporting Information.  This discussion leans heavily on the paper by Suh et al (2010) entitled ``Generalized Make and Use Framework for Allocation in Life Cycle Assessment.''  I concur with the authors of that paper when they state that the two major approaches for handling multi-functionality in LCA are ``equivalent'' to two major constructs developed in input-output analysis.  The authors further state, quite authoritatively, that ``the two data types can be hybridized without the loss of methodological consistency'' and that the make-and-use formulation presents ``a coherent mathematical framework'' for solving the general LCI problem.

Hopefully that article, upon review, and accompanied by my exposition in the Supporting Information, will put to rest the lingering concerns of Reviewer \#1 that the selection of the input-output formalism somehow constrains the applicability of my work, or that it somehow introduces inconsistency.  I have made a number of minor edits to the main text to support and clarify this argument.  I have also removed a footnote that previously provided a perfunctory explanation.

I have also added (to the supporting information) a table of mathematical symbols used in the paper for reference.  If it is the editor's preference, I would be happy for this table to be reproduced in the main body as well.

Finally, I have substituted the term ``originality'' for the term ``primacy'' from the prior draft, based on external feedback.

Below I add a few additional thoughts in response to the reviewers' specific comments.

\subsection*{Reviewer \#1}

\emph{I do regret that you choose for keeping the "I-A" structure. While I do understand your point of departure, the supply-use tables and the constructs derived from that, the final step to the IO-format is unfortunate and inconssistent.}\par
\emph{  a) It is unfortunate because the standard computational stucture of LCA is process-based using A\^{}-1, and not IO-based using (I-A)\^{}-1. Just read SETAC and ISO: classical LCA is process-based. This is not only in the Heijungs \& Suh book, but also in many subsequent articles. Of course there are articles on the basis of IO-based (I-A)\^{}-1, but this is a minority, and almost never a reference to classical LCA.}

As I discuss above, the two approaches are equivalent.  The simplest way to see this is to simply \textit{reinterpret} the Leontief expression $(I-A)$ to be a formulation of a technology matrix, in which the columns have been normalized to a unit output.  More importantly, reading the ISO specification it is clear that there is no preference placed upon one mathematical formulation over another.  Indeed, the only salient characteristic of a ``unit process'' in Section 3.34 of ISO 14044 (2006) is that it has both inputs and outputs, as certainly do all entities in the supply-and-use framework.

\emph{b) It is inconsistent because, indeed as is written on page 6 line 8, the IO-framework is either "commodity-by-commodity or industry-by-industry". But your system  is commodity-by-process. This is for instance clear from page 9 lines 12-26, where columns are processes and rows are commodities.}

The comments regarding inconsistency are more relevant and have led me to make clarifying changes to the text. To the point, the $A$ matrix must be symmetric to be invertible, and in the case where every process has a single output, the distinction between processes and commodities is really only semantic.  In the present text, it is more accurate to say the technology matrix is allocated-commodity by allocated-commodity.  I have made a few changes on page 12 and throughout the text to highlight this distinction.

\emph{Indeed, classical/ISO/SETAC/process-based LCA uses A\^{}-1, not (I-A)\^{}-1, and its format is commodity-by-process. In the first version version I commenented on an inconsistent treatment, in the second version you chose for the wrong repair, and now you seem to go even deeper in this wrong repair. I simply can't understand this choice, and it creates a major limitation for the usefulness and acceptance of the rest of the paper.}

The ISO standards make no reference to Heijungs' formulation.  Please do read the new section in the Supplementary Information that addresses this.

\emph{I am still puzzled about the canonical functional unit. Although it is clearer now, I am in particular concerned if (and how) your approach would work if the functional unit of a study is not 1 bread or 1 kWh electricity, but 1000 breads of 100000 kWh electricity. Please discuss and demonstrate that you can do this. Otherwise, your approach will again be useless.}

I have added a footnote to the paragraph in which the canonical functional unit is defined, reading thus:
\begin{quote}Of course, the `functional unit' does not need to be limited to a unit of any given output---instead, it \textit{defines} a unit of output. For instance, if the functional unit of a study is 1,000 loaves of bread, and one loaf of bread weighs 0.454 kg, and the unit output from column 567 of the A matrix is 1 kg of bread, then the canonical functional unit would be realized by entering \texttt{454} in column 1, row 567, leaving the rest of that column zero.
\end{quote}

\emph{I would highly appreciate a list of symbols. This list could contain the symbol, a description if its meaning, and possibly its size(s), e.g. products*processes or so.}

This has been added to the supplementary material.

\emph{There are still several places where the notation is not consistent. }

Thank you for highlighting these errors.  My failure to properly use a bold 0 and transpose symbols around the 0-vector in the SI has been corrected.  Note, however, that in seeking consistency with the rest of the notation, the bold zero and bold zero transpose notation were only used for zero \emph{vectors} and not for zero \emph{sub-matrices}.

\subsection*{Reviewer \#2}

\emph{1) The examples provided for multi-output foreground processes in Section 3.2 in the SI are very instructive, however, their description lacks conciseness. There are a number of phrases like ‘arbitrarily-specific allocated inventories’, ‘explicitly allocated’, ‘canonical functional unit’, etc., that should be better explained or avoided. SI section 3.2 would benefit from more concise language and a few more descriptions.}

Thank you for this observation.  I have attempted to clarify the text in Section 5.2 (formerly section 3.2) of the supplementary information.  Note, however, that the notion of ``arbitrarily-specified'' allocated processes is drawn from Majeau-Bettez et al (2014) who elucidated it.  Also the canonical functional unit is a term defined in the main text.

\emph{2) In the response letter, I saw that reviewer 1 emphasized the difference between the mathematical framework for process and IO-based LCA, referring to Suh and Heijung’s A\^{}-1 and Leontief’s (I-A)\^{}-1. (note that the two A’s are different matrices with a different meaning). I am not aware of a reference that shows that both approaches are equivalent, maybe it can be shown in the SI, while at the same time explaining to the reader that the A matrix used here is not the S\&H A matrix but the Leontief A-matrix. There seems to be a gap here both in the literature and in general understanding, but filling it may be too much for this paper.}

Please review the new section in the Supplementary Information. I am of the opinion that the equivalence of the two models was shown convincingly by Suh et al (2010).

I have added a clarification that the $A$ matrix introduced in Eq.1 is the Leontief matrix, and also this has been noted in the Table of Symbols.

\end{document}

### 2018-04-23 aggregation/agg-agg.tex
Reviewer response
Late nite work
Revision 3

@@
 
-\subsubsection{Unit Process Inventory}
+\subsubsection{Single-Output Unit Process}
 
-A unit process makes up a single shared column of $A_f$, $A_d$, and $B_f$, as depicted in Figure~\ref{fig:aggregation}a.
+A single-output unit process makes up a single shared column of $A_f$, $A_d$, and $B_f$, as depicted in Figure~\ref{fig:aggregation}a.
 Documentation of unit processes is specified in ISO 14048, and several compliant formats exist, the most well-known being the ILCD and Ecospold XML formats.
@@
 In a matrix representation, each unit process must have exactly one reference flow, which corresponds to its row and column in $A_f$.
+Please see the supporting information for more detail on preparing the $A_f$ matrix.
 
@@
 For examples of multi-output processes represented as foreground models, please see the supporting information.
-The example shown in Figure~\ref{fig:aggregation}b shows five nodes, including one cut-off flow
+The example shown in Figure~\ref{fig:aggregation}b shows five nodes, including one cut-off flow.
 
@@
  \tilde{\mathbf{a}}_d = \tilde{\mathbf{a}}_{d,priv} + \tilde{\mathbf{a}}_d'
+\label{eqn:partial}
 \end{equation}

### 2018-04-23 aggregation/agg-basic-lca.tex
Revision 3

@@
 Such an account can always be transformed into a \textit{symmetric} table of inputs and outputs on a commodity-by-commodity or industry-by-industry basis \citep{Eurostat_2008}.
-Doing so requires the application of various assumptions and transformations to the data, the details of which are not in scope of this paper, but which have recently been reviewed in depth \citep{Suh_JIE_2010, Majeau_Bettez_2014}.
+Doing so requires the application of various assumptions and transformations to the data, the details of which are not in scope of this paper, but which have recently been reviewed in depth \citep{Majeau_Bettez_2014}.
+Although they originated independently, the ``classical'' process-based approach and the input-output approach have been shown to be equivalent for the purposes of LCA computation \citep{Suh_JIE_2010}.
+Please see the Supporting information for a discussion about this equivalency.
 
-A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products), equivalent to a ``unit process'' in ISO LCA, except that the reference flow is implicit in the column's index.
+A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products), equivalent to a ``unit process'' in ISO LCA, except that the process must have exactly one reference flow, the identity of which is implicit in the column's index.
 The table of coefficients, commonly known as a \textit{direct requirements matrix}, can be used to compute the output from all activities that is necessary to produce a given commodity.
@@
 \end{equation}
-Here $A$ represents the direct requirements; $B$ is the emission matrix;  $\mathbf{c}$ is a column vector of characterization factors for the environmental emissions; $\mathbf{y}$ is the externally specified final demand; and $s$ is the numerical impact score or category indicator.
+Here $A$ represents the direct requirements, sometimes known as the \textit{Leontief matrix}; $B$ is the emission matrix;  $\mathbf{c}$ is a column vector of characterization factors for the environmental emissions; $\mathbf{y}$ is the externally specified final demand; and $s$ is the numerical impact score or category indicator.
 For simplicity, we will consider the computation of a single impact result.
@@
 
-The term $(I-A)$ is equivalent to a normalized \textit{technology matrix}, and in certain cases can be computed from the difference between the supply and use tables, $(V'-U)$, with equivalent results \citep{Suh_JIE_2010}.
-The coefficient matrices $A$ (or $I-A$) and $B$ together are called a \textit{life cycle inventory database} (LCIDB), and the functional unit of a study is given by the final demand vector $\mathbf{y}$, which is the product or service whose delivery is to be assessed.\footnote{%
-Differences in convention have led to both the direct requirements matrix $A$, and the technology matrix, analogous to $(I-A)$, being denoted by the same symbol in different contexts.
+The term $(I-A)$ is equivalent to a \textit{technology matrix} in traditional LCA, normalized by column so that each column represents a unit output of a single reference product, and each row represents the use of the corresponding product.
+The coefficient matrices $A$ (or $I-A$) and $B$ together are called a \textit{life cycle inventory database} (LCIDB), and the functional unit of a study is given by the final demand vector $\mathbf{y}$, which is the product or service whose delivery is to be assessed.
+%%\footnote{Differences in convention have led to both the direct requirements matrix $A$, and the technology matrix, analogous to $(I-A)$, being denoted by the same symbol in different contexts.
 If the data contained in the LCIDB is sufficiently rich, either approach may be implemented with equivalent computational results.
-The direct requirement formulation has the advantage of permitting iterative solution, which can dramatically reduce the computational resources required \citep{Peters_JLCA_2007}.}  Intermediate results of this computation include:
+The direct requirement formulation has the advantage of permitting iterative solution, which can dramatically reduce the computational resources required \citep{Peters_JLCA_2007}.}
+Intermediate results of this computation include:
 \begin{itemize}
@@
 However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are not usually modified by study authors.
-Thus exporting $A$ and $B$ does not meet our authority and primacy objectives.
+Thus exporting $A$ and $B$ does not meet our authority and originality objectives.
 Additionally, replicating the entire LCIDB is cumbersome because of the large size of the matrices.

### 2018-04-23 aggregation/agg-disclosure.tex
Revision 3

@@
   \item The disclosure includes only information for which the author can claim \textbf{authority};
-  \item The disclosure includes all information specific to the study, or for which the study is the \textbf{primary} source, including re-implementations of already-published data. 
+  \item The disclosure includes all information specific to the study, or for which the study is the \textbf{original} source, including re-implementations of already-published data. 
 \end{itemize}

### 2018-04-23 aggregation/agg-lca-study.tex
Revision 3

@@
 A process for brewing coffee could be created in a third column that requires coffee beans, water, and electricity.
-Finally, a fourth column could be added that invokes the first three columns in proportion to describe the activity of traveling to a coffeeshop to work.%\footnote{No prizes for guessing where this passage was written.}.
+Finally, a fourth column could be added that invokes the products from the first three columns in proportion to describe the reference product of traveling to a coffeeshop to work.%\footnote{No prizes for guessing where this passage was written.}.
 
@@
 Without loss of generality, it is possible to construct the study such that the \textit{first} column of $\tilde{A}$ delivers the functional unit of the study.
-We thereby designate the vector $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ as the \textit{canonical functional unit} of a foreground study, which allows us to introduce convenient notation later on.
+We thereby designate the vector $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ as the \textit{canonical functional unit} of a foreground study, which allows us to introduce convenient notation later on.\footnote{Of course, the ``functional unit'' does not need to be limited to a unit of any given output---instead, it \textit{defines} a unit of output.
+For instance, if the functional unit of a study is 1,000 loaves of bread, and one loaf of bread weighs 0.454 kg, and the unit output from column 567 of the A matrix is 1 kg of bread, then the canonical functional unit would be realized by entering \texttt{454} in column 1, row 567, leaving the rest of that column zero.}
 

### 2018-04-23 aggregation/agg-multioutput.tex
Revision 3

@@
 According to MWS14, Partitioning allocation (PA) is a ``production-balanced model that artificially splits the requirements of an activity across its different outputs.''  The exact manner of the partitioning need not be specified, as long as the parts sum up to the whole.
-In other words, for a process with $n$ outputs the modeler has $n-1$ degrees of freedom in specifying the allocated inventories, with the requirement that the $n$th inventory equals the remainder.
+In the disclosure framework, any manner of production-balanced partitioning can be represented by including the raw, un-allocated inventory as one foreground column, choosing any output as the reference.
+The other $n-1$ outputs are then included as allocated inventories in additional columns, for a total of $n$ columns.
+Each column must be normalized to a unit reference.
 
-In the disclosure framework, any manner of production-balanced partitioning can be represented by including the raw, un-allocated inventory as one foreground column, choosing any output as the reference.
-The other $n-1$ outputs are then included as arbitrarily-specified allocated inventories in additional columns, for a total of $n$ columns.
-Each column must be normalized to a unit reference.
-The reference flow not explicitly allocated will simply be computed as the remainder when the allocated inventories are subtracted from the un-allocated inventory.
+Note that, although it is customary for allocated process inventories to be derived as a uniform partition of the un-allocated process based on some parameter such as mass, price or energy content, this is not necessary.
+For a process with $n$ outputs the modeler has $n-1$ degrees of freedom in specifying the allocated inventories, with the requirement that the $n$th inventory equals the remainder.
+In the framework presented, the modeler must provide the $n-1$ allocated inventories explicitly.
+The reference flow for the $n$th process will simply be computed as the remainder when the allocated inventories are subtracted from the un-allocated inventory.
 
@@
 
-Using the foreground in Table~\ref{tbl:partition}, LCA computation using the canonical functional unit will automatically compute the self-consistent single-output result for 1 kg of chlorine gas because the allocated inventories for hydrogen and sodium hydroxide will be subtracted from the total.
+Using the foreground in Table~\ref{tbl:partition}, LCA computation using the canonical functional unit (see main text, Eq.~\ref{eqn:foreground}) will automatically compute the self-consistent single-output result for 1 kg of chlorine gas because the allocated inventories for hydrogen and sodium hydroxide will be subtracted from the total.
 The foreground activity levels for the system are:
@@
 
-In the classical system expansion case, applying the expanded functional unit of 1 kg chlorine, plus 0.028 kg hydrogen and 1.13 kg of sodium hydroxide, will result in a unit activity of the reference process, without any cut-off flows..
+In the classical system expansion case, applying the expanded functional unit of 1 kg chlorine, plus 0.028 kg hydrogen and 1.13 kg of sodium hydroxide, will result in a unit activity of the reference process, without any cut-off flows.
 Equivalently, the canonical functional unit, production of 1 kg of chlorine gas, will result in the same process activity level, and will also generate the co-products as cutoff flows.

### 2018-04-23 aggregation/agg-requirements.tex
Revision 3

@@
 
-\subsection{Authority and Primacy}
+\subsection{Authority and Originality}
 
@@
 
-Because of these conditions, in the vast majority of cases the LCA study disclosure must include the author's reimplementations in order to achieve both the aims for authority and primacy.
+Because of these conditions, in the vast majority of cases the LCA study disclosure must include the author's reimplementations in order to achieve both the aims for authority and originality.
 In so doing, the disclosure enables a critical reviewer to evaluate whether the author's implementations are generally correct and complete.

### 2018-04-23 aggregation/equivalence.tex
Revision 3

\section{On the Equivalence of Process and Input-Output LCA}

There is a level of confusion about the differing implications of process-based LCA inventory analysis as described in Heijungs and Suh (2002), and input-output analysis using a Leontief matrix.  In fact, these two techniques were shown to be equivalent in Suh et al, 2010\footnote{\doi{10.1111/j.1530-9290.2010.00235.x}; see full text for citation.}, whose abstract declares: ``the article shows that the partitioning method in LCA is equivalent to the industry-technology model in input-output economics, and system expansion in LCA is equivalent to the by-product-technology model in input-output output economics.''  The article does indeed show convincingly that ``the two data types can be hybridized without the loss of methodological consistency'' (p.345).

   Both the technology matrix $A$ in LCA and the Leontief matrix $(I-A)$ in input output analysis are ways of encoding the relationships between the products that some processes ``make'' and other processes ``use.''  In either the classical or input-output formulation, both matrices must be put into symmetric form to be invertible, and this means describing each column as having a single, reference output.  This is done either through disaggregation (i.e. dividing a process by its various outputs), aggregation (i.e. combining products to form a market or industry), or some combination of these.  While the task of transforming data is highly case-dependent, it is not constrained by the choice of mathematical formalism.

   Suh et al, after a rich investigation of the literature, a thorough mathematical development, and a worked example, conclude that ``even very large-scale LCI problems, such as those in commercial LCI databases, can be computed with a consistent mathematical framework$\ldots$'' that being the make-and-use framework (p. 349).  In effect, the different mathematical approaches can be made equivalent.  The equivalency can be made very plain by simply reinterpreting the input-output expressions \textit{as} technology matrices.  In other words, (in Suh et al) the expressions $(I-A_I)$ (eq. 6), $(I-A_C)$ (eq. 9), $(I-A_B)$ (eq. 12), and $(V'-U)$ (eq. 17) can simply be interpreted as \textit{different ways to formulate} the process-based technology matrix.  There is no inconsistency between this interpretation and the classical formulation.

   Once single-output processes are obtained, one substantive difference between the various approaches is column scaling: whether the entries describe real outputs, unit outputs, or unit activity levels.  The database creator must ensure that a consistent scaling is applied to both the technology matrix and the emission matrix $B$.  The convention in LCA is for a column to represent a unit activity level.  To use the input-output coefficient approach, as in the present article, the columns must be scaled to a \textit{unit of the reference product}.  This reference output is then omitted from the $A$ matrix, and is instead supplied by the $I$ in the expression $(I-A)$.  This formulation is completely consistent with any other matrix-based formulation of LCA, including classical process LCA as traditionally practiced.  In particular, the approaches described here apply equally well to LCA models using both process-based databases such as Ecoinvent, input-output databases such as Exiobase, and background databases such as the GaBi databases.

   Interested readers should note that the canonical reference for LCA, ISO 14044, has nothing to say about the mathematical techniques used to solve LCA problems, and the only preference for a technology matrix over an input-output matrix is borne of tradition or custom rather than theory.  SimaPro, the leading worldwide software system for scientific LCA, converted to using an iterative Leontief inverse for Version 8, as disclosed in a whitepaper published in 2014\footnote{\url{https://www.pre-sustainability.com/news/new-calculation-engine-simapro-8}}, realizing significant speed and memory usage improvements.  OpenLCA, the open-source alternative, also reported a dramatic reduction in memory usage and computation time around the release of version 1.4 in 2014, coincident with a revised algorithm ``optimized for huge databases.''\footnote{\url{http://www.openlca.org/openlca-beta-1-4-released-optimised-for-huge-databases/}} This is highly suggestive of a switch to an iterative approach.  GaBi, the more industrially oriented software, does not use a matrix formulation at all, but has always computed activity levels using iteration at the plan level.

### 2018-04-23 aggregation/proofs.tex
Revision 3

@@
 Given a life cycle inventory database (LCIDB) containing input-output matrix $A$ and emission matrix $B$, and a final demand vector $\mathbf{y}$, construct an augmented inventory database
-$\tilde{A} = \left[\begin{smallmatrix} 0 & \mathbf{0} \\ \mathbf{y} & A \end{smallmatrix}\right]$,
+$\tilde{A} = \left[\begin{smallmatrix} 0 & \mathbf{0}^T \\ \mathbf{y} & A \end{smallmatrix}\right]$,
 $\tilde{B} = [ \mathbf{0},\, B ]$, and canonical functional unit $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.  Show that
@@
   I &=&  (I-\tilde{A})^{-1}\times(I-\tilde{A})\\
-  \left[\begin{array}{cc} 1 & \mathbf{0} \\
+  \left[\begin{array}{cc} 1 & \mathbf{0}^T \\
     \mathbf{0} & I
   \end{array}\right] &=&
-  \left[\begin{array}{cc} 1 & \mathbf{0} \\
+  \left[\begin{array}{cc} 1 & \mathbf{0}^T \\
       \mathbf{q} & (I-A)^{-1}
   \end{array}\right]\times
-  \left[\begin{array}{cc} 1 & \mathbf{0} \\
+  \left[\begin{array}{cc} 1 & \mathbf{0}^T \\
       -\mathbf{y} & (I-A)
@@
 \end{eqnarray}
-It can be seen that $\mathbf{q}-(I-A)^{-1}\times\mathbf{y}=0$ is a necessary condition to satisfy the equality, and thus $\mathbf{q}=(I-A)^{-1}\times\mathbf{y}$.
+It can be seen that $\mathbf{q}-(I-A)^{-1}\times\mathbf{y}=\mathbf{0}$ is a necessary condition to satisfy the equality, and thus $\mathbf{q}=(I-A)^{-1}\times\mathbf{y}$.
 

### 2018-04-23 aggregation/symbols.tex
Revision 3

\section{List of Mathematical Symbols}

The symbols  used in the text are defined in Table~\ref{tbl:symbols}.

\begin{table}[p]
  \caption{List of Mathematical Symbols}
  \label{tbl:symbols}
\small\sffamily
  \def\gap{0.85ex}
\begin{tabularx}{\textwidth}{cclXl}
  \toprule
  \textbf{Symbol} & \textbf{Formula} & \textbf{Size} & \textbf{Description} & \textbf{ref.} \\
  \midrule
  \multicolumn{5}{l}{Dimensions} \\[\gap]
  $m$ &  & scalar & Number of emissions & Eq.~\ref{eqn:leontief} \\
  $n$ &  & scalar & Number of process outputs in background database & Eq.~\ref{eqn:leontief} \\
  $p$ &  & scalar & Number of nodes in foreground model & Fig.~\ref{fig:foreground}\\
  $\phantom{'}p'$ & & scalar & Number of cut-off nodes in foreground & Fig.~\ref{fig:aggregation}c \\
  $N$ & $n+p$ & scalar & size of augmented matrix & n.a. \\[\gap]
  \multicolumn{5}{l}{Classical LCA} \\[\gap]
  $A$ & & $n\oldtimes n$ & Direct requirements matrix (Leontief) & Eq.~\ref{eqn:leontief} \\
  $I-A$ & & $n\oldtimes n$ & Technology matrix & " \\%Eq.~\ref{eqn:leontief} \\
  $B$ & & $m\oldtimes n$ & Emission matrix & " \\%Eq.~\ref{eqn:leontief} \\
  $\mathbf{y}$ & & $n \oldtimes 1$ & Final demand vector & " \\%Eq.~\ref{eqn:leontief} \\
  $\mathbf{x}$ & $(I-A)^{-1}\times \mathbf{y}$ & $m \oldtimes 1$ & Activity level vector & Fig.~\ref{fig:lca} \\
  $\mathbf{b}$ & $B\times \mathbf{x}$ & $m \oldtimes 1$ & Life cycle inventory & " \\%Fig.~\ref{fig:lca} \\
  $\mathbf{c}$ & & $m \oldtimes 1$ & Characterization vector & Eq.~\ref{eqn:leontief} \\
  $s$ & $\mathbf{c}^T\times\mathbf{b}$ & scalar &  LCIA Result & " \\[\gap]%Eq.~\ref{eqn:leontief} \\[\gap]
  \multicolumn{5}{l}{Product System Model} \\[\gap]
  $\tilde{\mathbf{y}}$ & $[ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ & $N\oldtimes 1$ & Canonical functional unit vector & Eq.~\ref{eqn:canonical} \\
  $\tilde{A}$ & & $N\oldtimes N$ &  Augmented $A$ matrix & Eq.~\ref{eqn:foreground} \\ 
  $\tilde{B}$ & & $m\oldtimes N$ &  Augmented $B$ matrix & " \\%Eq.~\ref{eqn:foreground} \\
  $A_f$ & & $p\oldtimes p$ &  Foreground matrix & " \\%Eq.~\ref{eqn:foreground} \\ 
  $A_d$ & & $n\oldtimes p$ &  Background dependency matrix & " \\%Eq.~\ref{eqn:foreground} \\ 
  $B_f$ & & $m\oldtimes p$ &  Foreground emission matrix & " \\%Eq.~\ref{eqn:foreground} \\
  $B_x$ & $B\times(I-A)^{-1}$ & $m\oldtimes n$ &  Background LCI database matrix & Eq.~\ref{eqn:study} \\
  $\tilde{\mathbf{y}}_f$ & $[ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ & $p\oldtimes 1$ & Canonical foreground functional unit & " \\%Eq.~\ref{eqn:study} \\
  $\tilde{\mathbf{x}}$ & $(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f$ & $p\oldtimes 1$ & Foreground activity vector & Fig.~\ref{fig:foreground} \\[\gap]
  \multicolumn{5}{l}{Foreground Aggregation} \\[\gap]
  $\tilde{\mathbf{b}}_f$ & $B_f\times\tilde{\mathbf{x}}$ & $m\oldtimes 1$ & Aggregated foreground emissions & Eq.~\ref{eqn:agg}\\
  $\tilde{\mathbf{a}}_d$ & $A_d\times\tilde{\mathbf{x}}$ & $n\oldtimes 1$ & Aggregated dependencies & " \\%Eq.~\ref{eqn:agg}\\
  $\tilde{\mathbf{b}}_x$ & $B_x\times\tilde{\mathbf{a}}_d$ & $m\oldtimes 1$ & Aggregated background & " \\%Eq.~\ref{eqn:agg}\\
  $\tilde{\mathbf{a}}_f$ & & $p'\oldtimes 1$ & Vector of foreground cut-off flows & Fig.~\ref{fig:aggregation}c \\
  $\tilde{\mathbf{b}}$ & $\tilde{\mathbf{b}}_x+\tilde{\mathbf{b}}_f$ & $m\oldtimes 1$ & Aggregated life cycle inventory & Fig.~\ref{fig:aggregation}f\\[\gap]
  \multicolumn{5}{l}{Private Aggregation} \\[\gap]
  $\tilde{\mathbf{a}}_{d,priv}$ & & $n\oldtimes 1$ & Private dependencies & Eq.~\ref{eqn:partial} \\
  $\tilde{\mathbf{a}}_d'$ & $\tilde{\mathbf{a}}_d-\tilde{\mathbf{a}}_{d,priv}$& $n\oldtimes 1$ & Disclosed dependencies & " \\%Eq.~\ref{eqn:partial} \\
  $A_{d,priv}$ & & $n\oldtimes p$ & Private dependency matrix & Eq.~\ref{eqn:partition}\\
  $A_{d,pub}$ & $A_d - A_{d,priv}$ & $n\oldtimes p$ & Public dependency matrix & " \\%Eq.~\ref{eqn:partition}\\
  $B_{b,priv}$ & & $m\oldtimes p$ & Private foreground emission matrix & " \\%Eq.~\ref{eqn:partition}\\
  $B_{f,pub}$ & $B_f - B_{f,priv}$& $m\oldtimes p$ & Public foreground emission matrix & " \\%Eq.~\ref{eqn:partition}\\
  $\mathbf{b}_{priv}$ & $(B_{f,priv} + B_x\cdot A_{d,priv})\cdot\tilde{\mathbf{x}}$ & $ m\oldtimes 1$ & Private life cycle inventory & Eq.~\ref{eqn:bpriv}\\
  $\varphi$ & Eq.~\ref{eqn:metric} & scalar & Disclosure completeness metric & Eq.~\ref{eqn:metric}\\[\gap]
  \bottomrule
\end{tabularx}
\end{table}

### 2018-05-22 aggregation/agg-basic-lca.tex
make minor copyedits

@@
 
-A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products), equivalent to a ``unit process'' in ISO LCA, except that the process must have exactly one reference flow, the identity of which is implicit in the column's index.
+A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products).
+Each column is equivalent to a ``unit process'' in ISO LCA, except that the process must have exactly one reference flow, the identity of which is implicit in the column's index.
 The table of coefficients, commonly known as a \textit{direct requirements matrix}, can be used to compute the output from all activities that is necessary to produce a given commodity.

### 2018-05-22 aggregation/agg-community.tex
make minor copyedits

@@
 \item The transformation of multi-functional processes to single-function processes should be described using the framework (Milestones 1.2 and 1.4; see supporting information);
-\item Software makers should enable their users to automatically generate lists of foreground, background, and emission flows a model foreground (disclosure items \ref{itm:fg} through \ref{itm:em}) and to export foreground matrices (disclosure items \ref{itm:af} through \ref{itm:bf}) from a model in a concise format (Milestones 2.2 and 3.4);
+\item Software makers should enable their users to automatically generate lists of foreground, background, and emission flows in a model foreground (disclosure items \ref{itm:fg} through \ref{itm:em}) and to export foreground matrices (disclosure items \ref{itm:af} through \ref{itm:bf}) from a model in a concise format (Milestones 2.2 and 3.4);
 \item During ISO critical review, practitioners and reviewers should strive to come to agreement on the formal structure and contents of the model being reviewed;

### 2018-05-22 aggregation/agg-review.tex
make minor copyedits

@@
 
-Together, these lists can be considered a functional definition of the study scope that can be used to conduct a qualitative comparison of multiple studies by different authors.
+Together, these lists can be considered a \emph{functional definition of the study scope} that can be used to conduct a qualitative comparison of multiple studies by different authors.
 Background database choice, dataset selection, and version information can all be made available in disclosure item~\ref{itm:bg}, thereby streamlining what is presently an interpretive and study-specific process of extracting this information from written text.
@@
 If a study containing confidential information is split into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~\ref{fig:private}, then the results of the computation can be validated directly by the reviewer.
-To achieve this, the reviewer must be able to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly.
+To achieve this, the reviewer must be able to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly (Eq.~\ref{eqn:bpriv}).
 This requires that the reviewer be granted access to both the public and private partitions.

### 2018-05-22 post-accept/kuczenski-disclosure.tex
Post-accept wrangling (TeX native\!)

\documentclass[12pt]{article}

%% formatting
\pagestyle{plain}
\usepackage[margin=1in]{geometry}

%% font selection
\usepackage{mathptmx}
\renewcommand\sfdefault{phv}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{textcomp}

\usepackage{setspace}
\usepackage{textcase}
\usepackage{amsmath,amssymb,mathrsfs}
\usepackage{booktabs}
\usepackage{array}

\usepackage{enumitem}
\setlist[enumerate,2]{label=(\alph*)}
\setlist[enumerate,3]{label=\roman*.}

%% pdf foo
\usepackage[svgnames]{xcolor}

\definecolor{hyperblue}{rgb}{0.01 0.04 0.45}
\usepackage{thumbpdf}

\usepackage[hyphens]{url}

\RequirePackage[colorlinks,breaklinks=true,hyperindex,plainpages=false,%
linkcolor=hyperblue,urlcolor=hyperblue,anchorcolor=hyperblue,%
citecolor=black,%
pdfstartview=FitH,pdfpagemode=UseOutlines,%
bookmarksopen=false]{hyperref}
\def\pdfBorderAttrs{/Border [0 0 0] } % No border arround Links
\setlength{\pdfpagewidth}{\paperwidth}
\setlength{\pdfpageheight}{\paperheight}
\usepackage{breakurl}
\urlstyle{same}

%% Appearance

\usepackage{parskip}
\setlength{\parindent}{0pt}

\newcommand{\doi}[1]{\mbox{doi:~\href{http://dx.doi.org/#1}{\small\texttt{#1}}}}
\newcommand{\email}[1]{\texttt{\href{mailto:#1}{#1}}}
\newcommand{\overbar}[1]{\mkern
  1.5mu\overline{\mkern-1.5mu#1\mkern-1.5mu}\mkern 1.5mu}

\let\oldtimes\times
\let\times\cdot

%\lhead{\sffamily Disclosure of product system models}
%\rhead{\sffamily Kuczenski -- 2017}

\usepackage{tabularx}
\usepackage[round]{natbib}
\usepackage{siunitx}

%% To build supporting information:
%% - uncomment supportingfalse
%% - in the document uncomment both mainpaper and supporting
%% - latex agg-JIE
%% - in the document [bottom] inside \ifsupporting, comment mainpaper
%% - maketex -x -o supp.pdf agg-JIE

\let\oldsection\section
\makeatletter
\renewcommand\section{\@startsection {section}{1}{\z@}%
                                   {-3.5ex \@plus -1ex \@minus -.2ex}%
                                   {2.3ex \@plus.2ex}%
                                   {\normalfont\Large\bfseries <Level 1> }*}
\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
                                     {-3.25ex\@plus -1ex \@minus -.2ex}%
                                     {1.5ex \@plus .2ex}%
                                     {\normalfont\large\bfseries <Level 2> }*}
\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
                                     {-3.25ex\@plus -1ex \@minus -.2ex}%
                                     {1.5ex \@plus .2ex}%
                                     {\normalfont\normalsize\bfseries <Level 3> }*}

\makeatother

\usepackage{endnotes}
\let\footnote=\endnote

\bibliographystyle{authordate_bk}

\title{Disclosure of Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy}
\author{Brandon Kuczenski\\
Institute for Social, Behavioral, and Economic Research\\
University of California, Santa Barbara}
\date{Submitted after revision, 23 Apr 2018}

\usepackage{lineno}

\def\mainpaper{

  \maketitle
  %\emph{Submitted after revision to the Journal of Industrial Ecology; April 2018}\\
  \emph{Accepted for publication: 16 May 2018}

  \begin{abstract}
  Many of the challenges facing knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of the product system models used to perform LCA computations.
  This article presents a framework for formally disclosing the foreground of an LCA study in a way that permits the computations to be inspected, verified, and reproduced by a reader, provided that the reader has access to the same life cycle inventory and impact characterization resources as the author.
  The framework can also be used to partition a study into public and private portions, allowing both portions to be critically reviewed but omitting the private information from the disclosure.
  A disclosure is made up of six components, including three lists of entities in the model and three sparse matrices describing their interconnections.  The entity lists make reference to previously-published resources, including background inventory databases and characterized elementary flows, and the disclosure framework requires both author and reader to agree on the meaning of each of these references.
  The framework contributes to ongoing efforts within and beyond industrial ecology to improve the reproducibility and verifiability of scholarly works, and if implemented, plots a course toward distributed, platform-independent computation and validation of LCA results.
\end{abstract}

  \doublespacing
  %\linenumbers

\section{Introduction}

Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver 
them to a consumer.  The technique is well-established, widely practiced, and draws on an extensive body of standardization and scholarship.
Increasingly, the results of LCA studies are appearing in environmental declarations, corporate sustainability reports, and product marketing information.  LCA is also gaining prominence as a tool for developing and evaluating environmental policy.  
%
At the core of an LCA study is a model of a ``product system,'' which represents a collection of ``processes $\ldots$ performing one or more defined functions'', that describes the life cycle of a product \citep{iso14044}.  The product system model, or PSM, encodes information about how a product is manufactured, distributed, used by the consumer, and what happens to it after it is disposed.  A product system can be divided into a foreground, which denotes the portions of the life cycle whose operations are directly modeled by the study, and a background, which represents the global industrial system \citep{SETAC_inventory_1998}.  While the foreground is often modeled from direct observation or other primary data, background processes are typically drawn from a life cycle inventory (LCI) database prepared by a third party.

Although there has been considerable effort to normalize and harmonize 
LCI database design methodology and data interchange \citep{UNEP_2011, JRC_ILCD_ELCD_2013, Mila_e_Canals_2015, Ingwersen_JLCA_2015}, less attention has been paid to harmonizing PSMs.  These models are often highly complex and include countless modeling decisions, approximations, and assumptions made by a study author \citep{Lloyd2007, reap2008_I}.
Factors such as co-production strategy \citep{Finnveden_1999, Pelletier_2014} can also render study findings unreliable or highly contingent.  Consequently, while comparative evaluations can be made regarding alternative cases or scenarios within a single study, comparisons across studies are much more challenging \citep{Heath2012, Henriksson2014}.  Even when relatively simple and homogeneous systems are considered, wide variations can be found in the results of studies from different authors \citep{van_der_Harst_2013, Turconi_2013}.
A lack of transparency in reporting, particularly regarding definitions of study scope and system boundary,
is a crucial challenge to the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.  A further challenge to transparency can be found in the use of input data that are confidential or proprietary \citep{Kuczenski_ESD_2017}.  One consequence of all these challenges is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.

As LCA gains prominence, particularly in the policy realm, these problems become more acute.  If LCA is to be used in policy, it will necessarily have the effect of recommending certain product systems, technologies, or approaches at the expense of others; yet in that event it is critically important that there be a consensus among stakeholders that the quantitative results are credible and well-supported \citep{Rainville_2015, McManus_2015}.  As noted above, this level of consensus is hard to realize, in part because of the conflicting implications inherent in different analytic modes.
Plevin et al.\ famously observed in the strongest terms that careless reporting of attributional LCA study results can distort the significance of findings in a policy context \citep{Plevin_2013}, a shortcoming that can be found in other modes as well \citep{Brandao_2014}.

An LCA study result is ultimately an assertion that for some constructed PSM, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.  But if the PSM itself is ambiguously stated, then the significance of the result is ambiguous as well.  Often the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.  When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.  \cite{iso14071}, which provides guidelines on critical review, emphasizes that review of the PSM or individual datasets is optional.  
%A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.
Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the PSM and thus, no automatic mechanisms for interpretation or review of published LCA studies.  Moreover, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.

Because of the diversity of software and techniques for performing LCA, it is important to develop methods for communicating the structure of the PSM that is not dependent on any particular software or database \citep{Kuczenski_JLCA_2018}.    The purpose of this paper is to advance a framework for the formal description of PSM that would allow for sharing, revision, and critical review of models while also enabling authors to protect the privacy of confidential information.  This objective is also in line with the broader imperative throughout industrial ecology to improve transparency and reproducibility of computed results \citep{Hertwich_JIE_2018}.

\section{Disclosing the Product System Model}

\subsection{Objectives}

The task of computing life cycle impact assessment (LCIA) results from a product system model can be described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.  Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh2005a, Peters_JLCA_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.

Reporting the results of an LCA study implies the existence of two parties with different information: the study author has full knowledge of the product system model; and the reader only has knowledge of what the author has disclosed.  Obviously, the level of detail of the disclosure will determine how useful it is to the reader.  In this context, the reader will want to know the answers to various review objectives.  
At the same time the author may have constraints that limit what can be disclosed.  The main goal of the framework presented here is to give form to the disclosure so that the requirements of various review objectives, and restrictions associated with various disclosure contraints, can be evaluated clearly.

On the basis of transparency, the framework for disclosure of the PSM should meet the following objectives:
\begin{itemize}
\item \textbf{Computability}. The result should be computed using only the disclosure, plus resources available independently to the reader.  The computation should follow the consensus understanding of process LCA computation.
\item \textbf{Completeness / Minimality}.  The computation should use all of the information provided in the disclosure.
\item \textbf{Reproducibility}.  The result obtained by the reader should match the result provided by the author.
\end{itemize}
One key objective of disclosure is to inspect the work of the author.  Consequently, the disclosure should be made up of the author's own work product and should exclude, for instance, data that were drawn from a reference database or other independent source and used in an unmodified form.  The purpose of this is twofold: to focus the efforts of the reader on the work of the author, and to enable the reader to obtain and inspect the third-party data independently, thus ensuring its integrity.
\begin{itemize}
\item \textbf{Authority}.  The contents of the disclosure should authentically represent the outcomes of the author's labor in constructing the PSM.
\item \textbf{Originality}.  The disclosure should exclude information that is made available by a third party.  (It must, however, still include previously-published data that were re-implemented by the author in the study.)
\end{itemize}
In addition, because of the ubiquity of confidential and sensitive information in LCA, any operational disclosure framework must also meet the following:
\begin{itemize}
\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the author.
\end{itemize}

\subsection{Mathematical Formulation of LCA}

In the most general terms, the LCA problem concerns the interaction of a collection of activities and commodities described by a supply-and-use inventory, in which $U$ is a table of the use of each commodity by each activity (inputs) and $V$ is a table of the supply of each commodity by each activity (outputs) \citep{Pauliuk_2015_framework}.  Such an account can always be transformed into a \textit{symmetric} table of inputs and outputs on a commodity-by-commodity or industry-by-industry basis \citep{Eurostat_2008}. Doing so requires the application of various assumptions and transformations to the data, the details of which are not in scope of this paper, but which have recently been reviewed in depth \citep{Majeau_Bettez_2014}.  Although they originated independently, the ``classical'' process-based approach and the input-output approach have been shown to be equivalent for the purposes of LCA computation \citep{Suh_JIE_2010}.  Please see the Supporting information for a discussion about this equivalency.

A symmetric input output table can be written as a table of coefficients, in which each column of the matrix provides a ``recipe'' for creating the corresponding commodity by consuming other commodities (and, if applicable, producing other co-products). Each column is equivalent to a ``unit process'' in ISO LCA, except that the process must have exactly one reference flow, the identity of which is implicit in the column's index.  The table of coefficients, commonly known as a \textit{direct requirements matrix}, can be used to compute the output from all activities that is necessary to produce a given commodity.  That vector of outputs can then be combined with a second table of coefficients, known as an \textit{emission matrix}, that reports exchanges with the environment associated with the production of each commodity, to determine the cumulative environmental emissions associated with the production of the commodity over its life cycle.  Those emissions are individually \textit{characterized} with respect to an environmental impact category of interest, and the characterized emissions summed to report the \textit{impact score} or category indicator, which is the result of the study.  The formulation may be summarized:
\begin{equation}
s = \mathbf{c}^T \cdot B \cdot \left(I - A\right)^{-1} \cdot \mathbf{y}
\label{eqn:leontief}
\end{equation}
Here $A$ represents the direct requirements, sometimes known as the \textit{Leontief matrix}; $B$ is the emission matrix;  $\mathbf{c}$ is a column vector of characterization factors for the environmental emissions; $\mathbf{y}$ is the externally specified final demand; and $s$ is the numerical impact score or category indicator.  For simplicity, we will consider the computation of a single impact result.  However, it is straightforward to imagine the more typical case, in which $C$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.  Eq.~\ref{eqn:leontief} is visualized in Figure~1.
%\input{fig_lca}

The term $(I-A)$ is equivalent to a \textit{technology matrix} in traditional LCA, normalized by column so that each column represents a unit output of a single reference product, and each row represents the use of the corresponding product. 
The coefficient matrices $A$ (or $I-A$) and $B$ together are called a \textit{life cycle inventory database} (LCIDB), and the functional unit of a study is given by the final demand vector $\mathbf{y}$, which is the product or service whose delivery is to be assessed.
Intermediate results of this computation include:
\begin{itemize}
\item $\mathbf{x} = \left(I-A\right)^{-1}\cdot\mathbf{y}$, the activity level vector;
\item $\mathbf{b} = B\cdot\mathbf{x}$, the life cycle inventory;
\item $s = \mathbf{c}^T\cdot\mathbf{b}$, the life cycle impact category indicator, the study result.
\end{itemize}

Some LCA software systems have the capacity to export the $A$ and $B$ matrices for a given study model.  However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are not usually modified by study authors.  Thus exporting $A$ and $B$ does not meet our authority and originality objectives.  Additionally, replicating the entire LCIDB is cumbersome because of the large size of the matrices.   As a consequence, the matrix formulation in Eq.~\ref{eqn:leontief} is unsuitable as a disclosure framework.

\subsection{Foreground Study}

In practice, most LCA studies contain information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with their own modeling information, known as the study foreground.  To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.  Background databases are prepared independently of any study and do not typically vary across the studies in which they are used.  However, the results of a study with foreground content cannot generally be computed without knowledge of the background.

To understand how a foreground study is constructed on top of a background database, imagine a simple product system that can be fully described by only making reference to processes that exist in a background database.  For instance, the use of a computer may be modeled using reference processes for producing and disposing the computer, and a reference process for electricity consumption.  Assuming those reference processes are available in a background database, the product system can be represented entirely by specifying a final demand vector $\mathbf{y}$ whose nonzero values correspond to computer production, computer disposal, and electricity.  The LCIA results can be computed using Eq.~\ref{eqn:leontief}.

The same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
\begin{equation}
  \begin{array}{cc}
s =& \mathbf{c}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
 =& \mathbf{c}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
  \end{array}
  \label{eqn:canonical}
\end{equation}
where
$\tilde{A} = \left[\begin{smallmatrix} 0 & \mathbf{0}^T \\ \mathbf{y} & A  \end{smallmatrix}\right]$,
$\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$  (A proof is provided in the electronic supporting information).  This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.

This same workflow can be repeated to build a more elaborate foreground.  For instance, a model of private vehicle travel may make use of reference processes for vehicle production and decommissioning, road maintenance, fuel retail, and a model of direct emissions from fuel combustion.  This model could occupy an additional column of $\tilde{A}$, with the coefficients for fuel combustion emissions being contained in the corresponding column of $\tilde{B}$.  A process for brewing coffee could be created in a third column that requires coffee beans, water, and electricity. Finally, a fourth column could be added that invokes the products from the first three columns in proportion to describe the reference product of traveling to a coffeeshop to work.

In general, a product system model may be constructed through the successive augmentation of a background database with foreground content.  As long as the background database is not altered to depend on the newly created foreground, such a study can be written in block triangular form \citep{Kuczenski_JLCA_2015}:
\begin{equation}
\tilde{A} = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A
  \end{array}
\right];\;\;\;
  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right]
\label{eqn:foreground}
\end{equation}
Here, the submatrix $A_f$ represents the foreground; $A$ represents the background; the rectangular matrix $A_d$ represents the dependency of the foreground on the background; and the top right submatrix is zero.  The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$, which includes background emissions. 
The augmented $\tilde{A}$ and $\tilde{B}$ together make up an \emph{LCA foreground study}.  This formulation can be applied to the vast majority of LCA case studies, and very likely to all case studies prepared using commercial LCA software.\footnote{It does not apply to cases in which the background database must be altered to adjust for double counting, such as in so-called ``integrated hybrid'' models in which all quadrants contain non-zero cells \citep{Suh2004}.  These models are more complex than the models typically built by LCA practitioners, and require more sophisticated computation.  In conventional LCA practice it is inadvisable to alter background databases to adapt them to study conditions.  Instead, background processes can be replicated in the foreground, where they can be adapted as needed (cf. \cite{Bourgault_JLCA_2013}), without affecting the integrity of the background database.}

Without loss of generality, it is possible to construct the study such that the \textit{first} column of $\tilde{A}$ delivers the functional unit of the study.  We thereby designate the vector $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ as the \textit{canonical functional unit} of a foreground study, which allows us to introduce convenient notation later on.\footnote{Of course, the ``functional unit'' does not need to be limited to a unit of any given output---instead, it \textit{defines} a unit of output. For instance, if the functional unit of a study is 1,000 loaves of bread, and one loaf of bread weighs 0.454 kg, and the unit output from column 567 of the A matrix is 1 kg of bread, then the canonical functional unit would be realized by entering \texttt{454} in column 1, row 567, leaving the rest of that column zero.}

%\input{fig_foreground}
It is currently common practice for background database maintainers to pre-compute the life cycle inventory results for their databases.  In cases where the background database is used without modification, including the contents of $A$ in $\tilde{A}$ becomes unnecessary.
Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, Eq.~\ref{eqn:leontief} can be written so that the foreground matrix inversion is  performed separately from the computationally costly background computation:
\begin{equation}
s = \mathbf{c}^T \times (B_f + B_x\times A_d) \times (I - A_f)^{-1} \times \tilde{\mathbf{y}}_f
\label{eqn:study}
\end{equation}
where $\tilde{\mathbf{y}}_f$ is a canonical functional unit vector whose length is the same as the rank of $A_f$. Eq.~\ref{eqn:study} is derived in the supporting information.
Eq.~\ref{eqn:study} is visualized in Figure~2, with the background technology matrix $A$ subsumed into $B_x$ and replaced by an identity matrix $I$.

The activity levels of the foreground nodes are represented by $\tilde{\mathbf{x}}$, which is the result of $\tilde{\mathbf{y}}_f$ selecting the first column of $(I-A_f)^{-1}$.
This vector can be used to describe an \emph{aggregated foreground} in coefficient form, which delivers the same results as the foreground study given in Eq.~\ref{eqn:foreground}, but without disclosing any details of the construction of the foreground:
\begin{equation}
\begin{array}{rl}
    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
\end{array}
\label{eqn:agg}
\end{equation}
Here $\tilde{\mathbf{a}}_d$ reports the aggregated foreground's dependency on background processes in background flows per canonical functional unit;  $\tilde{\mathbf{b}}_f$ reports foreground emissions per functional unit; and $\tilde{\mathbf{b}}_x$ reports the aggregated background life cycle emissions per functional unit.

The aggregated result can also be computed from the aggregated foreground:
\begin{equation}
\begin{array}{rl}
s & = \mathbf{c}^T \times( B_f + B_x \times A_d) \times \tilde{\mathbf{x}} \\
  & = \mathbf{c}^T \times( \tilde{\mathbf{b}}_f + B_x \times\tilde{\mathbf{a}}_d) \\
  & = \mathbf{c}^T \times( \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x) \\
\end{array}
\end{equation}

  \subsection{Contents of a Disclosure}

The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.  Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{c}$ (or their product $\mathbf{c}\times B_x$) are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.  These three submatrices together can be considered a complete representation of the PSM.

%\input{fig_disclosure}
Thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.  Recognizing that all three matrices are likely to be sparse (that is, that most of the entries are zero), the most efficient disclosure would be to only report non-zero entries.  A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.  A formal disclosure of the PSM can be stated in six parts, including three lists of entities and three sets of data values:
\begin{enumerate}[label={\em d-\roman*}., ref={\em d-\roman*}]
\item\label{itm:fg} An ordered list of foreground nodes, beginning with the functional unit (rows/columns of $A_f$);
\item\label{itm:bg} An ordered list of background flows, each making reference to a particular background dataset (rows of $A_d$, mapping to columns of $B_x$);
\item\label{itm:em} An ordered list of exchanges with the environment (rows of $B_f$, mapping to entries in $\mathbf{c}$);
\item\label{itm:af} A set of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
\item\label{itm:ad} a set of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
\item\label{itm:bf} a set of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
\end{enumerate}
Each entity included in items \ref{itm:fg} through \ref{itm:em} describes a flow of some kind, and its unit of measure must be reported in the disclosure.
In both items~\ref{itm:bg} and~\ref{itm:em}, only rows containing non-zero entries need to be mentioned, since the index of each row is common information.  The contents of a disclosure are represented graphically in Fig.~3, and two example disclosures drawn from reference databases can be found in the supporting information.

A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):
\begin{itemize}
  \item The result is \textbf{computable} using Eq.~\ref{eqn:study};
  \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references in \ref{itm:bg} and \ref{itm:em};
  \item The disclosure includes only information for which the author can claim \textbf{authority};
  \item The disclosure includes all information specific to the study, or for which the study is the \textbf{original} source, including re-implementations of already-published data. 
\end{itemize}
As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{c}$.  These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.  However, the challenges of operationalizing this requirement are complex (see the discussion and supporting information).

\section{Reviewing Product System Models}

Objectives in critical review vary with the study and the reviewer, but they generally include consistency with the ISO 14044 standard, the validity of methods, appropriateness of data used, correctness of interpretations, and transparency of reporting \citep[Section 6.1]{iso14044}.  Although the object of critical review is typically a written report, this level of review is not sufficient to ensure the correctness of quantitative results.
The use of a structured disclosure of a PSM introduces the possibility that reviewers can inspect the model itself and even verify the computations and reported results.  This section discusses how a disclosure in the format described above can enable robust review of a product system model. Note that, for purposes of critical review, the objective for data privacy can be omitted because the reviewer may be granted privileged access to confidential data.  We return to the data privacy requirement later in this section.

\subsection{The Foreground System Boundary and Cut-off Flows}

The foreground of an LCA study is a collection of nodes that represent points in the inventory model where flows are exchanged.  The foreground matrix $A_f$ describes in precise terms what components are included in the product system and how they connect to one another.  The foreground is modelled as a weighted directed graph in which there is a 1:1 correspondence between nodes and product flows.  The foreground matrix $A_f$ reports the adjacency and weights of edges in the graph, and thus describes the relationships among these nodes.  Some simple foregrounds are described in the supporting information.

Generally, the direct requirements matrix can represent any process-flow model with a 1:1 correspondence between processes and flows.  Normally this is understood as a requirement to include only allocated single-output processes.  However, the foreground disclosure framework is also well suited to documenting how multi-output processes are transformed into single-output processes through either partitioning allocation or different forms of system expansion.  Different approaches for creating foregrounds to model multi-output process are presented in the supporting information.

An implicit ``foreground system boundary'' can be imagined to contain all the nodes in $A_f$.  Non-zero values in this matrix represent exchanges inside this boundary.  If a column in $A_f$ is nonempty, then the corresponding node has exchanges with other nodes in the foreground.  On the other hand, if a column in $A_f$ is empty, then that is an indication that the corresponding product flow is crossing the foreground system boundary.  In order for this flow to contribute any impacts, it must either be linked to a background process in the dependency matrix, or represented as an emission.

A flow whose entire column consists of 0s throughout $A_f$, $A_d$, and $B_f$ is a ``cut-off'' that exits the model with zero burdens.  Specifying cut-off criteria and reporting of cut-off flows is an important part of system boundary definition.  Thus, reviewing cut-off flows (including evaluating their significance) is a crucial part of review that is facilitated by having a structured disclosure of the model.

\subsection{Dependencies and Emissions}

The foreground matrix only describes the structure of the PSM; it does not provide any information about inventory or impacts.  The inventory requirements and environmental burdens associated with the model are all described in the dependency matrix $A_d$ and the foreground emission matrix $B_f$.  The dependency matrix shows how foreground nodes depend on activities provided by processes in a background database; the emission matrix shows direct exchanges with the environment.  In both cases, the respective columns of $A_d$ and $B_f$ correspond with columns in $A_f$ and denote dependencies and/or emissions of the corresponding nodes.  If columns of $A_d$ and/or $B_f$ are empty, then the node is simply a pass-through process that serves to connect different foreground elements.  However, if $A_d$ and/or $B_f$ have nonzero entries in that column, then that indicates some transformation process or emission is occurring at that point in the foreground.

Review of the dependencies of foreground nodes must include review of the selection of background processes to which they are connected (i.e. the identities of the columns of $B_x$ referred to in disclosure item~\ref{itm:bg}, the list of background flows). Likewise, the identities of the emissions (entries in disclosure item~\ref{itm:em}) must also be reviewed.

\section{Foreground Aggregation}

When an LCA study result is computed, the foreground and background matrices are aggregated into a scalar indicator score (or a small vector of a few scores).  The formulation in Eq.~\ref{eqn:study} and the illustration in Fig.~2 provides a framework for reporting results at varying levels of aggregation.  In this section, I present several different aggregation computations that provide equivalent results for an LCA study.   After that, I show how a foreground containing private data can be partitioned into public and private portions, each of which can be comprehensively checked by a reviewer having privileged access, and subsequently disclosed in a manner that protects the private data.

\subsection{Forms of Aggregation}

Figure~2 illustrates the LCA computation in various successive aggregation steps, from full matrices to vectors to a scalar value.  The level of aggregation provided in a disclosure will constrain what review objectives can be met by the reader of a study.  Different aggregation forms are illustrated in Figure~4.

\subsubsection{Single-Output Unit Process}

A single-output unit process makes up a single shared column of $A_f$, $A_d$, and $B_f$, as depicted in Figure~4a.  Documentation of unit processes is specified in ISO 14048, and several compliant formats exist, the most well-known being the ILCD and Ecospold XML formats.  ISO 14048 specifies a distinction between intermediate and elementary flows, but does not include the concept of a model foreground.  In a matrix representation, each unit process must have exactly one reference flow, which corresponds to its row and column in $A_f$.
Please see the supporting information for more detail on preparing the $A_f$ matrix.

%\input{fig_aggregation}

\subsubsection{Complete Foreground}

See Figure~4b.  A foreground can be viewed as a linked set of unit processes.  The collection of processes themselves make up the foreground of the study, and so exchanges among them are recorded in $A_f$, and exchanges outside the foreground are represented in $A_d$ or $B_f$.  A complete foreground represents
the full disclosure of a model, and a reader can replicate and extend it independently.
Any modeling technique used to treat a multi-functional process, such as an allocation or system expansion, can only be fully expressed as a foreground with at least as many columns as outputs.  For examples of multi-output processes represented as foreground models, please see the supporting information.  The example shown in Figure~4b shows five nodes, including one cut-off flow.

\subsubsection{Aggregated Foreground}

See Figure~4c. Any collection of foreground nodes can also be expressed in aggregated form by performing the computations in Eq.~\ref{eqn:agg}.  A disclosure of an aggregated foreground resembles the description of a unit process.  As in the unit process case, the aggregated foreground has exactly one reference flow, the canonical functional unit $\tilde{\mathbf{y}}$.  The aggregated foreground still reports explicit links between the model and a background database and can thus be used to review data set selection.

An aggregated foreground with cut-off flows resembles a multi-output process.  After aggregation, foreground flows that correspond to cut-offs must still be reported in a disclosure, while foreground flows interior to the model can be omitted.  This is done with the introduction of a new vector, $\tilde{\mathbf{a}}_f$, which includes the reference flow plus any entries from $\tilde{\mathbf{x}}$ that correspond to cut-off flows.

\subsubsection{Partial Background Aggregation}

See Figure~4d.  Using this approach, the
 aggregated dependency vector is split into two parts that sum to the original:
\begin{equation}
 \tilde{\mathbf{a}}_d = \tilde{\mathbf{a}}_{d,priv} + \tilde{\mathbf{a}}_d'
\label{eqn:partial}
\end{equation}
The disclosed dependencies $\tilde{\mathbf{a}}_d'$ are reported, and the private dependencies $\tilde{\mathbf{a}}_{d,priv}$ are replaced with an aggregated background inventory derived from the background database, $\tilde{\mathbf{b}}_{x,agg}$.  This approach can be used when a reader lacks access to the background inventory sets referred to in $\tilde{\mathbf{a}}_{d,priv}$, or when the author wishes not to disclose the private dependencies.  Foreground emissions $\tilde{\mathbf{b}}_f$ can be reported separately from the aggregation result, but often in current practice they are not distinguished from the background flows.

\subsubsection{Full Background Aggregation and LCI}

See Figure~4e and f. At this level of aggregation, the entire dependency vector is replaced with an aggregated life cycle inventory vector $\tilde{\mathbf{b}}_x$ derived from the background LCIDB.  The reader no longer requires access to any background database to perform the computation, but all dependency information is concealed.  As mentioned above, foreground emissions are often not distinguished from the aggregation result; if foreground and background emissions are added together, the disclosure is as shown in Figure~4f.  The life cycle inventory $\tilde{\mathbf{b}}$ provides the most aggregated form of the study that can still be independently validated with an external characterization vector $\mathbf{c}$.

\subsection{Reviewable Private Aggregation}

%\input{fig_private}

Often, a study author may wish to make a partial disclosure of a PSM that retains the privacy of confidential data, while still permitting comprehensive critical review of the complete model.  In this case it is necessary to grant a reviewer privileged access to the private data in order to meet the review objectives.  The study formulation in Eq.~\ref{eqn:study} may be used to partition a foreground into public and private portions in order to comply with any applicable constraints on what may be disclosed.  The approach is illustrated in Figure~5.

The first step is to identify the constraints on the disclosure.  In the framework presented here, constraints may include the locations or values of any subset of nonzero entries in the $A_d$ and $B_f$ matrices, as well as any subset of nodes in $A_f$.  Fig.~5a shows a PSM in which disclosure constraints are indicated by shaded regions of the matrices.  In this case the nodes represented by the three rightmost columns of $A_f$, as well as one entire row of $A_d$, are to remain secret.

The $A_d$ and $B_f$ matrices are then partitioned into two components that sum to the original:
\begin{equation}\begin{aligned}
  A_d =\; & A_{d,pub} + A_{d,priv} \\
  B_f =\; & B_{f,pub} + B_{f,priv}
  \label{eqn:partition}
  \end{aligned}
  \end{equation}
This is illustrated in Fig.~5b.  Substituted back into Eq.~\ref{eqn:study}, the private portions are aggregated into a private life cycle inventory vector $\mathbf{b}_{priv}$:
\begin{equation}
  \mathbf{b}_{priv} = (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}
  \label{eqn:bpriv}
\end{equation}
while the public portions remain disaggregated:
\begin{align}
  s =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
  =\; & \mathbf{c}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}
\end{align}
Any foreground nodes that are completely contained within the private partition can be omitted from the disclosed foreground.  The aggregation result can then be included as a separate foreground node with a unit weight, and the aggregated inventory vector included as the corresponding column of $B_f$.  This is illustrated in Fig.~5c.

The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values, or disclosing a range that includes the actual value.  This would enable a reader to understand \emph{that} a certain background process or emission was included in the model (and checked by the critical reviewer) without knowing how much, which may jointly satisfy review objectives and disclosure constraints in some cases.

\section{Discussion}

\subsection{Comparative Review: Formalizing Model Scope}

The disclosure of a PSM contains six parts, including three lists and three tables of sparse matrix data.  The graph structure encoded in the foreground matrices can be automatically rendered as a process-flow diagram, supplementing or supplanting the hand-crafted system diagrams that currently accompany ISO reports. While the numeric data are required to critically evaluate and reproduce the study results, substantial information about the study design can be found in the list of background dependencies (\ref{itm:bg}) and emissions (\ref{itm:em}).  These lists report all the distinct sources of environmental impacts in the model.  The cut-off flows, which are foreground nodes that have no nonzero entries in $A_f$, $A_d$, or $B_f$, report on the other hand what aspects were explicitly excluded from the model.

Together, these lists can be considered a \emph{functional definition of the study scope} that can be used to conduct a qualitative comparison of multiple studies by different authors.  Background database choice, dataset selection, and version information can all be made available in disclosure item~\ref{itm:bg}, thereby streamlining what is presently an interpretive and study-specific process of extracting this information from written text.  Similarly, the set of elementary flows explicitly modeled in the study foreground (item~\ref{itm:em}) expose whether any study-specific modeling of direct emissions was performed and what flows were included.

Moreover, because the disclosure of the PSM includes only references to background data without including the data themselves, it is trivial for item~\ref{itm:bg} to include references to a variety of background databases without adding any additional requirements.  For instance, a study that makes reference to a set of EPDs published by one firm, while using a background LCI database from another firm, can express both dependencies in an analogous way, even if the two sources are published independently.

\subsection{Privacy and Partial Aggregation}

Partial aggregation is already used in practice to conceal private data, but current methods do not provide a way for a reviewer to validate the result.  If a study containing confidential information is split into public and private partitions as in Eq.~\ref{eqn:private} and Fig.~5, then the results of the computation can be validated directly by the reviewer.  To achieve this, the reviewer must be able to confirm that the partitioning is valid according to Eq.~\ref{eqn:partition}, and that $\mathbf{b}_{priv}$ was computed correctly (Eq.~\ref{eqn:bpriv}).  This requires that the reviewer be granted access to both the public and private partitions.  However, the private information used in the computation can still be withheld from the disclosure.

A further advantage to this approach is that the reader can evaluate easily what fraction of the overall impact score is accounted for by the public versus the private portions of the model.  A ``disclosure completeness'' metric $\varphi$ can be defined as the fraction of the overall score which is accounted for by the disclosed portion of the PSM:
\begin{equation}
  \varphi =  \frac{\mathbf{c}^T\cdot(\mathbf{b} - \mathbf{b}_{priv})}{\mathbf{c}^T\cdot\mathbf{b}} = 1 - \frac{\mathbf{c}^T\cdot\mathbf{b}_{priv}}{\mathbf{c}^T\cdot\mathbf{b}}
  \label{eqn:metric}
\end{equation}
where $\mathbf{b}$ is the complete life cycle inventory.  The value of this metric can provide an indication of the level of transparency of the disclosure with respect to a given impact category indicator.

\subsection{Attacks on Privacy}

It is not thought possible for a reader to discern any information about a disaggregated system solely from an aggregated disclosure such as $\mathbf{b}_{priv}$ \citep[Ch. 3]{UNEP_2011}, although that assumption has never been rigorously tested.  Compressed sensing is a signal processing technique that seeks solutions to an underdetermined linear system, in which there are more unknowns than equations, based on the assumption that the solution is sparse \citep{Donoho_2006}.  A form of linear optimization called ``basis pursuit'' can be used to try to detect the signal.

In the LCA case, if the private LCI result $\mathbf{b}_{priv}$ includes inventory data from a background database but no foreground emissions, then it can be written as:
\begin{equation}
  \mathbf{b}_{priv} = B_x\times \mathbf{a}_{d,priv}
\end{equation}
Typically this equation is highly underdetermined (In ecoinvent, for instance, $B_x$ has roughly 1,800 rows and over 12,000 columns).  If the private input $\mathbf{a}_{d,priv}$ contains only a few nonzero elements, and the identity of $B_x$ is known, then $\mathbf{a}_{d,priv}$ may be vulnerable to a basis pursuit attack.  Further research is required to evaluate whether this attack can be implemented in a real situation, and how it may be defended against.  
%10.1109/TIT.2006.871582

\subsection{Stable Semantic References to Study Elements}

In practice, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity in identifying the background datasets and emissions to which those matrix entries correspond.
In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of the references contained in items~\ref{itm:bg} and \ref{itm:em}.
The FAIR guiding principles, which were designed to help scientists organize data for Findability, Accessibility, Interoperability, and Reusability \citep{Wilkinson_2016}, provide valuable perspective on this goal.

Central to many of the FAIR guidelines is the use of linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.  This link serves as a unique identifier and a resolvable reference to metadata, and the linked content can be  curated by the study authors and other members of the community \citep{Khan_2011}.

Before the goal of easily reproducible foreground models can be realized, this linked data foundation must be laid.  Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that denote a particular activity in a particular database version and configuration.  These references should take the form of URIs that can be accessed using any Internet browser and can provide both author and reader with documentary information describing the referred dataset, including version information; database configuration; the available reference flows and their dimensions (quantitative units of measure); and a mechanism for the reader to obtain access to the exchange data or LCI/LCIA computations for the purposes of model validation.  A dependency reference in disclosure item~\ref{itm:bg} should include both an activity and a reference flow.

Similarly, 
elementary flows and contexts should be given stable references that are held in commomn across data providers. Current research has revealed widespread challenges to finding consistency on the identities of elementary flows despite the community's longstanding awareness of the problem \citep{Speck_2015,Herrmann_2015}, and several mutually inconsistent reference flow sets now exist \citep{Edelen_2017}.  The conventional understanding of a ``flow'' as comprising a substance and a context together (e.g. methane, emissions to air) causes a combinatorial increase in the number of flows in a database, and also multiplies the potential points of disagreement across sources.  I reiterate Edelen et al.'s (ibid.) suggestion to explicitly regard substances (which can easily include non-material ``flowables'' such as land occupation and transformation) and contexts as independent semantic classes.  In this case, an elementary flow entry in the disclosure item~\ref{itm:em} would require both a flowable and a context.

\subsection{Recommendations for Advancing LCA Practice}

The most transformative aspect of the disclosure framework is its ability to describe the productive output of LCA modeling in a way that transcends differences in software environments and databases.  A product system model described in the disclosure framework can %thus
be implemented in any software system, though the manner of doing so, as well as availability of advanced features for analysis and visualization, varies widely.

To advance LCA practice, the community should embrace the notion of a software-independent PSM description and consider how to best achieve it.  The SETAC North America roadmap module for PSM description and revision \citep{Kuczenski_JLCA_2018} includes milestones toward this goal.  The disclosure framework presented here is offered as a candidate for Milestone 2.1, ``a minimal description of a PSM.''  The community should give this proposal a robust critical evaluation, and researchers should consider how to conceptualize their own models in terms of the framework.  Other milestones and cross-cutting issues are of crucial interest:
\begin{enumerate}[label=(\alph*)]
\item LCI database providers must provide their users with unambiguous, static, resolvable references to background datasets (Milestone 1.1) so that their use in a model may be documented precisely (Milestone 3.1);
\item The transformation of multi-functional processes to single-function processes should be described using the framework (Milestones 1.2 and 1.4; see supporting information);
\item Software makers should enable their users to automatically generate lists of foreground, background, and emission flows in a model foreground (disclosure items \ref{itm:fg} through \ref{itm:em}) and to export foreground matrices (disclosure items \ref{itm:af} through \ref{itm:bf}) from a model in a concise format (Milestones 2.2 and 3.4);
\item During ISO critical review, practitioners and reviewers should strive to come to agreement on the formal structure and contents of the model being reviewed;
\item The LCA research community, including interested parties within the International Society for Industrial Ecology, the UNEP/SETAC Life Cycle Initiative, and other organizations, should collectively pursue consistent, shared definitions of quantities of measure, flows, and contexts used in LCIA.  LCIA method developers should immediately begin using these consensus definitions instead of their own internal lists when publishing flow characterization data, so that software makers can interpret the publications correctly.  
\end{enumerate}
As these milestones are pursued and achieved, it will be possible for a standalone description of a PSM to be computed and extended by another party, regardless of the software used to construct the model originally.  This will tremendously improve the capacity for critical review and reuse of LCA models.

\section{Conclusion}

The current practice of documenting a PSM with a written report sharply limits the ability of readers to interpret and reuse the data.  It also results in the model being frozen to a particular configuration or set of configurations used to generate the published results.  Any review objectives involving sensitivity of the model to parameter variations and/or alternative scenarios are limited to the selections considered by the author.
When a study result is published in an aggregated form, often it is beyond the capacity of the critical reviewer to validate the aggregation result, even if the reviewer has privileged access to confidential materials used in preparing the report.

This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.  The non-quantitative portions of the disclosure make up a formal definition of study scope and the boundary of the foreground system.  The disclosure framework also allows modelers to express precisely how multi-functionality is handled through allocation or system expansion.  A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.  Once equipped with the PSM, the reviewer or data user can potentially go much further.  Practitioners who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, applying other impact assessment methods, and considering alternative allocation methods. Portions of the model which have been pre-aggregated can remain private in the disclosure. Further research is required to determine whether or under what conditions aggregation results can be reverse-engineered.

If data providers also make available stable semantic references to background data sets and elementary flow characterizations, then it becomes possible for a stand-alone description of a PSM to be used to reproduce an LCA result.  The same description can be extended or incorporated into subsequent studies by other authors. From this prescription, a framework for distributed, platform-independent LCA computation can be imagined.

\let\section\oldsection
    \section*{Acknowledgments}

Many thanks to the anonymous reviewers for significant improvement of the manuscript. This work was partially supported by the US National Science Foundation (CCF-1442966).

 \singlespacing

 \theendnotes

 \bibliography{kuczenski-disclosure}
}

\begin{document}

\mainpaper

\end{document}

### 2018-05-22 post-accept/kuczenski-disclosure_supp.tex
Post-accept wrangling (TeX native\!)

\documentclass[12pt]{article}

%% formatting
\pagestyle{plain}
\usepackage[margin=1in]{geometry}

%% font selection
\usepackage{mathptmx}
\renewcommand\sfdefault{phv}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{textcomp}

\usepackage{setspace}
\usepackage{textcase}
\usepackage{amsmath,amssymb,mathrsfs}
\usepackage{booktabs}
\usepackage{array}

\usepackage{enumitem}
\setlist[enumerate,2]{label=(\alph*)}
\setlist[enumerate,3]{label=\roman*.}

%% pdf foo
\usepackage[svgnames]{xcolor}

\definecolor{hyperblue}{rgb}{0.01 0.04 0.45}
\usepackage{thumbpdf}

\usepackage[hyphens]{url}

\RequirePackage[colorlinks,breaklinks=true,hyperindex,plainpages=false,%
linkcolor=hyperblue,urlcolor=hyperblue,anchorcolor=hyperblue,%
citecolor=black,%
pdfstartview=FitH,pdfpagemode=UseOutlines,%
bookmarksopen=false]{hyperref}
\def\pdfBorderAttrs{/Border [0 0 0] } % No border arround Links
\setlength{\pdfpagewidth}{\paperwidth}
\setlength{\pdfpageheight}{\paperheight}
\usepackage{breakurl}
\urlstyle{same}

%% Appearance

\usepackage{parskip}
\setlength{\parindent}{0pt}

\newcommand{\doi}[1]{\mbox{doi:~\href{http://dx.doi.org/#1}{\small\texttt{#1}}}}
\newcommand{\email}[1]{\texttt{\href{mailto:#1}{#1}}}
\newcommand{\overbar}[1]{\mkern
  1.5mu\overline{\mkern-1.5mu#1\mkern-1.5mu}\mkern 1.5mu}

\let\oldtimes\times
\let\times\cdot

%\lhead{\sffamily Disclosure of product system models}
%\rhead{\sffamily Kuczenski -- 2017}

\usepackage{tabularx}
\usepackage[round]{natbib}
\usepackage{siunitx}

%% To build supporting information:
%% - uncomment supportingfalse
%% - in the document uncomment both mainpaper and supporting
%% - latex agg-JIE
%% - in the document [bottom] inside \ifsupporting, comment mainpaper
%% - maketex -x -o supp.pdf agg-JIE

\bibliographystyle{authordate_bk}

\renewcommand{\topfraction}{0.7}
\renewcommand{\bottomfraction}{0.7}

%\thispagestyle{empty}

\input{psstyle}
\input{../manuscript/draw-matrices}
\let\showgrid\relax

\title{Disclosure of Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy}
\author{Brandon Kuczenski\\
Institute for Social, Behavioral, and Economic Research\\
University of California, Santa Barbara}
\date{Submitted after revision, 23 Apr 2018}

\usepackage{lineno}

\def\supporting{
  \setcounter{figure}{0}
  \setcounter{table}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}
  \renewcommand{\thetable}{S\arabic{table}}

  \maketitle

  \begin{center}
    \Large Supporting Information
  \end{center}

  \section{List of Mathematical Symbols}

The symbols  used in the text are defined in Table~\ref{tbl:symbols}.

\begin{table}[p]
  \caption{List of Mathematical Symbols}
  \label{tbl:symbols}
\small\sffamily
  \def\gap{0.85ex}
\begin{tabularx}{\textwidth}{cclXl}
  \toprule
  \textbf{Symbol} & \textbf{Formula} & \textbf{Size} & \textbf{Description} & \textbf{ref.} \\
  \midrule
  \multicolumn{5}{l}{Dimensions} \\[\gap]
  $m$ &  & scalar & Number of emissions & Eq.~1 \\
  $n$ &  & scalar & Number of process outputs in background database & Eq.~1 \\
  $p$ &  & scalar & Number of nodes in foreground model & Fig.~2\\
  $\phantom{'}p'$ & & scalar & Number of cut-off nodes in foreground & Fig.~4c \\
  $N$ & $n+p$ & scalar & size of augmented matrix & n.a. \\[\gap]
  \multicolumn{5}{l}{Classical LCA} \\[\gap]
  $A$ & & $n\oldtimes n$ & Direct requirements matrix (Leontief) & Eq.~1 \\
  $I-A$ & & $n\oldtimes n$ & Technology matrix & " \\%Eq.~\ref{eqn:leontief} \\
  $B$ & & $m\oldtimes n$ & Emission matrix & " \\%Eq.~\ref{eqn:leontief} \\
  $\mathbf{y}$ & & $n \oldtimes 1$ & Final demand vector & " \\%Eq.~\ref{eqn:leontief} \\
  $\mathbf{x}$ & $(I-A)^{-1}\times \mathbf{y}$ & $n \oldtimes 1$ & Activity level vector & Fig.~1 \\
  $\mathbf{b}$ & $B\times \mathbf{x}$ & $m \oldtimes 1$ & Life cycle inventory & " \\%Fig.~\ref{fig:lca} \\
  $\mathbf{c}$ & & $m \oldtimes 1$ & Characterization vector & Eq.~1 \\
  $s$ & $\mathbf{c}^T\times\mathbf{b}$ & scalar &  LCIA Result & " \\[\gap]%Eq.~\ref{eqn:leontief} \\[\gap]
  \multicolumn{5}{l}{Product System Model} \\[\gap]
  $\tilde{\mathbf{y}}$ & $[ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ & $N\oldtimes 1$ & Canonical functional unit vector & Eq.~2 \\
  $\tilde{A}$ & & $N\oldtimes N$ &  Augmented $A$ matrix & Eq.~3 \\ 
  $\tilde{B}$ & & $m\oldtimes N$ &  Augmented $B$ matrix & " \\%Eq.~\ref{eqn:foreground} \\
  $A_f$ & & $p\oldtimes p$ &  Foreground matrix & " \\%Eq.~\ref{eqn:foreground} \\ 
  $A_d$ & & $n\oldtimes p$ &  Background dependency matrix & " \\%Eq.~\ref{eqn:foreground} \\ 
  $B_f$ & & $m\oldtimes p$ &  Foreground emission matrix & " \\%Eq.~\ref{eqn:foreground} \\
  $B_x$ & $B\times(I-A)^{-1}$ & $m\oldtimes n$ &  Background LCI database matrix & Eq.~4 \\
  $\tilde{\mathbf{y}}_f$ & $[ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$ & $p\oldtimes 1$ & Canonical foreground functional unit & " \\%Eq.~\ref{eqn:study} \\
  $\tilde{\mathbf{x}}$ & $(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f$ & $p\oldtimes 1$ & Foreground activity vector & Fig.~2 \\[\gap]
  \multicolumn{5}{l}{Foreground Aggregation} \\[\gap]
  $\tilde{\mathbf{b}}_f$ & $B_f\times\tilde{\mathbf{x}}$ & $m\oldtimes 1$ & Aggregated foreground emissions & Eq.~5\\
  $\tilde{\mathbf{a}}_d$ & $A_d\times\tilde{\mathbf{x}}$ & $n\oldtimes 1$ & Aggregated dependencies & " \\%Eq.~\ref{eqn:agg}\\
  $\tilde{\mathbf{b}}_x$ & $B_x\times\tilde{\mathbf{a}}_d$ & $m\oldtimes 1$ & Aggregated background & " \\%Eq.~\ref{eqn:agg}\\
  $\tilde{\mathbf{a}}_f$ & & $p'\oldtimes 1$ & Vector of foreground cut-off flows & Fig.~4c \\
  $\tilde{\mathbf{b}}$ & $\tilde{\mathbf{b}}_x+\tilde{\mathbf{b}}_f$ & $m\oldtimes 1$ & Aggregated life cycle inventory & Fig.~4f\\[\gap]
  \multicolumn{5}{l}{Private Aggregation} \\[\gap]
  $\tilde{\mathbf{a}}_{d,priv}$ & & $n\oldtimes 1$ & Private dependencies & Eq.~7 \\
  $\tilde{\mathbf{a}}_d'$ & $\tilde{\mathbf{a}}_d-\tilde{\mathbf{a}}_{d,priv}$& $n\oldtimes 1$ & Disclosed dependencies & " \\%Eq.~\ref{eqn:partial} \\
  $A_{d,priv}$ & & $n\oldtimes p$ & Private dependency matrix & Eq.~8\\
  $A_{d,pub}$ & $A_d - A_{d,priv}$ & $n\oldtimes p$ & Public dependency matrix & " \\%Eq.~\ref{eqn:partition}\\
  $B_{b,priv}$ & & $m\oldtimes p$ & Private foreground emission matrix & " \\%Eq.~\ref{eqn:partition}\\
  $B_{f,pub}$ & $B_f - B_{f,priv}$& $m\oldtimes p$ & Public foreground emission matrix & " \\%Eq.~\ref{eqn:partition}\\
  $\mathbf{b}_{priv}$ & $(B_{f,priv} + B_x\cdot A_{d,priv})\cdot\tilde{\mathbf{x}}$ & $ m\oldtimes 1$ & Private life cycle inventory & Eq.~9\\
  $\varphi$ & Eq.~12 & scalar & Disclosure completeness metric & Eq.~12\\[\gap]
  \bottomrule
\end{tabularx}
\end{table}

\section{Mathematical Relations}

\subsection{Final Demand Augmentation}

This is a proof of Equation~2 in the manuscript.

Given a life cycle inventory database (LCIDB) containing input-output matrix $A$ and emission matrix $B$, and a final demand vector $\mathbf{y}$, construct an augmented inventory database
$\tilde{A} = \left[\begin{smallmatrix} 0 & \mathbf{0}^T \\ \mathbf{y} & A \end{smallmatrix}\right]$,
$\tilde{B} = [ \mathbf{0},\, B ]$, and canonical functional unit $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.  Show that
\begin{equation} B\times\left(I-A\right)^{-1}\times\mathbf{y} = \tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
\end{equation}

We begin by constructing $(I-\tilde{A})^{-1}$.  Because the matrix is block triangular, we only need to determine the value of $\mathbf{q}$ in the matrix below:
\begin{eqnarray}
  I &=&  (I-\tilde{A})^{-1}\times(I-\tilde{A})\\
  \left[\begin{array}{cc} 1 & \mathbf{0}^T \\
    \mathbf{0} & I
  \end{array}\right] &=&
  \left[\begin{array}{cc} 1 & \mathbf{0}^T \\
      \mathbf{q} & (I-A)^{-1}
  \end{array}\right]\times
  \left[\begin{array}{cc} 1 & \mathbf{0}^T \\
      -\mathbf{y} & (I-A)
        \end{array}\right]
\end{eqnarray}
It can be seen that $\mathbf{q}-(I-A)^{-1}\times\mathbf{y}=\mathbf{0}$ is a necessary condition to satisfy the equality, and thus $\mathbf{q}=(I-A)^{-1}\times\mathbf{y}$.

We know that $\tilde{\mathbf{y}}$ selects the first column of its argument, so:
\begin{equation}
  (I-\tilde{A})^{-1}\times\tilde{\mathbf{y}} = \left[\begin{array}{c}
      1\\
      \mathbf{q}
    \end{array}\right]
\end{equation}

And therefore

\begin{eqnarray}
  \tilde{B}\times(I-\tilde{A})^{-1}\times\tilde{\mathbf{y}} &=& \left[\begin{array}{cc}\mathbf{0}&B\end{array}\right]\times\left[\begin{array}{c}
      1\\
      \mathbf{q}
    \end{array}\right]\\
   &=&B\times\mathbf{q}\\
   &=&B\times(I-A)^{-1}\times\mathbf{y}
\end{eqnarray}
thus completing the proof.

\subsection{The Study Foreground Equation}

This section contains the derivation of Equation~4 in the manuscript, in two steps.

\subsubsection{``Flattening'' the background database}

First, information about the background database is removed from $\tilde{A}$ and replaced by an aggregated LCI database in a process I call ``flattening.''

Given a canonical LCA foreground study, show that the following formulations are equivalent:
\begin{eqnarray}
\tilde{A} = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A
  \end{array}
\right];&
  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right] \\
 \tilde{A}_{flat} = \left[\begin{array}{cc} 
A_f &  0 \\
A_d & 0 
   \end{array}\right];&  \tilde{B}_{flat} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
\end{eqnarray}
where $B_x = B\times(I-A)^{-1}$.

It is sufficient to show that $\tilde{B}\times(I-\tilde{A})^{-1} = \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}$.

Using the shorthand $Z=I-A$, we define the following identity:
\begin{equation}
  I* = \left[\begin{array}{cc}
      I & 0 \\
      0 & Z
    \end{array}\right]^{-1} \times
  \left[\begin{array}{cc}
      I & 0 \\
      0 & Z
    \end{array}\right]
\end{equation}

We introduce the identity in between the two terms of the LCIDB:
\begin{align}
  \tilde{B}\times(I-\tilde{A})^{-1} &=
  \left[\begin{array}{cc}B_f & B\end{array}\right]\times I*  \times
  \left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I-A\end{array}\right]^{-1}\\
      &=  \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\times
      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right] \times
      \left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]^{-1}
\end{align}
Using the associative property of matrix multiplication, it can be shown that $M\times N^{-1} = (N\times M^{-1})^{-1}$ for any invertible $M$ and $N$ of the same rank, so:
\begin{eqnarray}
  \tilde{B}\times(I-\tilde{A})^{-1}
  &=& \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
      \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\times
\left(
\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]\times
    \left[\begin{array}{cc}I & 0 \\ 0 & Z\end{array}\right]^{-1}\right)^{-1}\\
  &=&   \left[\begin{array}{cc}B_f & B\end{array}\right]\times 
      \left[\begin{array}{cc}I & 0 \\ 0 & Z^{-1}\end{array}\right]\times
\left(
\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & Z\end{array}\right]\times
\left[\begin{array}{cc}I & 0 \\ 0 & Z^{-1}\end{array}\right]\right)^{-1}\\
 &=& \left[\begin{array}{cc}B_f & B\times Z^{-1}\end{array}\right]\times
\left(\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I\end{array}\right]\right)^{-1}
\end{eqnarray}
But these terms are equivalent to the flattened LCIDB:
\begin{eqnarray}
  \tilde{B}\times(I-\tilde{A})^{-1} &=&
  \left[\begin{array}{cc}B_f & B\times (I-A)^{-1}\end{array}\right]\times
\left(\left[\begin{array}{cc}I-A_f & 0 \\ -A_d & I\end{array}\right]\right)^{-1} \\
&=& \left[\begin{array}{cc}B_f & B_x\end{array}\right]\times
\left(I-\left[\begin{array}{cc}A_f & 0 \\ A_d & 0\end{array}\right]\right)^{-1}\\
&=& \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}\label{eqn:si-flat}
\end{eqnarray}
thus completing the proof.

\subsubsection{Separating Background from Foreground}

Equation~4 in the manuscript is derived from Equation~\ref{eqn:si-flat} above.  It was also derived in Kuczenski (\citeyear{Kuczenski_JLCA_2015}).

We begin by computing $(I-\tilde{A}_{flat})^{-1}$, again taking advantage of the matrix's block triangularity.  Find $Q$ such that:
\begin{equation}
  \left[\begin{array}{cc} (I-A_f) & 0 \\ -A_d & I\end{array}\right]\times
  \left[\begin{array}{cc} (I-A_f)^{-1} & 0 \\ Q & I\end{array}\right] =
  \left[\begin{array}{cc} I & 0 \\ 0 & I\end{array}\right]
\end{equation}

It can be seen that $Q=A_d\times(I-A_f)^{-1}$ is a necessary condition to satisfy the equality, and thus:
\begin{equation}
  (I-\tilde{A})^{-1} = \left[\begin{array}{cc} (I-A_f)^{-1} & 0 \\ A_d\times(I-A_f)^{-1} & I\end{array}\right]
\end{equation}.

Substituting this into the LCA system equation:
\begin{align}
  s =\;& \mathbf{c}^T\times \tilde{B}_{flat}\times(I-\tilde{A}_{flat})^{-1}\tilde{\mathbf{y}}\\
  =\;& \mathbf{c}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
  \left[\begin{array}{cc} (I-A_f)^{-1} & 0 \\ A_d\times(I-A_f)^{-1} & I \end{array}\right]
  \times\left[\begin{array}{c} \tilde{\mathbf{y}}_f \\ \mathbf{0}\end{array}\right]
\end{align}
where $\tilde{\mathbf{y}}_f$ is a canonical functional unit having the same dimension as $A_f$.  Thus:
\begin{align}
s =\;& \mathbf{c}^T\times \left[\begin{array}{cc} B_f & B_x \end{array}\right]\times
 \left[\begin{array}{c} (I-A_f)^{-1}\times\tilde{\mathbf{y}}_f \\ A_d\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f\end{array}\right]\\
 =\;& \mathbf{c}^T\times \left(B_f\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f
 + B_x\times A_d\times(I-A_f)^{-1}\times\tilde{\mathbf{y}}_f\right)\\
 =\;& \mathbf{c}^T\times \left(B_f + B_x\times A_d\right)\times(I-A_f)^{-1}\times \tilde{\mathbf{y}}_f\\
 =\;& \mathbf{c}^T\times \left(B_f + B_x\times A_d\right)\times \tilde{\mathbf{x}}
\end{align}
thus completing the derivation.

\section{On the Equivalence of Process and Input-Output LCA}

There is a level of confusion about the differing implications of process-based LCA inventory analysis as described in Heijungs and Suh (\citeyear{Heijungs2002}), and input-output analysis using a Leontief matrix.  In fact, these two techniques were shown to be equivalent in Suh et al (\citeyear{Suh_JIE_2010}), whose abstract declares: ``the article shows that the partitioning method in LCA is equivalent to the industry-technology model in input-output economics, and system expansion in LCA is equivalent to the by-product-technology model in input-output output economics.''  The article does indeed show convincingly that ``the two data types can be hybridized without the loss of methodological consistency'' (p.345).

   Both the technology matrix $A$ in LCA and the Leontief matrix $(I-A)$ in input output analysis are ways of encoding the relationships between the products that some processes ``make'' and other processes ``use.''  In either the classical or input-output formulation, both matrices must be put into symmetric form to be invertible, and this means describing each column as having a single, reference output.  This is done either through disaggregation (i.e. dividing a process by its various outputs), aggregation (i.e. combining products to form a market or industry), or some combination of these.  While the task of transforming data is highly case-dependent, it is not constrained by the choice of mathematical formalism.

   Suh et al, after a rich investigation of the literature, a thorough mathematical development, and a worked example, conclude that ``even very large-scale LCI problems, such as those in commercial LCI databases, can be computed with a consistent mathematical framework$\ldots$'' that being the make-and-use framework (p. 349).  In effect, the different mathematical approaches can be made equivalent.  The equivalency can be made very plain by simply reinterpreting the input-output expressions \textit{as} technology matrices.  In other words, (in Suh et al) the expressions $(I-A_I)$ (eq. 6), $(I-A_C)$ (eq. 9), $(I-A_B)$ (eq. 12), and $(V'-U)$ (eq. 17) can simply be interpreted as \textit{different ways to formulate} the process-based technology matrix.  There is no inconsistency between this interpretation and the classical formulation.

   Once single-output processes are obtained, one substantive difference between the various approaches is column scaling: whether the entries describe real outputs, unit outputs, or unit activity levels.  The database creator must ensure that a consistent scaling is applied to both the technology matrix and the emission matrix $B$.  The convention in LCA is for a column to represent a unit activity level.  To use the input-output coefficient approach, as in the present article, the columns must be scaled to a \textit{unit of the reference product}.  This reference output is then omitted from the $A$ matrix, and is instead supplied by the $I$ in the expression $(I-A)$.  This formulation is completely consistent with any other matrix-based formulation of LCA, including classical process LCA as traditionally practiced.  In particular, the approaches described here apply equally well to LCA models using both process-based databases such as Ecoinvent, input-output databases such as Exiobase, and background databases such as the GaBi databases.

   Interested readers should note that the canonical reference for LCA, ISO 14044, has nothing to say about the mathematical techniques used to solve LCA problems, and the only preference for a technology matrix over an input-output matrix is borne of tradition or custom rather than theory.  SimaPro, the leading worldwide software system for scientific LCA, converted to using an iterative Leontief inverse for Version 8, as disclosed in a whitepaper published in 2014\footnote{\url{https://www.pre-sustainability.com/news/new-calculation-engine-simapro-8}}, realizing significant speed and memory usage improvements.  OpenLCA, the open-source alternative, also reported a dramatic reduction in memory usage and computation time around the release of version 1.4 in 2014, coincident with a revised algorithm ``optimized for huge databases.''\footnote{\url{http://www.openlca.org/openlca-beta-1-4-released-optimised-for-huge-databases/}} This is highly suggestive of a switch to an iterative approach.  GaBi, the more industrially oriented software, does not use a matrix formulation at all, but has always computed activity levels using iteration at the plan level.

\section{Meeting the Disclosure Objectives}

\subsection{Transparency}

The objective for transparency was identified as requiring computability, completeness, and reproducibility.  These requirements can be met by a disclosure that clearly describes the identities of each row or column of $A_f$, $A_d$, and $B_f$ containing non-zero entries, and the locations and values of those entries.  In principle, a reader with this information would have the capability to construct the augmented LCIDB and perform the computation in Eq.~4.  In actuality, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity in stating the identities of the rows and columns of those matrices.

The foreground nodes, which make up the columns of $A_f$, can be chosen freely by the study author according to the objectives of the study.  Large studies may contain hundreds of foreground nodes, and the nodes can correspond to physical activities, logical operations, unit conversions, accumulation or distribution points, or any other aspect of model construction that can be reflected in a process-flow diagram.  The only requirement on their disclosure is that the identity of each node's reference flow, including its unit of measure, is clearly stated.

For each row in $A_d$ containing a non-zero entry, the author must unambiguously identify the exact dataset used, including the version of the database, as well as the exact process and reference flow selected; the dimension (reference quantity or unit) of each reference flow must be specified; and the sign of the numeric entry in $A_d$ must be consistent with the implementation of the process in the background database.  Similarly, for each row in $B_f$ containing a non-zero entry, the author must unambiguously identify the substance being exchanged with the environment, the compartment or context into which it is being exchanged, and the reference quantity or unit associated with the flow.  Sign consistency must also be assured.  If LCIA indicator results are included, the author must also unambiguously identify the method computed (identity of $\mathbf{c}$).

\subsection{Authority and Originality}

It may be observed that many LCA studies make use of data sources that have been previously published but that have not been included in any LCI reference or background database.  As long as these data sources can be integrated seamlessly into an LCA computation, it is not necessary to reproduce them in a disclosure.  However, in most applications, previously published data must be re-implemented by the author in the LCA software context, and often this re-implementation requires a re-interpretation of the data source as a unit process inventory, where one exchange is recognized as a reference flow and other exchange values are reported in proportionality to the reference flow.

Because of these conditions, in the vast majority of cases the LCA study disclosure must include the author's reimplementations in order to achieve both the aims for authority and originality.  In so doing, the disclosure enables a critical reviewer to evaluate whether the author's implementations are generally correct and complete.

In the future, the inclusion of external data automatically can be accomplished in the same way that reference LCI data could conceivably be included automatically: by the data providers making their information available  using a stable semantic reference to a specialized Web-based application programming interface (API).  This would have the benefits of enabling downstream users to access the information without having to re-implement it, thus reducing the size of the disclosure necessary to describe the PSM and simplifying the task of the modeler.

\section{Preparing the Foreground Matrix}

The development in the main text assumes the product system model is already described as a symmetric collection of single-output processes (activities) and flows (commodities) related by coefficients.  In concrete terms, this means that:
\begin{itemize}
\item each node in the foreground (column shared by $A_f$, $A_d$, and $B_f$) corresponds to a particular flow in the model's foreground (row of $A_f$);
\item a unit activity of each node in the foreground corresponds to a unit output of each flow;
\item the coefficients in $A_f$, $A_d$, and $B_f$ are ratios expressing the quantity of the flow (row), in its native unit, exchanged per unit of the reference flow (column). 
\end{itemize}

Each foreground column has both an intrinsic reference flow and a unit of measure, and the use of this flow by other processes must be made with regard to this measure.  The activity level of each foreground node necessarily equals the magnitude of the total reference flow emanating from the node.  For background dependencies and emissions (rows of $A_d$ and $B_f$, respectively), the modeler must take care to use the units established by the references, but in the foreground case the modeler must ensure to be self-consistent in preparing the $A_f$ matrix.

\subsection{Simple Foreground Networks}

\begin{figure}[p]
  \begin{center}
    \begin{pspicture}(0,-7)(14,11)\showgrid
      \small
      \rput[tl](0,11){\input{../figures/fig3a-sequential.fig}}
      \rput[bl](0,0){\input{../figures/fig3b-additive.fig}}
      \rput[bl](0,-5.5){\input{../figures/fig3n-cyclic.fig}}
      %\rput[tl](6,12){\input{fig3c-unitprocess.fig}}
      \rput[br](14,-3){\input{../figures/fig3c-fragment.fig}}
    \end{pspicture}
    \caption{Equivalent matrix representations and graphs for different foregrounds.}
    \label{fig:fragments}
  \end{center}
\end{figure}

Figure~\ref{fig:fragments} illustrates the structure of some basic foreground configurations.  Most product system models would include many of these elements mixed together.  All the activities in these examples are single-output.
The first (a) is a sequential model, in which each node requires one foreground input and generates one output. This model is equivalent to a ``gate to gate'' model.  Here the weights $k_i$ indicate the amount of the preceding reference flow that is required by the subsequent node.  Figure~\ref{fig:fragments}(b) shows an additive model, in which the outputs of several foreground nodes are added together, equivalent to a ``mixer'' or a horizontal average.  In this arrangement the weights %represent the relative weights of each input, and
should add up to a unit output of the reference node.  Figure~\ref{fig:fragments}(c) shows a foreground model with a cyclic dependency, where some of the reference output is consumed by another foreground node.

A typical PSM may contain multiple modules or fragments that are interconnected.  An example of foreground composed of several fragments is illustrated in Figure~\ref{fig:fragments}(d).  Here, the nodes labeled 1--5 represent one fragment, which generates the foreground's canonical reference flow $\tilde{y}$. This fragment requires two interior flows from separate fragments ($y_0$ and $y_1$), and has two unconnected flows (4 and 5).  The reference $y_0$ is supplied by a second fragment, made up of nodes 6-8.  The reference $y_1$ is supplied by another fragment made of only one node (9). The reference flow $y_1$ is consumed in two different places by the other fragments.

As noted, a reviewer with access to items \emph{d-i} and \emph{d-iv} in the disclosure would be able to construct $A_f$ and automatically create a process-flow diagram, annotated with information about each node.

\subsection{Multi-Output Foreground Processes}

The treatment of multi-functional activities in LCA is a topic of constant discussion among practitioners and theorists.  While the question of the correct, appropriate, or ideal method of apportioning impacts among co-products cannot be resolved in general, the utility of a disclosure framework must be to \textit{express} transparently how co-production was resolved within a given foreground model.  Majeau-Bettez et al. (\citeyear{Majeau_Bettez_2014}) recently provided a general framework (referred to hereafter as MWS14) for describing the treatment of co-products, and I adopt that framework here to show how the technique used to handle co-production can be disclosed transparently for foreground processes.

\subsubsection{Example System}

In all strategies, an abbreviated example system drawn from Ecoinvent, reporting chlor-alkali electrolysis using a membrane cell, is used.  The un-allocated system is shown in Table~\ref{tbl:chlor-alkali}.

\begin{table}
  \begin{center}
  \caption{Example: Chlor-alkali electrolysis, membrane cell [RER] (Ecoinvent 3.2; excerpt)}
  \label{tbl:chlor-alkali}
  \begin{tabular}{rccl}
    \toprule
    \bf Flow & \bf Unit & \bf Direction & \bf Value \\
    \midrule
    \multicolumn{4}{l}{Reference:} \\
    Chlorine, gaseous & kg & Output & 1 \\
    Hydrogen, liquid & kg & Output & 0.028 \\
    Sodium hydroxide & kg dry & Output & 1.13 \\
    \midrule
    \multicolumn{4}{l}{Requirements:} \\
    Electricity, medium voltage & kWh & Input & 2.97 \\
    Sodium chloride, powder & kg & Input & 1.75 \\
    Chemical factory, organics & unit & Input & 4e-10 \\
    Sludge, NaCl electrolysis & kg & Output & 0.0153 \\[1ex]
    \multicolumn{4}{l}{Emissions:} \\
    Chloride [16887-00-6], surface water & kg & Output & 0.015 \\
    Carbon dioxide [124-38-9], urban air & kg & Output & 0.0031 \\
    \bottomrule
  \end{tabular}
  \end{center}
\end{table}

The objective is to disclose explicitly how the multi-output system was resolved to single-output activities for the purposes of review, reproduction, or modification.  Therefore, the raw (un-allocated) process inventory is included in all cases.  Examples are provided for partitioning allocation, classical system expansion, and consequential system expansion.

\subsubsection{Partitioning Allocation}

According to MWS14, Partitioning allocation (PA) is a ``production-balanced model that artificially splits the requirements of an activity across its different outputs.''  The exact manner of the partitioning need not be specified, as long as the parts sum up to the whole.  In the disclosure framework, any manner of production-balanced partitioning can be represented by including the raw, un-allocated inventory as one foreground column, choosing any output as the reference.  The other $n-1$ outputs are then included as allocated inventories in additional columns, for a total of $n$ columns.  Each column must be normalized to a unit reference.

Note that, although it is customary for allocated process inventories to be derived as a uniform partition of the un-allocated process based on some parameter such as mass, price or energy content, this is not necessary.  For a process with $n$ outputs the modeler has $n-1$ degrees of freedom in specifying the allocated inventories, with the requirement that the $n$th inventory equals the remainder.  In the framework presented, the modeler must provide the $n-1$ allocated inventories explicitly.  The reference flow for the $n$th process will simply be computed as the remainder when the allocated inventories are subtracted from the un-allocated inventory.

\begin{table}
  \begin{center}
  \caption{Chlor-alkali electrolysis as a foreground model, partitioning allocation}
  \label{tbl:partition}
  \footnotesize\sffamily
  \begin{tabular}{l|ccc}
    \midrule
    \bf $A_f$ & 0 & 1 & 2 \\
    \midrule
    \textbf{Chlorine, gaseous (kg)} & $\ast$ & & \\
    Hydrogen, liquid (kg) & -0.028 & $\ast$ & \\
    Sodium hydroxide (kg dry) & -1.13 & & $\ast$ \\
    \midrule
    \bf $A_d$ & 0 & 1 & 2 \\
    \midrule
    Electricity, medium voltage (kWh) & 2.97 & 1.38 & 1.38 \\
    Sodium chloride, powder (kg) & 1.75 & 0.811 & 0.811 \\
    Chemical factory, organics (unit) & 4e-10 & 1.9e-10 & 1.9e-10 \\
    Sludge, NaCl electrolysis (kg) & 0.0153 & 0.00709 & 0.00709 \\
    \midrule
    \bf $B_f$ & 0 & 1 & 2 \\
    \midrule
    Chloride, to surface water (kg) & 0.015 & 0.00695 & 0.00695 \\
    Carbon dioxide, to urban air (kg) & 0.0031 & 0.00144 & 0.00144 \\
    \midrule
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tbl:partition} illustrates this for the chlor-alkali example, for conventional allocation by mass.  The un-altered inventory is shown in column $0$, corresponding to the reference product of gaseous chlorine.  Co-products are shown as additional foreground flows, with negative signs indicating that the flows are outputs.  The reference for each column is indicated by an asterisk.

The inventories themselves are shown as entries in the $A_d$ and $B_f$ matrices. Because all co-products are already reported in mass, the allocated inventories for the two co-products are both the same and represent a scaling-down of the un-allocated inventory by the total mass of outputs (about 2.16 kg).

Using the foreground in Table~\ref{tbl:partition}, LCA computation using the canonical functional unit (see main text, Eq.~3) will automatically compute the self-consistent single-output result for 1 kg of chlorine gas because the allocated inventories for hydrogen and sodium hydroxide will be subtracted from the total.  The foreground activity levels for the system are:

\begin{equation}
  \tilde{\mathbf{x}} = [1,\,-0.028,\,-1.13]^T
\end{equation}

with the resulting aggregated dependency and emission vectors being, again, the same as the normalized and allocated inventories:

\begin{eqnarray}
  \tilde{\mathbf{a}}_d = A_d\times \tilde{\mathbf{x}} = & [1.38,\, 0.811,\,\num{1.9e-10},\,0.00709]^T \\
  \tilde{\mathbf{b}}_f = B_f\times \tilde{\mathbf{x}} = & [0.00695,\, 0.00144]^T
\end{eqnarray}

At the same time, LCA computation using a functional unit of one of the co-products will simply make use of the allocated inventory.  In this example, all processes are identical on a mass basis, but any other partitioning allocation can also be automatically rendered self-consistent through this approach.  A reviewer would be able to inspect the foreground model to determine whether the allocation was performed as described in the report.

\subsubsection{``Classical'' System Expansion}

Although the term `system expansion' is often used to refer to product substitution and/or alternate activity allocation, these modeling techniques are distinct and should be referred to with different names, as noted in GWS14.  In the original definition of system expansion, the functional unit is simply enlarged to include all co-products.  Representation of such a system is trivial and is shown in Table~\ref{tbl:cse}.

\begin{table}[h]
  \begin{center}
  \caption{Chlor-alkali electrolysis as a foreground model, classical system expansion}
  \label{tbl:cse}
  \footnotesize\sffamily
  \begin{tabular}{l|ccc}
    \midrule
    \bf $A_f$ & 0 & 1 & 2 \\
    \midrule
    \textbf{Chlorine, gaseous (kg)} & $\ast$ & & \\
    Hydrogen, liquid (kg) & -0.028 & $\ast$ & \\
    Sodium hydroxide (kg dry) & -1.13 & & $\ast$ \\
    \midrule
    \bf $A_d$ & 0 & 1 & 2 \\
    \midrule
    Electricity, medium voltage (kWh) & 2.97 &  &  \\
    Sodium chloride, powder (kg) & 1.75 &  & \\
    Chemical factory, organics (unit) & 4e-10 &  &  \\
    Sludge, NaCl electrolysis (kg) & 0.0153 &  &  \\
    \midrule
    \bf $B_f$ & 0 & 1 & 2 \\
    \midrule
    Chloride, to surface water (kg) & 0.015 &  &  \\
    Carbon dioxide, to urban air (kg) & 0.0031 &  &  \\
    \midrule
  \end{tabular}
  \end{center}
\end{table}

In the classical system expansion case, applying the expanded functional unit of 1 kg chlorine, plus 0.028 kg hydrogen and 1.13 kg of sodium hydroxide, will result in a unit activity of the reference process, without any cut-off flows.  Equivalently, the canonical functional unit, production of 1 kg of chlorine gas, will result in the same process activity level, and will also generate the co-products as cutoff flows.

\subsubsection{Consequential System Expansion}

The foreground disclosure framework is also well suited to describe consequential system expansion through product substitution or ``alternate activity allocation,'' which are the terms used in MWS14.  An example illustrating these methods is shown in Table~\ref{tbl:exp}

\begin{table}[t]
  \begin{center}
  \caption{Chlor-alkali electrolysis as a foreground model, consequential system expansion.  An inventory for the electrolysis of water (shown in blue) is provided as an alternate activity to produce hydrogen, while the sodium hydroxide product is assumed to substitute for for magnesium hydroxide (shown in red).}
  \label{tbl:exp}
  \footnotesize\sffamily
  \psset{unit=0.5cm}
  \begin{tabular}{l|cccc}
    \midrule
    \bf $A_f$ & 0 & 1 & 2 & \blue\textit{3} \\
    \midrule
    \textbf{Chlorine, gaseous (kg)} & $\ast$ & & &\\
    Hydrogen, liquid (kg) & -0.028 & $\ast$ & &\\
    Sodium hydroxide (kg dry) & -1.13 & & $\ast$ & \\
    \blue\textit{Hydrogen, electrolysis (kg)} & & \blue\textit{1} & & $\ast$ \\
    \midrule
    \bf $A_d$ & 0 & 1 & 2 & \blue\textit{3}\\
    \midrule
    Electricity, medium voltage (kWh) & 2.97 &  & & \blue\textit{50} \\
    Sodium chloride, powder (kg) & 1.75 &  & & \\
    Chemical factory, organics (unit) & 4e-10 &  & &  \\
    Sludge, NaCl electrolysis (kg) & 0.0153 &  & & \\
    \blue\textit{Water, distilled} (kg) & & & & \blue\textit{9} \\
    \red\textit{Magnesium hydroxide (kg dry)} & & & \red\textit{0.8} & \\
    \midrule
    \bf $B_f$ & 0 & 1 & 2 & \blue\textit{3}\\
    \midrule
    Chloride, to surface water (kg) & 0.015 &  & & \\
    Carbon dioxide, to urban air (kg) & 0.0031 &  & & \\
    \midrule
  \end{tabular}
  \end{center}
\end{table}

Here an alternate activity, namely electrolysis of water, is included for liquid hydrogen production and is specified at index 3 in the foreground  (shown in blue).  The inventory converts 9 kg of water to 1 kg of hydrogen using 50 kWh of electricity.  The hydrogen produced from the multi-output reference process is (somewhat dubiously) taken to displace hydrogen produced through the alternate activity by introducing a unity coefficient in row 3, column 1 of $A_f$.

Meanwhile, sodium hydroxide from the reference process is modeled to substitute the production of magnesium hydroxide through the introduction of a new row and a new nonzero coefficient in column 2 of $A_d$ (shown in red). Here the match is inexact, and because magnesium hydroxide is more effective in some applications, 1 kg of sodium hydroxide is taken to displace only 0.8 kg of magnesium hydroxide.

\section{Examples}

LCI databases often contain product systems that can be modeled as foreground studies because they describe products that are not required by the background.  In this section, two product systems selected from LCI databases are used to illustrate the concept of structured publication.  The systems selected because are complex enough to illustrate the premise but simple enough to review easily.

Each system is illustrated as a table that shows the foreground model, cutoff flows, background dependencies and foreground emissions included in the system.  Aggregation results $\tilde{\mathbf{x}}$, $\tilde{\mathbf{a}}_d$, and $\tilde{\mathbf{b}}_f$ are also reported.  The table omits numeric data from most of the table for clarity. Instead, a black square indicates the presence of a nonzero value.  %The systems are also reproduced as structured publications in the supplementary materials, formatted as Excel spreadsheets.

\subsection{US LCI: Secondary Aluminum}

\begin{figure}
  \input{../figures/aluminum-secondary}
  \caption{A structured product system model for secondary aluminum, drawn from US LCI. Exchange values are replaced with black squares for clarity.}
  \label{table:aluminum}
\end{figure}

The US LCI database contains a small background system of 39 processes, as well as 395 foreground product systems that range in size from one to 83 foreground flows.
%The largest, ``paper, freesheet, coated, at mill'' makes reference to 34 of the 39 background processes.

The example product system  reports the production of secondary aluminum from automotive scrap (Figure~\ref{table:aluminum}).   The aluminum production process has direct requirements for two other foreground systems, including transportation services and quicklime production.  The quicklime production itself depends on limestone extraction, which is also part of the foreground.  The aluminum production requires two background systems, electricity and natural gas combustion, and reports 10 cutoff flows.  The main input to the process, ``Aluminum scrap, automotive,'' appears as a cutoff flow.  Overall, the four foreground processes require the background for only combustion (five fuels), transport (three modes) and grid electricity.  The same natural gas combustion and electricity models were used in all three non-transport foreground nodes.

The system also reports several direct emissions, mainly from the aluminum process.  Only one resource consumption (input flow) is reported -- the extraction of limestone. The direct emission modeling is limited in scope and includes a number of ``unspecified'' flows that may not be well characterized in many impact methods.

%The structured publication of the secondary aluminum model includes a total of 47 data points about 36 entities.  The $A_f$, $A_d$, and $B_f$ matrices are reported in sparse format, along with the aggregation vectors. In the publication, LCIA scores are reported for several impact indicators drawn from the Ecoinvent reference LCIA implementation.\footnote{The Ecoinvent LCIA factors are not published, but are available from the Ecoinvent Centre. The name of the file used in the current study is \texttt{LCIA implementation v3.1 2014\_08\_13.xlsx}.}    The $E$ matrix is also partially reproduced: entries are only reported if they match flows that appear in the foreground.

\subsection{Ecoinvent: Organic Potatoes}

\begin{figure}
  \input{../figures/potato-organic}
  \caption{A structured product system model for organic potato production, drawn from Ecoinvent v3.2 (APOS). Exchange values are replaced with black squares for clarity.}
  \label{table:potato}
\end{figure}

Ecoinvent version 3 is provided in three different system models that reflect different linking strategies.  The example is drawn from the ``Allocation at the point of substitution'' or APOS model of Ecoinvent 3.2, which includes 11,420 processes that produce 12,966 product flows.  Of these, 10,282 are background flows and the rest are foreground flows.

The example system reports production of organic potatoes supplied to the global market (Figure \ref{table:potato}).  The foreground includes nine nodes, of which six (nodes 3 through 8) contain cyclic dependencies associated with the production of potato seeds.  The Ecoinvent database's use of ``markets'' as mixer processes is evident in the table: node 0 is a mixer process that combines Swiss (``CH'') production (node 1) weighted at 2.4\%  with rest-of-World (``RoW'') production (node 2) weighted at 97.6\%.  That same market split (2.4 / 97.6) can also be seen in nodes 4 and 8.

Nodes 5 and 6, which make seeds ready ``for setting,'' are mixed by node 8. Although nodes 5 and 6 are geographically distinguished (CH vs RoW), both can be seen to consume potato seed from the global market.

Looking at the dependency and emission lists, the different ``signatures'' of different kinds of processes can be seen: nodes 1, 2, 3, and 7 are clearly agricultural processes that require irrigation, tillage, manure and so on.  Nodes 0 and 4 are visible as market processes, their only requirements being transport processes.  Nodes 5 and 6 each consume electricity and require use of a ``multi-storey building.''  Taken together, the CH-locality processes appear to use CH irrigation and electricity supply but are otherwise similar to their RoW counterparts.  The product model includes no cutoff flows.

%The structured publication of the organic potato model reports the $A_f$, $A_d$, and $B_f$ matrices in full rather than sparse format to enable a more visually immediate review.  In addition, the $E$ matrix is reported in full, for all emissions found in the model, rather than only the emissions found in the foreground.  This enables the publication to be used to visually evaluate different implementations of the same impact assessment method in the context of the inventory model (Ecoinvent vs ILCD implementations of ReCiPe Midpoint marine eutrophication; see supplementary materials for details).

 \bibliography{kuczenski-disclosure}
}

\begin{document}

\supporting

\end{document}

### 2018-07-09 aggregation/agg-multioutput.tex
backport post-accept changes

@@
 Majeau-Bettez et al.
-(2014) recently provided a general framework (referred to hereafter as MWS14) for describing the treatment of co-products, and I adopt that framework here to show how the technique used to handle co-production can be disclosed transparently for foreground processes.
+(\citeyear{Majeau_Bettez_2014}) recently provided a general framework (referred to hereafter as MWS14) for describing the treatment of co-products, and I adopt that framework here to show how the technique used to handle co-production can be disclosed transparently for foreground processes.
 

### 2018-07-09 aggregation/equivalence.tex
backport post-accept changes

@@
 
-There is a level of confusion about the differing implications of process-based LCA inventory analysis as described in Heijungs and Suh (2002), and input-output analysis using a Leontief matrix.
-In fact, these two techniques were shown to be equivalent in Suh et al, 2010\footnote{\doi{10.1111/j.1530-9290.2010.00235.x}; see full text for citation.}, whose abstract declares: ``the article shows that the partitioning method in LCA is equivalent to the industry-technology model in input-output economics, and system expansion in LCA is equivalent to the by-product-technology model in input-output output economics.''  The article does indeed show convincingly that ``the two data types can be hybridized without the loss of methodological consistency'' (p.345).
+There is a level of confusion about the differing implications of process-based LCA inventory analysis as described in Heijungs and Suh (\citeyear{Heijungs2002}), and input-output analysis using a Leontief matrix.
+In fact, these two techniques were shown to be equivalent in Suh et al, (\citeyear{Suh_JIE_2010}), whose abstract declares: ``the article shows that the partitioning method in LCA is equivalent to the industry-technology model in input-output economics, and system expansion in LCA is equivalent to the by-product-technology model in input-output output economics.''  The article does indeed show convincingly that ``the two data types can be hybridized without the loss of methodological consistency'' (p.345).
 
